---
title: "R Notebook"
output: html_notebook
---

<!-- Single-cell analysis of Lymph Node data: from raw count matrices to an
     annotated Seurat object, followed by cell-cell communication (CCC)
     analysis on the same object. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

```{r}
# additional packages needed for QC and doublet detection
library(DropletUtils)
library(plyr)
library(scDblFinder)
library(BiocParallel)
```

## 2. Load raw data

```{r}
# read the raw count matrices for each sample
m1 <- Read10X("data/raw/temp/exp3/m1/MN1")
m2 <- Read10X("data/raw/temp/exp3/m2/MN2")
m3 <- Read10X("data/raw/temp/exp3/m3/MN3")
m4 <- Read10X("data/raw/temp/exp3/m4/MN4")
m5 <- Read10X("data/raw/temp/exp3/m5/MN5")
m6 <- Read10X("data/raw/temp/exp3/m6/MN6")
m7 <- Read10X("data/raw/temp/exp3/m7/MN7")
m8 <- Read10X("data/raw/temp/exp3/m8/MN8")
m9 <- Read10X("data/raw/temp/exp3/m9/MN9")
```

## 3. Create Seurat objects and merge

```{r}
data_list <- list(m1, m2, m3, m4, m5, m6, m7, m8, m9)

# build a Seurat object from one sample's matrix: keep real cells only
# (via defaultDrops) and tag the object with its sample of origin
process_matrix <- function(mat, sample_name, expected_cells = 5000) {

  # handle multimodal data (e.g. GEX + ADT); keep Gene Expression only
  if (is.list(mat)) mat <- mat[["Gene Expression"]]

  # identify real cells using defaultDrops
  em <- defaultDrops(mat, expected = expected_cells)
  keep <- which(em)

  if (length(keep) == 0) {
    stop(paste("No cells retained for sample:", sample_name))
  }

  mat_filtered <- mat[, keep, drop = FALSE]

  Obj <- CreateSeuratObject(
    counts       = mat_filtered,
    project      = "Lymph Nodes",
    min.cells    = 3,
    min.features = 200
  )
  Obj$sample <- sample_name

  return(Obj)
}

# expected number of real cells per sample, used by defaultDrops
expected_cells_vec <- c(5000, 5000, 5000, 5000, 5000, 5000, 5000, 5000, 5000)

obj_list <- lapply(
  seq_along(data_list),
  function(i) process_matrix(data_list[[i]],
                              sample_name    = paste0("Obj", i),
                              expected_cells = expected_cells_vec[i])
)
names(obj_list) <- paste0("Obj", seq_along(obj_list))

Obj <- merge(
  x            = obj_list[[1]],
  y            = obj_list[2:length(obj_list)],
  add.cell.ids = names(obj_list),
  project      = "LymphNodes"
)

saveRDS(Obj, file = "data/process/SeuratObjects/Lymph_data/LN_raw.rds")
```

## 4. Quality control

```{r}
Obj <- PercentageFeatureSet(Obj, col.name = "percent.mt", assay = "RNA", pattern = "^MT-")
VlnPlot(Obj, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC thresholds set based on this dataset's own distributions
Obj <- subset(Obj, subset = nFeature_RNA > 300 & nFeature_RNA < 8000 & percent.mt < 10)
```

## 5. Normalization and PCA

```{r}
Obj <- NormalizeData(Obj, verbose = FALSE)
Obj <- FindVariableFeatures(Obj)
Obj <- ScaleData(Obj, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = Obj))
Obj <- RunPCA(Obj, features = VariableFeatures(object = Obj))
ElbowPlot(Obj, ndims = 50, reduction = "pca")
```

## 6. Harmony integration and clustering resolutions

```{r}
library(harmony)
Obj <- RunHarmony(Obj, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:30)
Obj[["RNA"]] <- JoinLayers(Obj[["RNA"]])
Obj <- RunUMAP(Obj, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")
DimPlot(Obj, reduction = "umap.harmony", group.by = c("sample"), label.size = 2, shuffle = TRUE)

# scan a range of resolutions before picking one
Obj <- FindNeighbors(Obj, dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  Obj <- FindClusters(Obj, resolution = res)
  colname <- paste0("res_", res)
  Obj[[colname]] <- Obj$seurat_clusters
}

res_cols <- grep("^res_", colnames(Obj@meta.data), value = TRUE)
for (res in res_cols) {
  p <- DimPlot(Obj, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}

# resolution 0.4 was chosen for downstream clustering
p_res04 <- DimPlot(Obj, group.by = "res_0.4", label = TRUE, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.4") +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/Lymph_node/resolution.pdf", plot = p_res04, width = 10, height = 8, dpi = 300)
```

## 7. Clustering and marker genes

```{r}
Obj <- FindNeighbors(Obj, reduction = "harmony", dims = 1:30)
Obj <- FindClusters(Obj, resolution = 0.4, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
Obj_markers <- FindAllMarkers(Obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
Obj_markers <- Obj_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 8. Cell type annotation

```{r}
library(openxlsx)

clusters <- unique(Obj_markers$cluster)

# top 30 marker genes per cluster, by log fold-change
top_genes_list <- list()
for (clust in clusters) {
  top_genes <- Obj_markers %>%
    filter(cluster == clust) %>%
    arrange(desc(avg_log2FC)) %>%
    slice_head(n = 30) %>%
    pull(gene)

  top_genes_list[[paste0("cluster_", clust)]] <- top_genes
}

for (clust in names(top_genes_list)) {
  cat("=== ", clust, " ===\n")
  print(top_genes_list[[clust]])
  cat("\n")
}

# combine into one table, one column per cluster
top_genes_df <- data.frame(
  matrix(
    NA,
    nrow = 30,
    ncol = length(clusters),
    dimnames = list(NULL, paste0("cluster_", clusters))
  )
)

for (i in seq_along(clusters)) {
  clust <- clusters[i]
  genes <- top_genes_list[[paste0("cluster_", clust)]]
  top_genes_df[1:length(genes), i] <- genes
}
```

```{r}
# curated marker genes for lymph node stromal, endothelial and immune cell types
genes <- c(
  # Endothelial - Arterial EC
  "SEMA3G", "HEY1", "ACE",

  # Endothelial - Capillary EC
  "CA4", "RGCC", "CD36", "PLVAP", "CD300LG",

  # Endothelial - Venous EC
  "ACKR1", "MADCAM1", "CH25H", "SELE", "SELP", "VCAM1",

  # LEC - bLEC (boundary)
  "ACKR4", "SERPINE1", "BMP2",

  # LEC - fLEC (floor)
  "CXCL5", "MARCO", "MFAP4",

  # LEC - msLEC (mantle sinus)
  "PTX3", "STAB2", "LYVE1",

  # LEC - Valve
  "CLDN11", "GJA4", "PROX1",

  # Stromal - Adventitial SC (advSC)
  "FBN1", "PCOLCE2", "CD34", "PI16", "DCN",

  # Stromal - ATG SC
  "APOE", "APOC1", "AGT", "LPL", "APOD",

  # Stromal - TRC (T cell zone Reticular)
  "CCL19", "CCL21", "CCL8", "CXCL13", "ICAM1",

  # Stromal - Pericytes (PC)
  "RGS5", "NOTCH3", "ACTA2", "PDGFRB", "MCAM",

  # Stromal - Smooth Muscle Cells (SMC)
  "MYH11", "TAGLN", "MYL9", "CNN1", "ACTG2",

  # Stromal - Fibroblasts
  "SFRP2", "SFRP4", "MFAP5", "SVEP1", "FBLN1",

  # Cytotoxic T CD8
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T CD4
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A"
)
```

```{r}
# use the marker genes to check expression on the UMAP
FeaturePlot(Obj, features = genes, combine = FALSE, order = TRUE, reduction = "umap.harmony")
```

```{r}
p <- DotPlot(
  Obj,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "harmony_clusters"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)

ggsave(filename = "output/Lymph_node/LN_cluster.pdf", plot = p, width = 15, height = 10)
```

```{r}
# map cluster numbers to cell type labels
Idents(Obj) <- "harmony_clusters"
cluster_name <- c(
  "0"  = "Fibroblasts",
  "1"  = "Endothelial Cells",
  "2"  = "Endothelial Cells",
  "3"  = "Endothelial Cells",
  "4"  = "Plasma Cells",
  "5"  = "B Cells",
  "6"  = "Mural Cells",
  "7"  = "Endothelial Cells",
  "8"  = "Endothelial Cells",
  "9"  = "Fibroblasts",
  "10" = "CD8",
  "11" = "Fibroblasts",
  "12" = "Smooth Muscle Cells",
  "13" = "Endothelial Cells",
  "14" = "CD4",
  "15" = "Mono/Mac",
  "16" = "Endothelial Cells",
  "17" = "Endothelial Cells",
  "18" = "Fibroblasts",
  "19" = "Endothelial Cells"
)

Obj <- RenameIdents(Obj, cluster_name)
Obj$General <- Idents(Obj)

p <- DotPlot(
  Obj,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "General"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)

ggsave(filename = "output/Lymph_node/LN_cluster_name.pdf", plot = p, width = 15, height = 10)
```

```{r}
celltype_cols <- c(
  "Fibroblasts"           = "cyan",
  "CD8"                   = "tomato",
  "CD4"                   = "olivedrab2",
  "Mono/Mac"              = "darkblue",
  "Endothelial Cells"     = "#E31A1C",
  "Plasma Cells"          = "paleturquoise3",
  "Mural Cells"           = "#1F78B4",
  "B Cells"               = "gold",
  "Smooth Muscle Cells"   = "lightpink2"
)

# UMAP with a label placed at each cell type's centroid
umap_df <- as.data.frame(Embeddings(Obj, "umap.harmony"))
umap_df$General <- Obj$General

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ General,
  data = umap_df,
  FUN = median
)

library(ggrepel)
fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = General)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = General),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4.5,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

ggsave(filename = "output/Lymph_node/CellType.pdf", plot = fig1a, width = 12, height = 10)
```

## 9. Marker expression heatmap

```{r}
# bubble heatmap: dot size = percent of cells expressing, color = scaled expression
Idents(Obj) <- "General"
Obj$General <- gsub("-", "_", Obj$General)
Idents(Obj) <- "General"

DefaultAssay(Obj) <- "RNA"
expr_mat <- GetAssayData(Obj, layer = "data")

genes <- genes[genes %in% rownames(expr_mat)]
if (length(genes) < 2) stop("Too few genes found")

celltypes <- levels(factor(Idents(Obj)))

# percent of cells expressing each marker, per cell type
percent_mat <- sapply(celltypes, function(ct) {
  cells <- WhichCells(Obj, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  Obj,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA
colnames(avg_exp) <- gsub("-", "_", colnames(avg_exp))

common_ct <- intersect(celltypes, colnames(avg_exp))
if (length(common_ct) < 2) stop("Cell type mismatch still exists")

percent_mat <- percent_mat[, common_ct, drop = FALSE]
exp_mat <- as.matrix(avg_exp[genes, common_ct, drop = FALSE])

# clean and scale
exp_mat[is.infinite(exp_mat)] <- NA
exp_mat <- exp_mat[rowSums(is.na(exp_mat)) < ncol(exp_mat), , drop = FALSE]
percent_mat <- percent_mat[rownames(exp_mat), , drop = FALSE]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
if (all(exp_mat_scaled == 0)) stop("No expression signal")

exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# color scale
q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95), na.rm = TRUE)
q <- unique(as.numeric(q))
q <- q[is.finite(q)]
if (length(q) < 2) q <- range(exp_mat_scaled, na.rm = TRUE)
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# custom cell function drawing a dot per cluster/gene combination
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(common_ct))
names(cols) <- common_ct

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = common_ct,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf("output/Lymph_node/heatmap.pdf", width = 12, height = 12)
ComplexHeatmap::draw(ht)
dev.off()
```

## 10. Save processed Seurat object

```{r}
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mt",
  "sample",
  "harmony_clusters",
  "General"
)
Obj@meta.data <- Obj@meta.data[, keep_cols]

saveRDS(Obj, file = "data/process/SeuratObjects/Lymph_data/LN_process.rds")
Obj <- readRDS("data/process/SeuratObjects/Lymph_data/LN_process.rds")
```

## 11. Cell-cell communication setup

```{r}
suppressMessages({
  library(dplyr, quietly = TRUE)
  library(tidyr, quietly = TRUE)
  library(purrr, quietly = TRUE)
  library(forcats, quietly = TRUE)
  library(textshape, quietly = TRUE)
  library(ggpubr, quietly = TRUE)
  library(rstatix, quietly = TRUE)
  library(reshape2, quietly = TRUE)
  library(tibble, quietly = TRUE)

  library(scater, quietly = TRUE)
  library(scuttle, quietly = TRUE)

  library(ggplot2, quietly = TRUE)

  library(liana, quietly = TRUE)
  library(decoupleR, quietly = TRUE)
  library("ExperimentHub")
  c2c <- reticulate::import(module = "cell2cell")
})

# load the processed Seurat object and convert to SingleCellExperiment
LN <- readRDS("data/process/SeuratObjects/Lymph_data/LN_process.rds")
sce <- as.SingleCellExperiment(LN)
saveRDS(sce, file = "data/process/SeuratObjects/Lymph_data/sce_raw.rds")
```

## 12. Quality control (SingleCellExperiment)

```{r}
# check the threshold before filtering
hist(
  colSums(counts(sce) > 0),
  breaks = 100,
  main = "Detected genes per cell",
  xlab = "Number of genes"
)

# basic filters
sce <- scater::addPerCellQC(sce)
min.features <- 500
min.cells <- 5

sce <- sce[rowSums(counts(sce) >= 1) >= min.cells, ]
sce <- sce[, colSums(counts(sce) >= 1) >= min.features]

# mitochondrial percentage
is.mito <- grep("MT-", rownames(sce))
per.cell <- scuttle::perCellQCMetrics(sce, subsets = list(Mito = is.mito))
colData(sce)[["subsets_Mito_percent"]] <- per.cell$subsets_Mito_percent

hist(
  colData(sce)$subsets_Mito_percent,
  breaks = 100,
  main = "Mitochondrial percentage per cell",
  xlab = "Percent mitochondrial reads"
)
sce <- sce[, colData(sce)$subsets_Mito_percent < 15]
```

## 13. Normalization (SingleCellExperiment)

```{r}
scale.factor <- 1e4
lib.sizes <- colSums(counts(sce)) / scale.factor
assay(sce, "logcounts") <- scuttle::normalizeCounts(sce, size.factors = lib.sizes, log = FALSE, center.size.factors = FALSE)
assay(sce, "logcounts") <- log1p(assay(sce, "logcounts"))
head(colData(sce))
```

## 14. Cell-cell communication analysis (LIANA)

```{r}
liana_res <- liana_bysample(
  sce = sce,
  sample_col = "sample", # context dimension column from SCE metadata
  idents_col = "General", # cell types column from SCE metadata
  resource = "Consensus",
  expr_prop = 0.1, # must be expressed in expr_prop fraction of cells
  min_cells = 5,
  assay.type = "logcounts", # run on log- and library-normalized counts
  aggregate_how = "both", # aggregate magnitude and specificity
  verbose = TRUE,
  inplace = FALSE, # keep result separate from the sce object
  permutation.params = list(nperms = 100),
  return_all = TRUE
)

# convert list of per-sample results into one data frame
names(liana_res) <- unique(colData(sce)$sample)
liana_res <- bind_rows(liana_res, .id = "sample")
liana_res$sample <- as.character(liana_res$sample)
table(liana_res$sample)
write.csv(liana_res, file = "CCC/Lymph_node/liana_res.csv", row.names = FALSE)

# filter for strong, significant interactions
liana_filt <- liana_res %>%
  filter(
    magnitude_rank <= 0.05,
    cellphonedb.pvalue <= 0.01
  )
table(liana_filt$sample)
write.csv(liana_filt, file = "CCC/Lymph_node/liana_filt.csv", row.names = FALSE)
```

## 15. Save processed SCE object

```{r}
saveRDS(sce, file = "data/process/SeuratObjects/Lymph_data/sce_processed.rds")
```
