---
title: "Lymph Node - Autoimmune Fusion and Comparison"
output: html_notebook
---

<!-- LN_Au: fusion of the Lymph Node reference data with the merged
     autoimmune disease ("Au") object, then comparison of cell type
     composition and cell-cell communication between the two. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

## 2. Load data

```{r}
LN <- readRDS("data/process/SeuratObjects/Lymph_data/LN_process.rds")
Au <- readRDS("data/process/SeuratObjects/Fusion/F_process.rds")

names(Au@assays) # Au has both SCT and RNA
DefaultAssay(Au) <- "RNA"
Au <- DietSeurat(Au, assays = "RNA") # drop the SCT assay
```

## 3. Merge Lymph Node and autoimmune fusion data

```{r}
LN_Au <- merge(
  x = Au,
  y = LN,
  add.cell.ids = c("AuDiseases", "LymphNodes")
)
LN_Au <- JoinLayers(LN_Au)

saveRDS(LN_Au, file = "data/process/SeuratObjects/LN_Au/LN_Au_raw.rds")
# LN_Au <- readRDS("data/process/SeuratObjects/LN_Au/LN_Au_raw.rds")
```

## 4. Normalization, scaling, and PCA

```{r}
LN_Au <- NormalizeData(LN_Au)
LN_Au <- FindVariableFeatures(LN_Au)
LN_Au[["percent.mt"]] <- PercentageFeatureSet(LN_Au, pattern = "^MT-") # Human
LN_Au <- ScaleData(LN_Au, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = LN_Au))
LN_Au <- RunPCA(LN_Au, features = VariableFeatures(object = LN_Au))
ElbowPlot(LN_Au, ndims = 50, reduction = "pca")
LN_Au <- RunUMAP(LN_Au, dims = 1:30)
p <- DimPlot(LN_Au, reduction = "umap", label = TRUE, raster = FALSE)
```

## 5. Harmony integration

```{r}
library(harmony)
LN_Au <- RunHarmony(LN_Au, group.by.vars = "orig.ident", orig.reduction = "pca", dims.use = 1:30)
LN_Au[["RNA"]] <- JoinLayers(LN_Au[["RNA"]])
LN_Au <- RunUMAP(LN_Au, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")
```

## 6. UMAP plot colored by disease

```{r}
# harmonize the Lymph Node label and drop the factor's legend title
LN_Au@meta.data$orig.ident <- gsub("Lymph Nodes", "LymphNodes", LN_Au@meta.data$orig.ident)
LN_Au$orig.ident <- factor(LN_Au$orig.ident)
names(attributes(LN_Au$orig.ident)$levels) <- NULL

Idents(LN_Au) <- "orig.ident"

disease_colors <- c(
  "MS"         = "darkolivegreen3",
  "RA"         = "red3",
  "SS"         = "orangered",
  "IPF.U"      = "cyan4",
  "IPF.L"      = "cyan",
  "T1D"        = "darkred",
  "SScILD.U"   = "gold",
  "SScILD.L"   = "magenta",
  "SjD"        = "darkblue",
  "LymphNodes" = "green"
)

P <- DimPlot(
  LN_Au,
  reduction = "umap.harmony",
  group.by = "orig.ident",
  label = FALSE,
  cols = disease_colors,
  raster = TRUE
) +
  theme(
    aspect.ratio = 1,
    legend.text = element_text(size = 12),
    legend.position = "right"
  ) +
  labs(title = NULL)

P <- P + labs(x = "UMAP.h", y = "UMAP.h")

ggsave(filename = "output/LN_Au/UMAP.pdf", plot = P, height = 8, width = 12)
```

## 7. Shared cell types across diseases

```{r}
library(dplyr)
library(tidyr)
library(pheatmap)
library(grid)

#0. exclude the same cell types used throughout the project, starting
# directly from the merged object's metadata
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
                         "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
                         "Plasma Cells")
meta_filtered <- LN_Au@meta.data %>%
  filter(!General %in% excluded_celltypes)
cat("Cell types remaining after exclusion:\n")
print(sort(unique(meta_filtered$General)))

#1. count how many diseases per cell type (using the FILTERED metadata)
tab <- table(meta_filtered$General, meta_filtered$orig.ident)
n_diseases <- rowSums(tab > 0)
celltype_disease_table <- data.frame(
  General = rownames(tab),
  n_diseases = n_diseases,
  diseases = apply(tab > 0, 1, function(x) {
    paste(colnames(tab)[x], collapse = ", ")
  }),
  row.names = NULL
)

#2. extract only shared cell types (present in >=2 diseases)
shared_celltypes <- celltype_disease_table$General[
  celltype_disease_table$n_diseases >= 2
]
shared_celltype_table <- celltype_disease_table[
  celltype_disease_table$n_diseases >= 2,
]
write.csv(
  shared_celltype_table,
  file = "output/LN_Au/shared_celltype.csv",
  row.names = FALSE
)
cat("Shared cell types (>=2 diseases):\n")
print(shared_celltypes)

#3. count cells per shared cell type per disease
count_tbl_shared <- meta_filtered %>%
  filter(General %in% shared_celltypes) %>%
  dplyr::count(General, orig.ident) %>%
  filter(n >= 2)

#4. convert to matrix
count_mat_shared <- count_tbl_shared %>%
  pivot_wider(
    names_from = orig.ident,
    values_from = n,
    values_fill = 0
  )
count_mat_shared <- as.data.frame(count_mat_shared)
rownames(count_mat_shared) <- count_mat_shared$General
count_mat_shared$General <- NULL
count_mat_shared <- as.matrix(count_mat_shared)

#5. convert counts to PERCENTAGE of cells per disease
# (each column = one disease; total is the FULL filtered cell count for
# that disease, across all kept cell types - not just the shared ones -
# so percentages reflect true composition, not just the shared subset)
total_per_disease <- meta_filtered %>%
  dplyr::count(orig.ident) %>%
  tibble::deframe()
# align to the same disease columns as count_mat_shared, in the same order
total_per_disease <- total_per_disease[colnames(count_mat_shared)]
percent_mat_shared <- sweep(count_mat_shared, 2, total_per_disease, FUN = "/") * 100
percent_mat_shared <- round(percent_mat_shared, 1)

#6. legend breaks/labels so the color scale ALSO reads as a percentage
# (in addition to - not instead of - the numbers printed inside cells)
max_val <- max(percent_mat_shared)
color_breaks  <- seq(0, max_val, length.out = 101) # 101 breaks -> 100 colors
legend_breaks <- pretty(c(0, max_val))
legend_breaks <- legend_breaks[legend_breaks >= 0 & legend_breaks <= max_val]
legend_labels <- paste0(legend_breaks, "%")
legend_title  <- "% of cells"

out_width  <- 10
out_height <- 12

#7. build the heatmap on a THROWAWAY scratch device first, sized to
# match the final output. pheatmap(silent = TRUE) still does an
# internal calibration draw on whatever device is active (that's what
# was producing the extra blank page) - so we let that calibration
# draw land on a disposable temp file, discard it, then draw the
# already-built gtable exactly once on the real output device below.
scratch_file <- tempfile(fileext = ".pdf")
pdf(scratch_file, width = out_width, height = out_height)
p <- pheatmap::pheatmap(
  percent_mat_shared,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  display_numbers = percent_mat_shared,
  number_color = "white",
  color = colorRampPalette(c("grey", "red", "red3"))(100),
  breaks = color_breaks,
  legend_breaks = legend_breaks,
  legend_labels = legend_labels,
  border_color = NA,
  fontsize_row = 12,
  fontsize_col = 12,
  fontsize_number = 12,
  silent = TRUE
)
dev.off()
unlink(scratch_file)

gt <- p$gtable

#8. open the REAL output device and locate the layout boundaries we
# need (heatmap body panel, and the color-scale legend panel)
pdf("output/LN_Au/HEATMAP_shared_celltype.pdf", width = out_width, height = out_height)
grid.newpage() # IMPORTANT: start the page BEFORE any convertWidth/Height
                # call below. convertWidth() on a device with no page
                # started yet silently creates its own blank page - that
                # was the cause of the extra page in the PDF.

w <- convertWidth(gt$widths, "npc", valueOnly = TRUE)
h <- convertHeight(gt$heights, "npc", valueOnly = TRUE)

mat_row <- gt$layout[gt$layout$name == "matrix", ]
x0 <- sum(w[seq_len(mat_row$l - 1)])
x1 <- sum(w[seq_len(mat_row$r)])
y1 <- 1 - sum(h[seq_len(mat_row$t - 1)])
y0 <- 1 - sum(h[seq_len(mat_row$b)])

leg_row <- gt$layout[gt$layout$name == "legend", ]
leg_x0 <- sum(w[seq_len(leg_row$l - 1)])
leg_x1 <- sum(w[seq_len(leg_row$r)])
leg_y1 <- 1 - sum(h[seq_len(leg_row$t - 1)])

#9. find where "LymphNodes" ended up after column clustering
col_order <- if (!is.null(p$tree_col)) p$tree_col$order else seq_len(ncol(percent_mat_shared))
lymph_pos <- which(colnames(percent_mat_shared)[col_order] == "LymphNodes")
n_col <- ncol(percent_mat_shared)
col_width_npc <- (x1 - x0) / n_col
rect_x0 <- x0 + (lymph_pos - 1) * col_width_npc
rect_x1 <- rect_x0 + col_width_npc

#10. draw the final figure, ONCE: heatmap (with numbers) + a square
# outline around ONLY the LymphNodes column + a title above the legend
grid.draw(gt)
grid.rect(
  x = unit((rect_x0 + rect_x1) / 2, "npc"),
  y = unit((y0 + y1) / 2, "npc"),
  width = unit(rect_x1 - rect_x0, "npc"),
  height = unit(y1 - y0, "npc"),
  gp = gpar(col = "black", fill = NA, lwd = 3)
)
grid.text(
  legend_title,
  # right-justified, anchored at the legend's right edge: the color
  # bar column is narrower than most titles, so centering it let the
  # text run past the page's right edge and get clipped. Anchoring at
  # leg_x1 (right edge) and growing leftward keeps it fully on the page.
  x = unit(leg_x1, "npc"),
  y = unit(leg_y1 + 0.015, "npc"),
  just = c("right", "bottom"),
  gp = gpar(fontsize = 10, fontface = "bold")
)
dev.off()
```

## 8. Correlation between cell types

```{r}
Idents(LN_Au) <- "orig.ident"

#1. average expression per disease of origin (rows = genes, columns = disease)
av.exp <- AverageExpression(
  LN_Au,
  assays = "RNA",
  slot = "data"
)$RNA

#2. correlation matrix across diseases
correlation_matrix <- cor(
  as.matrix(av.exp),
  use = "pairwise.complete.obs"
)
correlation_matrix[is.na(correlation_matrix)] <- 0

#3. plot
mcol <- colorRampPalette(c("white", "darkblue", "red"))(100)
library(pheatmap)
pheatmap(
  correlation_matrix,
  clustering_method = "ward.D2",
  color = mcol,
  border_color = NA,
  fontsize_row = 10,
  fontsize_col = 10,
  display_numbers = FALSE,
  filename = "output/LN_Au/correlation.pdf",
  width = 10,
  height = 12
)
```

## 9. Compare cell-cell interactions between autoimmune diseases and Lymph Node

```{r}
library(dplyr)

#1. load the LIANA output for every disease, plus Lymph Node
MS         <- read.csv("CCC/MS/output/liana_filt.csv")
RA         <- read.csv("CCC/RA/output/liana_filt.csv")
SS         <- read.csv("CCC/SS/output/liana_filt.csv")
T1D        <- read.csv("CCC/T1D/output/liana_filt.csv")
SjD        <- read.csv("CCC/SjD/output/liana_filt.csv")
SScILD.U   <- read.csv("CCC/SScILD/Upper/output/liana_filt.csv")
SScILD.L   <- read.csv("CCC/SScILD/Low/output/liana_filt.csv")
IPF.U      <- read.csv("CCC/IPF/Upper/output/liana_filt.csv")
IPF.L      <- read.csv("CCC/IPF/Low/output/liana_filt.csv")
Lymph_Node <- read.csv("CCC/Lymph_node/liana_filt.csv")

diseases_list <- list(
  MS = MS, RA = RA, SS = SS, T1D = T1D, SjD = SjD,
  SScILD.U = SScILD.U, SScILD.L = SScILD.L,
  IPF.U = IPF.U, IPF.L = IPF.L,
  "Lymph Node" = Lymph_Node
)

# sanity check: confirm every dataset - including Lymph Node - has the
# columns the rest of this chunk depends on, before building "interaction"
# from them (a differently-structured liana_filt.csv is the most likely
# cause if this warning fires)
required_cols <- c("ligand.complex", "receptor.complex", "magnitude_rank", "source", "target")
for (d in names(diseases_list)) {
  cols <- names(diseases_list[[d]])
  missing <- setdiff(required_cols, cols)
  if (length(missing) > 0) {
    cat("[WARNING]", d, "is missing columns:", paste(missing, collapse = ", "), "\n")
    cat("  actual columns in", d, ":", paste(cols, collapse = ", "), "\n")
  }
}

#2. rename cell types in source/target, same mapping used for RSS/TLS,
# and drop any interaction where source or target is Glial Cells, Neurons, or Immune Cells
celltype_rename_map <- c(
  "Astrocytes" = "Glial Cells",
  "Oligodendrocytes" = "Glial Cells",
  "OPC" = "Glial Cells",
  "Schwann Cells" = "Glial Cells",
  "Neurons" = "Neurons",
  "Endothelial Cells" = "Endothelial Cells",
  "Arterial Endothelial" = "Endothelial Cells",
  "Venous Endothelial" = "Endothelial Cells",
  "Lymphatic Endothelial" = "Endothelial Cells",
  "Capillary Endothelial" = "Endothelial Cells",
  "Peribronchial Endothelial" = "Endothelial Cells",
  "Macrophages" = "Mono/Mac",
  "Monocytes" = "Mono/Mac",
  "Myeloid Cells" = "Mono/Mac",
  "Mono/Mac" = "Mono/Mac",
  "Dendritic Cells" = "Mono/Mac",
  "Plasmacytoid Dendritic" = "Mono/Mac",
  "Microglia" = "Mono/Mac",
  "Alveolar Cells" = "Epithelial Cells",
  "Ciliated Cells" = "Epithelial Cells",
  "Goblet Cells" = "Epithelial Cells",
  "Basal Cells" = "Epithelial Cells",
  "Secretory Cells" = "Epithelial Cells",
  "Acinar Cells" = "Epithelial Cells",
  "Ductal Cells" = "Epithelial Cells",
  "Basal Keratinocyte" = "Keratinocytes",
  "Suprabasal Keratinocyte" = "Keratinocytes",
  "Alpha Cells" = "Endocrine Cells",
  "Delta Cells" = "Endocrine Cells",
  "CD4" = "CD4",
  "CD8" = "CD8",
  "B Cells" = "B Cells",
  "Fibroblasts" = "Fibroblasts",
  "Mast Cells" = "Mast Cells",
  "Melanocytes" = "Melanocytes",
  "Mural Cells" = "Mural Cells",
  "Smooth Muscle Cells" = "Smooth Muscle Cells",
  "Immune Cells" = "Immune Cells",
  "NK Cells" = "NK Cells",
  "Plasma Cells" = "Plasma Cells",
  "Proliferating Cells" = "Proliferating Cells"
)
# additional granular names found in SjD - appended on top of the base map
celltype_rename_map <- c(celltype_rename_map, c(
  "Neuronal cells" = "Neurons",
  "Lymphatic Endothelial cells" = "Endothelial Cells",
  "CCL14 Venous Endothelial" = "Endothelial Cells",
  "Endothelial Progenitors" = "Endothelial Cells",
  "cDC1" = "Mono/Mac",
  "cDC2" = "Mono/Mac",
  "inflammatory cDC2" = "Mono/Mac",
  "Mo-DC" = "Mono/Mac",
  "M1 Macrophages" = "Mono/Mac",
  "M2 Macrophages" = "Mono/Mac",
  "Basal Epithelial cells" = "Epithelial Cells",
  "Epithelial Progenitor cells" = "Epithelial Cells",
  "Intercalated Ductal Epithelial cells" = "Epithelial Cells",
  "Ionocytes" = "Epithelial Cells",
  "Mucous Acinar Epithelial cells" = "Epithelial Cells",
  "Myoepithelial cells" = "Epithelial Cells",
  "Seromucous Acinar Epithelial cells" = "Epithelial Cells",
  "Serous Acinar Epithelial cells" = "Epithelial Cells",
  "Striated Ductal Epithelial cells" = "Epithelial Cells",
  "Activated Treg cells" = "CD4",
  "Resting Treg cells" = "CD4",
  "Cytotoxic Th cells" = "CD4",
  "DLGAP1+ Th cells" = "CD4",
  "LTB+ Th17 cells" = "CD4",
  "Memory Th cells" = "CD4",
  "Migrating Th17 cells" = "CD4",
  "Naive Tfh cells" = "CD4",
  "Naive Th cells" = "CD4",
  "Secretory Tph cells" = "CD4",
  "Tph17" = "CD4",
  "CD62L+ TPEx" = "CD8",
  "CD8+ Tcm" = "CD8",
  "CD8+ Tem" = "CD8",
  "CD8+ Trm" = "CD8",
  "Exhausted CD8+ T cells" = "CD8",
  "Exhausted IEG CD8+ T cells" = "CD8",
  "Infiltrating CD8+ T cells" = "CD8",
  "Inflammatory CD8+ T cells" = "CD8",
  "Naive CD8+ T cells" = "CD8",
  "Recruited Activated CD8+ T cells" = "CD8",
  "Recruiting Effector CD8+ Trm" = "CD8",
  "GC B cells" = "B Cells",
  "Naive B cells" = "B Cells",
  "Immunofibroblasts" = "Fibroblasts",
  "SPARC+ COL3A1+ Fibroblasts" = "Fibroblasts",
  "Capillary pericytes" = "Mural Cells",
  "Immunopericytes" = "Mural Cells",
  "VSMC" = "Smooth Muscle Cells",
  "gd-T cells" = "Immune Cells",
  "ILCs" = "Immune Cells",
  "CD16+ NK cells" = "NK Cells",
  "CD27+ NK cells" = "NK Cells",
  "CD56+ NK cells" = "NK Cells",
  "IgA Plasma cells" = "Plasma Cells",
  "IgG1 Plasma cells" = "Plasma Cells",
  "IgG3/4 Plasma cells" = "Plasma Cells",
  "IgM Plasma cells" = "Plasma Cells",
  "TNF+ Plasma cells" = "Plasma Cells",
  "Proliferating cells" = "Proliferating Cells"
))
rename_map_df <- data.frame(
  CellType     = names(celltype_rename_map),
  CellType_new = unname(celltype_rename_map),
  stringsAsFactors = FALSE
)
rename_vec <- function(x, map_df) {
  # strip ALL trailing cluster-number groups, e.g.
  # "Fibroblasts_18" -> "Fibroblasts", "Mural Cells_11_2" -> "Mural Cells"
  x <- gsub("(_[0-9]+)+$", "", x, perl = TRUE)
  x <- trimws(gsub("[[:space:] ]+", " ", x, perl = TRUE))
  idx <- match(x, map_df$CellType)
  ifelse(is.na(idx), x, map_df$CellType_new[idx])
}
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
                         "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
                         "Plasma Cells")
diseases_list <- lapply(diseases_list, function(df) {
  df$source <- rename_vec(df$source, rename_map_df)
  df$target <- rename_vec(df$target, rename_map_df)
  df <- df[!(df$source %in% excluded_celltypes | df$target %in% excluded_celltypes), ]
  df
})

#3. build interaction column and combine all diseases (now including Lymph Node)
all_data <- dplyr::bind_rows(lapply(names(diseases_list), function(d) {
  df <- diseases_list[[d]]
  df$disease <- d
  df$interaction <- paste(df$ligand.complex, df$receptor.complex, sep = "^")
  df
}))
stopifnot("`interaction` did not survive bind_rows - check the [WARNING] output above for the offending dataset" =
            "interaction" %in% names(all_data))
cat("Rows per disease:\n")
print(table(all_data$disease))

#4. find interactions present in at least 2 conditions (diseases + Lymph Node)
# using magnitude_rank instead of specificity_rank (which was tied/uninformative)
# lower magnitude_rank = stronger interaction
shared_int <- all_data %>%
  dplyr::group_by(interaction, ligand.complex, receptor.complex) %>%
  dplyr::summarise(
    n_diseases     = dplyr::n_distinct(disease),
    best_magnitude = min(magnitude_rank, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::filter(n_diseases >= 2) %>%
  dplyr::arrange(dplyr::desc(n_diseases), best_magnitude)
stopifnot("`interaction` missing from shared_int" = "interaction" %in% names(shared_int))
cat("Total unique shared interactions:", nrow(shared_int), "\n")

#5. build long format table
# picks the SINGLE best source->target pair per interaction per disease/condition
# (avoids collapsing multiple source/target combos into one string)
long_int <- all_data %>%
  dplyr::filter(interaction %in% shared_int$interaction) %>%
  dplyr::group_by(interaction, ligand.complex, receptor.complex, disease) %>%
  dplyr::slice_min(order_by = magnitude_rank, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::rename(best_magnitude = magnitude_rank) %>%
  dplyr::left_join(
    shared_int %>% dplyr::select(interaction, n_diseases, group_best = best_magnitude),
    by = "interaction"
  )
stopifnot("`interaction` missing after left_join" = "interaction" %in% names(long_int))

long_int <- long_int %>%
  dplyr::arrange(dplyr::desc(n_diseases), group_best, interaction, best_magnitude) %>%
  dplyr::select(interaction, ligand.complex, receptor.complex, n_diseases, best_magnitude,
                disease, source, target)
# put values on a comparable, readable scale: -log10(magnitude_rank)
long_int$score <- -log10(long_int$best_magnitude)

#6. blank out repeated ligand/receptor/n_diseases for readability
# (best_magnitude/score are NOT blanked - they differ per disease)
interaction_key <- paste(long_int$ligand.complex, long_int$receptor.complex, sep = "^")
is_duplicate <- duplicated(interaction_key)
long_int_fmt <- long_int
long_int_fmt$ligand.complex   <- ifelse(is_duplicate, "", long_int$ligand.complex)
long_int_fmt$receptor.complex <- ifelse(is_duplicate, "", long_int$receptor.complex)
long_int_fmt$n_diseases       <- ifelse(is_duplicate, "", as.character(long_int$n_diseases))
long_int_fmt$score            <- round(long_int_fmt$score, 2)

#7. save
dir.create("CCC/LN_Au", recursive = TRUE, showWarnings = FALSE)
write.csv(long_int_fmt, "CCC/LN_Au/shared_interactions_LN_Au.csv", row.names = FALSE)
```

## 10. Venn diagram of shared interactions

```{r}
library(dplyr)
library(ggVennDiagram)
library(ggplot2)

#1. build two sets: interactions present in the autoimmune diseases vs Lymph Node
# uses all_data (the FULL, unfiltered interaction universe from every disease
# file), NOT shared_int/long_int (already restricted to >=2 conditions)
au_interactions <- unique(all_data$interaction[all_data$disease != "Lymph Node"])
ln_interactions <- unique(all_data$interaction[all_data$disease == "Lymph Node"])

venn_list <- list(
  "Lymph Node\nInteractions"      = ln_interactions,
  "Autoimmune Data\nInteractions" = au_interactions
)

cat("Unique interactions in Lymph Node:", length(ln_interactions), "\n")
cat("Unique interactions in AuD (all diseases combined):", length(au_interactions), "\n")
cat("Overlap (shared between Lymph Node and AuD):", length(intersect(ln_interactions, au_interactions)), "\n")

#2. draw the Venn diagram - shows COUNT and PERCENTAGE in every region
p <- ggVennDiagram(
  venn_list,
  label = "both",            # count + percentage together, e.g. "17 (0.5%)"
  label_alpha = 0,
  label_percent_digit = 1,   # 1 decimal place, so small % never rounds to "0%"
  set_size = 4
) +
  scale_fill_gradient(low = "white", high = "#D6604D") +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(20, 40, 20, 40))

#3. save
dir.create("CCC/LN_Au", recursive = TRUE, showWarnings = FALSE)
ggsave("CCC/LN_Au/venn_Diagramme.pdf", plot = p,
       width = 9, height = 7, dpi = 300)
```

## 11. Top shared interactions plot

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(ggtext)

disease_levels <- c("MS", "RA", "SS", "T1D", "SjD", "IPF.U", "IPF.L", "SScILD.U", "SScILD.L", "Lymph Node")

#1. fixed color palette per cell type (same categories used across the project)
celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

#2. top N shared LR pairs, present in 7-9 conditions (absent in at least one)
top_n <- 15
top_lr <- shared_int %>%
  distinct(interaction, ligand.complex, receptor.complex, n_diseases, best_magnitude) %>%
  filter(n_diseases >= 7, n_diseases <= 9) %>%
  arrange(desc(n_diseases), best_magnitude) %>%
  slice_head(n = top_n) %>%
  mutate(lr = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  pull(lr)
cat("Keeping top", length(top_lr), "LR pairs (present in 7-9 conditions)\n")

#3. lr x disease table
lr_disease <- long_int %>%
  mutate(lr = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  filter(lr %in% top_lr) %>%
  mutate(
    lr      = factor(lr, levels = rev(top_lr)),
    disease = factor(disease, levels = disease_levels)
  )
stopifnot(!any(duplicated(paste(lr_disease$lr, lr_disease$disease))))

# separate dataframes with a common column name ("celltype") for the
# color mapping - required so both plots' legends are truly identical
# and can be merged into one
lr_disease_snd <- lr_disease %>% mutate(celltype = source)
lr_disease_rcv <- lr_disease %>% mutate(celltype = target)

#4. shared size scale for BOTH plots - since score = -log10(magnitude_rank),
# lower magnitude_rank = stronger = higher score = bigger dot
score_range <- range(lr_disease$score, na.rm = TRUE)
score_breaks <- seq(score_range[1], score_range[2], length.out = 5)
magnitude_labels <- signif(10^(-score_breaks), 2) # convert back to real magnitude_rank for display
shared_size_scale <- scale_size(
  name   = "magnitude_rank\n(smaller = stronger)",
  range  = c(1, 6),
  limits = score_range,
  breaks = score_breaks,
  labels = as.character(magnitude_labels)
)

#4b. highlight rectangle for the "Lymph Node" column
ln_pos <- match("Lymph Node", disease_levels)
highlight_ln <- annotate(
  "rect",
  xmin = ln_pos - 0.5, xmax = ln_pos + 0.5,
  ymin = -Inf, ymax = Inf,
  fill = "#B2182B", alpha = 0.25
)

#4c. colored x-axis labels - only "Lymph Node" gets a distinct color
axis_label_colors <- ifelse(disease_levels == "Lymph Node",
                             "<span style='color:#B2182B;'>**Lymph Node**</span>",
                             disease_levels)
names(axis_label_colors) <- disease_levels

#4d. dummy row per cell type, placed at an existing x/y so no axis expands
dummy_celltype <- data.frame(
  celltype = factor(names(celltype_cols), levels = names(celltype_cols)),
  disease  = lr_disease_snd$disease[1],
  lr       = lr_disease_snd$lr[1]
)

#5. Senders dotplot - this is the ONLY plot that shows the Cell type
# color legend. LR pair labels (y-axis) explicitly set to black.
p_snd <- ggplot(lr_disease_snd, aes(x = disease, y = lr)) +
  highlight_ln +
  geom_point(aes(size = score, color = celltype),
             show.legend = c(size = TRUE, colour = FALSE)) +
  geom_point(data = dummy_celltype, aes(x = disease, y = lr, color = celltype),
             size = 5, alpha = 0, inherit.aes = FALSE) +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols), drop = FALSE) +
  scale_x_discrete(labels = axis_label_colors) +
  shared_size_scale +
  guides(size  = guide_legend(order = 1),
         color = guide_legend(order = 2, override.aes = list(alpha = 1, size = 5))) +
  theme_bw() +
  theme(axis.text.x = ggtext::element_markdown(angle = 90, hjust = 1),
        axis.text.y = element_text(size = 10, face = "bold", color = "black")) +
  labs(x = "Disease", y = "Ligand-Receptor pair",
       color = "Cell type",
       title = "Senders")

#6. Receivers dotplot - color legend suppressed (guide="none"), since
# Senders already shows the identical Cell type legend. Same explicit
# black LR pair label color as Senders.
p_rcv <- ggplot(lr_disease_rcv, aes(x = disease, y = lr,
                                     size = score, color = celltype)) +
  highlight_ln +
  geom_point() +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols),
                      drop = FALSE, guide = "none") +
  scale_x_discrete(labels = axis_label_colors) +
  shared_size_scale +
  guides(size = guide_legend(order = 1)) +
  theme_bw() +
  theme(axis.text.x = ggtext::element_markdown(angle = 90, hjust = 1),
        axis.text.y = element_text(size = 10, face = "bold", color = "black")) +
  labs(x = "Disease", y = "Ligand-Receptor pair",
       color = "Cell type",
       title = "Receivers")

#7. combine side by side, ONE merged legend
final_plot <- (p_snd | p_rcv) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

dir.create("CCC/LN_Au", recursive = TRUE, showWarnings = FALSE)
ggsave("CCC/LN_Au/top_interactions_LN_au.pdf",
       final_plot, width = 18, height = max(10, top_n * 0.3), limitsize = FALSE)
```

## 12. TLS and GC-associated interactions table

```{r}
library(dplyr)

#1. define gene sets (same TLS marker groups used throughout the project)
germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
                      "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")
inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
                   "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
                   "ICAM1", "VCAM1", "SELE")
lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12",
                          "CXCR5", "CCR7", "CXCR4")
hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")
fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
         "CXCL13", "CD54", "ICAM1", "FCER2")

all_genes <- unique(c(germinal_center, inflammation, lymphoid_chemokines, hev, fdc))

#2. exclude these cell types from source/target before building the table
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
                         "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
                         "Plasma Cells")
all_data <- all_data[!(all_data$source %in% excluded_celltypes |
                        all_data$target %in% excluded_celltypes), ]

#3. filter from all_data directly (not long_int), so we can pick the
# SINGLE best source->target pair per interaction per disease
shared_tls_gc <- all_data %>%
  filter(
    ligand.complex %in% all_genes |
    receptor.complex %in% all_genes
  ) %>%
  filter(interaction %in% shared_int$interaction) %>%
  group_by(interaction, ligand.complex, receptor.complex, disease) %>%
  slice_min(order_by = magnitude_rank, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(shared_int %>% select(interaction, n_diseases), by = "interaction") %>%
  mutate(category = case_when(
    ligand.complex %in% germinal_center | receptor.complex %in% germinal_center ~ "Germinal Center",
    ligand.complex %in% inflammation | receptor.complex %in% inflammation ~ "Inflammation",
    ligand.complex %in% lymphoid_chemokines | receptor.complex %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
    ligand.complex %in% hev | receptor.complex %in% hev ~ "HEV",
    ligand.complex %in% fdc | receptor.complex %in% fdc ~ "FDC",
    TRUE ~ "Other"
  )) %>%
  dplyr::rename(best_magnitude = magnitude_rank) %>%
  mutate(score = -log10(best_magnitude)) %>%
  select(ligand.complex, receptor.complex, n_diseases, best_magnitude, score,
         disease, source, target, category)

cat("Total shared TLS/GC interactions:", nrow(shared_tls_gc), "\n")
print(table(shared_tls_gc$category))
print(table(shared_tls_gc$disease))

#4. format - blank repeated ligand/receptor for readability
interaction_key <- paste(shared_tls_gc$ligand.complex, shared_tls_gc$receptor.complex, sep = "^")
is_duplicate <- duplicated(interaction_key)

shared_tls_gc_fmt <- shared_tls_gc
shared_tls_gc_fmt$ligand.complex   <- ifelse(is_duplicate, "", shared_tls_gc$ligand.complex)
shared_tls_gc_fmt$receptor.complex <- ifelse(is_duplicate, "", shared_tls_gc$receptor.complex)
shared_tls_gc_fmt$n_diseases       <- ifelse(is_duplicate, "", as.character(shared_tls_gc$n_diseases))
shared_tls_gc_fmt$category         <- ifelse(is_duplicate, "", shared_tls_gc$category)

#5. save
dir.create("CCC/LN_Au", recursive = TRUE, showWarnings = FALSE)
write.csv(shared_tls_gc_fmt, "CCC/LN_Au/shared_TLS_GC_interactions.csv", row.names = FALSE)
```

## 13. TLS and GC interactions heatmap

```{r}
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(gridtext)

# add extra padding
ht_opt$TITLE_PADDING <- unit(c(6, 6), "mm")

#1. keep only the TOP N LR pairs per category (ranked by best score =
# lowest best_magnitude across diseases) - this is what cuts down the
# oversized Inflammation category specifically
top_n_per_category <- 8

lr_best_score <- shared_tls_gc %>%
  mutate(lr_pair = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  group_by(lr_pair, category) %>%
  summarise(best_score = max(score), .groups = "drop")

lr_order <- lr_best_score %>%
  group_by(category) %>%
  arrange(desc(best_score), .by_group = TRUE) %>%
  slice_head(n = top_n_per_category) %>%
  ungroup()

category_order <- c("FDC", "Germinal Center", "HEV", "Inflammation", "Lymphoid Chemokines")
lr_order$category <- factor(lr_order$category, levels = category_order)
lr_order <- lr_order %>% arrange(category)

lr_pairs <- lr_order$lr_pair
lr_category <- setNames(as.character(lr_order$category), lr_order$lr_pair)

cat("LR pairs kept per category:\n")
print(table(lr_order$category))

shared_tls_gc$lr_pair <- paste(shared_tls_gc$ligand.complex, shared_tls_gc$receptor.complex, sep = "^")
shared_tls_gc_top <- shared_tls_gc %>% filter(lr_pair %in% lr_pairs)

disease_levels <- c("MS", "RA", "SS", "T1D", "SjD", "IPF.U", "IPF.L", "SScILD.U", "SScILD.L", "Lymph Node")
diseases_present <- intersect(disease_levels, unique(shared_tls_gc_top$disease))

# diagnostic: confirm no duplicate disease names, which would make that
# column render wider than the others
stopifnot(!any(duplicated(diseases_present)))

#1b. build rich-text row labels: bold, colored text for whichever side
# (ligand or receptor) is an actual TLS marker gene; the other side stays
# plain. NOTE: background-color is not reliably rendered by gridtext's
# markdown parser, so font color + bold is used instead (this does render)
make_rich_label <- function(lr_pair_str) {
  parts <- strsplit(lr_pair_str, "\\^")[[1]]
  styled <- ifelse(
    parts %in% all_genes,
    paste0("**<span style='color:black;'>", parts, "</span>**"),
    parts
  )
  paste(styled, collapse = " ^ ")
}
rich_row_labels <- vapply(lr_pairs, make_rich_label, character(1))

#1c. rich-text column labels: bold + colored for "Lymph Node", plain
# for every disease column
make_col_label <- function(col_name) {
  if (col_name == "Lymph Node") {
    paste0("**<span style='color:#B2182B;'>", col_name, "</span>**")
  } else {
    col_name
  }
}
rich_col_labels <- vapply(diseases_present, make_col_label, character(1))

#1d. shared sizing - both panels use these EXACT same values, so there
# is no possibility of them ending up visually different sizes
panel_width  <- unit(length(diseases_present) * 12, "mm")
panel_height <- unit(length(lr_pairs) * 6, "mm")
row_label_fontsize <- 9

#2. build matrices: rows = LR pairs (top-N per category), columns = diseases
source_mat <- matrix(NA, nrow = length(lr_pairs), ncol = length(diseases_present),
                      dimnames = list(lr_pairs, diseases_present))
target_mat <- matrix(NA, nrow = length(lr_pairs), ncol = length(diseases_present),
                      dimnames = list(lr_pairs, diseases_present))
score_mat  <- matrix(NA, nrow = length(lr_pairs), ncol = length(diseases_present),
                      dimnames = list(lr_pairs, diseases_present))

for (i in seq_len(nrow(shared_tls_gc_top))) {
  r <- shared_tls_gc_top[i, ]
  if (r$disease %in% diseases_present) {
    source_mat[r$lr_pair, r$disease] <- r$source
    target_mat[r$lr_pair, r$disease] <- r$target
    score_mat[r$lr_pair, r$disease]  <- r$score
  }
}

#3. colors
category_cols <- c(
  "Germinal Center"      = "#2166AC",
  "Inflammation"         = "#D6604D",
  "Lymphoid Chemokines"  = "#35978F",
  "HEV"                  = "#8073AC",
  "FDC"                  = "#E08214"
)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

#4. normalize score for dot size
min_score <- min(score_mat, na.rm = TRUE)
max_score <- max(score_mat, na.rm = TRUE)
normalize_score <- function(s) {
  0.15 + 0.85 * (s - min_score) / (max_score - min_score)
}

row_split_vec <- factor(lr_category[lr_pairs], levels = category_order)

#5. Source-colored plot (LR pair left, category color+text on the RIGHT)
# row_labels uses gt_render() for rich-text (bold/colored TLS gene)
cell_fun_source <- function(j, i, x, y, w, h, fill) {
  # highlight the Lymph Node column background before drawing the dot
  if (trimws(diseases_present[j]) == "Lymph Node") {
    grid::grid.rect(x = x, y = y, width = w, height = h,
                     gp = grid::gpar(fill = "#B2182B", alpha = 0.35, col = NA))
  }
  ct <- source_mat[i, j]
  if (is.na(ct)) return(invisible(NULL))
  s <- score_mat[i, j]
  grid::grid.circle(
    x = x, y = y,
    r = normalize_score(s) * 0.42 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = celltype_cols[ct], col = celltype_cols[ct])
  )
}

ht_source <- Heatmap(
  matrix(0, nrow = length(lr_pairs), ncol = length(diseases_present),
         dimnames = list(lr_pairs, diseases_present)),
  name = "Source",
  col = colorRamp2(c(0, 1), c("white", "white")),
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun_source,
  show_heatmap_legend = FALSE,
  row_labels = gt_render(rich_row_labels, gp = gpar(fontsize = row_label_fontsize)),
  row_names_side = "left",
  column_labels = gt_render(rich_col_labels, gp = gpar(fontsize = 11)),
  column_names_rot = 45,
  column_names_side = "bottom",
  column_title = "Senders",
  column_title_gp = gpar(fontsize = 13, fontface = "bold"),
  border = TRUE,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = row_split_vec,
  row_title = rep("", nlevels(row_split_vec)),
  row_gap = unit(3, "mm"),
  width = panel_width,
  height = panel_height
)

#6. Target-colored plot (category color+text on the LEFT, LR pair on default right side)
cell_fun_target <- function(j, i, x, y, w, h, fill) {
  # highlight the Lymph Node column background before drawing the dot
  if (trimws(diseases_present[j]) == "Lymph Node") {
    grid::grid.rect(x = x, y = y, width = w, height = h,
                     gp = grid::gpar(fill = "#B2182B", alpha = 0.35, col = NA))
  }
  ct <- target_mat[i, j]
  if (is.na(ct)) return(invisible(NULL))
  s <- score_mat[i, j]
  grid::grid.circle(
    x = x, y = y,
    r = normalize_score(s) * 0.42 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = celltype_cols[ct], col = celltype_cols[ct])
  )
}

ht_target <- Heatmap(
  matrix(0, nrow = length(lr_pairs), ncol = length(diseases_present),
         dimnames = list(lr_pairs, diseases_present)),
  name = "Target",
  col = colorRamp2(c(0, 1), c("white", "white")),
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun_target,
  show_heatmap_legend = FALSE,
  row_labels = gt_render(rich_row_labels, gp = gpar(fontsize = row_label_fontsize)),
  row_names_side = "right",
  column_labels = gt_render(rich_col_labels, gp = gpar(fontsize = 11)),
  column_names_rot = 45,
  column_names_side = "bottom",
  column_title = "Receivers",
  column_title_gp = gpar(fontsize = 13, fontface = "bold"),
  border = TRUE,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = row_split_vec,
  row_title = rep("", nlevels(row_split_vec)), # keep the gap for alignment, hide the duplicate text
  row_gap = unit(3, "mm"),
  width = panel_width,
  height = panel_height
)

#7. legends - each panel gets its own full copy
all_present_ct <- intersect(names(celltype_cols),
                             unique(c(na.omit(as.vector(source_mat)), na.omit(as.vector(target_mat)))))

make_celltype_legend <- function() {
  Legend(
    labels = all_present_ct,
    legend_gp = gpar(fill = celltype_cols[all_present_ct]),
    title = "Cell type",
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 9)
  )
}

size_breaks <- seq(min_score, max_score, length.out = 5)
magnitude_breaks <- signif(10^(-size_breaks), 2)

make_size_legend <- function() {
  Legend(
    labels = as.character(magnitude_breaks),
    title = "magnitude_rank\n(smaller = stronger)",
    type = "points",
    pch = 16,
    size = unit(normalize_score(size_breaks) * 0.42 * 4, "mm"),
    legend_gp = gpar(col = "grey40", fill = "grey70"),
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 9)
  )
}

# legend explaining the highlighted gene
make_highlight_legend <- function() {
  Legend(
    labels = "TLS marker gene",
    type = "points",
    pch = 15,
    legend_gp = gpar(col = "black", fontsize = 12),
    title = "",
    labels_gp = gpar(fontsize = 9, col = "black", fontface = "bold")
  )
}

# legend explaining the category color bands (FDC, Germinal Center, HEV,
# Inflammation, Lymphoid Chemokines) - replaces trying to center text
# labels directly in the plot, which ComplexHeatmap can't do reliably
make_category_legend <- function() {
  Legend(
    labels = category_order,
    legend_gp = gpar(fill = category_cols[category_order]),
    title = "TLS category",
    title_gp = gpar(fontsize = 10, fontface = "bold"),
    labels_gp = gpar(fontsize = 9)
  )
}

#8. combine: Senders -> colored band -> colored band -> Receivers.
# Category names are no longer drawn in the plot itself (ComplexHeatmap
# can't reliably center row_title text between two panels) - instead,
# the meaning of each color is explained in the "TLS category" legend
band_right <- rowAnnotation(
  Category = row_split_vec,
  col = list(Category = category_cols),
  annotation_width = unit(6, "mm"),
  show_annotation_name = FALSE,
  show_legend = FALSE
)

band_left <- rowAnnotation(
  Category = row_split_vec,
  col = list(Category = category_cols),
  annotation_width = unit(6, "mm"),
  show_annotation_name = FALSE,
  show_legend = FALSE
)

ht_list <- ht_source + band_right + band_left + ht_target

plot_h <- max(7, length(lr_pairs) * 0.3)
plot_w <- 20

dir.create("CCC/LN_Au", recursive = TRUE, showWarnings = FALSE)

pdf("CCC/LN_Au/TLS_GC_interactions.pdf", width = plot_w, height = plot_h)
draw(ht_list,
     annotation_legend_list = list(make_celltype_legend(), make_size_legend(),
                                    make_highlight_legend(), make_category_legend()),
     heatmap_legend_side = "right", annotation_legend_side = "right",
     ht_gap = unit(c(1, 6, 1), "mm"))
dev.off()
```

## 14. Save final Seurat object

```{r}
saveRDS(LN_Au, file = "data/process/SeuratObjects/LN_Au/LN_Au_process.rds")
LN_Au <- readRDS("data/process/SeuratObjects/LN_Au/LN_Au_process.rds")
```
