---
title: "Sjogren's Disease - Processing and Annotation"
output: html_notebook
---

<!-- Step 2: Seurat processing for Sjogren's Disease - cleans and annotates
     the CD45-negative and CD45-positive compartments (annotated upstream
     by Tomas Gomes) separately, then merges them into one combined Seurat
     object. The cell-cell communication (CCC) analysis is a separate step,
     in SjD.sce.Rmd. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

# CD45-negative compartment

## 2. Load CD45-negative object

```{r}
cd45neg <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45neg_annotated_clean.RDS")
DefaultAssay(cd45neg)
Assays(cd45neg)

View(cd45neg@meta.data)
```

## 3. Clean metadata and annotation labels

```{r}
keep_cols <- c(
  "orig.ident",
  "sample_name",
  "nCount_RNA",
  "nFeature_RNA",
  "seurat_clusters",
  "major_annot",
  "major_annot_joint",
  "specific_annot",
  "specific_annot_subcl"
)
cd45neg@meta.data <- cd45neg@meta.data[, keep_cols]

colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "major_annot"] <- "Major_Celltype"
cd45neg$orig.ident <- "SjD"
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "sample_name"] <- "sample"

# clean major annotation labels
cd45neg@meta.data$Major_Celltype <- dplyr::recode(
  cd45neg@meta.data$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Neuronal_cells" = "Neurons",
  "Endothelial_cells" = "Endothelial Cells",
  "Immune_cells" = "Immune Cells",
  "Mural" = "Mural Cells",
  "Acinal_cells" = "Acinar Cells",
  "Ductal_cells" = "Ductal Cells"
)

# coarser "General" annotation grouping epithelial subtypes together
cd45neg@meta.data$General <- dplyr::recode(
  cd45neg@meta.data$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Neurons" = "Neurons",
  "Endothelial Cells" = "Endothelial Cells",
  "Immune Cells" = "Immune Cells",
  "Mural Cells" = "Mural Cells",
  "Acinar Cells" = "Epithelial Cells",
  "Ductal Cells" = "Epithelial Cells"
)

# remove the single mis-annotated cell and drop the now-empty factor level
cd45neg <- subset(cd45neg, subset = Major_Celltype != "Immune Cells")
cd45neg$Major_Celltype <- droplevels(as.factor(cd45neg$Major_Celltype))
table(cd45neg$Major_Celltype)
```

cd45neg carries two annotation levels: `Major_Celltype` (broad) and `specific_annot` (fine-grained). Both are explored below.

## 4. UMAP by major cell type

```{r}
umap_df <- as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")
umap_df$Major_Celltype <- cd45neg$Major_Celltype

centers <- aggregate(
  cbind(UMAP_1, UMAP_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

celltype_cols_sjd <- c(
  "Acinar Cells"      = "#1f78b4",
  "Ductal Cells"      = "#66a61e",
  "Endothelial Cells" = "#2ca02c",
  "Fibroblasts"       = "#D73027",
  "Mural Cells"       = "#984EA3",
  "Neurons"           = "#a6761d"
)

library(ggrepel)
suppressWarnings({
  fig_sjd <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Major_Celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols_sjd) +
    geom_label_repel(
      data = centers,
      aes(x = UMAP_1, y = UMAP_2, label = Major_Celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 3.5,
      box.padding = 0.4,
      label.size = 0.40,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "UMAP 1", y = "UMAP 2")
})

fig_sjd
ggsave(filename = "output/SjD/cd45neg_cellType.png", plot = fig_sjd, width = 12, height = 8)
```

## 5. Major cell type marker heatmap

```{r}
markers <- c(
  "PTPRC", "MZB1", "PRDM1", "CCL19", "VCAM1", "CXCL10", "TNFSF13B", "CXCL9", "ICAM1", "SPARC",
  "COL3A1", "DCN", "LUM", "COL14A1", "DPT", "PECAM1", "CD34", "PDGFRB", "DLL4", "EFNB2",
  "SOX17", "SOX18", "EPHB4", "ACKR1", "NR2F2", "ZNF385D", "MEOX1", "LYVE1", "PROX1", "HES1",
  "ACTA2", "MCAM", "RGS5", "ABCC9", "CCL21", "TNC", "CD36", "MYH11", "NGFR", "CDH3",
  "TP63", "DLK1", "CFTR", "ASCL3", "FOXI1", "AQP5", "KRT5", "KRT14", "SOX2", "MUC19",
  "LPO", "FGFR2", "MUC1", "MUC5B", "ANO1", "KIT", "PIP"
)

DefaultAssay(cd45neg) <- "SCT"
markers <- markers[markers %in% rownames(cd45neg)]

Idents(cd45neg) <- "Major_Celltype"
expr_mat <- LayerData(cd45neg, assay = "SCT", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = markers,
  assays = "SCT",
  layer = "data",
  return.seurat = FALSE
)$SCT

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42) # row_km uses k-means (stochastic), seed for reproducibility
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD/cd45neg", recursive = TRUE, showWarnings = FALSE)
png("output/SjD/Major_Celltype_heatmap.png",
  width = 3200, height = 3500, res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 6. Correlation between major cell types

```{r}
agg.exp <- AggregateExpression(cd45neg, assays = "RNA", layers = "counts", group.by = "Major_Celltype")$RNA
dim(agg.exp)
colnames(agg.exp)

agg.exp.mat <- as.matrix(agg.exp)
agg.exp.mat <- log1p(agg.exp.mat)

cor_mat <- cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)
cor_mat[is.na(cor_mat)] <- 0

# tiny jitter to avoid a degenerate (all-equal) correlation matrix
set.seed(1)
cor_mat <- cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)

png(
  filename = "output/SjD/Major_correlation.png",
  width = 3000,
  height = 2600,
  res = 300
)
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()
```

## 7. Specific annotation marker heatmap

```{r}
genes <- c(
  "PTPRC", "MZB1", "PRDM1", "CCL19", "VCAM1", "CXCL10", "TNFSF13B", "CXCL9", "ICAM1", "SPARC",
  "COL3A1", "DCN", "LUM", "COL14A1", "DPT", "PECAM1", "CD34", "PDGFRB", "DLL4", "EFNB2",
  "SOX17", "SOX18", "EPHB4", "ACKR1", "NR2F2", "ZNF385D", "MEOX1", "LYVE1", "PROX1", "HES1",
  "ACTA2", "MCAM", "RGS5", "ABCC9", "CCL21", "TNC", "CD36", "MYH11", "NGFR", "CDH3",
  "TP63", "DLK1", "CFTR", "ASCL3", "FOXI1", "AQP5", "KRT5", "KRT14", "SOX2", "MUC19",
  "LPO", "FGFR2", "MUC1", "MUC5B", "ANO1", "KIT", "PIP"
)

cd45neg[["RNA"]] <- JoinLayers(cd45neg[["RNA"]])

if (!"data" %in% Layers(cd45neg[["RNA"]])) {
  cd45neg <- NormalizeData(cd45neg, assay = "RNA")
}

genes <- genes[genes %in% rownames(cd45neg[["RNA"]])]

Idents(cd45neg) <- "specific_annot"
cd45neg$specific_annot <- droplevels(as.factor(cd45neg$specific_annot))
Idents(cd45neg) <- "specific_annot"

expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD", recursive = TRUE, showWarnings = FALSE)
png(
  "output/SjD/Specific_Celltype_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 8. TLS marker heatmap

```{r}
tls <- c(
  "AID", "AICDA", "BCL6", "PAX5", "MYC", "IRF8", "CD40LG", "ICOS", "S1PR2", "CXCR4", "CXCR5", "CCR7",
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", "IFNG", "TGFB",
  "CCL2", "CCL3", "CCL4", "CCL5", "CXCL12", "CXCL13", "CCL19", "CCL21",
  "ICAM1", "VCAM1", "SELE", "CD54", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C", "CD21", "CD23", "CR1", "CR2", "CD35",
  "FDCSP", "FCER2", "NTAN1"
)

genes <- tls[tls %in% rownames(cd45neg)]

cd45neg[["RNA"]] <- JoinLayers(cd45neg[["RNA"]])
Idents(cd45neg) <- "specific_annot"

expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png(
  "output/SjD/cd45neg/TLS_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 9. UMAP by specific annotation

```{r}
Idents(cd45neg) <- "specific_annot"
cd45neg$celltype <- Idents(cd45neg)

celltype_cols <- c(
  "VSMC"                                 = "#D73027",
  "Capillary pericytes"                  = "#4575B4",
  "Immunopericytes"                      = "#1B9E77",
  "Neuronal cells"                       = "#762A83",
  "Endothelial Progenitors"              = "#FDAE61",
  "Arterial Endothelial"                 = "#E7298A",
  "Lymphatic Endothelial cells"          = "#66A61E",
  "CCL14 Venous Endothelial"             = "#E6AB02",
  "Venous Endothelial"                   = "#A6761D",
  "Fibroblasts_3"                        = "#666666",
  "Fibroblasts_4"                        = "#1F78B4",
  "Serous Acinar Epithelial cells"       = "#B2DF8A",
  "Intercalated Ductal Epithelial cells" = "#33A02C",
  "Seromucous Acinar Epithelial cells"   = "#FB9A99",
  "Striated Ductal Epithelial cells"     = "#E31A1C",
  "Basal Epithelial cells"               = "#FDBF6F",
  "Mucous Acinar Epithelial cells"       = "#FF7F00",
  "Epithelial Progenitor cells"          = "#CAB2D6",
  "Ionocytes"                            = "#6A3D9A",
  "Myoepithelial cells"                  = "#B15928",
  "Fibroblasts_1"                        = "#8DA0CB",
  "Fibroblasts_2"                        = "#66C2A5",
  "Fibroblasts_5"                        = "#FC8D62",
  "SPARC+ COL3A1 + Fibroblasts"          = "#E78AC3"
)

setdiff(unique(cd45neg$specific_annot), names(celltype_cols))
Reductions(cd45neg)

umap_df <- as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$specific_annot <- cd45neg$specific_annot

centers <- aggregate(cbind(umap_1, umap_2) ~ specific_annot, data = umap_df, FUN = median)

library(ggrepel)
library(ggplot2)

fig <- ggplot(umap_df, aes(umap_1, umap_2, color = specific_annot)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = specific_annot),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")

fig

dir.create("output/SjD/cd45neg", recursive = TRUE, showWarnings = FALSE)
ggsave("output/SjD/SjD_specific_UMAP.png", plot = fig, width = 12, height = 8)
```

## 10. Save CD45-negative object

```{r}
saveRDS(cd45neg, file = "data/process/SeuratObjects/SjD/cd45(-).rds")

# reload to confirm the save
cd45neg <- readRDS("data/process/SeuratObjects/SjD/cd45(-).rds")
View(cd45neg@meta.data)
```

# CD45-positive compartment

## 11. Load CD45-positive object

```{r}
cd45pos <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45pos_annotated_clean.RDS")

DefaultAssay(cd45pos)
colnames(cd45pos@meta.data)
View(cd45pos@meta.data)
```

## 12. Clean metadata and annotation labels

```{r}
keep_cols <- c(
  "orig.ident",
  "sample_name",
  "nCount_RNA",
  "nFeature_RNA",
  "seurat_clusters",
  "major_annot",
  "major_annot_cl",
  "major_annot_joint",
  "predicted_labels",
  "majority_voting",
  "specific_annot"
)
cd45pos@meta.data <- cd45pos@meta.data[, keep_cols]

colnames(cd45pos@meta.data)[colnames(cd45pos@meta.data) == "major_annot_joint"] <- "Major_Celltype"
cd45pos$orig.ident <- "SjD"
colnames(cd45pos@meta.data)[colnames(cd45pos@meta.data) == "sample_name"] <- "sample"

# clean major annotation labels
cd45pos@meta.data$Major_Celltype <- dplyr::recode(
  cd45pos@meta.data$Major_Celltype,
  "Mono/Mac"            = "Mono/Mac",
  "gd/NK"               = "NK Cells",
  "B cells"             = "B Cells",
  "Proliferating cells" = "Proliferating Cells",
  "CD8"                 = "CD8",
  "Plasma cells"        = "Plasma Cells",
  "CD4"                 = "CD4"
)

# coarser "General" annotation
cd45pos@meta.data$General <- dplyr::recode(
  cd45pos@meta.data$Major_Celltype,
  "Mono/Mac"             = "Mono/Mac",
  "gd/NK"                = "NK Cells",
  "B Cells"              = "B Cells",
  "Proliferating Cells"  = "Proliferating Cells",
  "CD8"                  = "CD8",
  "Plasma Cells"         = "Plasma Cells",
  "CD4"                  = "CD4"
)

unique(cd45pos$Major_Celltype)
```

## 13. Major cell type marker heatmap

```{r}
markers <- c(
  "ZG16B", "IGHM", "IGHD", "CCR7", "GPR183", "TEX9", "MYC", "KLF6", "MIR155HG", "ITGB1",
  "FOXP1", "BACH2", "CLEC9A", "WDFY4", "IRF8", "CLEC4C", "THBD", "BATF3", "CD1C",
  "FCER1A", "FCGR3A", "CD14", "CX3CR1", "CD163", "CD36", "FCGR1A", "MERTK", "CD80",
  "ZEB2", "LAMP3", "MRC1", "SIRPA", "CLEC10A", "C1QA", "IL1B", "IGHA1", "IGHG1",
  "IGHG3", "IGHG4", "TNF", "NCAM1", "NKG7", "CD27", "TRDC", "KLRG1", "FXYD7", "KIT",
  "AHR", "FGFBP2", "FCRL6", "CD8A", "CD8B", "MYB", "SELL", "TOX", "IKZF2", "HAVCR2",
  "CTLA4", "ZNF683", "ITGAE", "CXCR6", "CCL4L2", "CCL4", "CCL3", "PRDM1", "IFNG",
  "CD69", "CCR6", "XCL1", "XCL2", "CCL20", "CLEC2B", "NR4A1", "HSPA6", "BAG3",
  "GZMH", "LGALS1", "LEF1", "TCF7", "ANXA2", "CD44", "GZMA", "CCL5", "DLGAP1",
  "RORA", "RORC", "IL17A", "FOXP3", "CCR8", "TIGIT", "TNFRSF9", "IL7R", "CXCR5",
  "BCL6", "SH2D1A", "MAF", "CXCL13", "IL21", "IL2RG", "ITGA4", "ITGB7"
)

markers <- markers[markers %in% rownames(cd45pos)]
Idents(cd45pos) <- "Major_Celltype"

expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45pos,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png("output/SjD/cd45pos/Major_Celltype_heatmap.png",
  width = 3200, height = 3500, res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 14. UMAP by major cell type

```{r}
Idents(cd45pos) <- "Major_Celltype"
cd45pos$Major_Celltype <- Idents(cd45pos)

celltype_cols <- c(
  "B Cells"             = "magenta",
  "CD8"                 = "#FC8D62",
  "CD4"                 = "#1B9E77",
  "NK Cells"            = "darkblue",
  "Plasma Cells"        = "#FDAE61",
  "Proliferating Cells" = "yellow",
  "Mono/Mac"            = "red3"
)

names(cd45pos@reductions)
Reductions(cd45pos)
umap_df <- as.data.frame(Embeddings(cd45pos, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$Major_Celltype <- cd45pos$Major_Celltype

centers <- aggregate(cbind(umap_1, umap_2) ~ Major_Celltype,
  data = umap_df, FUN = median
)

library(ggplot2)
library(ggrepel)

fig_major <- ggplot(umap_df, aes(umap_1, umap_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")

fig_major

ggsave("output/SjD/cd45pos/Major_heatmap.png", plot = fig_major, width = 12, height = 8)
```

## 15. Correlation between major cell types

```{r}
agg.exp <- AggregateExpression(cd45pos, assays = "RNA", layers = "counts", group.by = "Major_Celltype")$RNA
dim(agg.exp)
colnames(agg.exp)

agg.exp.mat <- as.matrix(agg.exp)
agg.exp.mat <- log1p(agg.exp.mat)

cor_mat <- cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)
cor_mat[is.na(cor_mat)] <- 0

set.seed(1)
cor_mat <- cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)

png(
  filename = "output/SjD/cd45pos/Major_correlation.png",
  width = 3000,
  height = 2600,
  res = 300
)
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()
```

## 16. Specific annotation marker heatmap

```{r}
markers <- c(
  "PTPRC", "MZB1", "PRDM1", "CCL19", "VCAM1", "CXCL10", "TNFSF13B", "CXCL9", "ICAM1", "SPARC",
  "COL3A1", "DCN", "LUM", "COL14A1", "DPT", "PECAM1", "CD34", "PDGFRB", "DLL4", "EFNB2",
  "SOX17", "SOX18", "EPHB4", "ACKR1", "NR2F2", "ZNF385D", "MEOX1", "LYVE1", "PROX1", "HES1",
  "ACTA2", "MCAM", "RGS5", "ABCC9", "CCL21", "TNC", "CD36", "MYH11", "NGFR", "CDH3",
  "TP63", "DLK1", "CFTR", "ASCL3", "FOXI1", "AQP5", "KRT5", "KRT14", "SOX2", "MUC19",
  "LPO", "FGFR2", "MUC1", "MUC5B", "ANO1", "KIT", "PIP"
)

cd45pos[["RNA"]] <- JoinLayers(cd45pos[["RNA"]])

if (!"data" %in% Layers(cd45pos[["RNA"]])) {
  cd45pos <- NormalizeData(cd45pos, assay = "RNA")
}

markers <- markers[markers %in% rownames(cd45pos[["RNA"]])]

Idents(cd45pos) <- "specific_annot"
cd45pos$specific_annot <- droplevels(as.factor(cd45pos$specific_annot))
Idents(cd45pos) <- "specific_annot"

expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45pos,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[markers, common, drop = FALSE])
percent_mat <- percent_mat[markers, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD/cd45pos", recursive = TRUE, showWarnings = FALSE)
png(
  "output/SjD/cd45pos/Specific_Celltype_Umap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 17. UMAP by specific annotation

```{r}
Idents(cd45pos) <- "specific_annot"
cd45pos$celltype <- Idents(cd45pos)

celltype_cols <- c(
  "Activated Treg cells"       = "#E76F51",
  "CD16+ NK cells"              = "#F4A261",
  "CD27+ NK cells"              = "#E9C46A",
  "CD56+ NK cells"              = "#D4A373",
  "CD62L+ TPEx"                 = "#CDB4DB",
  "CD8+ Tcm"                    = "#90DBF4",
  "CD8+ Tem"                    = "#48CAE4",
  "CD8+ Trm"                    = "#0096C7",
  "cDC1"                        = "#A3B18A",
  "cDC2"                        = "#588157",
  "Cytotoxic Th cells"          = "#6A994E",
  "DLGAP1+ Th cells"            = "#A7C957",
  "Exhausted CD8+ T cells"      = "#6D597A",
  "Exhausted IEG CD8+ T cells"  = "#B56576",
  "GC B cells_1"                = "#4CC9F0",
  "GC B cells_2"                = "#4895EF",
  "GC B cells_3"                = "#4361EE",
  "GC B cells_4"                = "#3A0CA3",
  "gd-T cells"                  = "#2A9D8F",
  "IgA Plasma cells"            = "#F72585",
  "IgG1 Plasma cells"           = "#FF7096",
  "IgG3/4 Plasma cells"         = "#FF8FAB",
  "IgM Plasma cells"            = "#FFC2D1",
  "ILCs"                        = "#52B788",
  "Infiltrating CD8+ T cells"   = "#1B9AAA",
  "Inflammatory CD8+ T cells"   = "#0077B6",
  "inflammatory cDC2"           = "#6C757D",
  "LAMP3+ cDCs"                 = "#ADB5BD",
  "LTB+ Th17 cells"             = "#9D4EDD",
  "M1 Macrophages"              = "#E63946",
  "M2 Macrophages"              = "#A8DADC",
  "Memory Th cells"             = "#457B9D",
  "Migrating Th17 cells"        = "#5E60CE",
  "Mo-DC"                       = "#7400B8",
  "Monocytes"                   = "#B5179E",
  "Naive B cells"               = "#F1FAEE",
  "Naive CD8+ T cells"          = "#8ECAE6",
  "Naive Tfh cells"             = "#219EBC",
  "Naive Th cells"              = "#023047",
  "Proliferating cells"         = "#FFBE0B",
  "Recruited Activated CD8-"    = "#FB5607",
  "Recruiting Effector CD8+"    = "#FF006E",
  "Resting Treg cells"          = "#8338EC",
  "Secretory Tph cells"         = "#3A86FF",
  "TNF+ Plasma cells"           = "#D00000",
  "Tph17"                       = "#9B2226"
)

names(cd45pos@reductions)
Reductions(cd45pos)
umap_df <- as.data.frame(Embeddings(cd45pos, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$specific_annot <- cd45pos$specific_annot

centers <- aggregate(cbind(umap_1, umap_2) ~ specific_annot, data = umap_df, FUN = median)

library(ggrepel)
library(ggplot2)

fig <- ggplot(umap_df, aes(umap_1, umap_2, color = specific_annot)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = specific_annot),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")

fig

ggsave("output/SjD/cd45pos/Specific_UMAP.png", plot = fig, width = 12, height = 8)
```

## 18. TLS marker heatmap

```{r}
tls <- c(
  "AID", "AICDA", "BCL6", "PAX5", "MYC", "IRF8", "CD40LG", "ICOS", "S1PR2", "CXCR4", "CXCR5", "CCR7",
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", "IFNG", "TGFB",
  "CCL2", "CCL3", "CCL4", "CCL5", "CXCL12", "CXCL13", "CCL19", "CCL21",
  "ICAM1", "VCAM1", "SELE", "CD54", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C", "CD21", "CD23", "CR1", "CR2", "CD35",
  "FDCSP", "FCER2", "NTAN1"
)

genes <- tls[tls %in% rownames(cd45pos)]

cd45pos[["RNA"]] <- JoinLayers(cd45pos[["RNA"]])
Idents(cd45pos) <- "specific_annot"

expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45pos,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png(
  "output/SjD/cd45pos/TLS_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 19. Correlation between specific annotations

```{r}
agg.exp <- AggregateExpression(cd45pos, assays = "RNA", layers = "counts", group.by = "specific_annot")$RNA
dim(agg.exp)
colnames(agg.exp)

agg.exp.mat <- as.matrix(agg.exp)
agg.exp.mat <- log1p(agg.exp.mat)

cor_mat <- cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)
cor_mat[is.na(cor_mat)] <- 0

set.seed(1)
cor_mat <- cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)

png(
  filename = "output/SjD/cd45pos/Specific_correlation.png",
  width = 3600,
  height = 3600,
  res = 300
)
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()
```

# Merge and combined annotation

## 20. Save CD45-positive object and merge compartments

```{r}
saveRDS(cd45pos, file = "data/process/SeuratObjects/SjD/cd45(+).rds")

cd45pos <- readRDS("data/process/SeuratObjects/SjD/cd45(+).rds")
cd45neg <- readRDS("data/process/SeuratObjects/SjD/cd45(-).rds")

cd45neg$orig.ident <- "SjD"
cd45pos$orig.ident <- "SjD"

# tag compartment of origin before merging
cd45neg$compartment <- "CD45neg"
cd45pos$compartment <- "CD45pos"

SjD <- merge(
  cd45neg, y = cd45pos,
  add.cell.ids = c("CD45neg", "CD45pos"),
  project = "SjD"
)

saveRDS(SjD, file = "data/process/SeuratObjects/SjD/SjD(+).rds")
SjD <- readRDS("data/process/SeuratObjects/SjD/SjD(+).rds")
```

## 21. Export merged object for pyscenic (SCENIC) input

```{r}
SjD[["RNA"]] <- JoinLayers(SjD[["RNA"]]) # merge leaves per-compartment layers split

expr <- GetAssayData(SjD, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SjD/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SjD/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SjD/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

## 22. Major cell type dot plot for cluster naming

```{r}
markers <- c(
  "PTPRC", "MZB1", "PRDM1", "CCL19", "VCAM1", "CXCL10", "TNFSF13B", "CXCL9", "ICAM1", "SPARC",
  "COL3A1", "DCN", "LUM", "COL14A1", "DPT", "PECAM1", "CD34", "PDGFRB", "DLL4", "EFNB2",
  "SOX17", "SOX18", "EPHB4", "ACKR1", "NR2F2", "ZNF385D", "MEOX1", "LYVE1", "PROX1", "HES1",
  "ACTA2", "MCAM", "RGS5", "ABCC9", "CCL21", "TNC", "CD36", "MYH11", "NGFR", "CDH3",
  "TP63", "DLK1", "CFTR", "ASCL3", "FOXI1", "AQP5", "KRT5", "KRT14", "SOX2", "MUC19",
  "LPO", "FGFR2", "MUC1", "MUC5B", "ANO1", "KIT", "PIP"
)

DefaultAssay(SjD) <- "SCT"
SjD[["SCT"]] <- JoinLayers(SjD[["SCT"]])

markers <- markers[markers %in% rownames(SjD[["SCT"]])]

Idents(SjD) <- "specific_annot"
SjD$specific_annot <- droplevels(as.factor(SjD$specific_annot))
Idents(SjD) <- "specific_annot"

p <- DotPlot(
  SjD,
  features = markers,
  assay = "SCT",
  cols = c("lightgrey", "blue"),
  dot.scale = 6
) + RotatedAxis() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold")
  )

pdf("/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/output/SjD/SjD_cluster_name.pdf",
  width = 13, height = 13
)
print(p)
dev.off()
```

## 23. UMAP for merged object

```{r}
names(SjD@reductions)
SjD <- RunUMAP(SjD, reduction = "pca", dims = 1:30, seed.use = 42)
names(SjD@reductions)
SjD$Major_Celltype <- droplevels(as.factor(SjD$Major_Celltype))

celltype_cols_SjD <- c(
  "Acinar Cells"        = "coral1",
  "Ductal Cells"        = "mistyrose3",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Neurons"             = "#4B2EFF",
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "Mono/Mac"            = "darkblue",
  "NK Cells"            = "magenta",
  "Plasma Cells"        = "paleturquoise3",
  "Proliferating Cells" = "#B15928"
)

setdiff(unique(SjD$Major_Celltype), names(celltype_cols_SjD))

umap_df <- as.data.frame(Embeddings(SjD, "umap"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")
umap_df$Major_Celltype <- SjD$Major_Celltype

centers <- aggregate(
  cbind(UMAP_1, UMAP_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

library(ggplot2)
library(ggrepel)

suppressWarnings({
  fig_sjd <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Major_Celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols_SjD) +
    geom_label_repel(
      data = centers,
      aes(x = UMAP_1, y = UMAP_2, label = Major_Celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 6,
      box.padding = 0.4,
      label.size = 0.40,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "UMAP 1", y = "UMAP 2")
})

fig_sjd

pdf("/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/output/SjD/SjD_CellType.pdf",
  width = 12,
  height = 10
)
print(fig_sjd)
dev.off()
```

## 24. Major cell type marker heatmap

```{r}
markers <- c(
  # general / stromal / epithelial
  "PTPRC", "MZB1", "PRDM1", "CCL19", "VCAM1", "CXCL10", "TNFSF13B", "CXCL9", "ICAM1", "SPARC", "COL3A1", "DCN", "LUM", "COL14A1", "DPT",
  "PECAM1", "CD34", "PDGFRB", "DLL4", "EFNB2", "SOX17", "SOX18", "EPHB4", "ACKR1", "NR2F2", "ZNF385D", "MEOX1", "LYVE1", "PROX1", "HES1",
  "ACTA2", "MCAM", "RGS5", "ABCC9", "CCL21", "TNC", "CD36", "MYH11", "NGFR",
  "CDH3", "TP63", "DLK1", "CFTR", "ASCL3", "FOXI1", "AQP5", "KRT5", "KRT14", "SOX2", "MUC19", "LPO", "FGFR2", "MUC1", "MUC5B", "ANO1", "KIT", "PIP",
  # immune-specific additions for CD4/CD8/B/Mono-Mac/NK/Plasma/Proliferating
  "CD3D", "CD3E", "CD4", "CD8A", "CD8B", "IL7R", "CCR7", "MS4A1", "CD19", "CD79A",
  "CD14", "CD68", "CD163", "LYZ", "FCGR3A", "GNLY", "NKG7", "KLRD1", "MKI67", "TOP2A"
)

DefaultAssay(SjD) <- "SCT"
SjD[["SCT"]] <- JoinLayers(SjD[["SCT"]])

markers <- markers[markers %in% rownames(SjD[["SCT"]])]

Idents(SjD) <- "Major_Celltype"
SjD$Major_Celltype <- droplevels(as.factor(SjD$Major_Celltype))
Idents(SjD) <- "Major_Celltype"

expr_mat <- LayerData(SjD, assay = "SCT", layer = "data")

percent_mat <- sapply(levels(Idents(SjD)), function(ct) {
  cells <- WhichCells(SjD, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SjD,
  features = markers,
  assays = "SCT",
  layer = "data",
  return.seurat = FALSE
)$SCT

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[markers, common, drop = FALSE])
percent_mat <- percent_mat[markers, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD", recursive = TRUE, showWarnings = FALSE)
png(
  "output/SjD/SjD_Major_Celltype_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
```

## 25. TLS / immune microenvironment marker heatmap

```{r}
# marker groups from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)
inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)
lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)
hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)
fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(
  germinal_center, inflammation,
  lymphoid_chemokines, hev, fdc
))

DefaultAssay(SjD) <- "SCT"
SjD[["SCT"]] <- JoinLayers(SjD[["SCT"]])

markers <- markers[markers %in% rownames(SjD[["SCT"]])]

marker_groups <- case_when(
  markers %in% germinal_center     ~ "Germinal Center",
  markers %in% inflammation        ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev                 ~ "HEV",
  markers %in% fdc                 ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center"           = "#2166AC",
  "Inflammation & Cytokines"  = "#D6604D",
  "Lymphoid Chemokines"       = "#35978F",
  "HEV"                       = "#8073AC",
  "FDC"                       = "#E08214"
)

Idents(SjD) <- "Major_Celltype"
SjD$Major_Celltype <- droplevels(as.factor(SjD$Major_Celltype))
Idents(SjD) <- "Major_Celltype"

expr_mat <- GetAssayData(SjD, assay = "SCT", layer = "data")

percent_mat <- sapply(levels(Idents(SjD)), function(ct) {
  cells <- WhichCells(SjD, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SjD,
  features = markers,
  assays = "SCT",
  layer = "data",
  return.seurat = FALSE
)$SCT

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[markers, common, drop = FALSE])
percent_mat <- percent_mat[markers, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

dir.create("output/SjD", recursive = TRUE, showWarnings = FALSE)
png("output/SjD/TLS_markers.png",
  width = 3200, height = 3500, res = 300
)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```
