---
title: "R Notebook"
output: html_notebook
---

###Sjogrens Diseases############


##packages
```{r setup, include=FALSE}


library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")

```

###################################################################################

                      ###STEP_1Analysis for the CD45 NEGATIVE#############

##1-load the obj CD45 negative
```{r}
#-----

cd45neg=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45neg_annotated_clean.RDS")
DefaultAssay(cd45neg)
DefaultAssay(cd45neg) <- "RNA"
colnames(cd45neg@meta.data)
 cd45neg<- DietSeurat(cd45neg, assays = "RNA")
Assays(cd45neg)




```




```{r}
#------modify the meta data
keep_cols <- c(
  "orig.ident",
  "sample_name",
  "nCount_RNA",
  "nFeature_RNA",
  "seurat_clusters",
  "major_annot",
  "major_annot_joint",
  "specific_annot",
  "specific_annot_subcl"
)
cd45neg@meta.data <- cd45neg@meta.data[, keep_cols]
# Rename column (only if needed)
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "major_annot"] <- "Major_Celltype"

# Clean Major annotation
cd45neg@meta.data$Major_Celltype <- dplyr::recode(
  cd45neg@meta.data$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Neuronal_cells" = "Neurons",
  "Endothelial_cells" = "Endothelial Cells",
  "Immune_cells" = "Immune Cells",
  "Mural" = "Mural Cells",
  "Acinal_cells" = "Acinar Cells",
  "Ductal_cells" = "Ductal Cells"
)

# Create General annotation
cd45neg@meta.data$General <- dplyr::recode(
  cd45neg@meta.data$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Neurons" = "Neurons",
  "Endothelial Cells" = "Endothelial Cells",
  "Immune Cells" = "Immune Cells",
  "Mural Cells" = "Mural Cells",
  "Acinar Cells" = "Epithelial Cells",
  "Ductal Cells" = "Epithelial Cells"
)
View(cd45neg@meta.data)
```



```{r}
cd45neg <- NormalizeData(cd45neg)
cd45neg <- FindVariableFeatures(cd45neg)
cd45neg <- ScaleData(cd45neg)
cd45neg <- RunPCA(cd45neg, verbose = FALSE)
cd45neg <- RunUMAP(cd45neg, dims = 1:30)
```




We have two type of cell_annotattion the :major_annot and specif_annot(look for colnames(cd45neg@meta.data) to see the colnames)


###process for the Major Celltype#####

###2 Annotation 


```{r}

#markers________________



markers = c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")


markers <- markers[markers %in% rownames(cd45neg)]
cd45neg[["RNA"]]=JoinLayers(cd45neg[["RNA"]])
Idents(cd45neg) <- "Major_Celltype"


expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png("output/SjD/cd45neg/Major_Celltype_heatmap.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)

dev.off()




```







```{r}
##----umap---------
#1for the major annotation

Idents(cd45neg)="Major_Celltype"
cd45neg$Major_Celltype=Idents(cd45neg)  

#color for major annotation
celltype_cols=c(
"Fibroblasts"          = "#D73027",
"Endothelial Cells"    = "#4575B4",
"Mural Cells"          = "#1B9E77",
"Acinar Cells"         = "#762A83",
"Ductal Cells"         = "#FDAE61",
"Neurons"              = "#E7298A"
)
 #umap
names(cd45neg@reductions)
Reductions(cd45neg)
umap_df <- as.data.frame(Embeddings(cd45neg, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$Major_Celltype <- cd45neg$Major_Celltype

centers <- aggregate(cbind(umap_1, umap_2) ~ Major_Celltype,
                     data = umap_df, FUN = median)

library(ggplot2)
library(ggrepel)

fig_major <- ggplot(umap_df, aes(umap_1, umap_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")

fig_major
ggsave("output/SjD/cd45neg/Major_UMAP.png", plot = fig_major, width =12, height = 8 )


```








###4correlation between celltype
```{r}
#

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "Major_Celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45neg/Major_correlation.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()


```






########        ####process for the specific_annotation#####
specific_annot

##1annotation



```{r}
#markers________________
genes = c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")
colnames(cd45neg@meta.data)


genes <- genes[genes %in% rownames(cd45neg)]

cd45neg[["RNA"]] <- JoinLayers(cd45neg[["RNA"]])
Idents(cd45neg) <- "specific_annot"

expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {

  cells <- WhichCells(cd45neg, idents = ct)

  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})

percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))

exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))

col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {

  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png(
  "output/SjD/cd45neg/Specific_Celltype_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)

ComplexHeatmap::draw(ht)

dev.off()





```


##markers that we can have in TLS 

```{r}
tls <- c(
  "AID","AICDA","BCL6","PAX5","MYC","IRF8","CD40LG","ICOS","S1PR2","CXCR4","CXCR5","CCR7",
  "TNF","IL1B","IL6","IL7","IL8","IL12","IL17","IFNG","TGFB",
  "CCL2","CCL3","CCL4","CCL5","CXCL12","CXCL13","CCL19","CCL21",
  "ICAM1","VCAM1","SELE","CD54","MADCAM1","CHST4","GLYCAM1","FUT7","CHST2",
  "MEOX2","TSPAN7","ANKRD43","ZNF280C","CD21","CD23","CR1","CR2","CD35",
  "FDCSP","FCER2","NTAN1"
)

genes <- tls[tls %in% rownames(cd45neg)]

cd45neg[["RNA"]] <- JoinLayers(cd45neg[["RNA"]])
Idents(cd45neg) <- "General"

expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})

percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))

exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))

col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png(
  "output/SjD/cd45neg/TLS_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)

ComplexHeatmap::draw(ht)

dev.off()

```








```{r}
#-----------------
##plot for annotations
#1for the major annotation

Idents(cd45neg)="specific_annot"
cd45neg$celltype=Idents(cd45neg)  

#color for major annotation
celltype_cols = c(
  "VSMC"                                    = "#D73027",
  "Capillary pericytes"                     = "#4575B4",
  "Immunopericytes"                         = "#1B9E77",
  "Neuronal cells"                          = "#762A83",
  "Endothelial Progenitors"                 = "#FDAE61",
  "Arterial Endothelial"                    = "#E7298A",
  "Lymphatic Endothelial cells"             = "#66A61E",
  "CCL14 Venous Endothelial"                = "#E6AB02",
  "Venous Endothelial"                      = "#A6761D",
  "Fibroblasts_3"                           = "#666666",
  "Fibroblasts_4"                           = "#1F78B4",
  "Serous Acinar Epithelial cells"          = "#B2DF8A",
  "Intercalated Ductal Epithelial cells"    = "#33A02C",
  "Seromucous Acinar Epithelial cells"      = "#FB9A99",
  "Striated Ductal Epithelial cells"        = "#E31A1C",
  "Basal Epithelial cells"                  = "#FDBF6F",
  "Mucous Acinar Epithelial cells"          = "#FF7F00",
  "Epithelial Progenitor cells"             = "#CAB2D6",
  "Ionocytes"                               = "#6A3D9A",
  "Myoepithelial cells"                     = "#B15928",
  "Fibroblasts_1"                           = "#8DA0CB",
  "Fibroblasts_2"                           = "#66C2A5",
  "Fibroblasts_5"                           = "#FC8D62",
  "SPARC+ COL3A1 + Fibroblasts"             = "#E78AC3"
)

 #umap
#umap
names(cd45neg@reductions)
Reductions(cd45neg)
umap_df <- as.data.frame(Embeddings(cd45neg, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$specific_annot<- cd45neg$specific_annot

#center
centers=aggregate(cbind(umap_1, umap_2) ~ specific_annot,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = specific_annot)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = specific_annot),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP", y = "UMAP")
fig
ggsave("output/SjD/cd45neg/SjD_specific_UMAP.png", plot = fig, width =12, height = 8 )

```



###3correlation between cell_types
```{r}
#---

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "specific_annot")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45neg/Specific_correlation.png",
  width = 3600,
  height = 2600,
  res = 300
)
#plot
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()

```


##save
```{r}
#save
cd45neg$orig.ident= "SjD"
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "SjD"] <- "orig.ident"
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "sample_name"] <- "sample"
saveRDS(cd45neg, file = "data/process/SeuratObjects/SjD/cd45neg.rds")

#load later
#cd45neg <- readRDS("data/process/SeuratObjects/SjD/cd45neg.rds")




```









#####------------------------#################------------------------------------____----------------------------------------------- 

###STEP_2_Analysis for the CD45 Positive######################################################

#####----------------------------------------------------------------------------------------------------------------------------------------------------









####For  major_annot_joint#####
##1 load the data annotation for
```{r}
#---load the data
cd45pos=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45pos_annotated_clean.RDS")


View(cd45pos@meta.data)
DefaultAssay(cd45pos)
DefaultAssay(cd45pos) <- "RNA"
colnames(cd45pos@meta.data)
 cd45pos<- DietSeurat(cd45pos, assays = "RNA")

```




```{r}
#to have umapo
cd45pos <- NormalizeData(cd45pos)
cd45pos <- FindVariableFeatures(cd45pos)
cd45pos <- ScaleData(cd45pos)
cd45pos <- RunPCA(cd45pos, verbose = FALSE)
cd45pos <- RunUMAP(cd45pos, dims = 1:30)
```




```{r}



#------modify the meta data
keep_cols <- c(
  "orig.ident",
  "sample_name",
  "nCount_RNA",
  "nFeature_RNA",
  "seurat_clusters",
  "major_annot",
  "major_annot_joint",
  "major_annot_cl",
  "predicted_labels",
  "majority_voting",
  "specific_annot"
)
cd45pos@meta.data <- cd45pos@meta.data[, keep_cols]

# Rename column (only if needed)

colnames(cd45pos@meta.data)[colnames(cd45pos@meta.data) == "major_annot_joint"] <- "Major_Celltype"


# Clean Major annotation

unique(cd45pos$Major_Celltype)
cd45pos@meta.data$Major_Celltype <- dplyr::recode(
  cd45pos@meta.data$Major_Celltype,
  "Mono/Mac" = "Mono/Mac",
  "CD4" = "CD4",
  "gd/NK" = "gd/NK",
  "B cells" = "B Cells",
  "CD8" = "CD8",
  "Plasma cells" = "Plasma Cells",
  "Proliferating cells" = "Proliferating Cells"
)

# Create General annotation
cd45pos@meta.data$General <- dplyr::recode(
  cd45pos@meta.data$Major_Celltype,
 "Mono/Mac" = "Mono/Mac",
  "CD4" = "CD4",
  "gd/NK" = "gd/NK",
  "B Cells" = "B Cells",
  "CD8" = "CD8",
  "Plasma Cells" = "Plasma Cells",
  "Proliferating Cells" = "Proliferating Cells"
)
View(cd45pos@meta.data)

```









###2.annotation
```{r}

markers = c(
  "ZG16B","IGHM","IGHD","CCR7","GPR183","TEX9","MYC","KLF6","MIR155HG","ITGB1",
  "FOXP1","BACH2","CLEC9A","WDFY4","IRF8","CLEC4C","THBD","BATF3","CD1C",
  "FCER1A","FCGR3A","CD14","CX3CR1","CD163","CD36","FCGR1A","MERTK","CD80",
  "ZEB2","LAMP3","MRC1","SIRPA","CLEC10A","C1QA","IL1B","IGHA1","IGHG1",
  "IGHG3","IGHG4","TNF","NCAM1","NKG7","CD27","TRDC","KLRG1","FXYD7","KIT",
  "AHR","FGFBP2","FCRL6","CD8A","CD8B","MYB","SELL","TOX","IKZF2","HAVCR2",
  "CTLA4","ZNF683","ITGAE","CXCR6","CCL4L2","CCL4","CCL3","PRDM1","IFNG",
  "CD69","CCR6","XCL1","XCL2","CCL20","CLEC2B","NR4A1","HSPA6","BAG3",
  "GZMH","LGALS1","LEF1","TCF7","ANXA2","CD44","GZMA","CCL5","DLGAP1",
  "RORA","RORC","IL17A","FOXP3","CCR8","TIGIT","TNFRSF9","IL7R","CXCR5",
  "BCL6","SH2D1A","MAF","CXCL13","IL21","IL2RG","ITGA4","ITGB7"
)

markers <- markers[markers %in% rownames(cd45pos)]
cd45pos[["RNA"]] <- JoinLayers(cd45pos[["RNA"]])
Idents(cd45pos) <- "Major_Celltype"
expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "counts")
markers <- markers[markers %in% rownames(expr_mat)]

# only keep cells present in expr_mat
valid_cells <- colnames(cd45pos)[colnames(cd45pos) %in% colnames(expr_mat)]

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  cells <- cells[cells %in% valid_cells]
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# compute average expression directly from expr_mat
avg_exp <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  cells <- cells[cells %in% valid_cells]
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE])
})
exp_mat <- as.matrix(avg_exp)

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png("output/SjD/cd45pos/Major_Celltype_heatmap.png",
    width = 3200, height = 3500, res = 300)
ComplexHeatmap::draw(ht)
dev.off()



```




```{r}
##----umap---------
#1for the major annotation

Idents(cd45pos)="Major_Celltype"
cd45pos$Major_Celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols=c(
"Mono/Mac"             = "#D73027",
"CD4"                  = "#4575B4",
"CD8"                  = "#1B9E77",
"gd/NK"                = "#762A83",
"Proliferating Cells"  = "#FDAE61",
"B Cells"              = "#E7298A",
"Plasma Cells"         = "cyan"
)
 #umap
names(cd45pos@reductions)
Reductions(cd45pos)
umap_df <- as.data.frame(Embeddings(cd45pos, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$Major_Celltype <- cd45pos$Major_Celltype

centers <- aggregate(cbind(umap_1, umap_2) ~ Major_Celltype,
                     data = umap_df, FUN = median)

library(ggplot2)
library(ggrepel)

fig_major <- ggplot(umap_df, aes(umap_1, umap_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP", y = "UMAP")

fig_major
ggsave("output/SjD/cd45pos/Major_UMAP.png", plot = fig_major, width =12, height = 8 )


```























######"specific_annot"



####process for the specific_annotation in cd45_positive#################
specific_annot

###1annotation
```{r}

##plot for annotations
#1for the major annotation

Idents(cd45pos)="specific_annot"
cd45pos$celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols = c(
  "Activated Treg cells"                   = "#E76F51",
  "CD16+ NK cells"                         = "#F4A261",
  "CD27+ NK cells"                         = "#E9C46A",
  "CD56+ NK cells"                         = "#D4A373",
  "CD62L+ TPEx"                            = "#CDB4DB",
  "CD8+ Tcm"                               = "#90DBF4",
  "CD8+ Tem"                               = "#48CAE4",
  "CD8+ Trm"                               = "#0096C7",
  "cDC1"                                   = "#A3B18A",
  "cDC2"                                   = "#588157",
  "Cytotoxic Th cells"                     = "#6A994E",
  "DLGAP1+ Th cells"                       = "#A7C957",
  "Exhausted CD8+ T cells"                 = "#6D597A",
  "Exhausted IEG CD8+ T cells"             = "#B56576",
  "GC B cells_1"                           = "#4CC9F0",
  "GC B cells_2"                           = "#4895EF",
  "GC B cells_3"                           = "#4361EE",
  "GC B cells_4"                           = "#3A0CA3",
  "gd-T cells"                             = "#2A9D8F",
  "IgA Plasma cells"                       = "#F72585",
  "IgG1 Plasma cells"                      = "#FF7096",
  "IgG3/4 Plasma cells"                    = "#FF8FAB",
  "IgM Plasma cells"                       = "#FFC2D1",
  "ILCs"                                   = "#52B788",
  "Infiltrating CD8+ T cells"              = "#1B9AAA",
  "Inflammatory CD8+ T cells"              = "#0077B6",
  "inflammatory cDC2"                      = "#6C757D",
  "LAMP3+ cDCs"                            = "#ADB5BD",
  "LTB+ Th17 cells"                        = "#9D4EDD",
  "M1 Macrophages"                         = "#E63946",
  "M2 Macrophages"                         = "#A8DADC",
  "Memory Th cells"                        = "#457B9D",
  "Migrating Th17 cells"                   = "#5E60CE",
  "Mo-DC"                                  = "#7400B8",
  "Monocytes"                              = "#B5179E",
  "Naive B cells"                          = "#F1FAEE",
  "Naive CD8+ T cells"                     = "#8ECAE6",
  "Naive Tfh cells"                        = "#219EBC",
  "Naive Th cells"                         = "#023047",
  "Proliferating cells"                    = "#FFBE0B",
  "Recruited Activated CD8-"               = "#FB5607",
  "Recruiting Effector CD8+"               = "#FF006E",
  "Resting Treg cells"                     = "#8338EC",
  "Secretory Tph cells"                    = "#3A86FF",
  "TNF+ Plasma cells"                      = "#D00000",
  "Tph17"                                  = "#9B2226"
)


 #umap
umap_df=as.data.frame(Embeddings(cd45pos, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45pos$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45pos/SjD_specific_ann.png", plot = fig, width =12, height = 8 )

```


##markers that we can have in TLS 

```{r, fig.width=12, fig.height=8}
#----------
markers <- c(
  "AID", "AICDA", "BCL6", "PAX5", "MYC", "IRF8","CD40LG", "ICOS","S1PR2", "CXCR4", "CXCR5", "CCR7","TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", "IFNG", "TGFB","CCL2", "CCL3", "CCL4", "CCL5","CXCL12", "CXCL13", "CCL19", "CCL21","ICAM1", "VCAM1", "SELE", "CD54","MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2","MEOX2", "TSPAN7", "ANKRD43", "ZNF280C","CD21", "CD23", "CR1", "CR2", "CD35","FDCSP", "FCER2", "NTAN1")

p=DotPlot(
  cd45pos,
  features = markers,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "specific_annot") + RotatedAxis() 
p
#save
ggsave(filename = "output/SjD/cd45pos/TLS_markers.png", plot = p, width = 13, height = 8) 

```










###2proportions of cells
```{r}
####plot
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45pos_meta <- dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45pos_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45pos/SjD_specific_cell_proport.png", plot = fig1c, width = 12, height = 8)

```





###3correlation between celltype
```{r}
#---
#join layers(we have a lot of assays)
cd45pos = JoinLayers(
  cd45pos,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/SjD_specific_cell_cor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()
```





###4Donor proportion
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45pos@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```



```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45pos/SjD_specific_Donor.png", plot = fig, width = 12, height = 8)

##meta data

cd45pos$orig.ident="CD45 +"


```




##save new meta.dat
```{r}
#

# change_name
cd45pos_new <- cd45pos

# 
cd45pos_new$orig.ident <- "CD45+"

# keep col
cols_to_keep <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mt",
  "seurat_clusters",
  "specific_annot"
)

# keep
cd45pos_new@meta.data <- cd45pos_new@meta.data[, cols_to_keep, drop = FALSE]
colnames(cd45pos_new@meta.data)[
  colnames(cd45pos_new@meta.data) == "specific_annot"
] <- "celltype"

# Export CSV (optionnel)
write.csv(
  cd45pos_new@meta.data,
  file = "data/process/SeuratObjects/SjD/cd45pos_new.csv",
  row.names = TRUE
)

#duplicate the celltype col
cd45pos_new@meta.data$celltype_clust <- cd45pos_new@meta.data$celltype
View(cd45pos_new@meta.data)

```




###save
```{r}
###i add some modification to the initial meta data so load the original one to see the celltype annotation of both cd45pos and cd45neg
# save
saveRDS(cd45pos_new, file = "data/process/SeuratObjects/SjD/cd45pos_new.rds")

#load later
#
cd45pos_new <- readRDS("data/process/SeuratObjects/SjD/cd45pos_new.rds")
View(cd45pos_new@meta.data)

```













