---
title: "R Notebook"
output: html_notebook
---

###Sjogrens Diseases############


##packages
```{r setup, include=FALSE}


library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")

```

###################################################################################

                      ###STEP_1Analysis for the CD45 NEGATIVE#############

##1-load the obj CD45 negative
```{r}
#-----

cd45neg=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45neg_annotated_clean.RDS")
DefaultAssay(cd45neg)
#DefaultAssay(cd45neg) <- "RNA"
#colnames(cd45neg@meta.data)
 #cd45neg<- DietSeurat(cd45neg, assays = "RNA")
Assays(cd45neg)

View(cd45neg@meta.data)


```


```{r}
#------modify the meta data
keep_cols <- c(
  "orig.ident",
  "sample_name",
  "nCount_RNA",
  "nFeature_RNA",
  "seurat_clusters",
  "major_annot",
  "major_annot_joint",
  "specific_annot",
  "specific_annot_subcl"
)
cd45neg@meta.data <- cd45neg@meta.data[, keep_cols]
# Rename column (only if needed)
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "major_annot"] <- "Major_Celltype"
cd45neg$orig.ident= "SjD"
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "SjD"] <- "orig.ident"
colnames(cd45neg@meta.data)[colnames(cd45neg@meta.data) == "sample_name"] <- "sample"



```


```{r}

# Clean Major annotation
cd45neg@meta.data$Major_Celltype <- dplyr::recode(
  cd45neg@meta.data$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Neuronal_cells" = "Neurons",
  "Endothelial_cells" = "Endothelial Cells",
  "Immune_cells" = "Immune Cells",
  "Mural" = "Mural Cells",
  "Acinal_cells" = "Acinar Cells",
  "Ductal_cells" = "Ductal Cells"
)

# Create General annotation
cd45neg@meta.data$General <- dplyr::recode(
  cd45neg@meta.data$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Neurons" = "Neurons",
  "Endothelial Cells" = "Endothelial Cells",
  "Immune Cells" = "Immune Cells",
  "Mural Cells" = "Mural Cells",
  "Acinar Cells" = "Epithelial Cells",
  "Ductal Cells" = "Epithelial Cells"
)


## remove the single mis-annotated cell
cd45neg <- subset(cd45neg, subset = Major_Celltype != "Immune Cells")

## drop the now-empty factor level so it doesn't show up in the legend/plot
cd45neg$Major_Celltype <- droplevels(as.factor(cd45neg$Major_Celltype))

## confirm it's gone
table(cd45neg$Major_Celltype)




```


```{r}
umap_df <- as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")
umap_df$Major_Celltype <- cd45neg$Major_Celltype

centers <- aggregate(
  cbind(UMAP_1, UMAP_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

celltype_cols_sjd <- c(
  "Acinar Cells"      = "#1f78b4",
  "Ductal Cells"      = "#66a61e",
  "Endothelial Cells" = "#2ca02c",
  "Fibroblasts"       = "#D73027",
  "Mural Cells"       = "#984EA3",
  "Neurons"           = "#a6761d"
)

library(ggrepel)
suppressWarnings({
  fig_sjd <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Major_Celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols_sjd) +
    geom_label_repel(
      data = centers,
      aes(x = UMAP_1, y = UMAP_2, label = Major_Celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 3.5,
      box.padding = 0.4,
      label.size = 0.40,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "UMAP 1", y = "UMAP 2")
})

fig_sjd
ggsave(filename = "output/SjD/cd45neg_cellType.png", plot = fig_sjd, width = 12, height = 8)

```






We have two type of cell_annotattion the :major_annot and specif_annot(look for colnames(cd45neg@meta.data) to see the colnames)


###process for the Major Celltype#####

###2 Annotation 


```{r}

#markers________________
markers <- c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")

DefaultAssay(cd45neg) <- "SCT"   # make sure rownames check below is against SCT, not RNA
markers <- markers[markers %in% rownames(cd45neg)]

Idents(cd45neg) <- "Major_Celltype"
expr_mat <- LayerData(cd45neg, assay = "SCT", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = markers,
  assays = "SCT",
  layer = "data",
  return.seurat = FALSE
)$SCT   # <-- fixed: was $RNA, but you requested assays = "SCT"

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)   # row_km uses k-means, which is stochastic — set a seed for reproducibility
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD/cd45neg", recursive = TRUE, showWarnings = FALSE)  # avoids "cannot open file" if the folder doesn't exist yet
png("output/SjD/Major_Celltype_heatmap.png",
    width = 3200, height = 3500, res = 300)
ComplexHeatmap::draw(ht)
dev.off()

```












###4correlation between celltype
```{r}
#

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "Major_Celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/Major_correlation.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()


```






########        ####process for the specific_annotation#####
specific_annot

##1annotation



```{r}

#markers________________
genes <- c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")

cd45neg[["RNA"]] <- JoinLayers(cd45neg[["RNA"]])

# confirm RNA has a normalized "data" layer before pulling from it
if (!"data" %in% Layers(cd45neg[["RNA"]])) {
  cd45neg <- NormalizeData(cd45neg, assay = "RNA")
}

genes <- genes[genes %in% rownames(cd45neg[["RNA"]])]   # filter against RNA specifically, not whatever DefaultAssay is set to

Idents(cd45neg) <- "specific_annot"
cd45neg$specific_annot <- droplevels(as.factor(cd45neg$specific_annot))  # drop any leftover unused levels (e.g. from the removed Immune Cells)
Idents(cd45neg) <- "specific_annot"  # re-set after droplevels

expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)   # row_km uses k-means (stochastic) — seed for reproducibility
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD", recursive = TRUE, showWarnings = FALSE)
png(
  "output/SjD/Specific_Celltype_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()


```










##markers that we can have in TLS 

```{r}
tls <- c(
  "AID","AICDA","BCL6","PAX5","MYC","IRF8","CD40LG","ICOS","S1PR2","CXCR4","CXCR5","CCR7",
  "TNF","IL1B","IL6","IL7","IL8","IL12","IL17","IFNG","TGFB",
  "CCL2","CCL3","CCL4","CCL5","CXCL12","CXCL13","CCL19","CCL21",
  "ICAM1","VCAM1","SELE","CD54","MADCAM1","CHST4","GLYCAM1","FUT7","CHST2",
  "MEOX2","TSPAN7","ANKRD43","ZNF280C","CD21","CD23","CR1","CR2","CD35",
  "FDCSP","FCER2","NTAN1"
)

genes <- tls[tls %in% rownames(cd45neg)]

cd45neg[["RNA"]] <- JoinLayers(cd45neg[["RNA"]])
Idents(cd45neg) <- "specific_annot"

expr_mat <- LayerData(cd45neg, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45neg)), function(ct) {
  cells <- WhichCells(cd45neg, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})

percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45neg,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))

exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))

col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png(
  "output/SjD/cd45neg/TLS_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)

ComplexHeatmap::draw(ht)

dev.off()

```








```{r}
#-----------------
## plot for annotations
## 1 for the specific annotation
Idents(cd45neg) <- "specific_annot"
cd45neg$celltype <- Idents(cd45neg)

# color for specific annotation
celltype_cols <- c(
  "VSMC"                                    = "#D73027",
  "Capillary pericytes"                     = "#4575B4",
  "Immunopericytes"                         = "#1B9E77",
  "Neuronal cells"                          = "#762A83",
  "Endothelial Progenitors"                 = "#FDAE61",
  "Arterial Endothelial"                    = "#E7298A",
  "Lymphatic Endothelial cells"             = "#66A61E",
  "CCL14 Venous Endothelial"                = "#E6AB02",
  "Venous Endothelial"                      = "#A6761D",
  "Fibroblasts_3"                           = "#666666",
  "Fibroblasts_4"                           = "#1F78B4",
  "Serous Acinar Epithelial cells"          = "#B2DF8A",
  "Intercalated Ductal Epithelial cells"    = "#33A02C",
  "Seromucous Acinar Epithelial cells"      = "#FB9A99",
  "Striated Ductal Epithelial cells"        = "#E31A1C",
  "Basal Epithelial cells"                  = "#FDBF6F",
  "Mucous Acinar Epithelial cells"          = "#FF7F00",
  "Epithelial Progenitor cells"             = "#CAB2D6",
  "Ionocytes"                               = "#6A3D9A",
  "Myoepithelial cells"                     = "#B15928",
  "Fibroblasts_1"                           = "#8DA0CB",
  "Fibroblasts_2"                           = "#66C2A5",
  "Fibroblasts_5"                           = "#FC8D62",
  "SPARC+ COL3A1 + Fibroblasts"             = "#E78AC3"
)

## sanity checks before plotting
setdiff(unique(cd45neg$specific_annot), names(celltype_cols))
Reductions(cd45neg)

## umap
umap_df <- as.data.frame(Embeddings(cd45neg, "umap_h"))   # use "umap_h" for consistency with the Major_Celltype plot; swap back to "umap" if that's intentional
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$specific_annot <- cd45neg$specific_annot

## centers
centers <- aggregate(cbind(umap_1, umap_2) ~ specific_annot, data = umap_df, FUN = median)

library(ggrepel)
library(ggplot2)

fig <- ggplot(umap_df, aes(umap_1, umap_2, color = specific_annot)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = specific_annot),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")

fig

dir.create("output/SjD/cd45neg", recursive = TRUE, showWarnings = FALSE)
ggsave("output/SjD/SjD_specific_UMAP.png", plot = fig, width = 12, height = 8)

```






















##save
```{r}
#save

saveRDS(cd45neg, file = "data/process/SeuratObjects/SjD/cd45(-).rds")

#load later
cd45neg <- readRDS("data/process/SeuratObjects/SjD/cd45neg.rds")

View(cd45neg@meta.data)


```









#####------------------------#################------------------------------------____----------------------------------------------- 

###STEP_2_Analysis for the CD45 Positive######################################################

#####----------------------------------------------------------------------------------------------------------------------------------------------------











####For  major_annot_joint#####
##1 load the data annotation for
```{r}
#---load the data
cd45pos=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45pos_annotated_clean.RDS")



DefaultAssay(cd45pos)
#DefaultAssay(cd45pos) <- "RNA"
colnames(cd45pos@meta.data)
 #cd45pos<- DietSeurat(cd45pos, assays = "RNA")
View(cd45pos@meta.data)


```


```{r}
#------modify the meta data
keep_cols <- c(
  "orig.ident",
  "sample_name",
  "nCount_RNA",
  "nFeature_RNA",
  "seurat_clusters",
  "major_annot",
  "major_annot_cl",
  "major_annot_joint",
  "predicted_labels",
  "majority_voting",
  "specific_annot"
)
cd45pos@meta.data <- cd45pos@meta.data[, keep_cols]
# Rename column (only if needed)
colnames(cd45pos@meta.data)[colnames(cd45pos@meta.data) == "major_annot_joint"] <- "Major_Celltype"
cd45pos$orig.ident= "SjD"
colnames(cd45pos@meta.data)[colnames(cd45pos@meta.data) == "SjD"] <- "orig.ident"
colnames(cd45pos@meta.data)[colnames(cd45pos@meta.data) == "sample_name"] <- "sample"






# Clean Major annotation

cd45pos@meta.data$Major_Celltype <- dplyr::recode(
  cd45pos@meta.data$Major_Celltype,
  "Mono/Mac" = "Mono/Mac",
   "gd/NK"    ="NK Cells",
   "B cells"      ="B Cells",
   "Proliferating cells"= "Proliferating Cells",
  "CD8" = "CD8",
  "Plasma cells" ="Plasma Cells",
  "CD4"       ="CD4"       
)

# Create General annotation
cd45pos@meta.data$General <- dplyr::recode(
  cd45pos@meta.data$Major_Celltype,
  "Mono/Mac" = "Mono/Mac",
   "gd/NK"    ="NK Cells",
   "B Cells"      ="B Cells",
   "Proliferating_cells"= "Proliferating Cells",
  "CD8" = "CD8",
  "Plasma Cells" ="Plasma Cells",
  "CD4"       ="CD4" 
)

unique(cd45pos$Major_Celltype)


```




###2annotation for Major_Celltype


```{r}
#marker__________________________
#to see col to plot 


markers = c(
  "ZG16B","IGHM","IGHD","CCR7","GPR183","TEX9","MYC","KLF6","MIR155HG","ITGB1",
  "FOXP1","BACH2","CLEC9A","WDFY4","IRF8","CLEC4C","THBD","BATF3","CD1C",
  "FCER1A","FCGR3A","CD14","CX3CR1","CD163","CD36","FCGR1A","MERTK","CD80",
  "ZEB2","LAMP3","MRC1","SIRPA","CLEC10A","C1QA","IL1B","IGHA1","IGHG1",
  "IGHG3","IGHG4","TNF","NCAM1","NKG7","CD27","TRDC","KLRG1","FXYD7","KIT",
  "AHR","FGFBP2","FCRL6","CD8A","CD8B","MYB","SELL","TOX","IKZF2","HAVCR2",
  "CTLA4","ZNF683","ITGAE","CXCR6","CCL4L2","CCL4","CCL3","PRDM1","IFNG",
  "CD69","CCR6","XCL1","XCL2","CCL20","CLEC2B","NR4A1","HSPA6","BAG3",
  "GZMH","LGALS1","LEF1","TCF7","ANXA2","CD44","GZMA","CCL5","DLGAP1",
  "RORA","RORC","IL17A","FOXP3","CCR8","TIGIT","TNFRSF9","IL7R","CXCR5",
  "BCL6","SH2D1A","MAF","CXCL13","IL21","IL2RG","ITGA4","ITGB7"
)

markers <- markers[markers %in% rownames(cd45pos)]
#cd45pos[["RNA"]]=JoinLayers(cd45pos[["RNA"]])
Idents(cd45pos) <- "Major_Celltype"


expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45pos,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png("output/SjD/cd45pos/Major_Celltype_heatmap.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)

dev.off()


```








```{r}
##plot for annotations
#1for the major annotation

Idents(cd45pos)="Major_Celltype"
cd45pos$Major_Celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols=c(
"B Cells"                       = "magenta",
"CD8"                           = "#FC8D62",
"CD4"                           = "#1B9E77",
"NK Cells"                      = "darkblue",
"Plasma Cells"                  = "#FDAE61",
"Proliferating Cells"           = "yellow",
"Mono/Mac"                      = "red3"
)
#umap
names(cd45pos@reductions)
Reductions(cd45pos)
umap_df <- as.data.frame(Embeddings(cd45pos, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$Major_Celltype <- cd45pos$Major_Celltype

centers <- aggregate(cbind(umap_1, umap_2) ~ Major_Celltype,
                     data = umap_df, FUN = median)

library(ggplot2)
library(ggrepel)

fig_major <- ggplot(umap_df, aes(umap_1, umap_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")

fig_major

ggsave("output/SjD/cd45pos/Major_heatmap.png", plot = fig_major, width =12, height = 8 )

```




###4correlation between celltype
```{r}
#

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "Major_Celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/Major_correlation.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()


```










######"specific_annot" in cd45positive#########################



    



#1annotation
```{r}
###
markers <- c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")

cd45pos[["RNA"]] <- JoinLayers(cd45pos[["RNA"]])   # uncommented — needed if RNA has split layers (multiple samples); safe no-op if it doesn't

# confirm RNA has a normalized "data" layer before pulling from it
if (!"data" %in% Layers(cd45pos[["RNA"]])) {
  cd45pos <- NormalizeData(cd45pos, assay = "RNA")
}

markers <- markers[markers %in% rownames(cd45pos[["RNA"]])]   # check against RNA specifically, not whatever DefaultAssay happens to be

Idents(cd45pos) <- "specific_annot"
cd45pos$specific_annot <- droplevels(as.factor(cd45pos$specific_annot))  # drop any unused levels
Idents(cd45pos) <- "specific_annot"

expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45pos,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[markers, common, drop = FALSE])
percent_mat <- percent_mat[markers, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)   # row_km uses k-means (stochastic) — seed for reproducibility
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD/cd45pos", recursive = TRUE, showWarnings = FALSE)
png(
  "output/SjD/cd45pos/Specific_Celltype_Umap.png",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()



```








###1annotation
```{r}

##plot for annotations
#1for the major annotation

Idents(cd45pos)="specific_annot"
cd45pos$celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols = c(
  "Activated Treg cells"                   = "#E76F51",
  "CD16+ NK cells"                         = "#F4A261",
  "CD27+ NK cells"                         = "#E9C46A",
  "CD56+ NK cells"                         = "#D4A373",
  "CD62L+ TPEx"                            = "#CDB4DB",
  "CD8+ Tcm"                               = "#90DBF4",
  "CD8+ Tem"                               = "#48CAE4",
  "CD8+ Trm"                               = "#0096C7",
  "cDC1"                                   = "#A3B18A",
  "cDC2"                                   = "#588157",
  "Cytotoxic Th cells"                     = "#6A994E",
  "DLGAP1+ Th cells"                       = "#A7C957",
  "Exhausted CD8+ T cells"                 = "#6D597A",
  "Exhausted IEG CD8+ T cells"             = "#B56576",
  "GC B cells_1"                           = "#4CC9F0",
  "GC B cells_2"                           = "#4895EF",
  "GC B cells_3"                           = "#4361EE",
  "GC B cells_4"                           = "#3A0CA3",
  "gd-T cells"                             = "#2A9D8F",
  "IgA Plasma cells"                       = "#F72585",
  "IgG1 Plasma cells"                      = "#FF7096",
  "IgG3/4 Plasma cells"                    = "#FF8FAB",
  "IgM Plasma cells"                       = "#FFC2D1",
  "ILCs"                                   = "#52B788",
  "Infiltrating CD8+ T cells"              = "#1B9AAA",
  "Inflammatory CD8+ T cells"              = "#0077B6",
  "inflammatory cDC2"                      = "#6C757D",
  "LAMP3+ cDCs"                            = "#ADB5BD",
  "LTB+ Th17 cells"                        = "#9D4EDD",
  "M1 Macrophages"                         = "#E63946",
  "M2 Macrophages"                         = "#A8DADC",
  "Memory Th cells"                        = "#457B9D",
  "Migrating Th17 cells"                   = "#5E60CE",
  "Mo-DC"                                  = "#7400B8",
  "Monocytes"                              = "#B5179E",
  "Naive B cells"                          = "#F1FAEE",
  "Naive CD8+ T cells"                     = "#8ECAE6",
  "Naive Tfh cells"                        = "#219EBC",
  "Naive Th cells"                         = "#023047",
  "Proliferating cells"                    = "#FFBE0B",
  "Recruited Activated CD8-"               = "#FB5607",
  "Recruiting Effector CD8+"               = "#FF006E",
  "Resting Treg cells"                     = "#8338EC",
  "Secretory Tph cells"                    = "#3A86FF",
  "TNF+ Plasma cells"                      = "#D00000",
  "Tph17"                                  = "#9B2226"
)

 #umap
#umap
names(cd45pos@reductions)
Reductions(cd45pos)
umap_df <- as.data.frame(Embeddings(cd45pos, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$specific_annot<- cd45pos$specific_annot

#center
centers=aggregate(cbind(umap_1, umap_2) ~ specific_annot,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = specific_annot)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = specific_annot),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig

ggsave("output/SjD/cd45pos/Specific_UMAP.png", plot = fig, width =12, height = 8 )

```



##markers that we can have in TLS 

```{r}
#----------
tls <- c(
  "AID", "AICDA", "BCL6", "PAX5", "MYC", "IRF8","CD40LG", "ICOS","S1PR2", "CXCR4", "CXCR5", "CCR7","TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", "IFNG", "TGFB","CCL2", "CCL3", "CCL4", "CCL5","CXCL12", "CXCL13", "CCL19", "CCL21","ICAM1", "VCAM1", "SELE", "CD54","MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2","MEOX2", "TSPAN7", "ANKRD43", "ZNF280C","CD21", "CD23", "CR1", "CR2", "CD35","FDCSP", "FCER2", "NTAN1")


genes <- tls[tls %in% rownames(cd45pos)]

cd45pos[["RNA"]] <- JoinLayers(cd45pos[["RNA"]])
Idents(cd45pos) <- "specific_annot"

expr_mat <- LayerData(cd45pos, assay = "RNA", layer = "data")

percent_mat <- sapply(levels(Idents(cd45pos)), function(ct) {
  cells <- WhichCells(cd45pos, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})

percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  cd45pos,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

common <- intersect(colnames(percent_mat), colnames(avg_exp))

exp_mat <- as.matrix(avg_exp[genes, common, drop = FALSE])
percent_mat <- percent_mat[genes, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))

col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png(
  "output/SjD/cd45pos/TLS_heatmap.png",
  width = 3200,
  height = 3500,
  res = 300
)

ComplexHeatmap::draw(ht)

dev.off()


```



###3correlation between cell_types
```{r}
#---

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "specific_annot")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/Specific_correlation.png",
  width = 3600,
  height = 3600,
  res = 300
)
#plot
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = "90",
  border_color = NA
)
dev.off()

```


```{r}
#pyscenic
#####for pyscenic 
expr <- GetAssayData(SjD, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SjD/input/matrix.mtx")

write.table(rownames(expr),
            file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SjD/input/genes.tsv",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

write.table(colnames(expr),
            file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SjD/input/barcodes.tsv",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

```








###save
```{r}

saveRDS(cd45pos, file = "data/process/SeuratObjects/SjD/cd45(+).rds")


## load cd45pos
cd45pos <- readRDS("data/process/SeuratObjects/SjD/cd45(+).rds")
cd45neg <- readRDS("data/process/SeuratObjects/SjD/cd45(-).rds")
## set orig.ident to "SjD" on both objects
cd45neg$orig.ident <- "SjD"
cd45pos$orig.ident <- "SjD"

## tag compartment of origin before merging (so you can still tell them apart afterward)
cd45neg$compartment <- "CD45neg"
cd45pos$compartment <- "CD45pos"

## merge, leaving RNA as-is
SjD <- merge(
  cd45neg, y = cd45pos,
  add.cell.ids = c("CD45neg", "CD45pos"),
  project = "SjD"
)

saveRDS(SjD, file = "data/process/SeuratObjects/SjD/SjD(+).rds")
SjD=readRDS("data/process/SeuratObjects/SjD/SjD(+).rds")


```



```{r}


#table(SjD$specific_annot)

#DefaultAssay(SjD) <- "SCT"
#SjD[["SCT"]] <- JoinLayers(SjD[["SCT"]])   # safe no-op if already joined

markers <- c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")
#markers <- markers[markers %in% rownames(SjD[["SCT"]])]



Idents(SjD) <- "specific_annot"
SjD$specific_annot <- droplevels(as.factor(SjD$specific_annot))
Idents(SjD) <- "specific_annot"

p <- DotPlot(
  SjD,
  features = markers,
  assay = "SCT",
  cols = c("lightgrey", "blue"),
  dot.scale = 6
) + RotatedAxis() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold")
  )

pdf("/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/output/SjD/SjD_cluster_name.pdf", 
    width = 13, height = 13)
print(p)
dev.off()


        
       
      
           

```



```{r}
names(SjD@reductions)   # confirm what's available, e.g. "pca" "harmony" "umap"
SjD <- RunUMAP(SjD, reduction = "pca", dims = 1:30, seed.use = 42)
names(SjD@reductions)   # should now show "pca" "umap"
SjD$Major_Celltype <- droplevels(as.factor(SjD$Major_Celltype))

celltype_cols_SjD <- c(
  "Acinar Cells"        = "coral1",
  "Ductal Cells"        = "mistyrose3",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Neurons"             = "#4B2EFF",
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "Mono/Mac"            = "darkblue",
  "NK Cells"            = "magenta",
  "Plasma Cells"        = "paleturquoise3",
  "Proliferating Cells" = "#B15928"
)

setdiff(unique(SjD$Major_Celltype), names(celltype_cols_SjD))

umap_df <- as.data.frame(Embeddings(SjD, "umap"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")

umap_df$Major_Celltype <- SjD$Major_Celltype

centers <- aggregate(
  cbind(UMAP_1, UMAP_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

library(ggplot2)
library(ggrepel)

suppressWarnings({
  fig_sjd <- ggplot(umap_df, aes(UMAP_1, UMAP_2, color = Major_Celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols_SjD) +
    geom_label_repel(
      data = centers,
      aes(x = UMAP_1, y = UMAP_2, label = Major_Celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 6,
      box.padding = 0.4,
      label.size = 0.40,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "UMAP 1", y = "UMAP 2")
})

fig_sjd


pdf("/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/output/SjD/SjD_CellType.pdf", 
    width = 12,  
    height = 10)
print(fig_sjd)
dev.off()
  

```




```{r}
#markers________________
markers <- c(
  # general / stromal / epithelial (as before)
  "PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT",
  "PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1",
  "ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR",
  "CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP",
  # immune-specific additions for CD4/CD8/B/Mono-Mac/NK/Plasma/Proliferating
  "CD3D","CD3E","CD4","CD8A","CD8B","IL7R","CCR7","MS4A1","CD19","CD79A",
  "CD14","CD68","CD163","LYZ","FCGR3A","GNLY","NKG7","KLRD1","MKI67","TOP2A"
)

DefaultAssay(SjD) <- "SCT"

# join SCT layers if split (merge can leave them split per source object)
#SjD[["SCT"]] <- JoinLayers(SjD[["SCT"]])

markers <- markers[markers %in% rownames(SjD[["SCT"]])]

Idents(SjD) <- "Major_Celltype"
SjD$Major_Celltype <- droplevels(as.factor(SjD$Major_Celltype))
Idents(SjD) <- "Major_Celltype"

expr_mat <- LayerData(SjD, assay = "SCT", layer = "data")

percent_mat <- sapply(levels(Idents(SjD)), function(ct) {
  cells <- WhichCells(SjD, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SjD,
  features = markers,
  assays = "SCT",
  layer = "data",
  return.seurat = FALSE
)$SCT

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[markers, common, drop = FALSE])
percent_mat <- percent_mat[markers, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

set.seed(42)
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

dir.create("output/SjD", recursive = TRUE, showWarnings = FALSE)
png(
  "output/SjD/SjD_Major_Celltype_heatmap.pdf",
  width = 3200,
  height = 3500,
  res = 300
)
ComplexHeatmap::draw(ht)
dev.off()
#
```


```{r}
# Define markers per group from genes_TLS.csv
germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC", 
                     "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")
inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", 
                  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5", 
                  "ICAM1", "VCAM1", "SELE")
lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12", 
                         "CXCR5", "CCR7", "CXCR4")
hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2", 
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")
fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP", 
         "CXCL13", "CD54", "ICAM1", "FCER2")

# Combine all markers in order
markers <- unique(c(germinal_center, inflammation, 
                    lymphoid_chemokines, hev, fdc))

DefaultAssay(SjD) <- "SCT"
#SjD[["SCT"]] <- JoinLayers(SjD[["SCT"]])   # safe no-op if already joined

# Filter to genes present in your Seurat object (SCT specifically)
markers <- markers[markers %in% rownames(SjD[["SCT"]])]

# Build group vector in same order
marker_groups <- case_when(
  markers %in% germinal_center    ~ "Germinal Center",
  markers %in% inflammation       ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev                ~ "HEV",
  markers %in% fdc                ~ "FDC"
)
names(marker_groups) <- markers

# Group colors
group_cols <- c(
  "Germinal Center"         = "#2166AC",
  "Inflammation & Cytokines"= "#D6604D",
  "Lymphoid Chemokines"     = "#35978F",
  "HEV"                     = "#8073AC",
  "FDC"                     = "#E08214"
)

# Extract expression and percent data
Idents(SjD) <- "Major_Celltype"
SjD$Major_Celltype <- droplevels(as.factor(SjD$Major_Celltype))
Idents(SjD) <- "Major_Celltype"

expr_mat <- GetAssayData(SjD, assay = "SCT", layer = "data")

percent_mat <- sapply(levels(Idents(SjD)), function(ct) {
  cells <- WhichCells(SjD, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SjD,
  features = markers,
  assays = "SCT",
  layer = "data",
  return.seurat = FALSE
)$SCT   # matches assays = "SCT" above

common <- intersect(colnames(percent_mat), colnames(avg_exp))
exp_mat <- as.matrix(avg_exp[markers, common, drop = FALSE])
percent_mat <- percent_mat[markers, common, drop = FALSE]

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# Color function
q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# Bubble cell function
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# Column annotation
celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# Row annotation
row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

# Heatmap
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

dir.create("output/SjD", recursive = TRUE, showWarnings = FALSE)
png("output/SjD/TLS_markers.pdf",
    width = 3200, height = 3500, res = 300)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```






