---
title: "R Notebook"
output: html_notebook
---

###Sjogrens Diseases############


##packages
```{r setup, include=FALSE}
#-----
library(pacman)
  p_load(Seurat, SeuratData, Matrix, dplyr, DropletUtils, presto,gprofiler2, ggplot2, patchwork, gplots, cowplot, openxlsx,plyr, scDblFinder, BiocParallel,devtools,openxlsx,reticulate,hdf5r,DropletUtils)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")

```


###STEP_1Analysis for the CD45 NEGATIVE#############

##1-load the obj CD45 negative
```{r}
#-----

cd45neg=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45neg_annotated_clean.RDS")
#cd45neg
#to see all annotations
#colnames(cd45neg@meta.data)
#p=DimPlot(cd45neg, label = TRUE, reduction = "umap_h")
#p
#ggsave("output/SjD/cd45neg/SjD_Resolution.png",plot = p, width = 12, height = 8)
cd45neg$orig.ident="CD45 -"
View(cd45neg@meta.data)
colnames(cd45neg@meta.data)

```

We have two type of cell_annotattion the :major_annot and specif_annot(look for colnames(cd45neg@meta.data) to see the colnames)


###process for the major_annotation#####

###2 Annotation 


```{r, fig.width=14, fig.height=8}

#major_annot for the major celltype ,major_annot_cl for the clusters_number and major_annot joint for the major celltpe and the markers in a dimplot. just need to change alll time the "group.by" 

#major_annot= major_annot_joint


#markers________________
genes = c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")


p=DotPlot(
  cd45neg,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "major_annot_cl") + RotatedAxis() 
p
#save
ggsave(filename = "output/SjD/cd45neg/SjD_Major_cluster_name.png", plot = p, width = 14, height = 8) 



```





```{r}
##----umap---------
#1for the major annotation

Idents(cd45neg)="major_annot_joint"
cd45neg$celltype=Idents(cd45neg)  

#color for major annotation
celltype_cols=c(
"Fibroblasts"          = "#D73027",
"Endothelial_cells"    = "#4575B4",
"Mural"                = "#1B9E77",
"Acinar_cells"         = "#762A83",
"Ductal_cells"         = "#FDAE61",
"Neuronal_cells"       = "#E7298A"
)
 #umap
umap_df=as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45neg$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig_major=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig_major
ggsave("output/SjD/cd45neg/SjD_Major_ann.png", plot = fig_major, width =12, height = 8 )


```





###3 proportion of cells
```{r}

library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45neg_meta <- dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45neg_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45neg/SjD_Major_proport.png", plot = fig1c, width = 12, height = 8)

```




###4correlation between celltype
```{r}
#---
#join layers(we have a lot of assays)
cd45neg <- JoinLayers(
  cd45neg,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45neg/SjD_Major_cor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()


```



##5Donor proportions
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45neg@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

```




```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45neg/Sj_Don_Major.png", plot = fig, width = 12, height = 8)

```









###process for the specific_annotation#####
specific_annot

##1annotation



```{r,fig.width=12,fig.height=8}
#markers________________
genes = c("PTPRC","MZB1","PRDM1","CCL19","VCAM1","CXCL10","TNFSF13B","CXCL9","ICAM1","SPARC","COL3A1","DCN","LUM","COL14A1","DPT","PECAM1","CD34","PDGFRB","DLL4","EFNB2","SOX17","SOX18","EPHB4","ACKR1","NR2F2","ZNF385D","MEOX1","LYVE1","PROX1","HES1","ACTA2","MCAM","RGS5","ABCC9","CCL21","TNC","CD36","MYH11","NGFR","CDH3","TP63","DLK1","CFTR","ASCL3","FOXI1","AQP5","KRT5","KRT14","SOX2","MUC19","LPO","FGFR2","MUC1","MUC5B","ANO1","KIT","PIP")


p=DotPlot(
  cd45neg,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "specific_annot_subcl") + RotatedAxis()  #have to be change each time depend on what you want to plot
p
#save
#ggsave(filename = "output/SjD/cd45neg/SjD_specific_ann_cluster_name.png", plot = p, width = 16, height = 8) 

```






```{r}
#-----------------
##plot for annotations
#1for the major annotation

Idents(cd45neg)="specific_annot"
cd45neg$celltype=Idents(cd45neg)  

#color for major annotation
celltype_cols = c(
  "VSMC"                                    = "#D73027",
  "Capillary pericytes"                     = "#4575B4",
  "Immunopericytes"                         = "#1B9E77",
  "Neuronal cells"                          = "#762A83",
  "Endothelial Progenitors"                 = "#FDAE61",
  "Arterial Endothelial"                    = "#E7298A",
  "Lymphatic Endothelial cells"             = "#66A61E",
  "CCL14 Venous Endothelial"                = "#E6AB02",
  "Venous Endothelial"                      = "#A6761D",
  "Fibroblasts_3"                           = "#666666",
  "Fibroblasts_4"                           = "#1F78B4",
  "Serous Acinar Epithelial cells"          = "#B2DF8A",
  "Intercalated Ductal Epithelial cells"    = "#33A02C",
  "Seromucous Acinar Epithelial cells"      = "#FB9A99",
  "Striated Ductal Epithelial cells"        = "#E31A1C",
  "Basal Epithelial cells"                  = "#FDBF6F",
  "Mucous Acinar Epithelial cells"          = "#FF7F00",
  "Epithelial Progenitor cells"             = "#CAB2D6",
  "Ionocytes"                               = "#6A3D9A",
  "Myoepithelial cells"                     = "#B15928",
  "Fibroblasts_1"                           = "#8DA0CB",
  "Fibroblasts_2"                           = "#66C2A5",
  "Fibroblasts_5"                           = "#FC8D62",
  "SPARC+ COL3A1 + Fibroblasts"             = "#E78AC3"
)

 #umap
umap_df=as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45neg$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45neg/SjD_specific_ann.png", plot = fig, width =12, height = 8 )

```







###2 proportion of cells
```{r}
#------
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45neg_meta <- dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45neg_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45neg/SjD_specific_ann_Proport.png", plot = fig1c, width = 12, height = 8)

```



###3correlation between cell_types
```{r}
#---
#join layers(we have a lot of assays)
cd45neg = JoinLayers(
  cd45neg,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45neg/SjD_specific_cell_cor.png",
  width = 3600,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()

```




###4Donor proportions
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45neg@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```





```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45neg/SjD_specific_cell_Donor.png", plot = fig, width = 12, height = 8)
#

View(cd45neg@meta.data)
```
"orig.ident"                        "nCount_RNA"                        "nFeature_RNA"   "percent.mt"   "seurat_clusters"  "major_annot"    

##save the new meta data
```{r}
#-------
##check the information


# change_name
cd45neg_new <- cd45neg

# 
cd45neg_new$orig.ident <- "CD45-"

# keep col
cols_to_keep <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mt",
  "seurat_clusters",
  "specific_annot"
)

# keep
cd45neg_new@meta.data <- cd45neg_new@meta.data[, cols_to_keep, drop = FALSE]
colnames(cd45neg_new@meta.data)[
  colnames(cd45neg_new@meta.data) == "specific_annot"
] <- "celltype"
# save
saveRDS(cd45neg_new, file = "data/process/SeuratObjects/SjD/cd45neg_new.rds")

#
cd45neg_new <- readRDS("data/process/SeuratObjects/SjD/cd45neg_new.rds")

# Export CSV (optionnel)
write.csv(
  cd45neg_new@meta.data,
  file = "data/process/SeuratObjects/SjD/cd45neg_new.csv",
  row.names = TRUE
)

# Visualisation
View(cd45neg_new@meta.data)


```





##save
```{r}



#load later
cd45neg=readRDS("data/process/SeuratObjects/SjD/cd45neg_new.rds")



```









#####------------------------#################------------------------------------____----------------------------------------------- 

###STEP_2_Analysis for the CD45 Positive######################################################

#####----------------------------------------------------------------------------------------------------------------------------------------------------











####For  major_annot_joint#####
##1 load the data annotation for
```{r}
#---load the data
cd45pos=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45pos_annotated_clean.RDS")

#colnames
colnames(cd45pos@meta.data)
View(cd45pos@meta.data)
"orig.ident"                       "nCount_RNA"                       "nFeature_RNA"   "percent.mt"    "seurat_clusters"   "sample_name"      "major_annot"    
```



###2annotation


```{r, fig.width=20, fig.height=8}
#marker__________________________
#to see col to plot 
colnames(cd45pos@meta.data)

genes = c(
  "ZG16B","IGHM","IGHD","CCR7","GPR183","TEX9","MYC","KLF6","MIR155HG","ITGB1",
  "FOXP1","BACH2","CLEC9A","WDFY4","IRF8","CLEC4C","THBD","BATF3","CD1C",
  "FCER1A","FCGR3A","CD14","CX3CR1","CD163","CD36","FCGR1A","MERTK","CD80",
  "ZEB2","LAMP3","MRC1","SIRPA","CLEC10A","C1QA","IL1B","IGHA1","IGHG1",
  "IGHG3","IGHG4","TNF","NCAM1","NKG7","CD27","TRDC","KLRG1","FXYD7","KIT",
  "AHR","FGFBP2","FCRL6","CD8A","CD8B","MYB","SELL","TOX","IKZF2","HAVCR2",
  "CTLA4","ZNF683","ITGAE","CXCR6","CCL4L2","CCL4","CCL3","PRDM1","IFNG",
  "CD69","CCR6","XCL1","XCL2","CCL20","CLEC2B","NR4A1","HSPA6","BAG3",
  "GZMH","LGALS1","LEF1","TCF7","ANXA2","CD44","GZMA","CCL5","DLGAP1",
  "RORA","RORC","IL17A","FOXP3","CCR8","TIGIT","TNFRSF9","IL7R","CXCR5",
  "BCL6","SH2D1A","MAF","CXCL13","IL21","IL2RG","ITGA4","ITGB7"
)


p=DotPlot(
  cd45pos,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "major_annot_cl") + RotatedAxis() 
p
#save
ggsave(filename = "output/SjD/cd45pos/SjD_Major_ann_cluster_name.png", plot = p, width = 20, height = 8) 


```




```{r}
##plot for annotations
#1for the major annotation

Idents(cd45pos)="major_annot_joint"
cd45pos$celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols=c(
"B cells"                       = "#D73027",
"CD8"                           = "#FC8D62",
"CD4"                           = "#1B9E77",
"gd/NK"                         = "#762A83",
"Plasma cells"                  = "#FDAE61",
"Proliferating cells"           = "#E7298A",
"Mono/Mac"                      = "#FB9A99"
)
 #umap
umap_df=as.data.frame(Embeddings(cd45pos, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45pos$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45pos/SjD_Major_ann.png", plot = fig, width =12, height = 8 )

```



###3 Proportion of cells
```{r}
####plot
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45pos_meta <- dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45pos_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45pos/SjDMajorproport.png", plot = fig1c, width = 12, height = 8)

```




###4 correlation between cell_types

```{r}
#---
#join layers(we have a lot of assays)
cd45pos = JoinLayers(
  cd45pos,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/SjDMajorcor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()

```




###5Donor proportions
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45pos@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```




```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45pos/SjD_Major_Donor.png", plot = fig, width = 12, height = 8)


```



"specific_annot"



####process for the specific_annotation in cd45_positive#################
specific_annot

###1annotation
```{r}

##plot for annotations
#1for the major annotation

Idents(cd45pos)="specific_annot"
cd45pos$celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols = c(
  "Activated Treg cells"                   = "#E76F51",
  "CD16+ NK cells"                         = "#F4A261",
  "CD27+ NK cells"                         = "#E9C46A",
  "CD56+ NK cells"                         = "#D4A373",
  "CD62L+ TPEx"                            = "#CDB4DB",
  "CD8+ Tcm"                               = "#90DBF4",
  "CD8+ Tem"                               = "#48CAE4",
  "CD8+ Trm"                               = "#0096C7",
  "cDC1"                                   = "#A3B18A",
  "cDC2"                                   = "#588157",
  "Cytotoxic Th cells"                     = "#6A994E",
  "DLGAP1+ Th cells"                       = "#A7C957",
  "Exhausted CD8+ T cells"                 = "#6D597A",
  "Exhausted IEG CD8+ T cells"             = "#B56576",
  "GC B cells_1"                           = "#4CC9F0",
  "GC B cells_2"                           = "#4895EF",
  "GC B cells_3"                           = "#4361EE",
  "GC B cells_4"                           = "#3A0CA3",
  "gd-T cells"                             = "#2A9D8F",
  "IgA Plasma cells"                       = "#F72585",
  "IgG1 Plasma cells"                      = "#FF7096",
  "IgG3/4 Plasma cells"                    = "#FF8FAB",
  "IgM Plasma cells"                       = "#FFC2D1",
  "ILCs"                                   = "#52B788",
  "Infiltrating CD8+ T cells"              = "#1B9AAA",
  "Inflammatory CD8+ T cells"              = "#0077B6",
  "inflammatory cDC2"                      = "#6C757D",
  "LAMP3+ cDCs"                            = "#ADB5BD",
  "LTB+ Th17 cells"                        = "#9D4EDD",
  "M1 Macrophages"                         = "#E63946",
  "M2 Macrophages"                         = "#A8DADC",
  "Memory Th cells"                        = "#457B9D",
  "Migrating Th17 cells"                   = "#5E60CE",
  "Mo-DC"                                  = "#7400B8",
  "Monocytes"                              = "#B5179E",
  "Naive B cells"                          = "#F1FAEE",
  "Naive CD8+ T cells"                     = "#8ECAE6",
  "Naive Tfh cells"                        = "#219EBC",
  "Naive Th cells"                         = "#023047",
  "Proliferating cells"                    = "#FFBE0B",
  "Recruited Activated CD8-"               = "#FB5607",
  "Recruiting Effector CD8+"               = "#FF006E",
  "Resting Treg cells"                     = "#8338EC",
  "Secretory Tph cells"                    = "#3A86FF",
  "TNF+ Plasma cells"                      = "#D00000",
  "Tph17"                                  = "#9B2226"
)


 #umap
umap_df=as.data.frame(Embeddings(cd45pos, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45pos$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45pos/SjD_specific_ann.png", plot = fig, width =12, height = 8 )

```



###2proportions of cells
```{r}
####plot
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45pos_meta <- dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45pos_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45pos/SjD_specific_cell_proport.png", plot = fig1c, width = 12, height = 8)

```





###3correlation between celltype
```{r}
#---
#join layers(we have a lot of assays)
cd45pos = JoinLayers(
  cd45pos,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/SjD_specific_cell_cor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()
```





###4Donor proportion
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45pos@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```



```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45pos/SjD_specific_Donor.png", plot = fig, width = 12, height = 8)

##meta data

cd45pos$orig.ident="CD45 +"


```




##save new meta.dat
```{r}
#

# change_name
cd45pos_new <- cd45pos

# 
cd45pos_new$orig.ident <- "CD45+"

# keep col
cols_to_keep <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "percent.mt",
  "seurat_clusters",
  "specific_annot"
)

# keep
cd45pos_new@meta.data <- cd45pos_new@meta.data[, cols_to_keep, drop = FALSE]
colnames(cd45pos_new@meta.data)[
  colnames(cd45pos_new@meta.data) == "specific_annot"
] <- "celltype"

# save
saveRDS(cd45pos_new, file = "data/process/SeuratObjects/SjD/cd45pos_new.rds")

#
cd45pos_new <- readRDS("data/process/SeuratObjects/SjD/cd45pos_new.rds")

# Export CSV (optionnel)
write.csv(
  cd45pos_new@meta.data,
  file = "data/process/SeuratObjects/SjD/cd45pos_new.csv",
  row.names = TRUE
)

# Visualisation
View(cd45pos_new@meta.data)

```




###save
```{r}
###


#load later
#cd45pos=readRDS("data/process/SeuratObjects/SjD/cd45pos_new.rds")

```













