---
title: "R Notebook"
output: html_notebook
---

###Sjogrens Diseases############


##packages
```{r setup, include=FALSE}
#-----
library(pacman)
  p_load(Seurat, SeuratData, Matrix, dplyr, DropletUtils, presto,gprofiler2, ggplot2, patchwork, gplots, cowplot, openxlsx,plyr, scDblFinder, BiocParallel,devtools,openxlsx,reticulate,hdf5r,DropletUtils)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```


###STEP_1Analysis for the CD45 NEGATIVE#############

##1-load the obj CD45 negative
```{r}
#-----
cd45neg=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45neg_annotated_clean.RDS")
cd45neg
#to see all annotations
colnames(cd45neg@meta.data)
p=DimPlot(cd45neg, label = TRUE)
ggsave("output/SjD/cd45neg/SjDintegration.png",plot = p, width = 12, height = 8)
```

We have two type of cell_annotattion the :major_annot and specif_annot(look for colnames(cd45neg@meta.data) to see the colnames)


###process for the major_annotation#####

###2 Annotation 
```{r}
##plot for annotations
#1for the major annotation

Idents(cd45neg)="major_annot_joint"
cd45neg$celltype=Idents(cd45neg)  

#color for major annotation
celltype_cols=c(
"Fibroblasts"          = "#D73027",
"Endothelial_cells"    = "#4575B4",
"Mural"                = "#1B9E77",
"Acinar_cells"         = "#762A83",
"Ductal_cells"         = "#FDAE61",
"Neuronal_cells"       = "#E7298A"
)
 #umap
umap_df=as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45neg$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig_major=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig_major
ggsave("output/SjD/cd45neg/SjD_Major_ann.png", plot = fig_major, width =12, height = 8 )
```





###3 proportion of cells
```{r}

library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45neg_meta <- dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45neg_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45neg/SjD_Major_proport.png", plot = fig1c, width = 12, height = 8)

```



###4correlation between celltype
```{r}
#---
#join layers(we have a lot of assays)
cd45neg <- JoinLayers(
  cd45neg,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45neg/SjD_Major_cor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()


```



##5Donor proportions
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45neg@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

```


```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45neg/Sj_Don_Major.png", plot = fig, width = 12, height = 8)

```









###process for the specific_annotation#####
specific_annot

##1annotation

```{r}
##plot for annotations
#1for the major annotation

Idents(cd45neg)="specific_annot"
cd45neg$celltype=Idents(cd45neg)  

#color for major annotation
celltype_cols = c(
  "VSMC"                                    = "#D73027",
  "Capillary pericytes"                     = "#4575B4",
  "Immunopericytes"                         = "#1B9E77",
  "Neuronal cells"                          = "#762A83",
  "Endothelial Progenitors"                 = "#FDAE61",
  "Arterial Endothelial"                    = "#E7298A",
  "Lymphatic Endothelial cells"             = "#66A61E",
  "CCL14 Venous Endothelial"                = "#E6AB02",
  "Venous Endothelial"                      = "#A6761D",
  "Fibroblasts_3"                           = "#666666",
  "Fibroblasts_4"                           = "#1F78B4",
  "Serous Acinar Epithelial cells"          = "#B2DF8A",
  "Intercalated Ductal Epithelial cells"    = "#33A02C",
  "Seromucous Acinar Epithelial cells"      = "#FB9A99",
  "Striated Ductal Epithelial cells"        = "#E31A1C",
  "Basal Epithelial cells"                  = "#FDBF6F",
  "Mucous Acinar Epithelial cells"          = "#FF7F00",
  "Epithelial Progenitor cells"             = "#CAB2D6",
  "Ionocytes"                               = "#6A3D9A",
  "Myoepithelial cells"                     = "#B15928",
  "Fibroblasts_1"                           = "#8DA0CB",
  "Fibroblasts_2"                           = "#66C2A5",
  "Fibroblasts_5"                           = "#FC8D62",
  "SPARC+ COL3A1 + Fibroblasts"             = "#E78AC3"
)

 #umap
umap_df=as.data.frame(Embeddings(cd45neg, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45neg$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45neg/SjD_specific_ann.png", plot = fig, width =12, height = 8 )

```




###2 proportion of cells
```{r}
#------
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45neg_meta <- dplyr::as_tibble(cd45neg@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45neg_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45neg/SjD_specific_ann_Proport.png", plot = fig1c, width = 12, height = 8)
```



###3correlation between cell_types
```{r}
#---
#join layers(we have a lot of assays)
cd45neg = JoinLayers(
  cd45neg,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45neg,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45neg/SjD_specific_cell_cor.png",
  width = 3600,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()

```


###4Donor proportions
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45neg@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```


```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45neg/SjD_specific_cell_Donor.png", plot = fig, width = 12, height = 8)
#

```

##save
```{r}
###
#saveRDS(cd45neg, file = "data/process/SeuratObjects/SjD/cd45neg_process.rds")

#load later
cd45neg=readRDS("data/process/SeuratObjects/SjD/cd45neg_process.rds")
```









#####------------------------#################------------------------------------____----------------------------------------------- 

###STEP_2_Analysis for the CD45 Positive######################################################

#####----------------------------------------------------------------------------------------------------------------------------------------------------

####For  major_annot_joint#####
##1 load the data annotation for
```{r}
#---load the data
cd45pos=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SjD/cd45pos_annotated_clean.RDS")

#colnames
colnames(cd45pos@meta.data)

```



###2annotation
```{r}
##plot for annotations
#1for the major annotation

Idents(cd45pos)="major_annot_joint"
cd45pos$celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols=c(
"B cells"                       = "#D73027",
"CD8"                           = "#FC8D62",
"CD4"                           = "#1B9E77",
"gd/NK"                         = "#762A83",
"Plasma cells"                  = "#FDAE61",
"Proliferating cells"           = "#E7298A",
"Mono/Mac"                      = "#FB9A99"
)
 #umap
umap_df=as.data.frame(Embeddings(cd45pos, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45pos$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4,
    box.padding = 0.4,
    label.size = 0.30,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45pos/SjDMajor_ann.png", plot = fig_major, width =12, height = 8 )

```



###3 Proportion of cells
```{r}
####plot
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45pos_meta <- dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45pos_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45pos/SjDMajorproport.png", plot = fig1c, width = 12, height = 8)

```




###4 correlation between cell_types

```{r}
#---
#join layers(we have a lot of assays)
cd45pos = JoinLayers(
  cd45pos,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/SjDMajorcor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()

```


###5Donor proportions
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45pos@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```


```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45pos/SjD_Major_Donor.png", plot = fig, width = 12, height = 8)


```






####process for the specific_annotation in cd45_positive#################
specific_annot

###1annotation
```{r}

##plot for annotations
#1for the major annotation

Idents(cd45pos)="specific_annot"
cd45pos$celltype=Idents(cd45pos)  

#color for major annotation
celltype_cols = c(
  "Activated Treg cells"                   = "#E76F51",
  "CD16+ NK cells"                         = "#F4A261",
  "CD27+ NK cells"                         = "#E9C46A",
  "CD56+ NK cells"                         = "#D4A373",
  "CD62L+ TPEx"                            = "#CDB4DB",
  "CD8+ Tcm"                               = "#90DBF4",
  "CD8+ Tem"                               = "#48CAE4",
  "CD8+ Trm"                               = "#0096C7",
  "cDC1"                                   = "#A3B18A",
  "cDC2"                                   = "#588157",
  "Cytotoxic Th cells"                     = "#6A994E",
  "DLGAP1+ Th cells"                       = "#A7C957",
  "Exhausted CD8+ T cells"                 = "#6D597A",
  "Exhausted IEG CD8+ T cells"             = "#B56576",
  "GC B cells_1"                           = "#4CC9F0",
  "GC B cells_2"                           = "#4895EF",
  "GC B cells_3"                           = "#4361EE",
  "GC B cells_4"                           = "#3A0CA3",
  "gd-T cells"                             = "#2A9D8F",
  "IgA Plasma cells"                       = "#F72585",
  "IgG1 Plasma cells"                      = "#FF7096",
  "IgG3/4 Plasma cells"                    = "#FF8FAB",
  "IgM Plasma cells"                       = "#FFC2D1",
  "ILCs"                                   = "#52B788",
  "Infiltrating CD8+ T cells"              = "#1B9AAA",
  "Inflammatory CD8+ T cells"              = "#0077B6",
  "inflammatory cDC2"                      = "#6C757D",
  "LAMP3+ cDCs"                            = "#ADB5BD",
  "LTB+ Th17 cells"                        = "#9D4EDD",
  "M1 Macrophages"                         = "#E63946",
  "M2 Macrophages"                         = "#A8DADC",
  "Memory Th cells"                        = "#457B9D",
  "Migrating Th17 cells"                   = "#5E60CE",
  "Mo-DC"                                  = "#7400B8",
  "Monocytes"                              = "#B5179E",
  "Naive B cells"                          = "#F1FAEE",
  "Naive CD8+ T cells"                     = "#8ECAE6",
  "Naive Tfh cells"                        = "#219EBC",
  "Naive Th cells"                         = "#023047",
  "Proliferating cells"                    = "#FFBE0B",
  "Recruited Activated CD8-"               = "#FB5607",
  "Recruiting Effector CD8+"               = "#FF006E",
  "Resting Treg cells"                     = "#8338EC",
  "Secretory Tph cells"                    = "#3A86FF",
  "TNF+ Plasma cells"                      = "#D00000",
  "Tph17"                                  = "#9B2226"
)


 #umap
umap_df=as.data.frame(Embeddings(cd45pos, "umap_h"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=cd45pos$celltype

#center
centers=aggregate(cbind(umap_1, umap_2) ~ celltype,data = umap_df,FUN = median)
  
library(ggrepel)
library(ggplot2)

fig=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umap_1, y = umap_2, label = celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA,
    max.overlaps = Inf
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP 1", y = "UMAP 2")
fig
ggsave("output/SjD/cd45pos/SjD_specific_ann.png", plot = fig, width =12, height = 8 )
```



###2proportions of cells
```{r}
####plot
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::count(`sample_name`, celltype) %>%
  dplyr::group_by(`sample_name`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
cd45pos_meta <- dplyr::as_tibble(cd45pos@meta.data) %>%
  dplyr::rename(sample_id = sample_name)   # rename to avoid conflicts
prop_df=cd45pos_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SjD/cd45pos/SjD_specific_cell_proport.png", plot = fig1c, width = 12, height = 8)
```




###3correlation between celltype
```{r}
#---
#join layers(we have a lot of assays)
cd45pos = JoinLayers(
  cd45pos,
  assay = "RNA"
)

#average expression
agg.exp = AggregateExpression( cd45pos,assays = "RNA",layers = "counts", group.by = "celltype")$RNA
dim(agg.exp)
colnames(agg.exp)
 #correlation 
agg.exp.mat = as.matrix(agg.exp)
agg.exp.mat = log1p(agg.exp.mat)

cor_mat = cor(
  agg.exp.mat,
  method = "pearson",
  use = "pairwise.complete.obs"
)

cor_mat[is.na(cor_mat)] = 0

#avoid error
set.seed(1)
cor_mat = cor_mat + matrix(
  runif(length(cor_mat), 1e-8, 1e-7),
  nrow = nrow(cor_mat)
)
 #save 
png(
  filename = "output/SjD/cd45pos/SjD_specific_cell_cor.png",
  width = 3000,
  height = 2600,
  res = 300
)
#plot
library(pheatmap)
pheatmap(
  cor_mat,
  clustering_method = "ward.D2",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  fontsize_row = 12,
  fontsize_col = 12,
  angle_col = 45,
  border_color = NA
)
dev.off()
```



###4Donor proportion
```{r}
##------
library(dplyr)

prop_donor_celltype <- cd45pos@meta.data %>%
  as.data.frame() %>%
  dplyr::count(celltype, sample_name) %>%   # number of cells
  dplyr::group_by(celltype) %>%              # normalize within celltype
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
```



```{r}
#----plot
fig = ggplot(
  prop_donor_celltype,
  aes(x = celltype, y = proportion, fill = sample_name)
) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  ylab("Proportion of cells") +
  xlab("") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

fig

ggsave("output/SjD/cd45pos/SjD_specific_Donor.png", plot = fig, width = 12, height = 8)

##meta data
View(cd45pos@meta.data)
```



###save
```{r}
###
#saveRDS(cd45pos, file = "data/process/SeuratObjects/SjD/cd45_process.rds")

#load later
readRDS("data/process/SeuratObjects/SjD/cd45_process.rds")
```













