---
title: "SScILD - Processing and Annotation"
output: html_notebook
---

<!-- Step 2: Seurat processing for Systemic Sclerosis-associated Interstitial
     Lung Disease (SScILD) - from raw count matrices to annotated, merged
     Seurat objects, processed separately for the Upper and Lower lung lobes.
     The cell-cell communication (CCC) analysis is a separate step, in
     SScILD_sce.Rmd. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

## 2. Load raw data, create per-sample Seurat objects, and merge by lobe

```{r}
# load raw count matrices - A is one dataset (h5 format), B has 4 paired
# samples (10X directories). L = Lower lung, U = Upper lung
AL1 <- Read10X_h5("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/A/Lower lung/SSc-ILD-A-Low/feature_bc_matrix.h5", use.names = TRUE, unique.features = TRUE)
AU1 <- Read10X_h5("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/A/Upper lung/GSE156310_RAW/raw_feature_bc_matrix.h5", use.names = TRUE, unique.features = TRUE)
B1L <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/1L")
B1U <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/1U")
B2L <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/2L")
B2U <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/2U")
B3L <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/3L")
B3U <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/3U")
B4L <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/4L")
B4U <- Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/4U")

sce_list <- list(
  AL1 = AL1, AU1 = AU1,
  B1L = B1L, B1U = B1U,
  B2L = B2L, B2U = B2U,
  B3L = B3L, B3U = B3U,
  B4L = B4L, B4U = B4U
)

# empty-droplet detection with DropletUtils, per sample
library(DropletUtils)
droplet_results <- lapply(sce_list, function(x) {
  iscell_dd <- defaultDrops(x, expected = 5000)

  eout <- emptyDrops(x, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed <- eout$FDR <= 0.01

  list(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    meta = data.frame(iscell_dd = iscell_dd, iscell_ed = iscell_ed)
  )
})
names(droplet_results) <- names(sce_list)

# keep only cells called by emptyDrops, then build one Seurat object per sample
library(Seurat)
library(SingleCellExperiment)

filtered_counts <- lapply(names(sce_list), function(nm) {
  mat <- sce_list[[nm]]
  keep <- droplet_results[[nm]]$iscell_ed
  keep[is.na(keep)] <- FALSE
  mat[, keep, drop = FALSE]
})
names(filtered_counts) <- names(sce_list)

seurat_list <- lapply(names(filtered_counts), function(nm) {
  CreateSeuratObject(counts = filtered_counts[[nm]], project = nm)
})
names(seurat_list) <- names(filtered_counts)

# tag each object with its lung region, from the U/L suffix in the sample name
for (nm in names(seurat_list)) {
  seurat_list[[nm]]$lung_region <- ifelse(grepl("U$", nm), "Upper", "Lower")
}

# save each per-sample object to its lobe folder
for (nm in names(seurat_list)) {
  seurat_obj <- seurat_list[[nm]]
  if (seurat_obj$lung_region[1] == "Lower") {
    saveRDS(seurat_obj, file = paste0("data/process/SeuratObjects/SScILD/Low/", nm, ".rds"))
  } else if (seurat_obj$lung_region[1] == "Upper") {
    saveRDS(seurat_obj, file = paste0("data/process/SeuratObjects/SScILD/Upper/", nm, ".rds"))
  }
}

# merge Upper lung samples into one object
SS.U <- readRDS("data/process/SeuratObjects/SScILD/Upper/AU1.rds")
SS.U$sample <- "AU1"
upper_names <- c("B1U", "B2U", "B3U", "B4U")

for (nm in upper_names) {
  obj <- readRDS(paste0("data/process/SeuratObjects/SScILD/Upper/", nm, ".rds"))
  obj$sample <- nm
  SS.U <- merge(SS.U, y = obj)
}
saveRDS(SS.U, "data/process/SeuratObjects/SScILD/Upper/SS.U.rds")

# merge Lower lung samples into one object
SS.L <- readRDS("data/process/SeuratObjects/SScILD/Low/AL1.rds")
SS.L$sample <- "AL1"
lower_names <- c("B1L", "B2L", "B3L", "B4L")

for (nm in lower_names) {
  obj <- readRDS(paste0("data/process/SeuratObjects/SScILD/Low/", nm, ".rds"))
  obj$sample <- nm
  SS.L <- merge(SS.L, y = obj)
}
saveRDS(SS.L, "data/process/SeuratObjects/SScILD/Low/SS.L.rds")

# free the per-sample objects, no longer needed
rm(obj, seurat_list, seurat_obj, nm, upper_names, lower_names)
gc(full = TRUE)
```

################################## Upper lobe ##########################################

## 3. Quality control

```{r}
SS.U <- readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U.rds")
DefaultAssay(SS.U) # if not RNA, set it as RNA

SS.U[["percent.mt"]] <- PercentageFeatureSet(SS.U, pattern = "^MT-")

Idents(SS.U) <- "x"
SS.U$x <- "All"
VlnPlot(SS.U, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC thresholds set for this lobe, based on its own distribution
SS.U <- subset(SS.U, subset = nFeature_RNA > 300 & nFeature_RNA < 7000 & percent.mt < 15)
```

## 4. Normalization, scaling, and PCA

```{r}
SS.U <- NormalizeData(SS.U)
SS.U <- FindVariableFeatures(SS.U)
SS.U <- ScaleData(SS.U, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = SS.U))
SS.U <- RunPCA(SS.U, features = VariableFeatures(object = SS.U))
ElbowPlot(SS.U, ndims = 50, reduction = "pca")
SS.U <- RunUMAP(SS.U, dims = 1:30)
```

## 5. Harmony integration and clustering resolutions

```{r}
library(harmony)
SS.U <- RunHarmony(SS.U, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:30)
SS.U[["RNA"]] <- JoinLayers(SS.U[["RNA"]])
SS.U <- RunUMAP(SS.U, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")

# scan a range of resolutions before picking one
SS.U <- FindNeighbors(SS.U, reduction = "harmony", dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS.U <- FindClusters(SS.U, resolution = res)
  colname <- paste0("res_", res)
  SS.U[[colname]] <- SS.U$seurat_clusters
}
res_cols <- grep("^res_", colnames(SS.U@meta.data), value = TRUE)
for (res in res_cols) {
  p <- DimPlot(SS.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}

# save all resolutions to a single PDF for visual comparison
pdf("output/SScILD/Upper/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(SS.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}
dev.off()

# resolution 0.6 was chosen for downstream clustering
p <- DimPlot(SS.U, group.by = "res_0.6", label = TRUE, label.size = 5, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.6") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/SScILD/Upper/SSU_resolution.pdf", plot = p, width = 12, height = 8, dpi = 300)
```

## 6. Clustering and marker genes

```{r}
SS.U <- FindNeighbors(SS.U, reduction = "harmony", dims = 1:30)
SS.U <- FindClusters(SS.U, resolution = 0.6, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
SS.U_markers <- FindAllMarkers(SS.U, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
SS.U_markers <- SS.U_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 7. Cell type annotation

```{r}
genes <- c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T CD8
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T CD4
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Secretory
  # "SCGB1A1", "MUC5B", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  # Pericyte
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)
```

```{r}
l <- FeaturePlot(SS.U, features = unique(genes), combine = FALSE, order = TRUE, reduction = "umap.harmony")
names(l) <- unique(genes)

# look at CD4 markers to guide subclustering
FeaturePlot(SS.U, features = c("CD4", "COTL1", "CCR7", "LEF1"), combine = FALSE, order = TRUE, reduction = "umap.harmony")

Idents(SS.U) <- SS.U$harmony_clusters
SS.U <- FindSubCluster(SS.U, cluster = c(1), resolution = 0.1, graph.name = "RNA_snn")

P <- DimPlot(SS.U, group.by = "sub.cluster", reduction = "umap.harmony", label = TRUE, label.size = 5) +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/SScILD/Upper/SSU_sub.cluster.pdf", plot = P, width = 12, height = 8)
```

```{r}
p <- DotPlot(
  SS.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "sub.cluster"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SScILD/Upper/SSU.cluster.pdf", plot = p, width = 20, height = 8)
```

```{r}
# fine-grained cell type labels (cluster number kept in the name)
Idents(SS.U) <- "sub.cluster"
cluster_name_x <- c(
  "0" = "Macrophages_0",
  "1_0" = "CD4_1_0",
  "1_1" = "CD8_1_1",
  "2" = "Macrophages_2",
  "3" = "Monocytes_3",
  "4" = "Macrophages_4",
  "5" = "Capillary Endothelial_5",
  "6" = "Macrophages_6",
  "7" = "Basal Cells_7",
  "8" = "NK Cells_8",
  "9" = "Fibroblasts_9",
  "10" = "Peribronchial Endothelial_10",
  "11" = "Ciliated Cells_11",
  "12" = "Alveolar Cells_12",
  "13" = "Mural Cells_13",
  "14" = "Mast Cells_14",
  "15" = "Proliferating Cells_15",
  "16" = "Mural Cells_16",
  "17" = "Macrophages_17",
  "18" = "Lymphatic Endothelial_18",
  "19" = "Myeloid Cells_19",
  "20" = "B Cells_20",
  "21" = "Plasmacytoid Dendritic"
)
SS.U <- RenameIdents(SS.U, cluster_name_x)
SS.U$CellType <- Idents(SS.U)

p <- DotPlot(
  SS.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "CellType"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SScILD/Upper/SSU_cluster_name.pdf", plot = p, width = 20, height = 10)
```

```{r}
# major cell types (clusters of the same type merged, no cluster number)
Idents(SS.U) <- "sub.cluster"
cluster_name <- c(
  "0" = "Macrophages",
  "1_0" = "CD4",
  "1_1" = "CD8",
  "2" = "Macrophages",
  "3" = "Monocytes",
  "4" = "Macrophages",
  "5" = "Capillary Endothelial",
  "6" = "Macrophages",
  "7" = "Basal Cells",
  "8" = "NK Cells",
  "9" = "Fibroblasts",
  "10" = "Peribronchial Endothelial",
  "11" = "Ciliated Cells",
  "12" = "Alveolar Cells",
  "13" = "Mural Cells",
  "14" = "Mast Cells",
  "15" = "Proliferating Cells",
  "16" = "Mural Cells",
  "17" = "Macrophages",
  "18" = "Lymphatic Endothelial",
  "19" = "Myeloid Cells",
  "20" = "B Cells",
  "21" = "Plasmacytoid Dendritic"
)
SS.U <- RenameIdents(SS.U, cluster_name)
SS.U$Major_Celltype <- Idents(SS.U)

celltype_cols <- c(
  "Macrophages" = "orangered",
  "CD8" = "tomato",
  "CD4" = "olivedrab2",
  "Monocytes" = "#2ca02c",
  "Capillary Endothelial" = "#7f7f7f",
  "Basal Cells" = "indianred3",
  "NK Cells" = "magenta",
  "Fibroblasts" = "cyan",
  "Peribronchial Endothelial" = "darkgoldenrod2",
  "Ciliated Cells" = "saddlebrown",
  "Alveolar Cells" = "aquamarine",
  "Mural Cells" = "#1F78B4",
  "Mast Cells" = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Lymphatic Endothelial" = "mediumorchid1",
  "Myeloid Cells" = "#c5b0d5",
  "B Cells" = "gold",
  "Plasmacytoid Dendritic" = "peru"
)

# sanity check - confirm embedding column names before hardcoding them below
colnames(Embeddings(SS.U, "umap.harmony"))

# UMAP with a label placed at each cell type's centroid
umap_df <- as.data.frame(Embeddings(SS.U, "umap.harmony"))
umap_df$Major_Celltype <- SS.U$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4.5,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

ggsave(filename = "output/SScILD/Upper/SSU_cellType.pdf", plot = fig1a, width = 12, height = 10)
```

## 8. Marker expression heatmaps

```{r}
# fine-grained cell type heatmap
Idents(SS.U) <- "CellType"
SS.U$CellType <- gsub("-", "_", SS.U$CellType)
Idents(SS.U) <- "CellType"

DefaultAssay(SS.U) <- "RNA"
expr_mat <- GetAssayData(SS.U, layer = "data")

genes <- genes[genes %in% rownames(expr_mat)]
if (length(genes) < 2) stop("Too few genes found")

celltypes <- levels(factor(Idents(SS.U)))

# percent of cells expressing each gene, per cell type
percent_mat <- sapply(celltypes, function(ct) {
  cells <- WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per cell type
avg_exp <- AverageExpression(
  SS.U,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA
colnames(avg_exp) <- gsub("-", "_", colnames(avg_exp))

common_ct <- intersect(celltypes, colnames(avg_exp))
if (length(common_ct) < 2) stop("Cell type mismatch still exists")

percent_mat <- percent_mat[, common_ct, drop = FALSE]
exp_mat <- as.matrix(avg_exp[genes, common_ct, drop = FALSE])

# clean and scale expression
exp_mat[is.infinite(exp_mat)] <- NA
exp_mat <- exp_mat[rowSums(is.na(exp_mat)) < ncol(exp_mat), , drop = FALSE]
percent_mat <- percent_mat[rownames(exp_mat), , drop = FALSE]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
if (all(exp_mat_scaled == 0)) stop("No expression signal")
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# color scale from the expression quantiles
q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95), na.rm = TRUE)
q <- unique(as.numeric(q))
q <- q[is.finite(q)]
if (length(q) < 2) q <- range(exp_mat_scaled, na.rm = TRUE)
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# dot size encodes percent expressing, color encodes scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(common_ct))
names(cols) <- common_ct

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = common_ct,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf("output/SScILD/Upper/SSU_heatmap.pdf", width = 10, height = 15)
ComplexHeatmap::draw(ht)
dev.off()
```

```{r}
# same heatmap, grouped by the major (coarser) cell type annotation instead
Idents(SS.U) <- "Major_Celltype"

DefaultAssay(SS.U) <- "RNA"
expr_mat <- GetAssayData(SS.U, layer = "data")

percent_mat <- sapply(levels(Idents(SS.U)), function(ct) {
  cells <- WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SS.U,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA
exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf("output/SScILD/Upper/markers.pdf", width = 10, height = 15)
ComplexHeatmap::draw(ht)
dev.off()
```

## 9. Correlation between cell types

```{r}
Idents(SS.U) <- "Major_Celltype"
levels(Idents(SS.U))

# average expression per cell type
av.expSSU <- AverageExpression(SS.U, assays = "RNA", slot = "data")$RNA

correlation_matrixSSU <- cor(as.matrix(av.expSSU), use = "pairwise.complete.obs")
correlation_matrixSSU[is.na(correlation_matrixSSU)] <- 0

pdf(file = "output/SScILD/Upper/SSU_correlation.pdf", width = 10, height = 8)
pheatmap(
  correlation_matrixSSU,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)
dev.off()
```

## 10. Broader cell-type grouping (General)

```{r}
unique(SS.U$Major_Celltype)

SS.U$General <- recode(SS.U$Major_Celltype,
  "Monocytes" = "Mono/Mac",
  "Capillary Endothelial" = "Endothelial Cells",
  "Mast Cells" = "Mast Cells",
  "Basal Cells" = "Epithelial Cells",
  "Plasmacytoid Dendritic" = "Mono/Mac",
  "Macrophages" = "Mono/Mac",
  "Mural Cells" = "Mural Cells",
  "Myeloid Cells" = "Mono/Mac",
  "Alveolar Cells" = "Epithelial Cells",
  "CD8" = "CD8",
  "CD4" = "CD4",
  "B Cells" = "B Cells",
  "NK Cells" = "NK Cells",
  "Ciliated Cells" = "Epithelial Cells",
  "Fibroblasts" = "Fibroblasts",
  "Proliferating Cells" = "Proliferating Cells",
  "Peribronchial Endothelial" = "Endothelial Cells",
  "Lymphatic Endothelial" = "Endothelial Cells"
)
```

## 11. TLS (tertiary lymphoid structure) marker heatmap

```{r}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(germinal_center, inflammation, lymphoid_chemokines, hev, fdc))
markers <- markers[markers %in% rownames(SS.U)]

marker_groups <- case_when(
  markers %in% germinal_center ~ "Germinal Center",
  markers %in% inflammation ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev ~ "HEV",
  markers %in% fdc ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center" = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines" = "#35978F",
  "HEV" = "#8073AC",
  "FDC" = "#E08214"
)

Idents(SS.U) <- "General"
DefaultAssay(SS.U) <- "RNA"
expr_mat <- GetAssayData(SS.U, layer = "data")

percent_mat <- sapply(levels(Idents(SS.U)), function(ct) {
  cells <- WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SS.U,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/SScILD/Upper/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 12. Save the processed object

```{r}
SS.U$orig.ident <- "SScILD.U"
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General",
  "sub.cluster"
)
SS.U@meta.data <- SS.U@meta.data[, keep_cols]

write.csv(SS.U@meta.data,
  file = "data/process/SeuratObjects/SScILD/Upper/SS.U.metadata.csv",
  row.names = TRUE
)

saveRDS(SS.U, file = "data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")

# checkpoint: reload from here without rerunning the steps above
SS.U <- readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")
```

## 13. Export for gene regulatory network analysis (SCENIC)

```{r}
# export the raw counts as a sparse matrix + gene/barcode lists, so they can
# be imported by pySCENIC in the terminal
expr <- GetAssayData(SS.U, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SScILD/Upper/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SScILD/Upper/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SScILD/Upper/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

########################### Lower lobe #######################################

## 14. Quality control

```{r}
SS.L <- readRDS("data/process/SeuratObjects/SScILD/Low/SS.L.rds")
DefaultAssay(SS.L) # if not RNA, set it as RNA

SS.L[["percent.mt"]] <- PercentageFeatureSet(SS.L, pattern = "^MT-")

Idents(SS.L) <- "x"
SS.L$x <- "All"
VlnPlot(SS.L, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC thresholds set for this lobe, based on its own distribution
SS.L <- subset(SS.L, subset = nFeature_RNA > 300 & nFeature_RNA < 7000 & percent.mt < 15)
```

## 15. Normalization, scaling, and PCA

```{r}
SS.L <- NormalizeData(SS.L)
SS.L <- FindVariableFeatures(SS.L)
SS.L <- ScaleData(SS.L, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = SS.L))
SS.L <- RunPCA(SS.L, features = VariableFeatures(object = SS.L))
ElbowPlot(SS.L, ndims = 50, reduction = "pca")
SS.L <- RunUMAP(SS.L, dims = 1:30)
DimPlot(SS.L, reduction = "umap", label = TRUE)
```

## 16. Harmony integration and clustering resolutions

```{r}
library(harmony)
SS.L <- RunHarmony(SS.L, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:30)
SS.L[["RNA"]] <- JoinLayers(SS.L[["RNA"]])
SS.L <- RunUMAP(SS.L, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")

# scan a range of resolutions before picking one
SS.L <- FindNeighbors(SS.L, reduction = "harmony", dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS.L <- FindClusters(SS.L, resolution = res)
  colname <- paste0("res_", res)
  SS.L[[colname]] <- SS.L$seurat_clusters
}
res_cols <- grep("^res_", colnames(SS.L@meta.data), value = TRUE)
for (res in res_cols) {
  p <- DimPlot(SS.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}

# save all resolutions to a single PDF for visual comparison
pdf("output/SScILD/Low/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(SS.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}
dev.off()

# resolution 0.6 was chosen for downstream clustering
p <- DimPlot(SS.L, group.by = "res_0.6", label = TRUE, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.6") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/SScILD/Low/SSL.resolution.pdf", plot = p, width = 10, height = 8, dpi = 300)
```

## 17. Clustering and marker genes

```{r}
SS.L <- FindNeighbors(SS.L, reduction = "harmony", dims = 1:30)
SS.L <- FindClusters(SS.L, resolution = 0.6, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
SS.L_markers <- FindAllMarkers(SS.L, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
SS.L_markers <- SS.L_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 18. Cell type annotation

```{r}
genes <- c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Secretory
  # "SCGB1A1", "MUC5B", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  # Pericyte
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)
```

```{r}
l <- FeaturePlot(SS.L, features = unique(genes), combine = FALSE, order = TRUE, reduction = "umap.harmony")
names(l) <- unique(genes)

Idents(SS.L) <- SS.L$harmony_clusters
SS.L <- FindSubCluster(SS.L, cluster = c(3), resolution = 0.1, graph.name = "RNA_snn")

P <- DimPlot(SS.L, group.by = "sub.cluster", reduction = "umap.harmony", label = TRUE, label.size = 7) +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/SScILD/Low/SSL_sub.cluster.pdf", plot = P, width = 12, height = 8)
```

```{r}
p <- DotPlot(
  SS.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "sub.cluster"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SScILD/Low/SSL_cluster.pdf", plot = p, width = 20, height = 8)
```

```{r}
# fine-grained cell type labels (cluster number kept in the name)
Idents(SS.L) <- "sub.cluster"
cluster_name_x <- c(
  "0" = "Macrophages_0",
  "1" = "Macrophages_1",
  "2" = "Monocytes_2",
  "3_0" = "CD4_3_0",
  "3_1" = "CD8_3_1",
  "3_2" = "Macrophages_3_2",
  "4" = "Peribronchial Endothelial_4",
  "5" = "Ciliated Cells_5",
  "6" = "Alveolar Cells_6",
  "7" = "Proliferating Cells_7",
  "8" = "Goblet Cells_8",
  "9" = "Mural Cells_9",
  "10" = "Capillary Endothelial_10",
  "11" = "Fibroblasts_11",
  "12" = "Macrophages_12",
  "13" = "Basal Cells_13",
  "14" = "Mast Cells_14",
  "15" = "NK Cells_15",
  "16" = "Ciliated Cells_16",
  "17" = "B Cells_17",
  "18" = "Lymphatic Endothelial"
)
SS.L <- RenameIdents(SS.L, cluster_name_x)
SS.L$CellType <- Idents(SS.L)

p <- DotPlot(
  SS.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "CellType"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SScILD/Low/SSL_cluster_name.pdf", plot = p, width = 20, height = 10)
```

```{r}
# major cell types (clusters of the same type merged, no cluster number)
Idents(SS.L) <- "sub.cluster"
cluster_name <- c(
  "0" = "Macrophages",
  "1" = "Macrophages",
  "2" = "Monocytes",
  "3_0" = "CD4",
  "3_1" = "CD8",
  "3_2" = "Macrophages",
  "4" = "Peribronchial Endothelial",
  "5" = "Ciliated Cells",
  "6" = "Alveolar Cells",
  "7" = "Proliferating Cells",
  "8" = "Goblet Cells",
  "9" = "Mural Cells",
  "10" = "Capillary Endothelial",
  "11" = "Fibroblasts",
  "12" = "Macrophages",
  "13" = "Basal Cells",
  "14" = "Mast Cells",
  "15" = "NK Cells",
  "16" = "Ciliated Cells",
  "17" = "B Cells",
  "18" = "Lymphatic Endothelial"
)
SS.L <- RenameIdents(SS.L, cluster_name)
SS.L$Major_Celltype <- Idents(SS.L)

celltype_cols <- c(
  "Macrophages" = "orangered",
  "Monocytes" = "#2ca02c",
  "CD8" = "tomato",
  "CD4" = "olivedrab2",
  "Peribronchial Endothelial" = "darkgoldenrod2",
  "Ciliated Cells" = "saddlebrown",
  "Alveolar Cells" = "aquamarine",
  "Proliferating Cells" = "#B15928",
  "Goblet Cells" = "cornflowerblue",
  "Mural Cells" = "#1F78B4",
  "Capillary Endothelial" = "#7f7f7f",
  "Fibroblasts" = "cyan",
  "Basal Cells" = "indianred3",
  "Mast Cells" = "#FDBF6F",
  "NK Cells" = "magenta",
  "B Cells" = "gold",
  "Lymphatic Endothelial" = "mediumorchid1"
)

# UMAP with a label placed at each cell type's centroid
umap_df <- as.data.frame(Embeddings(SS.L, "umap.harmony"))
umap_df$Major_Celltype <- SS.L$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4.5,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

fig1a
ggsave(filename = "output/SScILD/Low/SSL_cellType.pdf", plot = fig1a, width = 12, height = 10)
```

## 19. Marker expression heatmap

```{r}
# heatmap grouped by the major (coarser) cell type annotation
Idents(SS.L) <- "Major_Celltype"

DefaultAssay(SS.L) <- "RNA"
expr_mat <- GetAssayData(SS.L, layer = "data")

percent_mat <- sapply(levels(Idents(SS.L)), function(ct) {
  cells <- WhichCells(SS.L, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SS.L,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA
exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf("output/SScILD/Low/SS.L_markers.pdf", width = 10, height = 15)
ComplexHeatmap::draw(ht)
dev.off()
```

## 20. Correlation between cell types

```{r}
Idents(SS.L) <- "Major_Celltype"
levels(Idents(SS.L))

# average expression per cell type
av.expSSL <- AverageExpression(SS.L, assays = "RNA", slot = "data")$RNA

correlation_matrixSSL <- cor(as.matrix(av.expSSL), use = "pairwise.complete.obs")
correlation_matrixSSL[is.na(correlation_matrixSSL)] <- 0

pdf(file = "output/SScILD/Low/SSL_correlation.pdf", width = 10, height = 8)
pheatmap(
  correlation_matrixSSL,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)
dev.off()
```

## 21. Broader cell-type grouping (General)

```{r}
unique(SS.L$Major_Celltype)

SS.L$General <- recode(SS.L$Major_Celltype,
  "Ciliated Cells" = "Epithelial Cells",
  "Macrophages" = "Mono/Mac",
  "Monocytes" = "Mono/Mac",
  "CD8" = "CD8",
  "CD4" = "CD4",
  "Capillary Endothelial" = "Endothelial Cells",
  "Mural Cells" = "Mural Cells",
  "Peribronchial Endothelial" = "Endothelial Cells",
  "Mast Cells" = "Mast Cells",
  "B Cells" = "B Cells",
  "Fibroblasts" = "Fibroblasts",
  "NK Cells" = "NK Cells",
  "Proliferating Cells" = "Proliferating Cells",
  "Alveolar Cells" = "Epithelial Cells",
  "Basal Cells" = "Epithelial Cells",
  "Goblet Cells" = "Epithelial Cells",
  "Lymphatic Endothelial" = "Endothelial Cells"
)
```

## 22. TLS (tertiary lymphoid structure) marker heatmap

```{r}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(germinal_center, inflammation, lymphoid_chemokines, hev, fdc))
markers <- markers[markers %in% rownames(SS.L)]

marker_groups <- case_when(
  markers %in% germinal_center ~ "Germinal Center",
  markers %in% inflammation ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev ~ "HEV",
  markers %in% fdc ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center" = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines" = "#35978F",
  "HEV" = "#8073AC",
  "FDC" = "#E08214"
)

Idents(SS.L) <- "General"
DefaultAssay(SS.L) <- "RNA"
expr_mat <- GetAssayData(SS.L, layer = "data")

percent_mat <- sapply(levels(Idents(SS.L)), function(ct) {
  cells <- WhichCells(SS.L, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SS.L,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/SScILD/Low/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 23. Export for gene regulatory network analysis (SCENIC)

```{r}
# export the raw counts as a sparse matrix + gene/barcode lists, so they can
# be imported by pySCENIC in the terminal
expr <- GetAssayData(SS.L, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SScILD/Low/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SScILD/Low/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SScILD/Low/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

## 24. Save the processed object

```{r}
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General",
  "sub.cluster"
)

SS.L$orig.ident <- "SScILD.L"
SS.L@meta.data <- SS.L@meta.data[, keep_cols]

write.csv(SS.L@meta.data,
  file = "data/process/SeuratObjects/SScILD/Low/SS.L.metadata.csv",
  row.names = TRUE
)

saveRDS(SS.L, file = "data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")

# checkpoint: reload from here without rerunning the steps above
SS.L <- readRDS("data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")
```
