---
title: "R Notebook"
output: html_notebook
---

################################################################################################################################################
######ANALYSIS OF SScILD.Up - Systemic Sclerosis-associated Interstitial Lung Disease in Upper Lung DATA ##########################################
##############################################################################################################################################


## packages
```{r setup, include=FALSE}
##################PACKAGES
library(pacman)
  p_load(Seurat, SeuratData, Matrix, dplyr, DropletUtils, presto,gprofiler2, ggplot2, patchwork, gplots, cowplot, openxlsx,plyr, scDblFinder, BiocParallel,devtools,openxlsx,reticulate,hdf5r,DropletUtils)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```



####STEP-1

###load the data
```{r}
##load data,create seurat obj for each one and merge theme in lower and upper lung
#1-L for Lower lung and U for Upper lung
AL1=Read10X_h5("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/A/Lower lung/SSc-ILD-A-Low/feature_bc_matrix.h5",use.names = TRUE, unique.features = TRUE)
AU1=Read10X_h5("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/A/Upper lung/GSE156310_RAW/raw_feature_bc_matrix.h5", use.names = TRUE, unique.features = TRUE)
B1L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/1L")
B1U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/1U")
B2L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/2L")
B2U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/2U")
B3L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/3L")
B3U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/3U")
B4L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/4L")
B4U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/4U")

#---Default dropletUtils
#1
sce_list=list(
  AL1 = AL1,
  AU1 = AU1,
  B1L = B1L,
  B1U = B1U,
  B2L = B2L,
  B2U = B2U,
  B3L = B3L,
  B3U = B3U,
  B4L = B4L,
  B4U = B4U
)
#2
library(DropletUtils)
droplet_results=lapply(sce_list, function(x) {
  # defaultDrops
  iscell_dd=defaultDrops(x, expected = 5000)

  # emptyDrops
  eout=emptyDrops(x, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed=eout$FDR <= 0.01

  # metadata
  meta=data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed
  )
  list(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    meta = meta
  )
})
names(droplet_results)=names(sce_list)

##--create seurat obj
#create filter counts
library(Seurat)
library(SingleCellExperiment)

# Assuming sce_list are raw count matrices (dgCMatrix)
filtered_counts=lapply(names(sce_list), function(nm) {
  mat=sce_list[[nm]]
  keep=droplet_results[[nm]]$iscell_ed
  keep[is.na(keep)] <- FALSE
  mat[, keep, drop = FALSE]  # subset columns (cells)
})
names(filtered_counts)=names(sce_list)

#seurat obj
seurat_list=lapply(names(filtered_counts), function(nm) {
  CreateSeuratObject(
    counts = filtered_counts[[nm]],
    project = nm
  )
})
names(seurat_list)=names(filtered_counts)

#save
# Add lung_region metadata first
for (nm in names(seurat_list)) {
  seurat_list[[nm]]$lung_region <- ifelse(grepl("U$", nm), "Upper", "Lower")
}

# Save each Seurat object in the correct folder
for (nm in names(seurat_list)) {
  seurat_obj=seurat_list[[nm]]  # assign correctly

  if (seurat_obj$lung_region[1] == "Lower") {
    saveRDS(seurat_obj, file = paste0("data/process/SeuratObjects/SScILD/Low/", nm, ".rds"))
  } else if (seurat_obj$lung_region[1] == "Upper") {
    saveRDS(seurat_obj, file = paste0("data/process/SeuratObjects/SScILD/Upper/", nm, ".rds"))
  }
}
#for space
rm(obj,seurat_list,seurat_obj,upper_samples,lower_names,nm,upper_names)
gc(full = TRUE)

##--merge each seurat groups 

#1Upper lung
SS.U=readRDS("data/process/SeuratObjects/SScILD/Upper/AU1.rds")
SS.U$sample="AU1"
upper_names=c("B1U", "B2U", "B3U", "B4U")

for (nm in upper_names) {
  obj=readRDS(
    paste0("data/process/SeuratObjects/SScILD/Upper/", nm, ".rds")
  )
  obj$sample=nm
  
  SS.U=merge(SS.U, y = obj)
  
}
saveRDS(SS.U, "data/process/SeuratObjects/SScILD/Upper/SS.U.rds")

#2Lower lung
SS.L=readRDS("data/process/SeuratObjects/SScILD/Low/AL1.rds")
SS.L$sample="AL1"
lower_names=c("B1L", "B2L", "B3L", "B4L")

for (nm in lower_names) {
  obj=readRDS(
    paste0("data/process/SeuratObjects/SScILD/Low/", nm, ".rds")
  )
  obj$sample <- nm
  
  SS.L=merge(SS.L, y = obj)
}
saveRDS(SS.L, "data/process/SeuratObjects/SScILD/Low/SS.L.rds")



```



###STEP-2-Analysis Upper lung------------------

###1-qc
```{r, fig.width=12, fig.height=8}
#load the data 
SS.U=readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U.rds")
SS.U[["percent.mt"]]=PercentageFeatureSet(SS.U , pattern="^MT-")
Idents(SS.U)="x"
SS.U$x="All"
VlnPlot(SS.U,features =c("nCount_RNA","nFeature_RNA","percent.mt"),ncol = 3,log = TRUE)
SS.U=subset(SS.U,subset=nFeature_RNA>300 & nFeature_RNA<7000 & percent.mt<15)
```


###2 statistics
```{r}
##norm, fvf,scale,runpca

SS.U=NormalizeData(SS.U)
SS.U=FindVariableFeatures(SS.U)
SS.U=ScaleData(SS.U,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = SS.U))
SS.U=RunPCA(SS.U, features = VariableFeatures(object = SS.U))
ElbowPlot(SS.U, ndims = 50, reduction = "pca")
SS.U=RunUMAP(SS.U, dims = 1:30)
#DimPlot(SS.U,reduction = "umap", label = TRUE)

```


###3-integration
```{r}
##integration
library(harmony)
SS.U =RunHarmony(SS.U,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:30  )
SS.U[["RNA"]]=JoinLayers(SS.U[["RNA"]])
SS.U = RunUMAP(SS.U, reduction = "harmony", dims= 1:30, reduction.name = "umap.harmony")
p=DimPlot(SS.U,reduction = "umap.harmony",group.by = c("sample"),label.size = 2, shuffle = T)
ggsave(filename = "output/SScILD/Upper/SS.Uintegration.png", plot=p, width = 12, height = 8)


#for the resolution
SS.U=FindNeighbors(SS.U,reduction= "harmony",dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS.U=FindClusters(SS.U, resolution = res)
  colname=paste0("res_", res)
  SS.U[[colname]]=SS.U$seurat_clusters
}
res_cols=grep("^res_", colnames(SS.U@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(SS.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}
#save 
p=DimPlot(SS.U,group.by = "res_0.5",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.5")
ggsave(filename = "output/SScILD/Upper/SS.Uresolution.png",plot = p,width = 12,height = 8,dpi = 300)
```



#4cluster and markers
```{r}

###clusters
SS.U=FindNeighbors(SS.U,reduction= "harmony",dims = 1:30)
SS.U=FindClusters(SS.U, resolution = 0.5,cluster.name = "harmony_clusters")
#markers
SS.U_markers=FindAllMarkers(SS.U,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
SS.U_markers=SS.U_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
write.xlsx(SS.U_markers,file = "output/SScILD/Upper/SS.U_markers.xlsx")
View(SS.U_markers)
```


"AIF1", "PTPRC":myloid x
"SPP1", "MERTK","AIF1": macrophages x
"FCN1", "IL1B", "CD14":Monocytes x
"CD1C", "PLD4", "FCER1A" Dendritic x
"TPSAB1", "CPA3","MS4A2":Mast cells x
"CD3E":Lymphoid x
"CD3D", "CD8A", "CD8B":Cytotoxic T x
"NKG7", "KLRF1":NK cells x
"MS4A1", "CD79A", "CD19":Blymphocyte x
"JCHAIN", "CD79A":plasma x
 "NRP1", "IL3RA" plasma dendritic x
 "AGER", "PDPN", "CLIC5","EPCAM":epithelial and alveolar type 1 x
 "SFTPC", "SFTPD", "MUC1": alveolar type 2   x
 KRT5, KRT15, KRT17, :basal x
 "KRT17", "MMP7", "COL1A1", "FN1":aberrant basaloid x
 "RSPH1", "CAPS", "FOXJ1": ciliated x
 "SCGB3A2":club x
 "MUC5B", "SCGB1A1", "SPDEF":Goblet x
 "SCGB1A1", "MUC5B","SPDEF": secretoty
 "VWF", "PECAM1", "CLDN5": endothelial x
 "IGFBP3", "GJA5", "BMX" arterial endo x
 "HPGD", "IL7R", "CA4", "FCN3hi", "SEMA3G":cap endo x
 "COL15A1", "PLVAP", "MPZL2", "POSTN": peribronchial endo x
 "ACKR1", "SELE": venous x
 "LYVE1", "CCL21", "PROX1": Lymphatic endo x
 "LUM", "PDGFRA", "COL1A2":Fibroblasts x
 "ACTA2",  "CTHRC1": myofybroblasts x
 "DES", "ACTA2", "MYH11"_smc x
 "PDGFRB", "RGS5", "FAM162B": Pericytes x
 "MSLN":Mesothelial
 "TOP2A", "BIRC5", "UBE2C", "MKI67":Proliferating cells
###5Annotation



###5Annotation
```{r}
##markers

genes=c("AIF1", "PTPRC","SPP1", "MERTK","FCN1", "IL1B", "CD14","CD1C", "PLD4", "FCER1A","TPSAB1", "CPA3","MS4A2","CD3E","CD8A", "CD8B", "NKG7", "KLRF1","MS4A1", "CD79A", "CD19","JCHAIN","NRP1", "IL3RA","AGER", "PDPN", "CLIC5","EPCAM","SFTPC", "SFTPD", "MUC1","KRT5", "KRT15", "KRT17","MMP7", "COL1A1","RSPH1", "CAPS", "FOXJ1","SCGB3A2","MUC5B", "SCGB1A1", "SPDEF","VWF", "PECAM1", "CLDN5","IGFBP3", "GJA5","HPGD", "IL7R", "CA4", "FCN3", "SEMA3G","COL15A1", "PLVAP", "MPZL2","ACKR1", "SELE","LYVE1", "CCL21", "PROX1","LUM", "PDGFRA", "COL1A2","ACTA2",  "CTHRC1","DES", "MYH11","PDGFRB", "RGS5", "FAM162B","MSLN","TOP2A", "BIRC5", "UBE2C", "MKI67")

Idents(SS.U)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages",
  "1"  = "Myeloid cells",
  "2"  = "Cytotoxic T",
  "3"  = "Monocytes",
  "4"  = "Capillary Endothelial",
  "5"  = "Alveolar Type 2",
  "6"  = "Pericytes",
  "7"  = "Fibroblasts",
  "8"  = "Venous Endothelial",
  "9"  = "Ciliated",
  "10" = "NK cells",
  "11" = "Dendritic cells",
  "12" = "Goblet",
  "13" = "Mast cells",
  "14" = "Proliferating cells",
  "15" = "Myeloid cells",
  "16" = "Myeloid cells",
  "17" = "Myeloid cells",
  "18" = "B Lymphocytes",
  "19" = "Lymphatic Endothelial"
  )
#-------------
SS.U=RenameIdents(SS.U, cluster_name)
SS.U$celltype=Idents(SS.U)
##2color
celltype_cols = c(
  "Alveolar Type 2"         = "#E41A1C",  
  "Ciliated"                = "#377EB8",  
  "Goblet"                  = "#80B1D3",  

  "Macrophages"             = "#F781BF",  
  "Monocytes"               = "#A65628",  
  "Myeloid cells"           = "#FF7F00",  
  "Dendritic cells"         = "#000000",  
  "Mast cells"              = "#FFD92F",  

  "Cytotoxic T"             = "#4DAF4A",  
  "B Lymphocytes"           = "#984EA3",  
  "Capillary Endothelial"   = "#66C2A5",  
  "Venous Endothelial"      = "#3288BD",  
  "Lymphatic Endothelial"   = "#A6D854",  

  "Pericytes"               = "#8C510A",  
  "Fibroblasts"             = "#B15928",  
  "Proliferating cells"     = "#FB9A99"   
)
##3 center

umap_df=as.data.frame(Embeddings(SS.U, "umap.harmony"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=SS.U$celltype
centers=aggregate(
  cbind(umap_1, umap_2) ~ celltype,
  data = umap_df,
  FUN = median
)


##supp duplicate and plot
library(ggrepel)
suppressWarnings({
  fig1a=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = umap_1, y = umap_2, label = celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 3,
      box.padding = 0.4,
      label.size = 0.25,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "umap 1", y = "umap 2")
})
fig1a
ggsave(ggsave(filename = "output/SScILD/Upper/SS.UcellT.png", plot = fig1a, width = 12, height = 8))
```






```{r}
###plot for markers
##complex heatmap
SS.U$celltype=mapvalues(as.character(Idents(SS.U)), names(cluster_name), cluster_name)
Idents(SS.U)="celltype"

# 3️⃣ Extract expression data
DefaultAssay(SS.U)="RNA"
expr_mat=GetAssayData(SS.U, layer = "data")

# Percent of cells expressing each gene per cell type
percent_mat=sapply(levels(Idents(SS.U)), function(ct) {
  cells=WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells] > 0) * 100
})
# Average expression per cell type
avg_exp=AverageExpression(SS.U, features = genes, assays = "RNA", layer = "data",
                             return.seurat = FALSE)$RNA
exp_mat=avg_exp[genes, colnames(percent_mat)]

# Convert to matrix
exp_mat=as.matrix(exp_mat)
percent_mat=as.matrix(percent_mat)

# 4️⃣ Flashy color function
library(circlize)
library(ComplexHeatmap)
library(circlize)
library(viridis)
col_fun=colorRamp2(c(min(exp_mat), 0, max(exp_mat)), c("navy", "yellow", "yellow"))

# 5️⃣ Dot drawing function

cell_fun <- function(j, i, x, y, w, h, fill) {
  if (!is.na(exp_mat[i, j])) {
    grid.circle(
      x = x, y = y,
      r = (percent_mat[i, j] / 100) * 0.45 * min(unit.c(w, h)),  
      gp = gpar(fill = col_fun(exp_mat[i, j]), col = "black")    
    )
  }
}

# 6️⃣ Top annotation using your cell types
celltypes=colnames(percent_mat)            
unique_celltypes=unique(celltypes)
palette_colors=viridis(length(unique_celltypes))  

column_ha=HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = setNames(palette_colors, unique_celltypes)),
  na_col = "grey"
)


# 7️⃣ Draw heatmap

ht=Heatmap(
  exp_mat,
  name = "Expression",
  col = col_fun,
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, rot = 45, just = "right"),
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

draw(ht)
# PNG

 #png(filename = "output/SScILD/Upper/SSUmarker.png",
    width = 3100, height = 2200, res = 300)
 draw(ht)   
 dev.off()  
```



###6proportion of cells
```{r}
###
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(SS.U@meta.data) %>%
  dplyr::count(`sample`, celltype) %>%
  dplyr::group_by(`sample`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
SS.U_meta <- dplyr::as_tibble(SS.U@meta.data) %>%
  dplyr::rename(sample_id = sample)   # rename to avoid conflicts
prop_df=SS.U_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SScILD/Upper/SS.Uproportion.png", plot = fig1c, width = 12, height = 10)
```



###7proportion of Dnonors
```{r}
library(dplyr)
library(ggplot2)
library(scales)

#change colname 
SS.U$Donors = SS.U$sample

#create the meta data
meta = SS.U@meta.data[, c("harmony_clusters", "Donors")]
table(meta$Donors)

# Make sure dplyr functions are used explicitly
meta = SS.U@meta.data

#the  percentage 
donors_counts = meta %>%
  as_tibble() %>%
  dplyr::group_by(harmony_clusters,Donors) %>% #count cells each donor use in each cluster
  dplyr::summarise( n= dplyr::n(),.groups = "drop") %>% #numb of cells
  dplyr::group_by(harmony_clusters) %>%
  dplyr::mutate(proportion = n /sum(n))  #fraction of cells per donor
```


```{r}
###plot
p = ggplot(donors_counts, aes(
  x = harmony_clusters,
  y = proportion,
  group = Donors,
  color = Donors
)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_viridis_d(option = "turbo", name = "Donor") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Donors Proportion",
    x = "Cluster",
    y = "Proportion (%)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
p

ggsave(filename = "output/SScILD/Upper/SS.UDonor.png", plot = p, width = 12, height = 8)

```



```{r}
##correlation with the cells
# 1️⃣ Set identities to cell types
Idents(SS.U) <- "celltype"

# sanity check
levels(Idents(SS.U))
# should show cell type names, NOT numbers

# 2️⃣ Average expression per cell type
av.expSS <- AverageExpression(
  SS.U,
  assays = "RNA",
  slot = "data"
)$RNA
# rows = genes, columns = cell types

# 3️⃣ Cell-type correlation matrix
correlation_matrixSS <- cor(
  as.matrix(av.expSS),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixSS[is.na(correlation_matrixSS)]=0

# 4️⃣ Save heatmap
png(
  filename = "output/SScILD/Upper/SS.Ucellcor.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixSS,
  clustering_method = "ward.D2",
  main = "Correlation Between Cell Types",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()
```



###save the final meta data
```{r}
##save
# Keep your existing metadata names
SS.U$orig.ident="SScILD in Upper lung"

# Remove unwanted columns 
cols_to_remove=c(
  grep("res_|RNA_snn_res", colnames(SS.U@meta.data), value = TRUE),
  "cell_type",
  "iscell_ed",
  "iscell_dd",
  "Donors",
  "x"
)
SS.U@meta.data=SS.U@meta.data[, !(colnames(SS.U@meta.data) %in% cols_to_remove)]

# Ensure active identity reflects your celltype
Idents(SS.U)=SS.U$celltype

# Save 
saveRDS(SS.U, file = "data/process/SeuratObjects/SScILD/Upper/SS.U_Final.rds")

# Save only the final metadata 
saveRDS(SS.U@meta.data, file = "data/process/SeuratObjects/SScILD/Upper/SS.U.metadata.rds")

# Optional CSV 
write.csv(SS.U@meta.data, 
          file = "data/process/SeuratObjects/SScILD/Upper/SS.U.metadata.csv",
          row.names = TRUE)
View(SS.U@meta.data)
```


##save
```{r}
#saveRDS(SS.U, file = "data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")

## run it later to load all 
SS.U = readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")
```














-------------------------------------------------------------------------------------------------------
------------------------------------------------------------------------------------------------------------















######STEP3-ANALYSIS OF SScILD in Low Lung

###1qc
```{r}
##load the data
SS.L=readRDS("data/process/SeuratObjects/SScILD/Low/SS.L.rds")
#
SS.L[["percent.mt"]]=PercentageFeatureSet(SS.L, pattern = "^MT-")
Idents(SS.L)="x"
SS.L$x="All"
VlnPlot(SS.L,features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log= TRUE)
SS.L=subset(SS.L,subset=nFeature_RNA >300 & nFeature_RNA <7000 & percent.mt<15)
```



###2statistics
```{r}
#####
SS.L=NormalizeData(SS.L)
SS.L=FindVariableFeatures(SS.L)
SS.L=ScaleData(SS.L,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = SS.L))
SS.L=RunPCA(SS.L, features = VariableFeatures(object = SS.L))
ElbowPlot(SS.L, ndims = 50, reduction = "pca")
SS.L=RunUMAP(SS.L, dims = 1:30)
DimPlot(SS.L,reduction = "umap", label = TRUE)
```



###3integration
```{r}
#integration
library(harmony)
SS.L=RunHarmony(SS.L,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:30  )
SS.L[["RNA"]]=JoinLayers(SS.L[["RNA"]])

SS.L = RunUMAP(SS.L, reduction = "harmony", dims= 1:30, reduction.name = "umap.harmony")
p=DimPlot(SS.L,reduction = "umap.harmony",group.by = c("sample"),label.size = 2, shuffle = T)
ggsave(filename = "output/SScILD/Low/SS.Lintegration.png", plot=p, width = 12, height = 8)


#for the resolution
SS.L=FindNeighbors(SS.L,reduction= "harmony",dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS.L=FindClusters(SS.L, resolution = res)
  colname=paste0("res_", res)
  SS.L[[colname]]=SS.L$seurat_clusters
}
res_cols=grep("^res_", colnames(SS.L@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(SS.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}
#save 
p=DimPlot(SS.L,group.by = "res_0.5",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.5")
ggsave(filename = "output/SScILD/Low/SS.Lresolution.png",plot = p,width = 12,height = 8,dpi = 300)

```


###4cluster and markers
```{r}
##cluster
###clusters
SS.L=FindNeighbors(SS.L,reduction= "harmony",dims = 1:30)
SS.L=FindClusters(SS.L, resolution = 0.5,cluster.name = "harmony_clusters")
#markers
SS.L_markers=FindAllMarkers(SS.L,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
SS.L_markers=SS.L_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
write.xlsx(SS.L_markers,file = "output/SScILD/Low/SS.L_markers.xlsx")
View(SS.L_markers)

```


 
###5annotation

```{r}
##markers
genes=c("AIF1","PTPRC","FCGR1A","SPP1", "MERTK","FABP4", "INHBA","FCN1", "IL1B", "CD14","CD1C", "PLD4",  "FCER1A","TPSAB1", "CPA3", "MS4A2","CD3E","CD8A", "CD8B", "GZMK","NKG7", "KLRF1","MS4A1", "CD79A","JCHAIN","NRP1", "IL3RA","AGER", "PDPN", "CLIC5","SFTPC", "SFTPD", "MUC1","KRT5", "KRT15", "KRT17","MMP7", "COL1A1","RSPH1", "CAPS", "FOXJ1","SCGB3A2","MUC5B", "SCGB1A1", "SPDEF","VWF", "PECAM1", "CLDN5","GJA5", "BMX","FCN3", "SEMA3G","COL15A1", "PLVAP", "MPZL2", "POSTN","ACKR1", "SELE","LYVE1", "CCL21", "PROX1","LUM", "PDGFRA", "COL1A2",  "CTHRC1","DES", "ACTA2", "MYH11","PDGFRB", "RGS5", "FAM162B","TOP2A", "BIRC5", "UBE2C", "MKI67")


Idents(SS.L)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages",
  "1"  = "Venous Endothelial",
  "2"  = "Macrophages",
  "3"  = "Monocytes",
  "4"  = "Ciliated",
  "5"  = "Cytotoxic T",
  "6"  = "Alveolar Type 2",
  "7"  = "Goblet",
  "8"  = "Plasma cells",
  "9"  = "Proliferating cells",
  "10" = "Pericytes",
  "11" = "Fibroblasts",
  "12" = "Basal",
  "13" = "Ciliated",
  "14" = "Mast cells",
  "15" = "NK cells",
  "16" = "Macrophages",
  "17" = "Lymphatic Endothelial"
  )
#-------------
SS.L=RenameIdents(SS.L, cluster_name)
SS.L$celltype=Idents(SS.L)
##2color
celltype_cols = c(
  "Venous Endothelial"          = "#E41A1C",  
  "Pericytes"                   = "#377EB8",  
  "Lymphatic Endothelial"       = "#80B1D3",  
  "Proliferating cells"         = "#F781BF",  
  "Plasma cells"                = "#FF0345",  
  "Monocytes"                   = "#FF7F00",  
  "Macrophages"                 = "#000000",  
  "Alveolar Type 2"             = "#FFD92F",  
  "Ciliated"                    = "#4DAF4A",  
  "Goblet"                      = "#984EA3",  
  "Basal"                       = "#66C2A5",  
  "Mast cells"                  = "#3288BD",  
  "Cytotoxic T"                 = "#A6D854",  
  "NK cells"                    = "#8C510A" 
        
  
)
##3 center

umap_df=as.data.frame(Embeddings(SS.L, "umap.harmony"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=SS.L$celltype
centers=aggregate(
  cbind(umap_1, umap_2) ~ celltype,
  data = umap_df,
  FUN = median
)


##supp duplicate and plot
library(ggrepel)
suppressWarnings({
  fig1a=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = umap_1, y = umap_2, label = celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 3,
      box.padding = 0.4,
      label.size = 0.25,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "umap 1", y = "umap 2")
})
fig1a
ggsave(ggsave(filename = "output/SScILD/Low/SS.LcellT.png", plot = fig1a, width = 12, height = 8))
```



```{r}
###with the markers

##complex heatmap
SS.L$celltype=mapvalues(as.character(Idents(SS.L)), names(cluster_name), cluster_name)
Idents(SS.L)="celltype"

# 3️⃣ Extract expression data
DefaultAssay(SS.L)="RNA"
expr_mat=GetAssayData(SS.L, layer = "data")

# Percent of cells expressing each gene per cell type
percent_mat=sapply(levels(Idents(SS.L)), function(ct) {
  cells=WhichCells(SS.L, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells] > 0) * 100
})
# Average expression per cell type
avg_exp=AverageExpression(SS.L, features = genes, assays = "RNA", layer = "data",
                             return.seurat = FALSE)$RNA
exp_mat=avg_exp[genes, colnames(percent_mat)]

# Convert to matrix
exp_mat=as.matrix(exp_mat)
percent_mat=as.matrix(percent_mat)

# 4️⃣ Flashy color function
library(circlize)
library(ComplexHeatmap)
library(circlize)
library(viridis)
col_fun=colorRamp2(c(min(exp_mat), 0, max(exp_mat)), c("navy", "yellow", "yellow"))

# 5️⃣ Dot drawing function

cell_fun <- function(j, i, x, y, w, h, fill) {
  if (!is.na(exp_mat[i, j])) {
    grid.circle(
      x = x, y = y,
      r = (percent_mat[i, j] / 100) * 0.45 * min(unit.c(w, h)),  
      gp = gpar(fill = col_fun(exp_mat[i, j]), col = "black")    
    )
  }
}

# 6️⃣ Top annotation using your cell types
celltypes=colnames(percent_mat)            
unique_celltypes=unique(celltypes)
palette_colors=viridis(length(unique_celltypes))  

column_ha=HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = setNames(palette_colors, unique_celltypes)),
  na_col = "grey"
)


# 7️⃣ Draw heatmap

ht=Heatmap(
  exp_mat,
  name = "Expression",
  col = col_fun,
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, rot = 45, just = "right"),
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

draw(ht)
# PNG

 png(filename = "output/SScILD/Low/SSLmarker.png",
    width = 3100, height = 2200, res = 300)
 draw(ht)   
 dev.off()  
```




###6proportion of cells
```{r}
###
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(SS.L@meta.data) %>%
  dplyr::count(`sample`, celltype) %>%
  dplyr::group_by(`sample`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
SS.U_meta <- dplyr::as_tibble(SS.L@meta.data) %>%
  dplyr::rename(sample_id = sample)   # rename to avoid conflicts
prop_df=SS.U_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/SScILD/Low/SS.Lproportion.png", plot = fig1c, width = 12, height = 10)

```



###7proportion of Donors
```{r}
library(dplyr)
library(ggplot2)
library(scales)

#change colname 
SS.L$Donors = SS.L$sample

#create the meta data
meta = SS.L@meta.data[, c("harmony_clusters", "Donors")]
table(meta$Donors)

# Make sure dplyr functions are used explicitly
meta = SS.L@meta.data

#the  percentage 
donors_counts = meta %>%
  as_tibble() %>%
  dplyr::group_by(harmony_clusters,Donors) %>% #count cells each donor use in each cluster
  dplyr::summarise( n= dplyr::n(),.groups = "drop") %>% #numb of cells
  dplyr::group_by(harmony_clusters) %>%
  dplyr::mutate(proportion = n /sum(n))  #fraction of cells per donor
```



```{r}
###plot
p = ggplot(donors_counts, aes(
  x = harmony_clusters,
  y = proportion,
  group = Donors,
  color = Donors
)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_viridis_d(option = "turbo", name = "Donor") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Donors Proportion",
    x = "Cluster",
    y = "Proportion (%)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
p

ggsave(filename = "output/SScILD/Low/SS.LDonor.png", plot = p, width = 12, height = 8)
```



```{r}
##correlation with the cells
# 1️⃣ Set identities to cell types
Idents(SS.L) <- "celltype"

# sanity check
levels(Idents(SS.L))
# should show cell type names, NOT numbers

# 2️⃣ Average expression per cell type
av.expSS <- AverageExpression(
  SS.L,
  assays = "RNA",
  slot = "data"
)$RNA
# rows = genes, columns = cell types

# 3️⃣ Cell-type correlation matrix
correlation_matrixSS <- cor(
  as.matrix(av.expSS),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixSS[is.na(correlation_matrixSS)]=0

# 4️⃣ Save heatmap
png(
  filename = "output/SScILD/Low/SS.Lcellcor.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixSS,
  clustering_method = "ward.D2",
  main = "Correlation Between Cell Types",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()
```



###save the final meta.data
```{r}
##save
# Keep your existing metadata names
SS.L$orig.ident="SScILD in Low lung"

# Remove unwanted columns 
cols_to_remove=c(
  grep("res_|RNA_snn_res", colnames(SS.L@meta.data), value = TRUE),
  "cell_type",
  "iscell_ed",
  "iscell_dd",
  "Donors",
  "x"
)
SS.L@meta.data=SS.L@meta.data[, !(colnames(SS.L@meta.data) %in% cols_to_remove)]

# Ensure active identity reflects your celltype
Idents(SS.L)=SS.L$celltype

# Save 
saveRDS(SS.L, file = "data/process/SeuratObjects/SScILD/Low/SS.L_Final.rds")

# Save only the final metadata 
saveRDS(SS.L@meta.data, file = "data/process/SeuratObjects/SScILD/Low/SS.L.metadata.rds")

# Optional CSV 
write.csv(SS.L@meta.data, 
          file = "data/process/SeuratObjects/SScILD/Low/SS.L.metadata.csv",
          row.names = TRUE)
View(SS.L@meta.data)
```



###save
```{r}
#saveRDS(SS.L, file = "data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")

## run it later to load all 
SS.L = readRDS("data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")
```
















