---
title: "R Notebook"
output: html_notebook
---

################################################################################################################################################
######ANALYSIS OF SScILD.Up - Systemic Sclerosis-associated Interstitial Lung Disease in Upper Lung DATA ##########################################
##############################################################################################################################################


## packages
```{r setup, include=FALSE}

##################PACKAGES
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")

```





###1load the data
```{r}
##load data,create seurat obj for each one and merge theme in lower and upper lung
#1-L for Lower lung and U for Upper lung
AL1=Read10X_h5("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/A/Lower lung/SSc-ILD-A-Low/feature_bc_matrix.h5",use.names = TRUE, unique.features = TRUE)
AU1=Read10X_h5("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/A/Upper lung/GSE156310_RAW/raw_feature_bc_matrix.h5", use.names = TRUE, unique.features = TRUE)
B1L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/1L")
B1U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/1U")
B2L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/2L")
B2U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/2U")
B3L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/3L")
B3U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/3U")
B4L=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/4L")
B4U=Read10X("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/SSc-ILD/B/X/4U")

#---Default dropletUtils
#1
sce_list=list(
  AL1 = AL1,
  AU1 = AU1,
  B1L = B1L,
  B1U = B1U,
  B2L = B2L,
  B2U = B2U,
  B3L = B3L,
  B3U = B3U,
  B4L = B4L,
  B4U = B4U
)
#2
library(DropletUtils)
droplet_results=lapply(sce_list, function(x) {
  # defaultDrops
  iscell_dd=defaultDrops(x, expected = 5000)

  # emptyDrops
  eout=emptyDrops(x, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed=eout$FDR <= 0.01

  # metadata
  meta=data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed
  )
  list(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    meta = meta
  )
})
names(droplet_results)=names(sce_list)

##--create seurat obj
#create filter counts
library(Seurat)
library(SingleCellExperiment)

# Assuming sce_list are raw count matrices (dgCMatrix)
filtered_counts=lapply(names(sce_list), function(nm) {
  mat=sce_list[[nm]]
  keep=droplet_results[[nm]]$iscell_ed
  keep[is.na(keep)] <- FALSE
  mat[, keep, drop = FALSE]  # subset columns (cells)
})
names(filtered_counts)=names(sce_list)

#seurat obj
seurat_list=lapply(names(filtered_counts), function(nm) {
  CreateSeuratObject(
    counts = filtered_counts[[nm]],
    project = nm
  )
})
names(seurat_list)=names(filtered_counts)

#save
# Add lung_region metadata first
for (nm in names(seurat_list)) {
  seurat_list[[nm]]$lung_region <- ifelse(grepl("U$", nm), "Upper", "Lower")
}

# Save each Seurat object in the correct folder
for (nm in names(seurat_list)) {
  seurat_obj=seurat_list[[nm]]  # assign correctly

  if (seurat_obj$lung_region[1] == "Lower") {
    saveRDS(seurat_obj, file = paste0("data/process/SeuratObjects/SScILD/Low/", nm, ".rds"))
  } else if (seurat_obj$lung_region[1] == "Upper") {
    saveRDS(seurat_obj, file = paste0("data/process/SeuratObjects/SScILD/Upper/", nm, ".rds"))
  }
}
#for space
rm(obj,seurat_list,seurat_obj,upper_samples,lower_names,nm,upper_names)
gc(full = TRUE)

##--merge each seurat groups 

#1Upper lung
SS.U=readRDS("data/process/SeuratObjects/SScILD/Upper/AU1.rds")
SS.U$sample="AU1"
upper_names=c("B1U", "B2U", "B3U", "B4U")

for (nm in upper_names) {
  obj=readRDS(
    paste0("data/process/SeuratObjects/SScILD/Upper/", nm, ".rds")
  )
  obj$sample=nm
  
  SS.U=merge(SS.U, y = obj)
  
}
saveRDS(SS.U, "data/process/SeuratObjects/SScILD/Upper/SS.U.rds")

#2Lower lung
SS.L=readRDS("data/process/SeuratObjects/SScILD/Low/AL1.rds")
SS.L$sample="AL1"
lower_names=c("B1L", "B2L", "B3L", "B4L")

for (nm in lower_names) {
  obj=readRDS(
    paste0("data/process/SeuratObjects/SScILD/Low/", nm, ".rds")
  )
  obj$sample <- nm
  
  SS.L=merge(SS.L, y = obj)
}
saveRDS(SS.L, "data/process/SeuratObjects/SScILD/Low/SS.L.rds")



```





###2-qc
```{r}
#load the data 
SS.U=readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U.rds")
DefaultAssay(SS.U) # if not RNA set it as RNA
SS.U[["percent.mt"]]=PercentageFeatureSet(SS.U , pattern="^MT-")
Idents(SS.U)="x"
SS.U$x="All"
VlnPlot(SS.U,features =c("nCount_RNA","nFeature_RNA","percent.mt"),ncol = 3,log = TRUE)
SS.U=subset(SS.U,subset=nFeature_RNA>300 & nFeature_RNA<7000 & percent.mt<15)

```


###3 statistics
```{r}
##norm, fvf,scale,runpca

SS.U=NormalizeData(SS.U)
SS.U=FindVariableFeatures(SS.U)
SS.U=ScaleData(SS.U,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = SS.U))
SS.U=RunPCA(SS.U, features = VariableFeatures(object = SS.U))
ElbowPlot(SS.U, ndims = 50, reduction = "pca")
SS.U=RunUMAP(SS.U, dims = 1:30)
#DimPlot(SS.U,reduction = "umap", label = TRUE)

```


###4-integration
```{r}
##integration
library(harmony)
SS.U =RunHarmony(SS.U,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:30  )
SS.U[["RNA"]]=JoinLayers(SS.U[["RNA"]])
SS.U = RunUMAP(SS.U, reduction = "harmony", dims= 1:30, reduction.name = "umap.harmony")



#for the resolution
SS.U=FindNeighbors(SS.U,reduction= "harmony",dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS.U=FindClusters(SS.U, resolution = res)
  colname=paste0("res_", res)
  SS.U[[colname]]=SS.U$seurat_clusters
}
res_cols=grep("^res_", colnames(SS.U@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(SS.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}

#save all
pdf("output/SScILD/Upper/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(SS.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
}
dev.off()


#save 
p = DimPlot(SS.U,group.by = "res_0.6",label = TRUE,label.size = 5, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.6") + NoLegend()
  
ggsave(filename = "output/SScILD/Upper/SSU_resolution.png",plot = p,width = 12,height = 8,dpi = 300)

```



###5cluster and markers
```{r}

###clusters
SS.U=FindNeighbors(SS.U,reduction= "harmony",dims = 1:30)
SS.U=FindClusters(SS.U, resolution = 0.6,cluster.name = "harmony_clusters")
#markers
SS.U_markers=FindAllMarkers(SS.U,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
SS.U_markers=SS.U_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)



```



###6.Annotation


```{r}

genes = c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Secretory
  # "SCGB1A1", "MUC5B", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  # Pericyte
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)
```

```{r}
l = FeaturePlot(SS.U, features = unique(genes), combine = F, order = T, reduction = "umap.harmony")
names(l) = unique(genes)
```



```{r}

##plots 

p=DotPlot(
  SS.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "harmony_clusters") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)


ggsave(filename = "output/SScILD/Upper/SSU.cluster.png", plot = p, width = 20, height = 8)
colnames(SS.U@meta.data)

```



```{r}

##cluster_specifique_annotation_________

Idents(SS.U)="harmony_clusters"
cluster_name_x=c(
  "0"  = "Macrophages_0",
  "1"  = "CD8_1",
  "2"  = "Macrophages_2",
  "3"  = "Monocytes_3",
  "4"  = "Macrophages_4",
  "5"  = "Capillary Endothelial_5",
  "6"  = "Macrophages_6",
  "7"  = "Basal Cells_7",
  "8"  = "NK Cells_8",
  "9"  = "Fibroblasts_9",
  "10" = "Peribronchial Endothelial_10",
  "11" = "Ciliated Cells_11",
  "12" = "Alveolar Cells_12",
  "13" = "Mural Cells_13",
  "14" = "Mast Cells_14",
  "15" = "Proliferating Cells_15",
  "16" = "Mural Cells_16",
  "17" = "Macrophages_17",
  "18" = "Lymphatic Endothelial_18",
  "19" = "Myeloid Cells_19",
  "20" = "B Cells_20",
  "21"= "Plasmacytoid Dendritic"
  )
#-------------
SS.U=RenameIdents(SS.U, cluster_name_x)
SS.U$CellType=Idents(SS.U)

##plots 

p=DotPlot(
  SS.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "CellType") + RotatedAxis()
p
ggsave(filename = "output/SScILD/Upper/SSU_cluster_name.png", plot = p, width = 20, height = 8)


```





```{r}


#--------umap


Idents(SS.U)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages",
  "1"  = "CD8",
  "2"  = "Macrophages",
  "3"  = "Monocytes",
  "4"  = "Macrophages",
  "5"  = "Capillary Endothelial",
  "6"  = "Macrophages",
  "7"  = "Basal Cells",
  "8"  = "NK Cells",
  "9"  = "Fibroblasts",
  "10" = "Peribronchial Endothelial",
  "11" = "Ciliated Cells",
  "12" = "Alveolar Cells",
  "13" = "Mural Cells",
  "14" = "Mast Cells",
  "15" = "Proliferating Cells",
  "16" = "Mural Cells",
  "17" = "Macrophages",
  "18" = "Lymphatic Endothelial",
  "19" = "Myeloid Cells",
  "20" = "B Cells",
  "21"= "Plasmacytoid Dendritic" 
  )
#-------------
SS.U=RenameIdents(SS.U, cluster_name)
SS.U$Major_Celltype=Idents(SS.U)
##2color
celltype_cols = c(
  "Macrophages"     = "#1f77b4",
  "CD8"             = "#ff7f0e",
  "Monocytes"       = "#2ca02c",
  "Capillary Endothelial" = "#d62728",
  "Basal Cells"           = "#9467bd",
  "NK Cells"              = "#8c564b",
  "Fibroblasts"           = "#e377c2",
  "Peribronchial Endothelial" = "#7f7f7f",
  "Ciliated Cells"            = "#bcbd22",
  "Alveolar Cells"             = "#17becf",
  "Mural Cells"                 = "#aec7e8",
  "Mast Cells"                   = "#ffbb78",
  "Proliferating Cells"           = "#98df8a",
  "Lymphatic Endothelial"           = "#ff9896",
  "Myeloid Cells"                    = "#c5b0d5",
  "B Cells"                            = "darkblue",
  "Plasmacytoid Dendritic"              = "gold"
)
##3 center
colnames(Embeddings(SS.U, "umap.harmony"))

##3 center
umap_df <- as.data.frame(Embeddings(SS.U, "umap.harmony"))
umap_df$Major_Celltype <- SS.U$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)
##plot
fig1a = ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "umapharmony_1", y = "umapharmony_2")

ggsave(ggsave(filename = "output/SScILD/Upper/SSU_cellType.png", plot = fig1a, width = 12, height = 8))

```



```{r}

#####for markers
# Identity

Idents(SS.U) <- "CellType"


SS.U$CellType <- gsub("-", "_", SS.U$CellType)
Idents(SS.U) <- "CellType"


# Expression matrix

DefaultAssay(SS.U) <- "RNA"
expr_mat <- GetAssayData(SS.U, layer = "data")


# genes in list

genes <- genes[genes %in% rownames(expr_mat)]
if (length(genes) < 2) stop("Too few genes found")


# Cell

celltypes <- levels(factor(Idents(SS.U)))


# Percent expression

percent_mat <- sapply(celltypes, function(ct) {
  cells <- WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)


# Average expression

avg_exp <- AverageExpression(
  SS.U,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA


colnames(avg_exp) <- gsub("-", "_", colnames(avg_exp))


# Align
common_ct <- intersect(celltypes, colnames(avg_exp))
if (length(common_ct) < 2) stop("Cell type mismatch still exists")

percent_mat <- percent_mat[, common_ct, drop = FALSE]
exp_mat <- as.matrix(avg_exp[genes, common_ct, drop = FALSE])


#Clean

exp_mat[is.infinite(exp_mat)] <- NA
exp_mat <- exp_mat[rowSums(is.na(exp_mat)) < ncol(exp_mat), , drop = FALSE]
percent_mat <- percent_mat[rownames(exp_mat), , drop = FALSE]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"


# Scaling

exp_mat_scaled <- log1p(exp_mat)

if (all(exp_mat_scaled == 0)) stop("No expression signal")

exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0


# Color

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95), na.rm = TRUE)
q <- unique(as.numeric(q))
q <- q[is.finite(q)]

if (length(q) < 2) q <- range(exp_mat_scaled, na.rm = TRUE)
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))


# dot
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}


# Cell colors

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(common_ct))

names(cols) <- common_ct

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = common_ct,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)


# plot
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)


# save
png("output/SScILD/Upper/SSU_heatmap.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)
dev.off()
```


```{r}
#----same methods
# gene with major annotation
Idents(SS.U) <- "Major_Celltype"


# -------------------------------
DefaultAssay(SS.U) <- "RNA"
expr_mat <- GetAssayData(SS.U, layer = "data")

# -------------------------------
# Percent expression matrix

percent_mat <- sapply(levels(Idents(SS.U)), function(ct) {
  cells <- WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# -------------------------------
# Average expression matrix

avg_exp <- AverageExpression(
  SS.U,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

# -------------------------------
# Clean data

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

# -------------------------------
# Scaling

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# -------------------------------
# Color function

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# -------------------------------
# Dot rendering

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# -------------------------------
# Cell type colors (annotation)

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# -------------------------------
# plot

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,

  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),

  column_names_gp = grid::gpar(
    fontsize = 10,
    fontface = "bold"
  ),

  column_names_rot = 90,

  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

# -------------------------------
# save

png("output/SScILD/Upper/markers.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)

dev.off()
```



##markers that we can have in TLS 


```{r}

markers <- c(
  "AID", "AICDA", "BCL6", "PAX5", "MYC", "IRF8","CD40LG", "ICOS","S1PR2", "CXCR4", "CXCR5", "CCR7",
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", "IFNG", "TGFB",
  "CCL2", "CCL3", "CCL4", "CCL5",
  "CXCL12", "CXCL13", "CCL19", "CCL21",
  "ICAM1", "VCAM1", "SELE", "CD54",
  "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C",
  "CD21", "CD23", "CR1", "CR2", "CD35",
  "FDCSP", "FCER2", "NTAN1"
)

markers <- markers[markers %in% rownames(SS.U)]

Idents(SS.U) <- "Major_Celltype"

DefaultAssay(SS.U) <- "RNA"
expr_mat <- GetAssayData(SS.U, layer = "data")

percent_mat <- sapply(levels(Idents(SS.U)), function(ct) {
  cells <- WhichCells(SS.U, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SS.U,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png("output/SScILD/Upper/TLS_markers.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)

dev.off()


```



###7correlation
```{r}

##correlation for celltype

# 1️⃣ Set identities to cell types
Idents(SS.U) <- "Major_Celltype"

# sanity check
levels(Idents(SS.U))
# should show cell type names

# 2️⃣ Average expression per cell type
av.expSSU <- AverageExpression(
  SS.U,
  assays = "RNA",
  slot = "data"
)$RNA

# 3️⃣ Cell-type correlation matrix
correlation_matrixSSU <- cor(
  as.matrix(av.expSSU),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixSSU[is.na(correlation_matrixSSU)] <- 0

# 4️⃣ Save heatmap as PNG
png(
  filename = "output/SScILD/Upper/SSU_correlation.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixSSU,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()
colnames(SS.U@meta.data)
```



###8 Class Major_Celltype with family


```{r}
#------
#extract
unique(SS.U$Major_Celltype)

#
SS.U$General <- recode(SS.U$Major_Celltype,
                      "Monoccytes"="Mono/Mac" ,
                      "Capillary Endothelial " ="Endothelial Cells",
                      "Mast Cells" ="Mast Cells" ,
                      "Basal Cells" ="Epithelial Cells",
                      "Plasmacytoid Dendritic "="Mono/Mac",
                      "Macrophages"  ="Mono/Mac",
                      "Mural Cells"  ="Mural Cells",
                      "Myeloid Cells" ="Mono/Mac",
                      "Alveolar Cells"= "Epithelial Cells",
                      "CD8" ="CD8",
                      "B Cells" = "B Cells" ,
                      "NK Cells" = "NK Cells",
                       "Ciliated Cells" ="Epithelial Cells",
                      "Fibroblasts"="Fibroblasts",
                      "Proliferating Cells"="Proliferating Cells",
                      "Peribronchial Endothelial"="Endothelial Cells",
                      "Lymphatic Endothelial"="Endothelial Cells"
)

```








###save the final meta data
```{r}
View(SS.U@meta.data)
##save

SS.U$orig.ident= "SScILD.U"
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General"
)
SS.U@meta.data <- SS.U@meta.data[, keep_cols]

#----------
saveRDS(SS.U, file = "data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")

## run it later to load all 
SS.U = readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")

#CSV 
write.csv(SS.U@meta.data, 
          file = "data/process/SeuratObjects/SScILD/Upper/SS.U.metadata.csv",
          row.names = TRUE)

```

























-------------------------------------------------------------------------------------------------------
------------------------------------------------------------------------------------------------------------















######STEP3-ANALYSIS OF SScILD in Low Lung

###1qc
```{r}
##load the data

SS.L=readRDS("data/process/SeuratObjects/SScILD/Low/SS.L.rds")
DefaultAssay(SS.L) # if not RNA set is as RNA
#
SS.L[["percent.mt"]]=PercentageFeatureSet(SS.L, pattern = "^MT-")

Idents(SS.L)="x"
SS.L$x="All"
VlnPlot(SS.L,features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log= TRUE)
SS.L=subset(SS.L,subset=nFeature_RNA >300 & nFeature_RNA <7000 & percent.mt<15)

```



###2statistics
```{r}
#####
SS.L=NormalizeData(SS.L)
SS.L=FindVariableFeatures(SS.L)
SS.L=ScaleData(SS.L,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = SS.L))
SS.L=RunPCA(SS.L, features = VariableFeatures(object = SS.L))
ElbowPlot(SS.L, ndims = 50, reduction = "pca")
SS.L=RunUMAP(SS.L, dims = 1:30)
DimPlot(SS.L,reduction = "umap", label = TRUE)

```



###3integration
```{r}
#integration
library(harmony)
SS.L=RunHarmony(SS.L,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:30  )
SS.L[["RNA"]]=JoinLayers(SS.L[["RNA"]])
SS.L = RunUMAP(SS.L, reduction = "harmony", dims= 1:30, reduction.name = "umap.harmony")



#for the resolution
SS.L=FindNeighbors(SS.L,reduction= "harmony",dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS.L=FindClusters(SS.L, resolution = res)
  colname=paste0("res_", res)
  SS.L[[colname]]=SS.L$seurat_clusters
}
res_cols=grep("^res_", colnames(SS.L@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(SS.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}

#save all
pdf("output/SScILD/Low/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(SS.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
}
dev.off()




#save 
p=DimPlot(SS.L,group.by = "res_0.6",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.6") +NoLegend()
ggsave(filename = "output/SScILD/Low/SSL.resolution.png",plot = p,width = 12,height = 8,dpi = 300)

```


###4cluster and markers
```{r}
##cluster
###clusters
SS.L=FindNeighbors(SS.L,reduction= "harmony",dims = 1:30)
SS.L=FindClusters(SS.L, resolution = 0.6,cluster.name = "harmony_clusters")
#markers
SS.L_markers=FindAllMarkers(SS.L,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
SS.L_markers=SS.L_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)


```


 
###5annotation


```{r}

genes = c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Secretory
  # "SCGB1A1", "MUC5B", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  # Pericyte
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)

```


```{r}

l = FeaturePlot(SS.L, features = unique(genes), combine = F, order = T, reduction = "umap.harmony")
names(l) = unique(genes)

```


```{r}

p=DotPlot(
  SS.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "harmony_clusters") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)

ggsave(filename = "output/SScILD/Low/SSL_cluster.png", plot = p, width =20, height = 8)


```







```{r}
##specifique_cluster_name

Idents(SS.L)="harmony_clusters"
cluster_name_x=c(
  "0"  = "Macrophages_0",
  "1"  = "Macrophages_1",
  "2"  = "Monocytes_2",
  "3"  = "CD8_3",
  "4"  = "Peribronchial Endothelial_4",
  "5"  = "Ciliated Cells_5",
  "6"  = "Alveolar Cells_6",
  "7"  = "Proliferating Cells_7",
  "8"  = "Goblet Cells_8",
  "9"  = "Mural Cells_9",
  "10" = "Capillary Endothelial_10",
  "11" = "Fibroblasts_11",
  "12" = "Macrophages_12",
  "13" = "Basal Cells_13",
  "14" = "Mast Cells_14",
  "15" = "NK Cells_15",
  "16" = "Ciliated Cells_16",
  "17" = "B Cells_17",
  "18" = "Lymphatic Endothelial"
  )
#-------------
SS.L=RenameIdents(SS.L, cluster_name_x)
SS.L$CellType = Idents(SS.L)

#plot
p=DotPlot(
  SS.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "CellType") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SScILD/Low/SSL_cluster_name.png", plot = p, width = 20, height = 8)

```








```{r}

#----------umap---------------


Idents(SS.L)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages",
  "1"  = "Macrophages",
  "2"  = "Monocytes",
  "3"  = "CD8",
  "4"  = "Peribronchial Endothelial",
  "5"  = "Ciliated Cells",
  "6"  = "Alveolar Cells",
  "7"  = "Proliferating Cells",
  "8"  = "Goblet Cells",
  "9"  = "Mural Cells",
  "10" = "Capillary Endothelial",
  "11" = "Fibroblasts",
  "12" = "Macrophages",
  "13" = "Basal Cells",
  "14" = "Mast Cells",
  "15" = "NK Cells",
  "16" = "Ciliated Cells",
  "17" = "B Cells",
  "18" = "Lymphatic Endothelial"
  )
#-------------
SS.L=RenameIdents(SS.L, cluster_name)
SS.L$Major_Celltype=Idents(SS.L)
##2color
celltype_cols = c(
 "Macrophages"                = "#1f77b4",
  "Monocytes"                 = "#2ca02c",
  "CD8"                       = "#ff7f0e",
  "Peribronchial Endothelial" = "darkblue",
  "Ciliated Cells"            = "#bcbd22",
  "Alveolar Cells"            = "#17becf",
  "Proliferating Cells"       = "#98df8a",
  "Goblet Cells"              = "gold",
  "Mural Cells"               = "#aec7e8",
  "Capillary Endothelial"     = "#d62728",
  "Fibroblasts"               = "#e377c2",
  "Basal Cells"               = "#9467bd",
  "Mast Cells"                = "#ffbb78",
  "NK Cells"                  = "#8c564b",
  "B Cells"                   = "magenta",
  "Lymphatic Endothelial"     = "tomato"
)

##3 center
umap_df <- as.data.frame(Embeddings(SS.L, "umap.harmony"))
umap_df$Major_Celltype <- SS.L$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)
##plot
fig1a = ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 3,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "umapharmony_1", y = "umapharmony_2")

fig1a
ggsave(ggsave(filename = "output/SScILD/Low/SSL_cellType.png", plot = fig1a, width = 12, height = 8))


```



```{r}

#####for markers
# Identity

Idents(SS.L) <- "CellType"


SS.L$CellType <- gsub("-", "_", SS.L$CellType)
Idents(SS.L) <- "CellType"


# Expression matrix

DefaultAssay(SS.L) <- "RNA"
expr_mat <- GetAssayData(SS.L, layer = "data")


# genes in list

genes <- genes[genes %in% rownames(expr_mat)]
if (length(genes) < 2) stop("Too few genes found")


# Cell

celltypes <- levels(factor(Idents(SS.L)))


# Percent expression

percent_mat <- sapply(celltypes, function(ct) {
  cells <- WhichCells(SS.L, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)


# Average expression

avg_exp <- AverageExpression(
  SS.L,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA


colnames(avg_exp) <- gsub("-", "_", colnames(avg_exp))


# Align
common_ct <- intersect(celltypes, colnames(avg_exp))
if (length(common_ct) < 2) stop("Cell type mismatch still exists")

percent_mat <- percent_mat[, common_ct, drop = FALSE]
exp_mat <- as.matrix(avg_exp[genes, common_ct, drop = FALSE])


#Clean

exp_mat[is.infinite(exp_mat)] <- NA
exp_mat <- exp_mat[rowSums(is.na(exp_mat)) < ncol(exp_mat), , drop = FALSE]
percent_mat <- percent_mat[rownames(exp_mat), , drop = FALSE]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"


# Scaling

exp_mat_scaled <- log1p(exp_mat)

if (all(exp_mat_scaled == 0)) stop("No expression signal")

exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0


# Color

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95), na.rm = TRUE)
q <- unique(as.numeric(q))
q <- q[is.finite(q)]

if (length(q) < 2) q <- range(exp_mat_scaled, na.rm = TRUE)
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))


# dot
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}


# Cell colors

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(common_ct))

names(cols) <- common_ct

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = common_ct,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)


# plot
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)


# save
png("output/SScILD/Low/SSL_heatmap.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)
dev.off()


```





```{r}
#----same methods
# gene with major annotation
Idents(SS.L) <- "Major_Celltype"


# -------------------------------
DefaultAssay(SS.L) <- "RNA"
expr_mat <- GetAssayData(SS.L, layer = "data")

# -------------------------------
# Percent expression matrix

percent_mat <- sapply(levels(Idents(SS.L)), function(ct) {
  cells <- WhichCells(SS.L, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# -------------------------------
# Average expression matrix

avg_exp <- AverageExpression(
  SS.L,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

# -------------------------------
# Clean data

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

# -------------------------------
# Scaling

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# -------------------------------
# Color function

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# -------------------------------
# Dot rendering

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# -------------------------------
# Cell type colors (annotation)

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# -------------------------------
# plot

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,

  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),

  column_names_gp = grid::gpar(
    fontsize = 10,
    fontface = "bold"
  ),

  column_names_rot = 90,

  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

# -------------------------------
# save

png("output/SScILD/Low/SS.L_markers.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)

dev.off()

```



##markers that we can have in TLS 


```{r}

markers <- c(
  "AID", "AICDA", "BCL6", "PAX5", "MYC", "IRF8","CD40LG", "ICOS","S1PR2", "CXCR4", "CXCR5", "CCR7",
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", "IFNG", "TGFB",
  "CCL2", "CCL3", "CCL4", "CCL5",
  "CXCL12", "CXCL13", "CCL19", "CCL21",
  "ICAM1", "VCAM1", "SELE", "CD54",
  "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C",
  "CD21", "CD23", "CR1", "CR2", "CD35",
  "FDCSP", "FCER2", "NTAN1"
)

markers <- markers[markers %in% rownames(SS.L)]

Idents(SS.L) <- "Major_Celltype"

DefaultAssay(SS.L) <- "RNA"
expr_mat <- GetAssayData(SS.L, layer = "data")

percent_mat <- sapply(levels(Idents(SS.L)), function(ct) {
  cells <- WhichCells(SS.L, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  SS.L,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

png("output/SScILD/Low/TLS_markers.png",
    width = 3200, height = 3500, res = 300)

ComplexHeatmap::draw(ht)

dev.off()


```



###7correlation
```{r}

##correlation for celltype

# 1️⃣ Set identities to cell types
Idents(SS.L) <- "Major_Celltype"

# sanity check
levels(Idents(SS.L))
# should show cell type names

# 2️⃣ Average expression per cell type
av.expSSL <- AverageExpression(
  SS.L,
  assays = "RNA",
  slot = "data"
)$RNA

# 3️⃣ Cell-type correlation matrix
correlation_matrixSSL <- cor(
  as.matrix(av.expSSL),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixSSL[is.na(correlation_matrixSSL)] <- 0

# 4️⃣ Save heatmap as PNG
png(
  filename = "output/SScILD/Low/SSL_correlation.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixSSL,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()

```


###7 Class Major_Celltype with family
```{r}
#------
#extract
unique(SS.L$Major_Celltype)

#
SS.L$General <- recode(SS.L$Major_Celltype,
                      "Ciliated Cells" = "Epithelial Cells",
                      "Macrophages" = "Mono/Mac",
                      "Monocytes" = "Mono/Mac",
                      "CD8" = "CD8",
                      "Capillary Endothelial" = "Endothelial Cells",
                      "Mural Cells" = "Mural Cells",
                      "Peribronchial Endothelial" = "Endothelial Cells",
                      "Mast Cells" = "Mast Cells",
                      "B Cells" = "B Cells",
                      "Fibroblasts" = "Fibroblasts",
                      "NK Cells" = "NK Cells",
                      "Proliferating Cells" = "Proliferating Cells",
                      "Alveolar Cells" = "Epithelial Cells",
                      "Basal Cells" = "Epithelial Cells",
                      "Goblet Cells" = "Epithelial Cells",
                      "Lymphatic Endothelial" = "Endothelial Cells"
                      
)
```









###save the final meta.data
```{r}
##save


##save


keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General"
)

SS.L@meta.data <- SS.L@meta.data[, keep_cols]

SS.L$orig.ident="SScILD.L"


#CSV 
write.csv(SS.L@meta.data, 
          file = "data/process/SeuratObjects/SScILD/Low/SS.L.metadata.csv",
          row.names = TRUE)

#-------
saveRDS(SS.L, file = "data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")

## run it later to load all 
#SS.L = readRDS("data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")



```



















