---
title: "R Notebook"
output: html_notebook
---
############################################################################################################################
###ANALYSIS OF IDIOPATIC PULMONARY FIBROSIS#############################################################################
##############################################################################################################################

#THE PACKAGES
```{r setup, include=FALSE}
##################PACKAGES
library(pacman)
  p_load(Seurat, SeuratData, Matrix, dplyr, DropletUtils, presto,gprofiler2, ggplot2, patchwork, gplots, cowplot, openxlsx,plyr, scDblFinder, BiocParallel,devtools,openxlsx,reticulate,hdf5r,DropletUtils,harmony)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")


```

#####STEP_1
###load the data
```{r}
#data for the Lower lung(L) and Upper(U)
library(Seurat)
library(DropletUtils)

data.dir="../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Idiopathic Pulmonary Fibrosis/GSE128033_RAW"

# List all sample folders
sample.dirs=list.dirs(data.dir, recursive = FALSE)

for (folder_path in sample.dirs) {
  folder_name=basename(folder_path)  
  message("Processing ", folder_name)
  counts=Read10X(data.dir = folder_path)
  # Droplet filtering
  iscell_dd=defaultDrops(counts, expected = 5000)
  eout=emptyDrops(counts, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed=eout$FDR <= 0.01
  # Metadata
  meta=data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    row.names = colnames(counts)
  )
  # Keep droplets
  keep=iscell_dd | iscell_ed
  
  # Create Seurat object
  IPF=CreateSeuratObject(
    counts = counts[, keep],
    meta.data = meta[keep, ],
    project = folder_name
  )
  # Add sample ID
  IPF$sample=folder_name
  # Save
  saveRDS(IPF, file = paste0("data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/", folder_name, ".rds"))
}
#merge in two seurat Upper and Lung
save.base="data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis"

groups=list(
  L = c("1L1","1L2","2L1","2L2","3L"),
  U = c("1U","2U","3U")
)
for (grp in names(groups)) {
  folders=groups[[grp]]
  
  seurat_list=lapply(folders, function(folder_name) {
    # Correct path: point to Lower or Upper subfolder
    subfolder=ifelse(grp == "L", "Lower", "Upper")
    file_path=file.path(save.base, subfolder, paste0(folder_name, ".rds"))
    message("Loading ", file_path)
    readRDS(file_path)
  })
  names(seurat_list)=folders
  
  # Merge all Seurat objects in the group
  if (length(seurat_list) > 1) {
    combined=Reduce(function(x, y) merge(x, y), seurat_list)
  } else {
    combined=seurat_list[[1]]
  }
  
  # Save merged 
  save_dir=file.path(save.base, ifelse(grp == "L", "Lower", "Upper"))
  saveRDS(combined, file = file.path(save_dir, paste0("IPF_", grp, ".rds")))
  message("Saved merged Seurat object: IPF_", grp, " in ", save_dir)
}
rm(IPF.Up)
gc(full = TRUE)
```



###STEP_2: process and analyse the IPF_Upper

###1-qc and threshold
```{r,fig.width=10,fig.height=8}
##qc
IPF.U=readRDS("data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Upper/IPF_U.rds")
IPF.U[["percent.mt"]]=PercentageFeatureSet(IPF.U,pattern = "^MT-")
Idents(IPF.U)="IPF_Upper"
VlnPlot(IPF.U, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log = TRUE)
#threshold
IPF.U=subset(IPF.U, subset=nFeature_RNA>200 & nFeature_RNA<7000 & percent.mt <10)

```


###2-statistics
```{r}
###
IPF.U=NormalizeData(IPF.U)
IPF.U=FindVariableFeatures(IPF.U)
IPF.U=ScaleData(IPF.U,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = IPF.U))
IPF.U=RunPCA(IPF.U, features = VariableFeatures(object = IPF.U))
ElbowPlot(IPF.U)
IPF.U=RunUMAP(IPF.U, dims = 1:20)
P=DimPlot(IPF.U,reduction = "umap", label = TRUE)

```

###3-integration

```{r}
##integration
library(harmony)
IPF.U=RunHarmony(IPF.U,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:20  )
IPF.U[["RNA"]]=JoinLayers(IPF.U[["RNA"]])
IPF.U = RunUMAP(IPF.U, reduction = "harmony", dims= 1:20, reduction.name = "umap.harmony")
p=DimPlot(IPF.U,reduction = "umap.harmony",group.by = c("sample"),label.size = 2, shuffle = T)
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPF.integration.png", plot=p, width = 12, height = 8)

#for the resolution
IPF.U=FindNeighbors(IPF.U,reduction = "harmony",dims = 1:20)
for (res in seq(0.2, 1, by = 0.1)) {
  IPF.U=FindClusters(IPF.U, resolution = res)
  colname=paste0("res_", res)
  IPF.U[[colname]]=IPF.U$seurat_clusters
}
res_cols=grep("^res_", colnames(IPF.U@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(IPF.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}
##save
p_res05=DimPlot(IPF.U,group.by = "res_0.5",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.5")
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPF.U_resolution_0.5.png",plot = p_res05,width = 12,height = 8,dpi = 300)

```



###4-cluster and markers
```{r}
###clusters

IPF.U=FindNeighbors(IPF.U,reduction = "harmony",dims = 1:20)
IPF.U=FindClusters(IPF.U, resolution = 0.5,cluster.name = "harmony_clusters")
#markers
IPF.Umarkers=FindAllMarkers(IPF.U,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
IPF.Umarkers = IPF.Umarkers%>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
write.xlsx(IPF.Umarkers,file = "output/Idiopathic Pulmonary Fibrosis/Upper/IPF.Umarkers.xlsx")
View(IPF.Umarkers)

##the average and correlation 
av.expIPF=AverageExpression(IPF.U)$RNA
correlation_matrixIPF=cor(as.matrix(av.expIPF), 
                          as.matrix(av.expIPF))
library(pheatmap)
pheatmap(correlation_matrixIPF,
         clustering_method = "ward.D2",
         main = "Correlation between Cluster Groups",
         filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPF_corr.png",
         fontsize_row = 8,
         fontsize_col = 8)
```



###5-annotation
"AIF1","PTPRC" : myeloid 
"FABP4","SPP1","FCGR1A","INHBA":macrophages x
"FCN1", "IL1B", "CD14":monocyte. x
"FCER1A": dendritic
"TPSAB1", "CPA3", "MS4A2":mast cells x
"CD3E":lymphoid
"CD3D", "CD8A", "CD8B":cytotoxic T    x
"CCR7", "LEF1":helper T more or less
Treg:"FOXP3"
Narural killer:"NKG7", "KLRF1"  x
"MS4A1", "CD79A":Blymphocyte    x
 CD79A: plasma
 "NRP1", "IL3RA":Plasmacytoid Dendritic more or less
 "EPCAM": epithelial
 Alveolar Type 1:"PDPN"
 Alveolar Type 2:"SFTPC", "SFTPD", "MUC1" x
 Basal:"KRT15", "KRT17"
 Aberrant Basaloid:"COL1A1","FN1"
 Ciliated : "RSPH1", "CAPS", "FOXJ1"   x 
 Club :SCGB3A2
 Goblet :"MUC5B", "SCGB1A1"            x
 Secretory :"SCGB1A1", "MUC5B"
 Endothelial :"VWF", "PECAM1", "CLDN5"
 Arterial Endothelial :"GJA5","DKK2"
 Capillary Endothelial :"FCN3", "SEMA3G", more or less
 Peribronchial Endothelial:"PLVAP","MPZL2", "POSTN" 
 Venous Endothelial :"ACKR1", "SELE"      x
 Lymphatic Endothelial :"CCL21", "PROX1" x
 Fibroblast :"LUM","COL1A2"
 Myofibroblast :"POSTN", "CTHRC1"
Smooth Muscle:"DES", "ACTA2", "MYH11"
Pericyte :"PDGFRB", "RGS5", "FAM162B" 
Mesothelial :"MSLN","UPK3B"
Proliferating Cells :"TOP2A", "BIRC5", "UBE2C", "MKI67" X

```{r,fig.width=12, fig.height=8}

genes=c("AIF1","PTPRC","FABP4","SPP1","FCGR1A","INHBA","FCN1", "IL1B", "CD14","FCER1A","TPSAB1", "CPA3", "MS4A2","CD3E","CD3D", "CD8A", "CD8B","CCR7", "LEF1","FOXP3","NKG7", "KLRF1","MS4A1", "CD79A","NRP1", "IL3RA","EPCAM","PDPN","SFTPC", "SFTPD", "MUC1","KRT15", "KRT17","COL1A1","FN1","RSPH1", "CAPS", "FOXJ1","SCGB3A2","MUC5B", "SCGB1A1", "VWF", "PECAM1", "CLDN5","GJA5","DKK2","FCN3", "SEMA3G","PLVAP","MPZL2","POSTN" ,"ACKR1", "SELE","CCL21", "PROX1","LUM","COL1A2","CTHRC1","DES", "ACTA2", "MYH11","PDGFRB", "FAM162B" ,"MSLN","UPK3B","TOP2A", "BIRC5", "UBE2C")


##1celltype
Idents(IPF.U)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages"                  ,
  "1"  = "Macrophages"                   ,
  "2"  = "Cytotoxic T"                   ,
  "3"  = "Venous Endothelial"            ,
  "4"  = "Monocytes"                     ,
  "5"  = "NK cells"                       ,
  "6"  = "B Lymphocytes"                    ,
  "7"  = "Arterial Endothelial"            ,
  "8"  = "Goblet"                          ,
  "9"  = "Fibroblasts"                     ,
  "10" = "Aveolar Type 2"                  ,          
  "11" = "Mast cells"                      ,
  "12" = "Macrophages"                     ,
  "13" = "Smooth Muscle cells"             ,
  "14" = "Proliferating cells"             ,
  "15" = "Aveolar Type 2"                  ,
  "16" = "Lymphatic Endothelial"          ,
  "17" = "Ciliated"      
  
)

IPF.U=RenameIdents(IPF.U, cluster_name)
IPF.U$celltype=Idents(IPF.U)
##2color
celltype_cols <- c(
  "Macrophages"           = "#6A3D9A",  
  "Monocytes"             = "#66C2A5", 
  "Mast cells"            = "#FFD92F",  
  "NK cells"              = "#4DAF4A",  
  "Cytotoxic T"           = "#377EB8",  
  "B Lymphocytes"         = "#000000",  
  "Fibroblasts"           = "#A65628",  
  "Smooth Muscle cells"   = "#904EA3",  
  "Venous Endothelial"    = "#FF7F00",  
  "Arterial Endothelial"  = "#1B9E77",  
  "Lymphatic Endothelial" = "#F781BF",  
  "Goblet"                = "#8DD3C7",
  "Aveolar Type 2"        = "#E41A1C",  
  "Ciliated"              = "#B15928",  
  "Proliferating cells"   = "#999999"   
)


##3 center
umap_df=as.data.frame(Embeddings(IPF.U, "umap"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=IPF.U$celltype
centers=aggregate(
  cbind(umap_1, umap_2) ~ celltype,
  data = umap_df,
  FUN = median
)

##supp duplicate and plot
library(ggrepel)
suppressWarnings({
  fig1a=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = umap_1, y = umap_2, label = celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 4,
      box.padding = 0.4,
      label.size = 0.25,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "umap 1", y = "umap 2")
})
fig1a
ggsave(ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFcellT.png", plot = fig1a, width = 20, height = 10))

```


```{r,fig.width=20,fig.height=8}
###with the markers
p=DotPlot(
  IPF.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T ,
  group.by = "celltype") + RotatedAxis() 
 p 
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFmarkers.png", plot = p, width = 20, height = 10)

```


```{r, fig.width=12,fig.height=8}
###with the markers another plot type


# Create DotPlot but prevent automatic flipping
p1=DotPlot(IPF.U, features = genes, assay = "RNA") + 
  labs(x = "Genes", y = "Cell types") +
  scale_color_gradientn(colours = c('#330066','#336699','#66CC66','#FFCC33')) +
  guides(size = guide_legend("Percent Expression")) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )
p1=p1 + aes(x = features.plot, y = id)  # 'id' = cell type, 'features.plot' = gene
p1
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFmarkers2.png", plot = p1, width = 20, height = 10)


```



```{r, fig.width=12,fig.height=8}

###complex heatmap
IPF.U$celltype=mapvalues(as.character(Idents(IPF.U)), names(cluster_name), cluster_name)
Idents(IPF.U)="celltype"

# 3️⃣ Extract expression data
DefaultAssay(IPF.U)="RNA"
expr_mat=GetAssayData(IPF.U, layer = "data")

# Percent of cells expressing each gene per cell type
percent_mat=sapply(levels(Idents(IPF.U)), function(ct) {
  cells=WhichCells(IPF.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells] > 0) * 100
})
# Average expression per cell type
avg_exp=AverageExpression(IPF.U, features = genes, assays = "RNA", layer = "data",
                             return.seurat = FALSE)$RNA
exp_mat=avg_exp[genes, colnames(percent_mat)]

# Convert to matrix
exp_mat=as.matrix(exp_mat)
percent_mat=as.matrix(percent_mat)

# 4️⃣ Flashy color function

col_fun=colorRamp2(c(min(exp_mat), 0, max(exp_mat)), c("navy", "yellow", "yellow"))

# 5️⃣ Dot drawing function

cell_fun <- function(j, i, x, y, w, h, fill) {
  if (!is.na(exp_mat[i, j])) {
    grid.circle(
      x = x, y = y,
      r = (percent_mat[i, j] / 100) * 0.45 * min(unit.c(w, h)),  
      gp = gpar(fill = col_fun(exp_mat[i, j]), col = "black")    
    )
  }
}

# 6️⃣ Top annotation using your cell types
celltypes=colnames(percent_mat)            
unique_celltypes=unique(celltypes)
palette_colors=viridis(length(unique_celltypes))  

column_ha=HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = setNames(palette_colors, unique_celltypes)),
  na_col = "grey"
)


# 7️⃣ Draw heatmap

ht=Heatmap(
  exp_mat,
  name = "Expression",
  col = col_fun,
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, rot = 45, just = "right"),
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

draw(ht)
# PNG

 png(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFhmap.png",
    width = 3100, height = 2200, res = 300)
 draw(ht)   
 dev.off()  
```



###6-proportion of cells
```{r}
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(IPF.U@meta.data) %>%
  dplyr::count(`sample`, celltype) %>%
  dplyr::group_by(`sample`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
IPF.U_meta <- dplyr::as_tibble(IPF.U@meta.data) %>%
  dplyr::rename(sample_id = sample)   # rename to avoid conflicts
prop_df=IPF.U_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFproportion.png", plot = fig1c, width = 12, height = 10)
```




###7-Proportion of DONORS
```{r}
library(dplyr)
library(ggplot2)
library(scales)

#change colname 
IPF.U$Donors = IPF.U$sample

#create the meta data
meta = IPF.U@meta.data[, c("harmony_clusters", "Donors")]
table(meta$Donors)

# Make sure dplyr functions are used explicitly
meta = IPF.U@meta.data

#the  percentage 
donors_counts = meta %>%
  as_tibble() %>%
  dplyr::group_by(harmony_clusters,Donors) %>% #count cells each donor use in each cluster
  dplyr::summarise( n= dplyr::n(),.groups = "drop") %>% #numb of cells
  dplyr::group_by(harmony_clusters) %>%
  dplyr::mutate(proportion = n /sum(n))  #fraction of cells per donor
```


```{r}
###
#####
p=ggplot(donors_counts, aes(
  x = harmony_clusters,
  y = proportion,
  group = Donors,
  color = Donors
)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  # Use a colorblind-friendly palette with distinct colors
  scale_color_brewer(palette = "Set1", name = "Donor") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Donors Proportion ",
    x = "Cluster",
    y = "Proportion (%)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
p
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFDonor.png", plot = p, width = 12, height = 8)
```



```{r}
##correlation with the celltype
# 1️⃣ Set identities to cell types
Idents(IPF.U) <- "celltype"

# sanity check
levels(Idents(IPF.U))
# should show cell type names

# 2️⃣ Average expression per cell type
av.expIPF <- AverageExpression(
  IPF.U,
  assays = "RNA",
  slot = "data"
)$RNA

# 3️⃣ Cell-type correlation matrix
correlation_matrixIPF <- cor(
  as.matrix(av.expIPF),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixIPF[is.na(correlation_matrixIPF)] <- 0

# 4️⃣ Save heatmap as PNG
png(
  filename = "output/Idiopathic Pulmonary Fibrosis/Upper/IPFcelltypecorr.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixIPF,
  clustering_method = "ward.D2",
  main = "Correlation Between Cell Types (IPF – Upper Lung)",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()
```






###Final meta.data with the cells names
```{r}
##save
# Keep your existing metadata names
IPF.U$orig.ident="Idiopathic P. Fibrosis in Upper Lung"

# Remove unwanted columns 
cols_to_remove=c(
  grep("res_|RNA_snn_res", colnames(IPF.U@meta.data), value = TRUE),
  "cell_type",
  "iscell_ed",
  "iscell_dd"
)
IPF.U@meta.data=IPF.U@meta.data[, !(colnames(IPF.U@meta.data) %in% cols_to_remove)]

# Ensure active identity reflects your celltype
Idents(IPF.U)=IPF.U$celltype

# Save 
saveRDS(IPF.U, file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Upper/IPF.U_Final.rds")

# Save only the final metadata 
saveRDS(IPF.U@meta.data, file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Upper/IPF.U_metadata.rds")

# Optional CSV 
write.csv(IPF.U@meta.data, 
          file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Upper/IPF.U_metadata.csv",
          row.names = TRUE)
View(IPF.U@meta.data)
```



###save
```{r}
#saveRDS(IPF.U, file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Upper/IPF_process.rds")

## run it later to load all 
IPF.U = readRDS("data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Upper/IPF_process.rds")
```

























----------------------------------------------------------___________---------------------------------------------------------------
----------------------------------------------------------------------------------------------------------------------------
#####STEP-3 ANALYSIS FOR IPF in LOWER lUNG
-----------------------------------------------------------------------------------------------------------------------



###1-qc
```{r}
##qc---------

IPF.L=readRDS("data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/IPF_L.rds")
IPF.L[["percent.mt"]]=PercentageFeatureSet(IPF.L,pattern = "^MT-")
Idents(IPF.L)="x"
IPF.L$x="All"
VlnPlot(IPF.L, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log = TRUE)
#threshold
IPF.L=subset(IPF.L, subset=nFeature_RNA>200 & nFeature_RNA<7000 & percent.mt <10)
```



####2-statistics
```{r}
IPF.L=NormalizeData(IPF.L)
IPF.L=FindVariableFeatures(IPF.L)
IPF.L=ScaleData(IPF.L,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = IPF.L))
IPF.L=RunPCA(IPF.L, features = VariableFeatures(object = IPF.L))
ElbowPlot(IPF.L)
#IPF.L=RunUMAP(IPF.L, dims = 1:20)
#P=DimPlot(IPF.L,reduction = "umap", label = TRUE)

```



###3-integration
```{r}
##integration
library(harmony)
IPF.L=RunHarmony(IPF.L,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:20  )
IPF.L[["RNA"]]=JoinLayers(IPF.L[["RNA"]])
IPF.L = RunUMAP(IPF.L, reduction = "harmony", dims= 1:20, reduction.name = "umap.harmony")
p=DimPlot(IPF.L,reduction = "umap.harmony",group.by = c("sample"),label.size = 2, shuffle = T)
p
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPFintegration.png", plot=p, width = 12, height = 8)

#for the resolution
IPF.L=FindNeighbors(IPF.L,reduction = "harmony",dims = 1:20)
for (res in seq(0.2, 1, by = 0.1)) {
  IPF.L=FindClusters(IPF.L, resolution = res)
  colname=paste0("res_", res)
  IPF.L[[colname]]=IPF.L$seurat_clusters
}
res_cols=grep("^res_", colnames(IPF.L@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(IPF.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}

##save
p=DimPlot(IPF.L,group.by = "res_0.5",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.5")
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPF.L_resolution.png",plot = p,width = 12,height = 8,dpi = 300)

```


###4 cluster and markers
```{r}
###clusters

IPF.L=FindNeighbors(IPF.L,reduction = "harmony",dims = 1:20)
IPF.L=FindClusters(IPF.L, resolution = 0.5,cluster.name = "harmony_clusters")
#markers
IPF.Lmarkers=FindAllMarkers(IPF.L,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
IPF.Lmarkers = IPF.Lmarkers%>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
write.xlsx(IPF.Lmarkers,file = "output/Idiopathic Pulmonary Fibrosis/Low/IPF.Lmarkers.xlsx")
View(IPF.Lmarkers)

##the average and correlation 
av.expIPF=AverageExpression(IPF.L)$RNA
correlation_matrixIPF=cor(as.matrix(av.expIPF), 
                          as.matrix(av.expIPF))
library(pheatmap)
pheatmap(correlation_matrixIPF,
         clustering_method = "ward.D2",
         main = "Correlation between Cluster Groups",
         filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPF_corr.png",
         fontsize_row = 8,
         fontsize_col = 8)
```



###5-annotation

```{r, fig.width=14, fig.height=8}

##marker------

genes=c("AIF1","PTPRC","FABP4","SPP1","FCGR1A","INHBA","FCN1", "IL1B", "CD14","FCER1A","TPSAB1", "CPA3", "MS4A2","CD3E","CD3D", "CD8A", "CD8B","CCR7", "LEF1","FOXP3","NKG7", "KLRF1","MS4A1", "CD79A","NRP1", "IL3RA","EPCAM","PDPN","SFTPC", "SFTPD", "MUC1","KRT15", "KRT17","COL1A1","FN1","RSPH1", "CAPS", "FOXJ1","SCGB3A2","MUC5B", "SCGB1A1", "VWF", "PECAM1", "CLDN5","GJA5","DKK2","FCN3", "SEMA3G","PLVAP","MPZL2","POSTN" ,"ACKR1", "SELE","CCL21", "PROX1","LUM","COL1A2","CTHRC1","DES", "ACTA2", "MYH11","PDGFRB", "FAM162B" ,"MSLN","UPK3B","TOP2A", "BIRC5", "UBE2C")




##1celltype
Idents(IPF.L)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages"                                     ,
  "1"  = "Venous Endothelial"                   ,
  "2"  = "Fibroblasts"                                     ,
  "3"  = "Cytotoxic T"                                     ,
  "4"  = "Secretory cells"                                     ,
  "5"  = "Monocytes"                                     ,
  "6"  = "Cytotoxic T"                          ,
  "7"  = "Mast cells"                           ,
  "8"  = "Aveolar Type 2"                                     ,
  "9"  = "Ciliated"                                     ,
  "10" = "Basal"                                     ,          
  "11" = "Macrophages"                                     ,
  "12" = "Macrophages"                                     ,
  "13" = "Smooth Muscle cells"                                     ,
  "14" = "Venous Endothelial"                   ,
  "15" = "Fibroblasts"                                     ,
  "16" = "B Lymphocytes"                                     ,
  "17" = "Proliferating cells"                                     ,    
  "18" = "Lymphatic Endothelial"  
)

IPF.L=RenameIdents(IPF.L, cluster_name)
IPF.L$celltype=Idents(IPF.L)
##2color
celltype_cols = c(
  "Venous Endothelial"    = "#1F78B4",  # strong blue
  "Lymphatic Endothelial" = "#33A02C",  # green
  
  "Cytotoxic T"           = "#E31A1C",  # red
  "B Lymphocytes"         = "#6A3D9A",  # purple
  "Monocytes"             = "#FF7F00",  # orange
  "Macrophages"           = "#B15928",  # brown
  "Mast cells"            = "#FFD92F",  # yellow
  
  "Aveolar Type 2"        = "#A6CEE3",  # light blue
  "Ciliated"              = "#000000",  # black
  "Secretory cells"       = "#FB9A99",  # pink
  "Basal"                 = "#CAB2D6",  # lavender
  "Proliferating cells"   = "#FFFF99",  # pale yellow
  
  "Smooth Muscle cells"   = "#8B0000",  # dark red
  "Fibroblasts"           = "#7F7F7F"   # gray
)



##3 center
umap_df=as.data.frame(Embeddings(IPF.L, "umap.harmony"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=IPF.L$celltype
centers=aggregate(
  cbind(umap_1, umap_2) ~ celltype,
  data = umap_df,
  FUN = median
)

##supp duplicate and plot
library(ggrepel)
suppressWarnings({
  fig1a=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = umap_1, y = umap_2, label = celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 4,
      box.padding = 0.4,
      label.size = 0.25,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "umap 1", y = "umap 2")
})
fig1a
ggsave(ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPFcellT.png", plot = fig1a, width = 12, height = 8))


```





```{r}
#------------------------------

###with the markers
p=DotPlot(
  IPF.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T ,
  group.by = "celltype") + RotatedAxis() 
 p 
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPFmarkers.png", plot = p, width = 16, height = 8)
```





```{r}
##proportion of cells-------------------

library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(IPF.L@meta.data) %>%
  dplyr::count(`sample`, celltype) %>%
  dplyr::group_by(`sample`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
IPF.L_meta <- dplyr::as_tibble(IPF.L@meta.data) %>%
  dplyr::rename(sample_id = sample)   # rename to avoid conflicts
prop_df=IPF.L_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPFproportion.png", plot = fig1c, width = 12, height = 10)

```



###6-proportion of Donors
```{r}
library(dplyr)
library(ggplot2)
library(scales)

#change colname 
IPF.L$Donors = IPF.L$orig.ident

#create the meta data
meta = IPF.L@meta.data[, c("harmony_clusters", "Donors")]
table(meta$Donors)

# Make sure dplyr functions are used explicitly
meta = IPF.L@meta.data

#the  percentage 
donors_counts = meta %>%
  as_tibble() %>%
  dplyr::group_by(harmony_clusters,Donors) %>% #count cells each donor use in each cluster
  dplyr::summarise( n= dplyr::n(),.groups = "drop") %>% #numb of cells
  dplyr::group_by(harmony_clusters) %>%
  dplyr::mutate(proportion = n /sum(n))  #fraction of cells per donor
```





```{r, fig.width=10, fig.height=}

#####
p=ggplot(donors_counts, aes(
  x = harmony_clusters,
  y = proportion,
  group = Donors,
  color = Donors
)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  # Use a colorblind-friendly palette with distinct colors
  scale_color_brewer(palette = "Set1", name = "Donor") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Donors Proportion ",
    x = "Cluster",
    y = "Proportion (%)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
p
ggsave(filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPFDonor.png", plot = p, width = 12, height = 8)
```




```{r}

##correlation for celltype

# 1️⃣ Set identities to cell types
Idents(IPF.L) <- "celltype"

# sanity check
levels(Idents(IPF.L))
# should show cell type names

# 2️⃣ Average expression per cell type
av.expIPF <- AverageExpression(
  IPF.L,
  assays = "RNA",
  slot = "data"
)$RNA

# 3️⃣ Cell-type correlation matrix
correlation_matrixIPF <- cor(
  as.matrix(av.expIPF),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixIPF[is.na(correlation_matrixIPF)] <- 0

# 4️⃣ Save heatmap as PNG
png(
  filename = "output/Idiopathic Pulmonary Fibrosis/Low/IPF_celltype_corr.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixIPF,
  clustering_method = "ward.D2",
  main = "Correlation Between Cell Types (IPF – Lower Lung)",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()

```








###save the final meta.data
```{r}

##save
# Keep your existing metadata names
IPF.L$orig.ident="Idiopathic P.Fibrosis in Lower Lung"

# Remove unwanted columns 
cols_to_remove=c(
  grep("res_|RNA_snn_res", colnames(IPF.L@meta.data), value = TRUE),
  "cell_type",
  "iscell_ed",
  "iscell_dd",
  "Donors",
  "x"
)
IPF.L@meta.data=IPF.L@meta.data[, !(colnames(IPF.L@meta.data) %in% cols_to_remove)]

# Ensure active identity reflects your celltype
Idents(IPF.L)=IPF.L$celltype

# Save 
saveRDS(IPF.L, file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/IPF.L_Final.rds")

# Save only the final metadata 
saveRDS(IPF.L@meta.data, file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/IPF.L_metadata.rds")

# Optional CSV 
write.csv(IPF.L@meta.data, 
          file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/IPF.L_metadata.csv",
          row.names = TRUE)
View(IPF.L@meta.data)

```




```{r}
##save
#saveRDS(IPF.L, file = "data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/IPF.L_process.rds")

## run it later to load all 
IPF.L = readRDS("data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/IPF.L_process.rds")
```

```





















