---
title: "Idiopathic Pulmonary Fibrosis - Processing and Annotation"
output: html_notebook
---

<!-- Step 2: Seurat processing for Idiopathic Pulmonary Fibrosis, run
     separately for the Upper and Lower lung lobes - from raw 10x count
     matrices to annotated, harmony-integrated Seurat objects. The
     cell-cell communication (CCC) analysis is a separate step, in
     IPF.sce.rmd. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

## 2. Load raw data and merge samples by lobe

```{r}
library(Seurat)
library(DropletUtils)

data.dir <- "../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Idiopathic Pulmonary Fibrosis/GSE128033_RAW"
save.base <- "data/process/SeuratObjects/IPF"

# sample names grouped by lung lobe (L = Lower, U = Upper)
groups <- list(
  L = c("1L1", "1L2", "2L1", "2L2", "3L"),
  U = c("1U", "2U", "3U")
)

sample.dirs <- list.dirs(data.dir, recursive = FALSE)

for (folder_path in sample.dirs) {
  folder_name <- basename(folder_path)
  message("Processing ", folder_name)
  counts <- Read10X(data.dir = folder_path)

  # droplet filtering
  iscell_dd <- defaultDrops(counts, expected = 5000)
  eout <- emptyDrops(counts, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed <- eout$FDR <= 0.01
  meta <- data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    row.names = colnames(counts)
  )
  keep <- iscell_dd | iscell_ed

  IPF <- CreateSeuratObject(
    counts = counts[, keep],
    meta.data = meta[keep, ],
    project = folder_name
  )
  IPF$sample <- folder_name

  # save each sample under its own lobe (Upper/Lower) subfolder
  subfolder <- ifelse(folder_name %in% groups$U, "Upper", "Lower")
  saveRDS(IPF, file = file.path(save.base, subfolder, paste0(folder_name, ".rds")))
}

# merge the per-sample objects into one Seurat object per lobe
for (grp in names(groups)) {
  folders <- groups[[grp]]
  subfolder <- ifelse(grp == "L", "Lower", "Upper")

  seurat_list <- lapply(folders, function(folder_name) {
    file_path <- file.path(save.base, subfolder, paste0(folder_name, ".rds"))
    message("Loading ", file_path)
    readRDS(file_path)
  })
  names(seurat_list) <- folders

  if (length(seurat_list) > 1) {
    combined <- Reduce(function(x, y) merge(x, y), seurat_list)
  } else {
    combined <- seurat_list[[1]]
  }

  saveRDS(combined, file = file.path(save.base, subfolder, paste0("IPF_", grp, ".rds")))
  message("Saved merged Seurat object: IPF_", grp, " in ", file.path(save.base, subfolder))
}

rm(combined)
gc(full = TRUE)
```

# Upper lobe

## 3. Quality control and filtering

```{r, fig.width=10, fig.height=8}
IPF.U <- readRDS("data/process/SeuratObjects/IPF/Upper/IPF_U.rds")
DefaultAssay(IPF.U) # confirm default assay is RNA

IPF.U[["percent.mt"]] <- PercentageFeatureSet(IPF.U, pattern = "^MT-")
Idents(IPF.U) <- "IPF_Upper" # single group, just for the QC plot
VlnPlot(IPF.U, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC threshold, set from this sample's own distributions
IPF.U <- subset(IPF.U, subset = nFeature_RNA > 200 & nFeature_RNA < 7000 & percent.mt < 10)
```

## 4. Normalization, scaling, and PCA

```{r}
IPF.U <- NormalizeData(IPF.U)
IPF.U <- FindVariableFeatures(IPF.U)
IPF.U <- ScaleData(IPF.U, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = IPF.U))
IPF.U <- RunPCA(IPF.U, features = VariableFeatures(object = IPF.U))
ElbowPlot(IPF.U)
IPF.U <- RunUMAP(IPF.U, dims = 1:20)
```

## 5. Harmony integration and clustering resolutions

```{r}
library(harmony)
IPF.U <- RunHarmony(IPF.U, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:20)
IPF.U[["RNA"]] <- JoinLayers(IPF.U[["RNA"]])
IPF.U <- RunUMAP(IPF.U, reduction = "harmony", dims = 1:20, reduction.name = "umap.harmony")

# scan a range of resolutions before picking one
IPF.U <- FindNeighbors(IPF.U, reduction = "harmony", dims = 1:20)
for (res in seq(0.2, 1, by = 0.1)) {
  IPF.U <- FindClusters(IPF.U, resolution = res)
  colname <- paste0("res_", res)
  IPF.U[[colname]] <- IPF.U$seurat_clusters
}
res_cols <- grep("^res_", colnames(IPF.U@meta.data), value = TRUE)

# preview each resolution
for (res in res_cols) {
  p <- DimPlot(IPF.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}

# save all resolutions to a single PDF for comparison
pdf("output/IPF/Upper/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(IPF.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste(res))
  print(p)
}
dev.off()

# resolution 0.5 was chosen for downstream clustering
p <- DimPlot(IPF.U, group.by = "res_0.5", label = TRUE, reduction = "umap.harmony", label.size = 6) +
  ggtitle("Resolution: 0.5") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/IPF/Upper/Resolution.pdf", plot = p, width = 10, height = 8, dpi = 300)
```

## 6. Clustering and marker genes

```{r}
IPF.U <- FindNeighbors(IPF.U, reduction = "harmony", dims = 1:20)
IPF.U <- FindClusters(IPF.U, resolution = 0.5, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
IPF.Umarkers <- FindAllMarkers(IPF.U, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
IPF.Umarkers <- IPF.Umarkers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 7. Marker genes for cell type annotation

```{r}
genes <- c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  # Mural cells
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)

p <- DotPlot(
  IPF.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "harmony_clusters"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/IPF/Upper/cluster.pdf", plot = p, width = 20, height = 8)
```

## 8. Subclustering

```{r}
l <- FeaturePlot(IPF.U, features = unique(genes), combine = FALSE, order = TRUE, reduction = "umap.harmony")
names(l) <- unique(genes)

# check cluster 3 for finer structure
FeaturePlot(IPF.U, features = c("CD4", "COTL1", "CCR7", "LEF1"), combine = FALSE, order = TRUE, reduction = "umap.harmony")

Idents(IPF.U) <- IPF.U$harmony_clusters
IPF.U <- FindSubCluster(IPF.U, cluster = c(3), resolution = 0.1, graph.name = "RNA_snn")
P <- DimPlot(IPF.U, group.by = "sub.cluster", reduction = "umap.harmony", label = TRUE, label.size = 7) +
  labs(x = "UMAP.h", y = "UMAP.h")
P
ggsave(filename = "output/IPF/Upper/sub.cluster.pdf", plot = P, width = 10, height = 8)
```

## 9.cell type labels

```{r}
Idents(IPF.U) <- "sub.cluster"
cluster_name_x <- c(
  "0" = "Macrophages_0",
  "1" = "Macrophages_1",
  "2" = "Macrophages_2",
  "3_0" = "CD4_3_0",
  "3_1" = "CD8_3_1",
  "4" = "Goblet Cells_4",
  "5" = "Monocytes_5",
  "6" = "NK Cells_6",
  "7" = "Peribronchial Endothelial_7",
  "8" = "Arterial Endothelial_8",
  "9" = "Mural Cells_9",
  "10" = "Fibroblasts_10",
  "11" = "Ciliated Cells_11",
  "12" = "Fibroblasts_12",
  "13" = "Alveolar Cells_13",
  "14" = "Mast Cells_14",
  "15" = "Proliferating Cells_15",
  "16" = "B Cells_16",
  "17" = "Macrophages_17",
  "18" = "Lymphatic Endothelial_18"
)

IPF.U <- RenameIdents(IPF.U, cluster_name_x)
IPF.U$CellType <- Idents(IPF.U)

p <- DotPlot(
  IPF.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "CellType"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)

pdf("/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/output/IPF/Upper/IPFU_cluster_name.pdf", width = 20, height = 10)
print(p)
dev.off()
```

## 10. Major cell types and UMAP

```{r, fig.width=12, fig.height=8}
Idents(IPF.U) <- "sub.cluster"
cluster_name <- c(
  "0" = "Macrophages",
  "1" = "Macrophages",
  "2" = "Macrophages",
  "3_0" = "CD4",
  "3_1" = "CD8",
  "4" = "Goblet Cells",
  "5" = "Monocytes",
  "6" = "NK Cells",
  "7" = "Peribronchial Endothelial",
  "8" = "Arterial Endothelial",
  "9" = "Mural Cells",
  "10" = "Fibroblasts",
  "11" = "Ciliated Cells",
  "12" = "Fibroblasts",
  "13" = "Alveolar Cells",
  "14" = "Mast Cells",
  "15" = "Proliferating Cells",
  "16" = "B Cells",
  "17" = "Macrophages",
  "18" = "Lymphatic Endothelial"
)

IPF.U <- RenameIdents(IPF.U, cluster_name)
IPF.U$Major_Celltype <- Idents(IPF.U)

celltype_cols <- c(
  "Macrophages" = "orangered",
  "CD8" = "tomato",
  "CD4" = "olivedrab2",
  "Venous Endothelial" = "chocolate1",
  "Monocytes" = "#2ca02c",
  "NK Cells" = "magenta",
  "B Cells" = "gold",
  "Arterial Endothelial" = "cadetblue3",
  "Goblet Cells" = "#e377c2",
  "Alveolar Cells" = "aquamarine",
  "Mast Cells" = "#FDBF6F",
  "Mural Cells" = "#1F78B4",
  "Proliferating Cells" = "#B15928",
  "Lymphatic Endothelial" = "mediumorchid1",
  "Ciliated Cells" = "saddlebrown",
  "Fibroblasts" = "cyan",
  "Peribronchial Endothelial" = "darkgoldenrod2"
)

# UMAP with a label placed at each cell type's centroid
umap_df <- as.data.frame(Embeddings(IPF.U, "umap.harmony"))
umap_df$Major_Celltype <- IPF.U$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

library(ggrepel)
fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 6,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

ggsave(filename = "output/IPF/Upper/IPFU_CellType.pdf", plot = fig1a, width = 12, height = 10, dpi = 300)
```

## 11. Marker expression heatmap by major cell type

```{r}
Idents(IPF.U) <- "Major_Celltype"
DefaultAssay(IPF.U) <- "RNA"
expr_mat <- GetAssayData(IPF.U, layer = "data")

# percent of cells expressing each marker, per cell type
percent_mat <- sapply(levels(Idents(IPF.U)), function(ct) {
  cells <- WhichCells(IPF.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  IPF.U,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf("output/IPF/Upper/markers.pdf", width = 10, height = 15)
ComplexHeatmap::draw(ht)
dev.off()
```

## 12. Broader cell-type grouping (General)

```{r}
unique(IPF.U$Major_Celltype)

IPF.U$General <- recode(IPF.U$Major_Celltype,
  "NK Cells" = "NK Cells",
  "Macrophages" = "Mono/Mac",
  "CD8" = "CD8",
  "CD4" = "CD4",
  "Mast Cells" = "Mast Cells",
  "Mural Cells" = "Mural Cells",
  "Alveolar Cells" = "Epithelial Cells",
  "Arterial Endothelial" = "Endothelial Cells",
  "Proliferating Cells" = "Proliferating Cells",
  "Monocytes" = "Mono/Mac",
  "Fibroblasts" = "Fibroblasts",
  "B Cells" = "B Cells",
  "Peribronchial Endothelial" = "Endothelial Cells",
  "Goblet Cells" = "Epithelial Cells",
  "Lymphatic Endothelial" = "Endothelial Cells",
  "Ciliated Cells" = "Epithelial Cells"
)
unique(IPF.U$General)
```

## 13. TLS (tertiary lymphoid structure) marker heatmap

```{r}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(
  germinal_center, inflammation,
  lymphoid_chemokines, hev, fdc
))
markers <- markers[markers %in% rownames(IPF.U)]

marker_groups <- case_when(
  markers %in% germinal_center ~ "Germinal Center",
  markers %in% inflammation ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev ~ "HEV",
  markers %in% fdc ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center" = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines" = "#35978F",
  "HEV" = "#8073AC",
  "FDC" = "#E08214"
)

# percent of cells expressing each marker, per cell type
Idents(IPF.U) <- "General"
DefaultAssay(IPF.U) <- "RNA"
expr_mat <- GetAssayData(IPF.U, layer = "data")

percent_mat <- sapply(levels(Idents(IPF.U)), function(ct) {
  cells <- WhichCells(IPF.U, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  IPF.U,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/IPF/Upper/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 14. Save the processed object

```{r}
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General",
  "sub.cluster"
)
IPF.U@meta.data <- IPF.U@meta.data[, keep_cols]

write.csv(IPF.U@meta.data, file = "data/process/SeuratObjects/IPF/Upper/IPF.U_metadata.csv", row.names = TRUE)

saveRDS(IPF.U, file = "data/process/SeuratObjects/IPF/Upper/IPF.U_process.rds")


```

## 15. Export for gene regulatory network analysis (SCENIC)

```{r}
# Export the raw counts as a sparse matrix + gene/barcode lists, so they can
# be imported by pySCENIC in the terminal.
expr <- GetAssayData(IPF.U, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Upper/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Upper/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Upper/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

################################          Lower lobe              ####################

## 16. Quality control and filtering

```{r}
IPF.L <- readRDS("data/process/SeuratObjects/IPF/Lower/IPF_L.rds")
DefaultAssay(IPF.L) # confirm default assay is RNA

IPF.L[["percent.mt"]] <- PercentageFeatureSet(IPF.L, pattern = "^MT-")
Idents(IPF.L) <- "x" # single group, just for the QC plot
IPF.L$x <- "All"
VlnPlot(IPF.L, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC threshold, set from this sample's own distributions
IPF.L <- subset(IPF.L, subset = nFeature_RNA > 200 & nFeature_RNA < 7000 & percent.mt < 10)

```

## 17. Normalization, scaling, and PCA

```{r}

IPF.L <- NormalizeData(IPF.L)
IPF.L <- FindVariableFeatures(IPF.L)
IPF.L <- ScaleData(IPF.L, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = IPF.L))
IPF.L <- RunPCA(IPF.L, features = VariableFeatures(object = IPF.L))
ElbowPlot(IPF.L)

```

## 18. Harmony integration and clustering resolutions

```{r}
library(harmony)
IPF.L <- RunHarmony(IPF.L, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:20)
IPF.L[["RNA"]] <- JoinLayers(IPF.L[["RNA"]])
IPF.L <- RunUMAP(IPF.L, reduction = "harmony", dims = 1:20, reduction.name = "umap.harmony")
p <- DimPlot(IPF.L, reduction = "umap.harmony", group.by = c("sample"), label.size = 2, shuffle = TRUE)
p
ggsave(filename = "output/IPF/Low/IPFintegration.png", plot = p, width = 12, height = 8)

# scan a range of resolutions before picking one
IPF.L <- FindNeighbors(IPF.L, reduction = "harmony", dims = 1:20)
for (res in seq(0.2, 1, by = 0.1)) {
  IPF.L <- FindClusters(IPF.L, resolution = res)
  colname <- paste0("res_", res)
  IPF.L[[colname]] <- IPF.L$seurat_clusters
}
res_cols <- grep("^res_", colnames(IPF.L@meta.data), value = TRUE)

# preview each resolution
for (res in res_cols) {
  p <- DimPlot(IPF.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}

# resolution 0.5 was chosen for downstream clustering
p <- DimPlot(IPF.L, group.by = "res_0.5", label = TRUE, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.5") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/IPF/Low/IPF.L_resolution.pdf", plot = p, width = 10, height = 8, dpi = 300)
```

## 19. Clustering and marker genes

```{r}
IPF.L <- FindNeighbors(IPF.L, reduction = "harmony", dims = 1:20)
IPF.L <- FindClusters(IPF.L, resolution = 0.5, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
IPF.Lmarkers <- FindAllMarkers(IPF.L, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
IPF.Lmarkers <- IPF.Lmarkers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 20. Marker genes for cell type annotation

```{r}
genes <- c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "MZB1",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  # Mural Cells
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)
```

```{r}
l <- FeaturePlot(IPF.L, features = unique(genes), combine = FALSE, order = TRUE, reduction = "umap.harmony")
names(l) <- unique(genes)
```

```{r}
p <- DotPlot(
  IPF.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "harmony_clusters"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)

ggsave(filename = "output/IPF/Low/IPFL_cluster.pdf", plot = p, width = 20, height = 8)
```

## 21. Fine-grained cell type labels

```{r}
Idents(IPF.L) <- "harmony_clusters"
cluster_name_x <- c(
  "0" = "Macrophages_0",
  "1" = "Venous Endothelial_1",
  "2" = "Fibroblasts_2",
  "3" = "CD4_3",
  "4" = "Goblet Cells_4",
  "5" = "CD8_5",
  "6" = "Mast Cells_6",
  "7" = "Alveolar Cells_7",
  "8" = "Monocytes_8",
  "9" = "Ciliated Cells_9",
  "10" = "Basal Cells_10",
  "11" = "Macrophages_11",
  "12" = "Mural Cells_12",
  "13" = "Venous Endothelial_13",
  "14" = "Macrophages_14",
  "15" = "Fibroblasts_15",
  "16" = "B Cells_16",
  "17" = "Proliferating Cells_17",
  "18" = "Lymphatic Endothelial_18"
)

IPF.L <- RenameIdents(IPF.L, cluster_name_x)
IPF.L$CellType <- Idents(IPF.L)

p <- DotPlot(
  IPF.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "CellType"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/IPF/Low/IPFL_cluster_name.pdf", plot = p, width = 20, height = 10)
```

## 22. Major cell types and UMAP

```{r, fig.width=14, fig.height=8}
Idents(IPF.L) <- "harmony_clusters"
cluster_name <- c(
  "0" = "Macrophages",
  "1" = "Venous Endothelial",
  "2" = "Fibroblasts",
  "3" = "CD4",
  "4" = "Goblet Cells",
  "5" = "CD8",
  "6" = "Mast Cells",
  "7" = "Alveolar Cells",
  "8" = "Monocytes",
  "9" = "Ciliated Cells",
  "10" = "Basal Cells",
  "11" = "Macrophages",
  "12" = "Mural Cells",
  "13" = "Venous Endothelial",
  "14" = "Macrophages",
  "15" = "Fibroblasts",
  "16" = "B Cells",
  "17" = "Proliferating Cells",
  "18" = "Lymphatic Endothelial"
)

IPF.L <- RenameIdents(IPF.L, cluster_name)
IPF.L$Major_Celltype <- Idents(IPF.L)

celltype_cols <- c(
  "Venous Endothelial" = "chocolate1",
  "Lymphatic Endothelial" = "mediumorchid1",
  "CD4" = "olivedrab2",
  "B Cells" = "gold",
  "Monocytes" = "#2ca02c",
  "Macrophages" = "orangered",
  "Mast Cells" = "#FDBF6F",
  "Alveolar Cells" = "aquamarine",
  "Ciliated Cells" = "saddlebrown",
  "Goblet Cells" = "cornflowerblue",
  "Basal Cells" = "indianred3",
  "Proliferating Cells" = "#1F78B4",
  "Mural Cells" = "#1F78B4",
  "Fibroblasts" = "cyan",
  "CD8" = "tomato"
)

# UMAP with a label placed at each cell type's centroid
umap_df <- as.data.frame(Embeddings(IPF.L, "umap.harmony"))
umap_df$Major_Celltype <- IPF.L$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 6,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

fig1a
ggsave(filename = "output/IPF/Low/IPFL_cellType.pdf", plot = fig1a, width = 12, height = 10)
```

## 23. Marker expression heatmap by major cell type

```{r}
Idents(IPF.L) <- "Major_Celltype"
DefaultAssay(IPF.L) <- "RNA"
expr_mat <- GetAssayData(IPF.L, layer = "data")

# percent of cells expressing each marker, per cell type
percent_mat <- sapply(levels(Idents(IPF.L)), function(ct) {
  cells <- WhichCells(IPF.L, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  IPF.L,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf("output/IPF/Low/IPF.L_markers.pdf", width = 10, height = 15)
ComplexHeatmap::draw(ht)
dev.off()
```

## 24. Correlation between cell types

```{r}
Idents(IPF.L) <- "Major_Celltype"
levels(Idents(IPF.L))

# average expression per cell type
av.expIPL <- AverageExpression(
  IPF.L,
  assays = "RNA",
  slot = "data"
)$RNA

correlation_matrixIPL <- cor(
  as.matrix(av.expIPL),
  use = "pairwise.complete.obs"
)
correlation_matrixIPL[is.na(correlation_matrixIPL)] <- 0

pdf(
  file = "output/IPF/Low/IPFL_correlation.pdf",
  width = 8,
  height = 10
)
pheatmap(
  correlation_matrixIPL,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)
dev.off()
```

## 25. Broader cell-type grouping (General)

```{r}
unique(IPF.L$Major_Celltype)

IPF.L$General <- recode(IPF.L$Major_Celltype,
  "Macrophages" = "Mono/Mac",
  "Fibroblasts" = "Fibroblasts",
  "Goblet Cells" = "Epithelial Cells",
  "CD4" = "CD4",
  "Ciliated Cells" = "Epithelial Cells",
  "Alveolar Cells" = "Epithelial Cells",
  "CD8" = "CD8",
  "Proliferating Cells" = "Proliferating Cells",
  "Venous Endothelial" = "Endothelial Cells",
  "Monocytes" = "Mono/Mac",
  "Mast Cells" = "Mast Cells",
  "Mural Cells" = "Mural Cells",
  "B Cells" = "B Cells",
  "Lymphatic Endothelial" = "Endothelial Cells",
  "Basal Cells" = "Epithelial Cells"
)
```

## 26. TLS (tertiary lymphoid structure) marker heatmap

```{r}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(
  germinal_center, inflammation,
  lymphoid_chemokines, hev, fdc
))
markers <- markers[markers %in% rownames(IPF.L)]

marker_groups <- case_when(
  markers %in% germinal_center ~ "Germinal Center",
  markers %in% inflammation ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev ~ "HEV",
  markers %in% fdc ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center" = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines" = "#35978F",
  "HEV" = "#8073AC",
  "FDC" = "#E08214"
)

# percent of cells expressing each marker, per cell type
Idents(IPF.L) <- "General"
DefaultAssay(IPF.L) <- "RNA"
expr_mat <- GetAssayData(IPF.L, layer = "data")

percent_mat <- sapply(levels(Idents(IPF.L)), function(ct) {
  cells <- WhichCells(IPF.L, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  IPF.L,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/IPF/Low/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 27. Export for gene regulatory network analysis (SCENIC)

```{r}
# Export the raw counts as a sparse matrix + gene/barcode lists, so they can
# be imported by pySCENIC in the terminal.
expr <- GetAssayData(IPF.L, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Low/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Low/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Low/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

## 28. Save the processed object

```{r}
write.csv(IPF.L@meta.data,
  file = "data/process/SeuratObjects/IPF/Lower/IPF.L_metadata.csv",
  row.names = TRUE
)

saveRDS(IPF.L, file = "data/process/SeuratObjects/IPF/Lower/IPF.L_Final.rds")
IPF.L <- readRDS("data/process/SeuratObjects/IPF/Lower/IPF.L_Final.rds")
```
