---
title: "R Notebook"
output: html_notebook
---
############################################################################################################################
###ANALYSIS OF IDIOPATIC PULMONARY FIBROSIS#############################################################################
##############################################################################################################################


#THE PACKAGES
```{r setup, include=FALSE}

##################PACKAGES
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")


```



###1load the data
```{r}
#data for the Lower lung(L) and Upper(U)
library(Seurat)
library(DropletUtils)

data.dir="../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Idiopathic Pulmonary Fibrosis/GSE128033_RAW"

# List all sample folders
sample.dirs=list.dirs(data.dir, recursive = FALSE)

for (folder_path in sample.dirs) {
  folder_name=basename(folder_path)  
  message("Processing ", folder_name)
  counts=Read10X(data.dir = folder_path)
  # Droplet filtering
  iscell_dd=defaultDrops(counts, expected = 5000)
  eout=emptyDrops(counts, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed=eout$FDR <= 0.01
  # Metadata
  meta=data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    row.names = colnames(counts)
  )
  # Keep droplets
  keep=iscell_dd | iscell_ed
  
  # Create Seurat object
  IPF=CreateSeuratObject(
    counts = counts[, keep],
    meta.data = meta[keep, ],
    project = folder_name
  )
  # Add sample ID
  IPF$sample=folder_name
  # Save
  saveRDS(IPF, file = paste0("data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis/Lower/", folder_name, ".rds"))
}
#merge in two seurat Upper and Lung
save.base="data/process/SeuratObjects/Idiopathic Pulmonary Fibrosis"

groups=list(
  L = c("1L1","1L2","2L1","2L2","3L"),
  U = c("1U","2U","3U")
)
for (grp in names(groups)) {
  folders=groups[[grp]]
  
  seurat_list=lapply(folders, function(folder_name) {
    # Correct path: point to Lower or Upper subfolder
    subfolder=ifelse(grp == "L", "Lower", "Upper")
    file_path=file.path(save.base, subfolder, paste0(folder_name, ".rds"))
    message("Loading ", file_path)
    readRDS(file_path)
  })
  names(seurat_list)=folders
  
  # Merge all Seurat objects in the group
  if (length(seurat_list) > 1) {
    combined=Reduce(function(x, y) merge(x, y), seurat_list)
  } else {
    combined=seurat_list[[1]]
  }
  
  # Save merged 
  save_dir=file.path(save.base, ifelse(grp == "L", "Lower", "Upper"))
  saveRDS(combined, file = file.path(save_dir, paste0("IPF_", grp, ".rds")))
  message("Saved merged Seurat object: IPF_", grp, " in ", save_dir)
}
rm(IPF.Up)
gc(full = TRUE)
```



##########IPF UPPER--------------###################




###2-qc and threshold
```{r,fig.width=10,fig.height=8}
##qc
IPF.U=readRDS("data/process/SeuratObjects/IPF/Upper/IPF_U.rds")
DefaultAssay(IPF.U) # if not RNA set it as RNA
IPF.U[["percent.mt"]]=PercentageFeatureSet(IPF.U,pattern = "^MT-")
Idents(IPF.U)="IPF_Upper"
VlnPlot(IPF.U, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log = TRUE)
#threshold
IPF.U=subset(IPF.U, subset=nFeature_RNA>200 & nFeature_RNA<7000 & percent.mt <10)

```




###2-statistics
```{r}
###
IPF.U=NormalizeData(IPF.U)
IPF.U=FindVariableFeatures(IPF.U)
IPF.U=ScaleData(IPF.U,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = IPF.U))
IPF.U=RunPCA(IPF.U, features = VariableFeatures(object = IPF.U))
ElbowPlot(IPF.U)
IPF.U=RunUMAP(IPF.U, dims = 1:20)

```



###3-integration

```{r}

##integration
library(harmony)
IPF.U=RunHarmony(IPF.U,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:20  )
IPF.U[["RNA"]]=JoinLayers(IPF.U[["RNA"]])
IPF.U = RunUMAP(IPF.U, reduction = "harmony", dims= 1:20, reduction.name = "umap.harmony")


#for the resolution
IPF.U=FindNeighbors(IPF.U,reduction = "harmony",dims = 1:20)
for (res in seq(0.2, 1, by = 0.1)) {
  IPF.U=FindClusters(IPF.U, resolution = res)
  colname=paste0("res_", res)
  IPF.U[[colname]]=IPF.U$seurat_clusters
}
res_cols=grep("^res_", colnames(IPF.U@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(IPF.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")

}

#save all
pdf("output/IPF/Upper/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(IPF.U, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste(res))
  print(p)
}
dev.off()





##save
p = DimPlot(IPF.U, group.by = "res_0.5", label = TRUE, reduction = "umap.harmony", label.size = 6) + 
    ggtitle("Resolution: 0.5") + 
    NoLegend() +
    labs(x = "UMAP.h", y = "UMAP.h")

ggsave(filename = "output/IPF/Upper/Resolution.pdf", plot = p, width = 10, height = 8, dpi = 300)

```






###4-cluster and markers
```{r}
###clusters

IPF.U=FindNeighbors(IPF.U,reduction = "harmony",dims = 1:20)
IPF.U=FindClusters(IPF.U, resolution = 0.5,cluster.name = "harmony_clusters")
#markers
IPF.Umarkers=FindAllMarkers(IPF.U,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
IPF.Umarkers = IPF.Umarkers%>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)


```





###5-annotation



```{r}

genes = c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "IGLL5",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Secretory
  # "SCGB1A1", "MUC5B", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  #Mural cells
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)

p=DotPlot(
  IPF.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "harmony_clusters") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/IPF/Upper/cluster.pdf", plot = p, width = 20, height = 8)


```






```{r}
l = FeaturePlot(IPF.U, features = unique(genes), combine = F, order = T, reduction = "umap.harmony")
names(l) = unique(genes)

#find subclusters
FeaturePlot(IPF.U, features = c("CD4","COTL1", "CCR7", "LEF1"), combine = F, order = T, reduction = "umap.harmony")



Idents(IPF.U) = IPF.U$harmony_clusters
IPF.U = FindSubCluster(IPF.U, cluster = c(3), resolution = 0.1, graph.name = "RNA_snn")
P=DimPlot(IPF.U, group.by = "sub.cluster", reduction = "umap.harmony", label = T,label.size = 7)+
  labs(x = "UMAP.h", 
         y = "UMAP.h")
P
ggsave(filename = "output/IPF/Upper/sub.cluster.pdf", plot = P, width = 10, height = 8)


```







```{r}

#----cluster_specifique_name
Idents(IPF.U)="sub.cluster"
cluster_name_x=c(
  "0"  = "Macrophages_0"                  ,
  "1"  = "Macrophages_1"                   ,
  "2"  = "Macrophages_2",
  "3_0"  = "CD4_3_0"            ,
  "3_1"  = "CD8_3_1",
  "4"  = "Goblet Cells_4"                     ,
  "5"  = "Monocytes_5"                       ,
  "6"  = "NK Cells_6"                    ,
  "7"  = "Peribronchial Endothelial_7"            ,
  "8"  = "Arterial Endothelial_8"                          ,
  "9"  = "Mural Cells_9"                     ,
  "10" = "Fibroblasts_10"                  ,          
  "11" = "Ciliated Cells_11"                      ,
  "12" = "Fibroblasts_12"                     ,
  "13" = "Alveolar Cells_13"             ,
  "14" = "Mast Cells_14"             ,
  "15" = "Proliferating Cells_15"                  ,
  "16" = "B Cells_16"          ,
  "17" = "Macrophages_17"   ,
  "18" = "Lymphatic Endothelial_18"
  
)

IPF.U=RenameIdents(IPF.U, cluster_name_x)
IPF.U$CellType=Idents(IPF.U)

#
p=DotPlot(
  IPF.U,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "CellType") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)


pdf("/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/output/IPF/Upper/IPFU_cluster_name.pdf",
    width = 20, height = 10)
   print(p)
dev.off()

```







```{r,fig.width=12, fig.height=8}

#---------umap--------


##1celltype
Idents(IPF.U)="sub.cluster"
cluster_name=c(
  "0"  = "Macrophages"                  ,
  "1"  = "Macrophages"                   ,
  "2"  = "Macrophages",
  "3_0"  = "CD4"            ,
  "3_1"  = "CD8",
  "4"  = "Goblet Cells"                     ,
  "5"  = "Monocytes"                       ,
  "6"  = "NK Cells"                    ,
  "7"  = "Peribronchial Endothelial"            ,
  "8"  = "Arterial Endothelial"                          ,
  "9"  = "Mural Cells"                     ,
  "10" = "Fibroblasts"                  ,          
  "11" = "Ciliated Cells"                      ,
  "12" = "Fibroblasts"                     ,
  "13" = "Alveolar Cells"             ,
  "14" = "Mast Cells"             ,
  "15" = "Proliferating Cells"                  ,
  "16" = "B Cells"          ,
  "17" = "Macrophages"   ,
  "18" = "Lymphatic Endothelial"   
  
)

IPF.U=RenameIdents(IPF.U, cluster_name)
IPF.U$Major_Celltype=Idents(IPF.U)
##2color
celltype_cols <- c(
  "Macrophages"           = "orangered",
  "CD8"                   = "tomato",
  "CD4"                  =  "olivedrab2",
  "Venous Endothelial"    = "chocolate1",
  "Monocytes"             = "#2ca02c",
  "NK Cells"              = "magenta",
  "B Cells"               = "gold",
  "Arterial Endothelial"  = "cadetblue3",
  "Goblet Cells"          = "#e377c2",
  "Alveolar Cells"        = "aquamarine",
  "Mast Cells"            = "#FDBF6F",
  "Mural Cells"           = "#1F78B4",
  "Proliferating Cells"   = "#B15928",
  "Lymphatic Endothelial" = "mediumorchid1",
  "Ciliated Cells"        = "saddlebrown",
  "Fibroblasts"           = "cyan",
  "Peribronchial Endothelial" = "darkgoldenrod2"
)

##3 center
umap_df <- as.data.frame(Embeddings(IPF.U, "umap.harmony"))
umap_df$Major_Celltype <- IPF.U$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)
##plot
library(ggrepel)
fig1a = ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 6,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

ggsave(filename = "output/IPF/Upper/IPFU_CellType.pdf", plot = fig1a, width = 12, height = 10, dpi = 300)

```









```{r}
#----same methods
# for markers
Idents(IPF.U) <- "Major_Celltype"


# -------------------------------
DefaultAssay(IPF.U) <- "RNA"
expr_mat <- GetAssayData(IPF.U, layer = "data")

# -------------------------------
# Percent expression matrix

percent_mat <- sapply(levels(Idents(IPF.U)), function(ct) {
  cells <- WhichCells(IPF.U, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# -------------------------------
# Average expression matrix

avg_exp <- AverageExpression(
  IPF.U,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

# -------------------------------
# Clean data

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

# -------------------------------
# Scaling

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# -------------------------------
# Color function

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# -------------------------------
# Dot rendering

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# -------------------------------
# Cell type colors (annotation)

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# -------------------------------
# plot

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,

  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),

  column_names_gp = grid::gpar(
    fontsize = 10,
    fontface = "bold"
  ),

  column_names_rot = 90,

  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

# -------------------------------
# save

pdf("output/IPF/Upper/markers.pdf",
    width = 10, height = 15)

ComplexHeatmap::draw(ht)

dev.off()
```






###7 Class Major_Celltype with family


```{r}
#------
#extract
unique(IPF.U$Major_Celltype)

#
IPF.U$General <- recode(IPF.U$Major_Celltype,
                    "NK Cells" = "NK Cells",
                    "Macrophages" = "Mono/Mac",
                    "CD8" = "CD8",
                    "CD4" = "CD4",
                    "Mast Cells" = "Mast Cells",
                    "Mural Cells" = "Mural Cells",
                    "Alveolar Cells" = "Epithelial Cells",
                    "Arterial Endothelial" = "Endothelial Cells",
                    "Proliferating Cells" = "Proliferating Cells",
                    "Monocytes" = "Mono/Mac",
                    "Fibroblasts" = "Fibroblasts",
                    "B Cells" = "B Cells",
                    "Peribronchial Endothelial" = "Endothelial Cells",
                    "Goblet Cells" = "Epithelial Cells",
                    "Lymphatic Endothelial" = "Endothelial Cells",
                    "Ciliated Cells" = "Epithelial Cells" 
                      
)
unique(IPF.U$General)
```


##markers that we can have in TLS 


```{r}


### TLS markers that we can have in TLS 


# Define markers per group from genes_TLS.csv
germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC", 
                     "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")

inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", 
                  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5", 
                  "ICAM1", "VCAM1", "SELE")

lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12", 
                         "CXCR5", "CCR7", "CXCR4")

hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2", 
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")

fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP", 
         "CXCL13", "CD54", "ICAM1", "FCER2")

# Combine all markers in order
markers <- unique(c(germinal_center, inflammation, 
                    lymphoid_chemokines, hev, fdc))

# Filter to genes present in your Seurat object
markers <- markers[markers %in% rownames(IPF.U)]

# Build group vector in same order
marker_groups <- case_when(
  markers %in% germinal_center    ~ "Germinal Center",
  markers %in% inflammation       ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev                ~ "HEV",
  markers %in% fdc                ~ "FDC"
)
names(marker_groups) <- markers

# Group colors
group_cols <- c(
  "Germinal Center"         = "#2166AC",
  "Inflammation & Cytokines"= "#D6604D",
  "Lymphoid Chemokines"     = "#35978F",
  "HEV"                     = "#8073AC",
  "FDC"                     = "#E08214"
)

# Extract expression and percent data
Idents(IPF.U) <- "General"
DefaultAssay(IPF.U) <- "RNA"
expr_mat <- GetAssayData(IPF.U, layer = "data")

percent_mat <- sapply(levels(Idents(IPF.U)), function(ct) {
  cells <- WhichCells(IPF.U, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  IPF.U,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# Color function
q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# Bubble cell function
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# Column annotation
celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# Row annotation
row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

# Heatmap
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/IPF/Upper/TLS_markers.pdf",
    width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()



```







###Save
```{r}

keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General",
  "sub.cluster"
  
)

IPF.U@meta.data <- IPF.U@meta.data[, keep_cols]
#CSV 
write.csv(IPF.U@meta.data,file = "data/process/SeuratObjects/IPF/Upper/IPF.U_metadata.csv",row.names = TRUE)

saveRDS(IPF.U, file = "data/process/SeuratObjects/IPF/Upper/IPF.U_process.rds")
## run it later to load all 
IPF.U = readRDS("data/process/SeuratObjects/IPF/Upper/IPF.U_process.rds")


```








```{r}
#####for pyscenic 
expr <- GetAssayData(IPF.U, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Upper/input/matrix.mtx")

write.table(rownames(expr),
            file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Upper/input/genes.tsv",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

write.table(colnames(expr),
            file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Upper/input/barcodes.tsv",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

```








----------------------------------------------------------___________---------------------------------------------------------------
----------------------------------------------------------------------------------------------------------------------------
#####STEP-3 ANALYSIS FOR IPF in LOWER lUNG
-----------------------------------------------------------------------------------------------------------------------






###1-qc
```{r}
##qc---------

IPF.L=readRDS("data/process/SeuratObjects/IPF/Lower/IPF_L.rds")
DefaultAssay(IPF.L) # if not RNA set is as RNA
IPF.L[["percent.mt"]]=PercentageFeatureSet(IPF.L,pattern = "^MT-")
Idents(IPF.L)="x"
IPF.L$x="All"
VlnPlot(IPF.L, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log = TRUE)
#threshold
IPF.L=subset(IPF.L, subset=nFeature_RNA>200 & nFeature_RNA<7000 & percent.mt <10)
```



####2-statistics
```{r}
IPF.L=NormalizeData(IPF.L)
IPF.L=FindVariableFeatures(IPF.L)
IPF.L=ScaleData(IPF.L,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = IPF.L))
IPF.L=RunPCA(IPF.L, features = VariableFeatures(object = IPF.L))
ElbowPlot(IPF.L)
#IPF.L=RunUMAP(IPF.L, dims = 1:20)
#P=DimPlot(IPF.L,reduction = "umap", label = TRUE)

```



###3-integration
```{r}
##integration
library(harmony)
IPF.L=RunHarmony(IPF.L,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:20  )
IPF.L[["RNA"]]=JoinLayers(IPF.L[["RNA"]])
IPF.L = RunUMAP(IPF.L, reduction = "harmony", dims= 1:20, reduction.name = "umap.harmony")
p=DimPlot(IPF.L,reduction = "umap.harmony",group.by = c("sample"),label.size = 2, shuffle = T)
p
ggsave(filename = "output/IPF/Low/IPFintegration.png", plot=p, width = 12, height = 8)

#for the resolution
IPF.L=FindNeighbors(IPF.L,reduction = "harmony",dims = 1:20)
for (res in seq(0.2, 1, by = 0.1)) {
  IPF.L=FindClusters(IPF.L, resolution = res)
  colname=paste0("res_", res)
  IPF.L[[colname]]=IPF.L$seurat_clusters
}
res_cols=grep("^res_", colnames(IPF.L@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(IPF.L, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}

##save
p=DimPlot(IPF.L,group.by = "res_0.5",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.5") +NoLegend()+
  labs(x = "UMAP.h", 
         y = "UMAP.h")
ggsave(filename = "output/IPF/Low/IPF.L_resolution.pdf",plot = p,width = 10,height = 8,dpi = 300)

```


###4 cluster and markers
```{r}
###clusters

IPF.L=FindNeighbors(IPF.L,reduction = "harmony",dims = 1:20)
IPF.L=FindClusters(IPF.L, resolution = 0.5,cluster.name = "harmony_clusters")
#markers
IPF.Lmarkers=FindAllMarkers(IPF.L,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
IPF.Lmarkers = IPF.Lmarkers%>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)

```



###5-annotation


```{r}

genes = c(
  # Myeloid
  "AIF1", "PTPRC",

  # Macrophage
  "FCGR1A", "CD163",

  # SPP1hi Macrophage
  "SPP1", "MERTK", "CCL2",

  # FABP4hi Macrophage
  "FABP4", "INHBA",

  # Monocyte
  "FCN1", "IL1B", "CD14",

  # Dendritic
  "CD1C", "PLD4", "CLEC9A", "FCER1A",

  # Mast
  "TPSAB1", "CPA3", "MS4A2", "KIT",

  # Lymphoid
  "CD3E",

  # Cytotoxic T
  "CD3D", "CD8A", "CD8B", "GZMK",

  # Helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # T Regulatory
  "FOXP3",

  # Natural Killer
  "NKG7", "KLRF1", "FCER1G", "TYROBP",

  # B Lymphocyte
  "MS4A1", "CD79A", "CD19",

  # Plasma
  "JCHAIN", "MZB1",

  # Plasmacytoid Dendritic
  "CLEC4C", "CLIC3", "NRP1", "IL3RA",

  # Epithelial
  "EPCAM",

  # Alveolar Type 1
  "AGER", "PDPN", "CLIC5",

  # Alveolar Type 2
  "SFTPC", "SFTPD", "MUC1",

  # Basal
  "KRT5", "KRT15", "KRT17", "TP63",

  # Aberrant Basaloid
  "MMP7", "COL1A1", "FN1",

  # Ciliated
  "RSPH1", "CAPS", "FOXJ1",

  # Club
  "SCGB3A2",

  # Goblet
  "MUC5B", "SCGB1A1", "MUC5AC", "SPDEF",

  # Secretory
  # "SCGB1A1", "MUC5B", "SPDEF",

  # Endothelial
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Peribronchial Endothelial
  "COL15A1", "PLVAP", "MPZL2", "POSTN",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE", "SULT1E1",

  # Lymphatic Endothelial
  "LYVE1", "CCL21", "PROX1",

  # Fibroblast
  "LUM", "PDGFRA", "COL1A2",

  # Myofibroblast
  "ACTA2", "CTHRC1",

  # MFAP5hi Fibroblast
  "MFAP5", "PLA2G2A", "CXCL14",

  # SPINT2hi Fibroblast
  "SPINT2", "LMCD1", "AOC3", "FGFR4",

  # Smooth Muscle
  "DES", "MYH11",

  #Mural Cells
  "PDGFRB", "RGS5", "FAM162B",

  # Mesothelial
  "UPK3B", "WT1", "MSLN",

  # Proliferating Cells
  "TOP2A", "BIRC5", "UBE2C", "MKI67"
)

```

```{r}
l = FeaturePlot(IPF.L, features = unique(genes), combine = F, order = T, reduction = "umap.harmony")
names(l) = unique(genes)
```





```{r}



###with the markers
p=DotPlot(
  IPF.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "harmony_clusters") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
  
ggsave(filename = "output/IPF/Low/IPFL_cluster.pdf", plot = p, width = 20, height = 8)


```



```{r}
#---cluster_specifique_annotation---

##1celltype
Idents(IPF.L)="harmony_clusters"
cluster_name_x=c(
  "0"  = "Macrophages_0"                                     ,
  "1"  = "Venous Endothelial_1"                   ,
  "2"  = "Fibroblasts_2"                                     ,
  "3"  = "CD4_3"                                     ,
  "4"  = "Goblet Cells_4"                                     ,
  "5"  = "CD8_5"                                     ,
  "6"  = "Mast Cells_6"                          ,
  "7"  = "Alveolar Cells_7"                           ,
  "8"  = "Monocytes_8"                                     ,
  "9"  = "Ciliated Cells_9"                                     ,
  "10" = "Basal Cells_10"                                     ,          
  "11" = "Macrophages_11"                                   ,
  "12" = "Mural Cells_12"                                     ,
  "13" = "Venous Endothelial_13"                                     ,
  "14" = "Macrophages_14"                   ,
  "15" = "Fibroblasts_15"                                     ,
  "16" = "B Cells_16"                                     ,
  "17" = "Proliferating Cells_17"                                     ,    
  "18" = "Lymphatic Endothelial_18"  
)

IPF.L=RenameIdents(IPF.L, cluster_name_x)
IPF.L$CellType = Idents(IPF.L)

#plot
p=DotPlot(
  IPF.L,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "CellType") + RotatedAxis()  
    p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/IPF/Low/IPFL_cluster_name.pdf", plot = p, width = 20, height = 10)


```




```{r, fig.width=14, fig.height=8}

##------umap--------


##1celltype
Idents(IPF.L)="harmony_clusters"
cluster_name=c(
  "0"  = "Macrophages"                                     ,
  "1"  = "Venous Endothelial"                   ,
  "2"  = "Fibroblasts"                                     ,
  "3"  = "CD4"                                     ,
  "4"  = "Goblet Cells"                                     ,
  "5"  = "CD8"                                     ,
  "6"  = "Mast Cells"                          ,
  "7"  = "Alveolar Cells"                           ,
  "8"  = "Monocytes"                                     ,
  "9"  = "Ciliated Cells"                                     ,
  "10" = "Basal Cells"                                     ,          
  "11" = "Macrophages"                                   ,
  "12" = "Mural Cells"                                     ,
  "13" = "Venous Endothelial"                                     ,
  "14" = "Macrophages"                   ,
  "15" = "Fibroblasts"                                     ,
  "16" = "B Cells"                                     ,
  "17" = "Proliferating Cells"                                     ,    
  "18" = "Lymphatic Endothelial"  
)

IPF.L=RenameIdents(IPF.L, cluster_name)
IPF.L$Major_Celltype=Idents(IPF.L)
##2color
celltype_cols = c(
  "Venous Endothelial"    = "chocolate1",  # strong blue
  "Lymphatic Endothelial" = "mediumorchid1",  # green
  
  "CD4"                   = "olivedrab2",  # red
  "B Cells"               = "gold",  # purple
  "Monocytes"             = "#2ca02c",  # orange
  "Macrophages"           = "orangered",  # brown
  "Mast Cells"            = "#FDBF6F",  # yellow
  
  "Alveolar Cells"        = "aquamarine",  # light blue
  "Ciliated Cells"        = "saddlebrown",  # black
  "Goblet Cells"          = "cornflowerblue",  # pink
  "Basal Cells"           = "indianred3",  # lavender
  "Proliferating Cells"   = "#1F78B4",  # pale yellow
  
  "Mural Cells"           = "#1F78B4",  # dark red
  "Fibroblasts"           = "cyan",   # gray
  "CD8"                   ="tomato"
)



##3 center
umap_df <- as.data.frame(Embeddings(IPF.L, "umap.harmony"))
umap_df$Major_Celltype <- IPF.L$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)
##plot
fig1a = ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 6,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")

fig1a
ggsave(ggsave(filename = "output/IPF/Low/IPFL_cellType.pdf", plot = fig1a, width = 12, height = 10))


```







```{r}
#----same methods
# gene with major annotation
Idents(IPF.L) <- "Major_Celltype"


# -------------------------------
DefaultAssay(IPF.L) <- "RNA"
expr_mat <- GetAssayData(IPF.L, layer = "data")

# -------------------------------
# Percent expression matrix

percent_mat <- sapply(levels(Idents(IPF.L)), function(ct) {
  cells <- WhichCells(IPF.L, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# -------------------------------
# Average expression matrix

avg_exp <- AverageExpression(
  IPF.L,
  features = genes,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

# -------------------------------
# Clean data

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)

exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

# -------------------------------
# Scaling

exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# -------------------------------
# Color function

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# -------------------------------
# Dot rendering

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# -------------------------------
# Cell type colors (annotation)

celltypes <- colnames(exp_mat_scaled)

cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))

names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# -------------------------------
# plot

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,

  row_names_gp = grid::gpar(fontsize = 7, fontface = "bold"),

  column_names_gp = grid::gpar(
    fontsize = 10,
    fontface = "bold"
  ),

  column_names_rot = 90,

  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

# -------------------------------
# save

pdf("output/IPF/Low/IPF.L_markers.pdf",
    width = 10, height = 15)

ComplexHeatmap::draw(ht)

dev.off()
```







###7correlation
```{r}

##correlation for celltype

# 1️⃣ Set identities to cell types
Idents(IPF.L) <- "Major_Celltype"

# sanity check
levels(Idents(IPF.L))
# should show cell type names

# 2️⃣ Average expression per cell type
av.expIPL <- AverageExpression(
  IPF.L,
  assays = "RNA",
  slot = "data"
)$RNA

# 3️⃣ Cell-type correlation matrix
correlation_matrixIPL <- cor(
  as.matrix(av.expIPL),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixIPL[is.na(correlation_matrixIPL)] <- 0

# 4️⃣ Save heatmap as PNG
pdf(
  file = "output/IPF/Low/IPFL_correlation.pdf",
  width = 8,
  height = 10
)

pheatmap(
  correlation_matrixIPL,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()

```


###7 Class Major_Celltype with family


```{r}
#------
#extract
unique(IPF.L$Major_Celltype)

#
IPF.L$General <- recode(IPF.L$Major_Celltype,
                      "Macrophages" = "Mono/Mac",
                      "Fibroblasts" = "Fibroblasts",
                      "Goblet Cells" = "Epithelial Cells",
                      "CD4" = "CD4",
                      "Ciliated Cells" = "Epithelial Cells",
                      "Alveolar Cells" = "Epithelial Cells",
                      "CD8" = "CD8",
                      "Proliferating Cells" = "Proliferating Cells",
                      "Venous Endothelial" = "Endothelial Cells",
                      "Monocytes" = "Mono/Mac",
                      "Mast Cells" = "Mast Cells",
                      "Mural Cells" = "Mural Cells",
                      "B Cells" = "B Cells",
                      "Lymphatic Endothelial" = "Endothelial Cells",
                      "Basal Cells" = "Epithelial Cells"
                      
)

```



##markers that we can have in TLS 


```{r}
### TLS markers that we can have in TLS 


# Define markers per group from genes_TLS.csv
germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC", 
                     "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")

inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17", 
                  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5", 
                  "ICAM1", "VCAM1", "SELE")

lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12", 
                         "CXCR5", "CCR7", "CXCR4")

hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2", 
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")

fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP", 
         "CXCL13", "CD54", "ICAM1", "FCER2")

# Combine all markers in order
markers <- unique(c(germinal_center, inflammation, 
                    lymphoid_chemokines, hev, fdc))

# Filter to genes present in your Seurat object
markers <- markers[markers %in% rownames(IPF.L)]

# Build group vector in same order
marker_groups <- case_when(
  markers %in% germinal_center    ~ "Germinal Center",
  markers %in% inflammation       ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev                ~ "HEV",
  markers %in% fdc                ~ "FDC"
)
names(marker_groups) <- markers

# Group colors
group_cols <- c(
  "Germinal Center"         = "#2166AC",
  "Inflammation & Cytokines"= "#D6604D",
  "Lymphoid Chemokines"     = "#35978F",
  "HEV"                     = "#8073AC",
  "FDC"                     = "#E08214"
)

# Extract expression and percent data
Idents(IPF.L) <- "General"
DefaultAssay(IPF.L) <- "RNA"
expr_mat <- GetAssayData(IPF.L, layer = "data")

percent_mat <- sapply(levels(Idents(IPF.L)), function(ct) {
  cells <- WhichCells(IPF.L, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(
  IPF.L,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

# Color function
q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# Bubble cell function
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

# Column annotation
celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

# Row annotation
row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

# Heatmap
ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/IPF/Low/TLS_markers.pdf",
    width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()




```




```{r}
#####for pyscenic 
expr <- GetAssayData(IPF.L, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Low/input/matrix.mtx")

write.table(rownames(expr),
            file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Low/input/genes.tsv",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

write.table(colnames(expr),
            file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/IPF/Low/input/barcodes.tsv",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)


```







```{r}
#CSV 
write.csv(IPF.L@meta.data, 
          file = "data/process/SeuratObjects/IPF/Lower/IPF.L_metadata.csv",
          row.names = TRUE)
##save
saveRDS(IPF.L, file = "data/process/SeuratObjects/IPF/Lower/IPF.L_Final.rds")

## run it later to load all 
IPF.L = readRDS("data/process/SeuratObjects/IPF/Lower/IPF.L_Final.rds")


```























