---
title: "Rheumatoid Arthritis - Cell-Cell Communication Analysis"
output: html_notebook
---

<!-- Step 3: cell-cell communication (CCC) analysis for Rheumatoid Arthritis,
     using LIANA for interaction scoring, followed by pathway enrichment and
     export for gene-regulatory-network (SCENIC) analysis. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

Tensor-cell2cell runs through the `cell2cell` Python package, so the matching
conda environment is set here before it is imported.

```{r}
use_gpu <- TRUE
library(reticulate, quietly = TRUE)
reticulate::use_condaenv(
  "lianaenv",
  conda = "/mnt/nfs/home/aminata.sow/miniforge3/bin/conda",
  required = TRUE
)
```

## 2. Load data and convert to SingleCellExperiment

```{r}
suppressMessages({
  library(dplyr, quietly = TRUE)
  library(tidyr, quietly = TRUE)
  library(purrr, quietly = TRUE)
  library(forcats, quietly = TRUE)
  library(textshape, quietly = TRUE)
  library(ggpubr, quietly = TRUE)
  library(rstatix, quietly = TRUE)
  library(reshape2, quietly = TRUE)
  library(tibble, quietly = TRUE)
  library(scater, quietly = TRUE)
  library(scuttle, quietly = TRUE)
  library(ggplot2, quietly = TRUE)
  library(liana, quietly = TRUE)
  library(decoupleR, quietly = TRUE)
  library(ExperimentHub, quietly = TRUE)
  c2c <- reticulate::import(module = "cell2cell")
})

# Load the processed Seurat object and convert it to SingleCellExperiment,
# which is the format LIANA expects.

RA <- readRDS("data/process/SeuratObjects/RA/RA_process.rds")
sce <- as.SingleCellExperiment(RA)
saveRDS(sce, file = "data/process/SeuratObjects/RA/sce_raw.rds")
```

## 3. Quality control

```{r}
# inspect detected genes per cell before filtering
hist(
  colSums(counts(sce) > 0),
  breaks = 100,
  main = "Detected genes per cell",
  xlab = "Number of genes"
)

# basic filters: minimum genes per cell, minimum cells per gene
sce <- scater::addPerCellQC(sce)
min.features <- 500
min.cells <- 5

qc <- scater::perCellQCMetrics(sce)
sce <- sce[rowSums(counts(sce) >= 1) >= min.cells, ]
sce <- sce[, colSums(counts(sce) >= 1) >= min.features]

# mitochondrial content filter
is.mito <- grep("MT-", rownames(sce))
per.cell <- scuttle::perCellQCMetrics(sce, subsets = list(Mito = is.mito))
colData(sce)[["subsets_Mito_percent"]] <- per.cell$subsets_Mito_percent

class(colData(sce)$subsets_Mito_percent)
hist(
  colData(sce)$subsets_Mito_percent,
  breaks = 100,
  main = "Mitochondrial percentage per cell",
  xlab = "Percent mitochondrial reads"
)
sce <- sce[, colData(sce)$subsets_Mito_percent < 15]
rm(RA) # free memory - counts are still available via `sce`
```

## 4. Normalization

```{r}
scale.factor <- 1e4
lib.sizes <- colSums(counts(sce)) / scale.factor
assay(sce, "logcounts") <- scuttle::normalizeCounts(
  sce,
  size.factors = lib.sizes, log = FALSE, center.size.factors = FALSE
)
assay(sce, "logcounts") <- log1p(assay(sce, "logcounts"))
head(colData(sce))
```

## 5. Cell-cell communication (LIANA)

```{r}
liana_res <- liana_bysample(
  sce = sce,
  sample_col = "sample", # context dimension column from SCE metadata
  idents_col = "CellType", # cell types column from SCE metadata
  resource = "Consensus",
  expr_prop = 0.1, # must be expressed in expr_prop fraction of cells
  min_cells = 5,
  assay.type = "logcounts", # run on log- and library-normalized counts
  aggregate_how = "both", # aggregate magnitude and specificity
  verbose = TRUE,
  inplace = FALSE, # save it out of the sce obj
  permutation.params = list(nperms = 100),
  return_all = TRUE
)

# convert the list result to a single data frame
names(liana_res) <- unique(colData(sce)$sample) # all sample names
library(dplyr)
liana_res <- bind_rows(liana_res, .id = "sample")
liana_res$sample <- as.character(liana_res$sample)
table(liana_res$sample)
write.csv(liana_res, file = "CCC/RA/output/liana_res.csv", row.names = FALSE)

# keep only significant, high-confidence interactions
liana_filt <- liana_res %>%
  filter(
    magnitude_rank <= 0.05,
    cellphonedb.pvalue <= 0.01
  )
table(liana_filt$sample)
write.csv(liana_filt, file = "CCC/RA/output/liana_filt.csv", row.names = FALSE)

liana_filt <- read.csv("CCC/RA/output/liana_filt.csv")
liana_res <- read.csv("CCC/RA/output/liana_res.csv")
```

## 6. Top senders and receivers

```{r}
# count interactions where each cell type acts as sender / receiver
sender_df <- liana_filt %>%
  dplyr::count(source, name = "source_count") %>%
  dplyr::rename(celltype_clust = source)

receiver_df <- liana_filt %>%
  dplyr::count(target, name = "target_count") %>%
  dplyr::rename(celltype_clust = target)

activity_df <- full_join(sender_df, receiver_df, by = "celltype_clust") %>%
  replace_na(list(source_count = 0, target_count = 0))
```

```{r}
# rank cell types by total interaction count
activity_df <- activity_df %>%
  mutate(
    total_count = source_count + target_count,
    sender_rank = rank(-source_count, ties.method = "min"),
    receiver_rank = rank(-target_count, ties.method = "min")
  ) %>%
  arrange(desc(total_count))

write.csv(activity_df, file = "CCC/RA/output/Top_sender_receivers.csv", row.names = FALSE)
```

```{r}
# order cell types for the plot, most active first
activity_df <- activity_df %>%
  arrange(desc(total_count))

activity_df$celltype_clust <- factor(
  activity_df$celltype_clust,
  levels = rev(unique(activity_df$celltype_clust))
)

# sender/receiver counts side by side
plot_df <- activity_df %>%
  select(celltype_clust, source_count, target_count, total_count) %>%
  pivot_longer(
    cols = c(source_count, target_count),
    names_to = "role",
    values_to = "count"
  ) %>%
  mutate(
    role = recode(role, source_count = "source", target_count = "target"),
    role = factor(role, levels = c("source", "target"))
  )

p <- ggplot(
  plot_df,
  aes(x = count, y = celltype_clust, fill = role)
) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE), width = 0.65) +
  scale_fill_manual(
    name = NULL,
    values = c("source" = "#2166AC", "target" = "#D6604D"),
    labels = c("Sender", "Receiver")
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = "Number of interactions",
    y = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = c(0.85, 0.15),
    legend.background = element_rect(fill = "white", color = "grey80"),
    legend.text = element_text(size = 12),
    panel.border = element_rect(fill = NA, linewidth = 0.8),
    axis.line = element_blank(),
    axis.text.y = element_text(size = 11, color = "black"),
    axis.text.x = element_text(size = 11, color = "black"),
    axis.title.x = element_text(size = 13, margin = margin(t = 8)),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    plot.margin = margin(10, 20, 10, 10)
  )

ggsave(filename = "CCC/RA/output/TopSR.pdf", plot = p, width = 10, height = 8, dpi = 300)
```

## 7. Unique and shared ligand-receptor pairs

```{r}
# LR pairs seen exactly once across all interactions
unique_lr <- liana_filt %>%
  dplyr::mutate(LR_pair = paste(ligand.complex, receptor.complex, sep = "_")) %>%
  dplyr::group_by(LR_pair) %>%
  dplyr::filter(dplyr::n() == 1) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    source, target, ligand.complex, receptor.complex, LR_pair,
    magnitude_rank, cellphonedb.pvalue, specificity_rank, magnitude_mean_rank
  ) %>%
  dplyr::arrange(magnitude_rank)
write.csv(unique_lr, file = "CCC/RA/output/Unique_lr.csv", row.names = TRUE)

# LR pairs seen more than once (shared across cell type pairs)
shared_lr <- liana_filt %>%
  dplyr::mutate(LR_pair = paste(ligand.complex, receptor.complex, sep = "_")) %>%
  dplyr::group_by(LR_pair) %>%
  dplyr::mutate(n_repetitions = dplyr::n()) %>%
  dplyr::filter(n_repetitions > 1) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    source, target, ligand.complex, receptor.complex, LR_pair,
    n_repetitions, magnitude_rank, cellphonedb.pvalue,
    specificity_rank, magnitude_mean_rank
  )
write.csv(shared_lr, file = "CCC/RA/output/shared_lr.csv", row.names = TRUE)
```

```{r}
# plot: shared LR pairs for one chosen sender cell type
library(dplyr)
library(ggplot2)

# LR_pair already exists in shared_lr from the chunk above - safe to redo, just namespaced
shared_lr <- shared_lr %>%
  dplyr::mutate(LR_pair = paste(ligand.complex, receptor.complex, sep = "_"))

# filter by source cell
source_cell <- "Fibroblasts_1"
shared_source <- shared_lr %>%
  dplyr::filter(source == source_cell)

# keep only pairs shared across at least 3 target cell types
shared_source_clean <- shared_source %>%
  dplyr::group_by(LR_pair) %>%
  dplyr::filter(dplyr::n_distinct(target) >= 3) %>%
  dplyr::ungroup()

# top 15 LR pairs by mean rank
top_lr_pairs <- shared_source_clean %>%
  dplyr::group_by(LR_pair) %>%
  dplyr::summarise(mean_rank = mean(magnitude_mean_rank, na.rm = TRUE)) %>%
  dplyr::arrange(mean_rank) %>%
  dplyr::slice_head(n = 15) %>%
  dplyr::pull(LR_pair)

shared_source_top <- shared_source_clean %>%
  dplyr::filter(LR_pair %in% top_lr_pairs)

p <- ggplot(
  shared_source_top,
  aes(x = LR_pair, y = target)
) +
  geom_point(aes(
    size = -magnitude_mean_rank,
    color = -magnitude_mean_rank
  )) +
  scale_color_viridis_c(
    name = "Interaction\nstrength",
    option = "plasma",
    guide = guide_colorbar(
      barwidth = 1,
      barheight = 8,
      ticks = TRUE,
      ticks.colour = "white",
      frame.colour = "grey50"
    )
  ) +
  scale_size_continuous(
    name = "Interaction\nstrength",
    range = c(2, 8),
    guide = "none"
  ) +
  theme_bw(base_size = 13) +
  labs(
    title = paste(source_cell, "as Sender"),
    y = "Receivers"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 60, hjust = 1,
      size = 16, color = "black"
    ),
    axis.text.y = element_text(size = 20, color = "black"),
    axis.title.x = element_text(size = 20, margin = margin(t = 8)),
    axis.title.y = element_text(size = 20, margin = margin(r = 8)),
    plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 13),
    plot.margin = margin(10, 15, 10, 10)
  )

ggsave(filename = "CCC/RA/output/shared_lr.pdf", plot = p, width = 20, height = 20, dpi = 300)
```

```{r}
# plot: unique LR pairs for the same sender cell type
library(dplyr)
library(ggplot2)

source_cell <- "Fibroblasts_1"
unique_source <- unique_lr %>%
  filter(source == source_cell)

# top 50 LR pairs by mean rank
top_unique <- unique_source %>%
  arrange(magnitude_mean_rank) %>%
  slice_head(n = 50)

top_unique <- top_unique %>%
  mutate(
    LR_pair = paste(ligand.complex, receptor.complex, sep = "_"),
    LR_pair = factor(LR_pair, levels = rev(unique(LR_pair)))
  )

p <- ggplot(
  top_unique,
  aes(x = LR_pair, y = target)
) +
  geom_point(aes(
    size = -magnitude_mean_rank,
    color = -magnitude_mean_rank
  )) +
  scale_color_viridis_c(
    name = "Interaction\nstrength",
    option = "plasma",
    guide = guide_colorbar(
      barwidth = 1,
      barheight = 8,
      ticks = TRUE,
      ticks.colour = "white",
      frame.colour = "grey50"
    )
  ) +
  scale_size_continuous(
    name = "Interaction\nstrength",
    range = c(1, 6),
    guide = "none"
  ) +
  theme_bw(base_size = 13) +
  labs(
    title = paste(source_cell, "- Unique LR Pairs"),
    x = "Ligand-Receptor Pairs",
    y = "Receivers"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45, hjust = 1,
      size = 9, color = "black"
    ),
    axis.text.y = element_text(size = 15, color = "black"),
    axis.title.x = element_text(size = 15, margin = margin(t = 8)),
    axis.title.y = element_text(size = 12, margin = margin(r = 8)),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 11),
    plot.margin = margin(10, 15, 10, 10)
  )

p

ggsave(
  filename = "CCC/RA/output/unique_lr.pdf", plot = p,
  width = 22, height = 10, dpi = 300
)
```

## 8. Pathway enrichment

```{r}
# pathway enrichment on ligand/receptor genes, using gprofiler2
library(gprofiler2)
library(dplyr)

# query set: ligand/receptor genes from the significant interactions
ligands_strong <- liana_filt$ligand.complex
ligands_strong <- ligands_strong[!is.na(ligands_strong)]
receptors_strong <- liana_filt$receptor.complex
receptors_strong <- receptors_strong[!is.na(receptors_strong)]
query_genes <- unique(c(ligands_strong, receptors_strong))
length(query_genes)

# background set: ligand/receptor genes from all tested interactions
ligands_all <- liana_res$ligand.complex
ligands_all <- ligands_all[!is.na(ligands_all)]
receptors_all <- liana_res$receptor.complex
receptors_all <- receptors_all[!is.na(receptors_all)]
background_genes <- unique(c(ligands_all, receptors_all))
length(background_genes)

# over-representation analysis (ORA) using g:Profiler
gprof_enr <- gost(
  query = query_genes,
  ordered_query = FALSE,
  custom_bg = background_genes,
  organism = "hsapiens"
)

# clean and save results
gprof_results_clean <- gprof_enr$result
nrow(gprof_results_clean)
gprof_results_clean$parents <- NULL
gprof_results_clean <- gprof_results_clean[order(gprof_results_clean$p_value), ]

write.csv(gprof_results_clean, "CCC/RA/output/Pathways.csv", row.names = FALSE)

# keep GO:BP and Reactome terms.
gprof_GO <- gprof_results_clean %>%
  filter(source %in% c("GO:BP", "REAC")) %>%
  filter(term_size < 500) %>%
  filter(intersection_size >= 3) %>%
  arrange(p_value)

write.csv(gprof_GO, "CCC/RA/output/PathwaysGO.csv", row.names = FALSE)
```

```{r}
# plot top enriched pathways per database
library(ggplot2)
library(dplyr)

top_pathways <- gprof_GO %>%
  group_by(source) %>%
  slice_head(n = 20) %>%
  ungroup() %>%
  mutate(
    term_name = factor(term_name, levels = rev(unique(term_name))),
    log10p = -log10(p_value)
  )

p <- ggplot(
  top_pathways,
  aes(x = log10p, y = term_name, fill = source)
) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_vline(
    xintercept = -log10(0.05),
    linetype = "dashed",
    color = "grey40",
    linewidth = 0.5
  ) +
  scale_fill_manual(
    name = "Database",
    values = c("GO:BP" = "#2166AC", "REAC" = "#D6604D")
  ) +
  facet_wrap(~source, scales = "free_y", ncol = 1) +
  theme_bw(base_size = 13) +
  labs(
    title = "Pathway Enrichment",
    x = "-log10(p-value)",
    y = NULL
  ) +
  theme(
    axis.text.y = element_text(size = 15, color = "black"),
    axis.text.x = element_text(size = 15, color = "black"),
    axis.title.x = element_text(size = 12, margin = margin(t = 8)),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    strip.background = element_rect(fill = "grey95"),
    strip.text = element_text(size = 12, face = "bold"),
    plot.margin = margin(10, 15, 10, 10)
  )

ggsave(filename = "CCC/RA/output/Pathways.png", plot = p, width = 10, height = 14, dpi = 300)
ggsave(filename = "CCC/RA/output/Pathways.pdf", plot = p, width = 10, height = 14, dpi = 300)
```

## 9. Gene regulatory network (SCENIC) export

```{r}
# See SCENIC_pipeline_wd (.py and .sh files) for the full GRN pipeline: run
# pySCENIC in the terminal on the exported data, then load its results below.
library(Matrix)
library(SingleCellExperiment)
```

## 10. GRN plots

```{r}
# plot regulon specificity scores (RSS) from SCENIC output
library(dplyr)
library(ggplot2)
library(tidyr)

rss <- read.csv("GRN_SCENIC/RA/output/rss_cellType.csv", row.names = 1)

rss_long <- rss %>%
  as.data.frame() %>%
  tibble::rownames_to_column("regulon") %>%
  pivot_longer(-regulon,
    names_to = "cell_type",
    values_to = "RSS"
  )

# top 5 regulons per cell type
top_regulons <- rss_long %>%
  group_by(cell_type) %>%
  arrange(desc(RSS)) %>%
  slice_head(n = 5) %>%
  ungroup()

p <- ggplot(
  top_regulons,
  aes(
    x = RSS,
    y = reorder(regulon, RSS),
    fill = cell_type
  )
) +
  geom_bar(stat = "identity", width = 0.7) +
  facet_wrap(~cell_type, scales = "free_y", ncol = 2) +
  scale_fill_viridis_d(option = "plasma", guide = "none") +
  theme_bw(base_size = 13) +
  labs(
    title = "Top Regulons per Cell Type (RSS)",
    x = "Regulon Specificity Score",
    y = "Regulon (TF)"
  ) +
  theme(
    axis.text.x = element_text(size = 11, color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    strip.background = element_rect(fill = "grey95"),
    strip.text = element_text(size = 12, face = "bold"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 15, 10, 10)
  )

p

ggsave(
  filename = "GRN_SCENIC/RSS_top_regulons.png", plot = p,
  width = 20, height = 20, dpi = 300
)
```

## 11. Save the processed object

```{r}
saveRDS(sce, file = "data/process/SeuratObjects/RA/sce_processed.rds")
sce <- readRDS("data/process/SeuratObjects/RA/sce_processed.rds")
```
