---
title: "Rheumatoid Arthritis - Processing and Annotation"
output: html_notebook
---



## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

## 2. Load raw data and create per-sample Seurat objects

```{r}
# load barcodes, genes and the count matrix
barcodes <- read.table("data/process/RA/barcodes_filtered.txt",
  header = FALSE, stringsAsFactors = FALSE
)$V1
genes <- read.table("data/process/RA/genes_filtered.txt",
  header = FALSE, stringsAsFactors = FALSE
)$V1

cat("Barcodes loaded:", length(barcodes), "\n")
cat("Genes loaded:", length(genes), "\n")

cat("Loading matrix...\n")
count_matrix <- readMM("data/process/RA/count_matrix_filtered.mtx")
cat("Matrix loaded!\n")

rownames(count_matrix) <- genes
colnames(count_matrix) <- barcodes
```

```{r}
# gene names must be unique before building the Seurat object
cat("Duplicated genes:", sum(duplicated(genes)), "\n")
genes_unique <- make.unique(genes)
rownames(count_matrix) <- genes_unique
colnames(count_matrix) <- barcodes

cat("Creating Seurat object...\n")
seurat_full <- CreateSeuratObject(counts = count_matrix, project = "E-MTAB-11791")
cat("Seurat object created!\n")
print(seurat_full)
```

```{r}
# this dataset combines several patient samples in one matrix - split them
# out by barcode prefix and save each sample separately
out_dir <- "../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Rheumatoid Arthritis/"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

samples <- list(
  RA26 = "Syn_Bio_026",
  RA28 = "Syn_Bio_028",
  RA49 = "Syn_Bio_049",
  RA50 = "Syn_Bio_050"
)

for (name in names(samples)) {
  cat("Processing", name, "...\n")
  cells <- grep(samples[[name]], barcodes, value = TRUE)
  seurat_obj <- subset(seurat_full, cells = cells)
  saveRDS(seurat_obj, file = paste0(out_dir, name, ".rds"))
  cat(name, ":", ncol(seurat_obj), "cells - saved as", paste0(name, ".rds\n"))
}

cat("All done!\n")
```

## 3. Reload per-sample objects, label, and save

```{r}
RA26 <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Rheumatoid Arthritis/RA26.rds")
RA28 <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Rheumatoid Arthritis/RA28.rds")
RA49 <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Rheumatoid Arthritis/RA49.rds")
RA50 <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Rheumatoid Arthritis/RA50.rds")
```

```{r}
# label each object with its sample of origin, and keep a copy of each
RA26$sample <- "RA26"
RA28$sample <- "RA28"
RA49$sample <- "RA49"
RA50$sample <- "RA50"

saveRDS(RA26, file = "data/process/SeuratObjects/RA/RA26.rds")
saveRDS(RA28, file = "data/process/SeuratObjects/RA/RA28.rds")
saveRDS(RA49, file = "data/process/SeuratObjects/RA/RA49.rds")
saveRDS(RA50, file = "data/process/SeuratObjects/RA/RA50.rds")
```

## 4. Quality control and merge

```{r}
Layers(RA26)
str(RA26@assays$RNA)

# mitochondrial percentage per sample
RA26[["percent.mt"]] <- PercentageFeatureSet(RA26, pattern = "^MT-")
RA28[["percent.mt"]] <- PercentageFeatureSet(RA28, pattern = "^MT-")
RA49[["percent.mt"]] <- PercentageFeatureSet(RA49, pattern = "^MT-")
RA50[["percent.mt"]] <- PercentageFeatureSet(RA50, pattern = "^MT-")

p1 <- VlnPlot(RA26, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE)
p2 <- VlnPlot(RA28, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE)
p3 <- VlnPlot(RA49, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE)
p4 <- VlnPlot(RA50, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC thresholds set per sample, based on each sample's own distributions
RA26 <- subset(RA26, subset = nFeature_RNA > 500 & nFeature_RNA < 8000 & percent.mt < 20)
RA28 <- subset(RA28, subset = nFeature_RNA > 1000 & nFeature_RNA < 8000 & percent.mt < 20)
RA49 <- subset(RA49, subset = nFeature_RNA > 500 & nFeature_RNA < 7000 & percent.mt < 10)
RA50 <- subset(RA50, subset = nFeature_RNA > 500 & nFeature_RNA < 7000 & percent.mt < 15)

# merge samples, then apply one more filter on the combined object
RA <- merge(RA26,
  y = c(RA28, RA49, RA50),
  add.cell.ids = c("RA26", "RA28", "RA49", "RA50")
)
RA <- subset(RA, subset = nFeature_RNA > 500 & nFeature_RNA < 8000 & percent.mt < 15)

saveRDS(RA, file = "data/process/SeuratObjects/RA/RA_raw.rds")

rm(RA26, RA28, RA49, RA50)
gc(full = TRUE)

DefaultAssay(RA)
```

## 5. Normalization, scaling, and PCA

```{r}
RA <- NormalizeData(RA)
RA <- FindVariableFeatures(RA)
RA <- ScaleData(RA, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = RA))
RA <- RunPCA(RA, features = VariableFeatures(object = RA))
ElbowPlot(RA, ndims = 50, reduction = "pca")
```

## 6. Harmony integration and clustering resolutions

```{r}
library(harmony)
RA <- RunHarmony(RA, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:30)
RA[["RNA"]] <- JoinLayers(RA[["RNA"]])
RA <- RunUMAP(RA, reduction = "harmony", dims = 1:30, reduction.name = "UHarmony")

# scan a range of resolutions before picking one
RA <- FindNeighbors(RA, reduction = "harmony", dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  RA <- FindClusters(RA, resolution = res)
  colname <- paste0("res_", gsub("\\.", "_", res))
  RA[[colname]] <- RA$seurat_clusters
}

res_cols <- grep("^res_", colnames(RA@meta.data), value = TRUE)

# save all resolutions to a single PDF for visual comparison
pdf("output/RA/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(RA, group.by = res, label = TRUE, reduction = "UHarmony") +
    ggtitle(paste(res))
  print(p)
}
dev.off()

# resolution 0.8 was chosen for downstream clustering
p <- DimPlot(RA, reduction = "UHarmony", group.by = "res_0_8", label = TRUE, label.size = 8, pt.size = 2) +
  ggtitle("Resolution_0.8") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/RA/RA_resolution.pdf", plot = p, width = 12, height = 8, dpi = 300)
```

## 7. Clustering and marker genes

```{r}
RA <- FindNeighbors(RA, reduction = "harmony", dims = 1:30)
RA <- FindClusters(RA, resolution = 0.8, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
RA_markers <- FindAllMarkers(RA, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
RA_markers <- RA_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 8. Cell type annotation

```{r, fig.width=10, fig.height=8}
genes <- c(
  # T cell
  "CD3E", "TRAC", "CD3D",

  # Cytotoxic
  "CD8A", "CD8B", "GZMK",

  # helper T
  "CD4", "COTL1", "CCR7", "LEF1",

  # B cells
  "CD79A", "CD79B", "MS4A1",

  # Macrophages
  "CD14", "CD68", "LYZ",

  # Plasmacytoid Dc
  "LILRA4", "GZMB", "ITM2C",

  # Mast cells
  "CD1C", "FCER1A", "CLEC9A", "CPA3",

  # Neutrophils
  "IFITM2",

  # Endothelial cells
  "VWF", "PECAM1", "CLDN5",

  # Arterial Endothelial
  "DKK2", "IGFBP3", "GJA5", "BMX",

  # Capillary Endothelial
  "HPGD", "IL7R", "CA4", "FCN3", "SEMA3G",

  # Venous Endothelial
  "HDAC9", "ACKR1", "SELE",

  # Mural
  "ACTA2", "TAGLN", "DCN",

  # Fibroblasts
  "COL1A1", "LUM", "PDGFRA"
)

p <- DotPlot(
  RA,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "harmony_clusters"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/RA/RA_cluster.pdf", plot = p, width = 12, height = 8)
```

```{r}
# cell type labels
Idents(RA) <- "harmony_clusters"
cluster_name_x <- c(
  "0" = "Fibroblasts_0",
  "1" = "Fibroblasts_1",
  "2" = "Macrophages_2",
  "3" = "CD4_3",
  "4" = "Fibroblasts_4",
  "5" = "Fibroblasts_5",
  "6" = "Fibroblasts_6",
  "7" = "Fibroblasts_7",
  "8" = "Venous Endothelial_8",
  "9" = "Macrophages_9",
  "10" = "CD8_10",
  "11" = "Fibroblasts_11",
  "12" = "Fibroblasts_12",
  "13" = "Fibroblasts_13",
  "14" = "B Cells_14",
  "15" = "Mast Cells_15",
  "16" = "Arterial Endothelial_16",
  "17" = "Mural Cells_17",
  "18" = "Arterial Endothelial_18"
)
RA <- RenameIdents(RA, cluster_name_x)
RA$CellType <- Idents(RA)

p <- DotPlot(
  RA,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "CellType"
) + RotatedAxis()
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/RA/RA_cluster_name.pdf", plot = p, width = 12, height = 10)
```

```{r}
# major cell types
Idents(RA) <- "harmony_clusters"
cluster_name <- c(
  "0" = "Fibroblasts",
  "1" = "Fibroblasts",
  "2" = "Macrophages",
  "3" = "CD4",
  "4" = "Fibroblasts",
  "5" = "Fibroblasts",
  "6" = "Fibroblasts",
  "7" = "Fibroblasts",
  "8" = "Venous Endothelial",
  "9" = "Macrophages",
  "10" = "CD8",
  "11" = "Fibroblasts",
  "12" = "Fibroblasts",
  "13" = "Fibroblasts",
  "14" = "B Cells",
  "15" = "Mast Cells",
  "16" = "Arterial Endothelial",
  "17" = "Mural Cells",
  "18" = "Arterial Endothelial"
)
RA <- RenameIdents(RA, cluster_name)
RA$Major_Celltype <- Idents(RA)

celltype_cols <- c(
  "Fibroblasts" = "cyan",
  "B Cells" = "gold",
  "Arterial Endothelial" = "cadetblue3",
  "Venous Endothelial" = "chocolate1",
  "Mast Cells" = "#FDBF6F",
  "Mural Cells" = "#1F78B4",
  "CD4" = "olivedrab2",
  "CD8" = "tomato",
  "Macrophages" = "orangered"
)

# UMAP with a label placed at each cell type's centroid
umap_df <- as.data.frame(Embeddings(RA, "UHarmony"))
colnames(umap_df) <- c("UHarmony_1", "UHarmony_2")
umap_df$Major_Celltype <- RA$Major_Celltype
centers <- aggregate(
  cbind(UHarmony_1, UHarmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

library(ggrepel)
suppressWarnings({
  fig1a <- ggplot(umap_df, aes(UHarmony_1, UHarmony_2, color = Major_Celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = UHarmony_1, y = UHarmony_2, label = Major_Celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 4.5,
      box.padding = 0.4,
      label.size = 0.40,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "UMAP.h", y = "UMAP.h")
})
fig1a
ggsave(filename = "output/RA/RA_cellType.pdf", plot = fig1a, width = 12, height = 10)
```

```{r, fig.width=10, fig.height=8}
p <- DotPlot(
  RA,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  group.by = "Major_Celltype"
) + RotatedAxis()

p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)

ggsave(filename = "output/RA/RA_markers.pdf", plot = p, width = 12, height = 8)
```

## 9. Correlation between cell types

```{r}
Idents(RA) <- "Major_Celltype"
levels(Idents(RA))

# use highly variable genes to compute average expression per cell type
RA <- FindVariableFeatures(RA, nfeatures = 10000)
hvg <- VariableFeatures(RA)[1:5000]

av_expRA <- AverageExpression(
  RA,
  features = hvg,
  assays = "RNA",
  slot = "data"
)$RNA

correlation_matrix <- cor(
  as.matrix(av_expRA),
  method = "spearman",
  use = "pairwise.complete.obs"
)
correlation_matrix[is.na(correlation_matrix)] <- 0

pdf(
  file = "output/RA/RA_correlation.pdf",
  width = 10,
  height = 8
)
library(pheatmap)
colors <- colorRampPalette(c("#2166AC", "white", "#B2182B"))(100)
pheatmap(
  correlation_matrix,
  clustering_method = "ward.D2",
  color = colors,
  fontsize_row = 9,
  fontsize_col = 9
)
dev.off()
```

## 10.cell-type grouping (General)

```{r}
unique(RA$Major_Celltype)

RA$General <- recode(RA$Major_Celltype,
  "Fibroblasts" = "Fibroblasts",
  "Arterial Endothelial" = "Endothelial Cells",
  "Mural Cells" = "Mural Cells",
  "CD8" = "CD8",
  "Macrophages" = "Mono/Mac",
  "Mast Cells" = "Mast Cells",
  "CD4" = "CD4",
  "Venous Endothelial" = "Endothelial Cells",
  "B Cells" = "B Cells"
)
```

## 11. TLS (tertiary lymphoid structure) marker heatmap

```{r}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(
  germinal_center, inflammation,
  lymphoid_chemokines, hev, fdc
))
markers <- markers[markers %in% rownames(RA)]

marker_groups <- case_when(
  markers %in% germinal_center ~ "Germinal Center",
  markers %in% inflammation ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev ~ "HEV",
  markers %in% fdc ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center" = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines" = "#35978F",
  "HEV" = "#8073AC",
  "FDC" = "#E08214"
)

# percent of cells expressing each marker, per cell type
Idents(RA) <- "General"
DefaultAssay(RA) <- "RNA"
expr_mat <- GetAssayData(RA, layer = "data")

percent_mat <- sapply(levels(Idents(RA)), function(ct) {
  cells <- WhichCells(RA, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  RA,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/RA/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 12. Export for pySCENIC

```{r}
expr <- GetAssayData(RA, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/RA/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/RA/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/RA/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

## 13. Save final metadata and processed object

```{r}
RA$orig.ident <- "RA"
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General"
)
RA@meta.data <- RA@meta.data[, keep_cols]

write.csv(RA@meta.data, file = "data/process/SeuratObjects/RA/RA.metadata.csv", row.names = TRUE)

saveRDS(RA, file = "data/process/SeuratObjects/RA/RA_process.rds")
RA <- readRDS("data/process/SeuratObjects/RA/RA_process.rds")
```
