---
title: "Fusion - Cell-Cell Communication Analysis"
output: html_notebook
---

<!-- Cell-cell communication (CCC) analysis across disease conditions,
     using LIANA for per-sample interaction scoring and Tensor-cell2cell
     for multi-condition factorization. -->

## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

Tensor-cell2cell runs through the `cell2cell` Python package, so the matching
conda environment is set here before it is imported.

```{r}
use_gpu <- TRUE
library(reticulate, quietly = TRUE)
reticulate::use_condaenv(
  "lianaenv",
  conda = "/mnt/nfs/home/aminata.sow/miniforge3/bin/conda",
  required = TRUE
)
```

## 2. Load data and convert to SingleCellExperiment

```{r}
suppressMessages({
  library(dplyr, quietly = TRUE)
  library(tidyr, quietly = TRUE)
  library(purrr, quietly = TRUE)
  library(forcats, quietly = TRUE)
  library(textshape, quietly = TRUE)
  library(ggpubr, quietly = TRUE)
  library(rstatix, quietly = TRUE)
  library(reshape2, quietly = TRUE)
  library(tibble, quietly = TRUE)
  library(scater, quietly = TRUE)
  library(scuttle, quietly = TRUE)
  library(ggplot2, quietly = TRUE)
  library(liana, quietly = TRUE)
  library(decoupleR, quietly = TRUE)
  library(ExperimentHub, quietly = TRUE)
  c2c <- reticulate::import(module = "cell2cell")
})

# Load the merged Seurat object produced by Fusion.rmd and convert it to
# SingleCellExperiment, which is the format LIANA expects.
Fu <- readRDS("data/process/SeuratObjects/Fusion/F_process.rds")

# Keep only the metadata columns needed downstream. After merging (see
# Fusion.rmd), `orig.ident` holds each cell's disease label, so it is
# renamed to `Condition` to match how it is used later in this script.
keep_cols <- c("orig.ident", "nCount_RNA", "nFeature_RNA", "percent.mt", "Major_Celltype", "General")
Fu@meta.data <- Fu@meta.data[, keep_cols]
colnames(Fu@meta.data)[colnames(Fu@meta.data) == "orig.ident"] <- "Condition"

Fu <- DietSeurat(Fu, assays = "RNA")
Fu[["RNA"]] <- JoinLayers(Fu[["RNA"]])
sce <- as.SingleCellExperiment(Fu)
Assays(Fu) # sanity check: RNA should be the only assay left

saveRDS(sce, file = "data/process/SeuratObjects/Fusion/sce_raw.rds")
sce <- readRDS("data/process/SeuratObjects/Fusion/sce_raw.rds")
```

## 3. Quality control

```{r}
# inspect detected genes per cell before filtering
colData(sce)
hist(
  colSums(counts(sce) > 0),
  breaks = 100,
  main = "Detected genes per cell",
  xlab = "Number of genes"
)

# basic filters: minimum genes per cell, minimum cells per gene
sce <- scater::addPerCellQC(sce)
min.features <- 500
min.cells <- 5

qc <- scater::perCellQCMetrics(sce)
sce <- sce[rowSums(counts(sce) >= 1) >= min.cells, ]
sce <- sce[, colSums(counts(sce) >= 1) >= min.features]

# mitochondrial content filter
is.mito <- grep("MT-", rownames(sce))
per.cell <- scuttle::perCellQCMetrics(sce, subsets = list(Mito = is.mito))
colData(sce)[["subsets_Mito_percent"]] <- per.cell$subsets_Mito_percent

class(colData(sce)$subsets_Mito_percent)
hist(
  colData(sce)$subsets_Mito_percent,
  breaks = 100,
  main = "Mitochondrial percentage per cell",
  xlab = "Percent mitochondrial reads"
)
sce <- sce[, colData(sce)$subsets_Mito_percent < 15]
```

## 4. Normalization

```{r}
scale.factor <- 1e4
lib.sizes <- colSums(counts(sce)) / scale.factor
assay(sce, "logcounts") <- scuttle::normalizeCounts(
  sce,
  size.factors = lib.sizes, log = FALSE, center.size.factors = FALSE
)
assay(sce, "logcounts") <- log1p(assay(sce, "logcounts"))
head(colData(sce))
```

## 5. Cell-cell communication (LIANA)

```{r}
liana_res <- liana_bysample(
  sce = sce,
  sample_col = "Condition",
  idents_col = "General",
  resource = "Consensus",
  expr_prop = 0.1,
  min_cells = 5,
  assay.type = "logcounts",
  aggregate_how = "both",
  verbose = TRUE,
  inplace = FALSE,
  permutation.params = list(nperms = 100),
  return_all = TRUE
)

# reshape the per-condition list into a single data frame
names(liana_res) <- unique(colData(sce)$Condition)
liana_res <- bind_rows(liana_res, .id = "Condition")
liana_res$Condition <- as.character(liana_res$Condition)

table(liana_res$Condition)
write.csv(liana_res, file = "CCC/Fusion/output/liana_res.csv", row.names = FALSE)

# keep only significant, high-confidence interactions
liana_filt <- liana_res %>%
  filter(
    magnitude_rank <= 0.05,
    cellphonedb.pvalue <= 0.01
  )
table(liana_filt$Condition)
write.csv(liana_filt, file = "CCC/Fusion/output/liana_filt.csv", row.names = FALSE)

liana_filt <- read.csv("CCC/Fusion/output/liana_filt.csv")
```

```{r}
library(dplyr)

# exclude the cell types excluded throughout the project (kept consistent
# across all diseases), then keep only interactions where NEITHER source
# NOR target is one of them
excluded_celltypes <- c(
  "Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
  "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
  "Plasma Cells"
)

liana_filt_twice <- liana_filt %>%
  filter(!source %in% excluded_celltypes & !target %in% excluded_celltypes)

cat("Interactions before:", nrow(liana_filt), "\n")
cat("Interactions after:", nrow(liana_filt_twice), "\n")

cat("Cell types remaining as source:\n")
print(sort(unique(liana_filt_twice$source)))
cat("Cell types remaining as target:\n")
print(sort(unique(liana_filt_twice$target)))

write.csv(liana_filt_twice, file = "CCC/Fusion/output/liana_filt_twice.csv", row.names = FALSE)
cat("Saved: CCC/Fusion/output/liana_filt_twice.csv\n")
```

## 6. Build the interaction tensor (Tensor-cell2cell)

Tensor-cell2cell infers context-dependent CCC patterns across multiple
samples/conditions simultaneously.

```{r}
# store filtered interactions by condition
sce@metadata[["liana_filt"]] <- split(liana_filt_twice, liana_filt_twice$Condition)
sorted_samples <- sort(names(sce@metadata[["liana_filt"]]))

context_df_dict <- liana:::preprocess_scores(
  context_df_dict = sce@metadata[["liana_filt"]],
  invert = TRUE,
  invert_fun = function(x) 1 - x,
  non_negative = TRUE,
  non_negative_fill = 0
)

tensor <- liana_tensor_c2c(
  context_df_dict = context_df_dict,
  sender_col = "source",
  receiver_col = "target",
  ligand_col = "ligand.complex",
  receptor_col = "receptor.complex",
  score_col = "magnitude_rank",
  how = "outer_cells",
  outer_fraction = 1 / 3,
  context_order = sorted_samples,
  conda_env = "/mnt/nfs/home/aminata.sow/miniforge3/envs/lianaenv",
  build_only = TRUE,
  device = use_gpu
)

tensor$shape
reticulate::py_save_object(tensor, "CCC/Fusion/output/tensor_bycondition_filt.pkl")

library(reticulate)
tensor <- py_load_object("CCC/Fusion/output/tensor_bycondition_filt.pkl")
```

## 7. Tensor factorization

```{r}
tensor <- liana::decompose_tensor(
  tensor = tensor,
  rank = 12,
  tf_optimization = "regular",
  seed = 0,
  factors_only = FALSE
)

print(paste0("The estimated tensor rank is: ", tensor$rank))
```

## 8. Factor overview and downstream plots

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(ggpubr)

sce@metadata$tensor_res <- liana:::format_c2c_factors(tensor$factors)
sce@metadata$tensor_res$contexts$Condition <- as.character(sce@metadata$tensor_res$contexts$context)

condition_col <- "Condition"
sample_col <- "Condition"

condition_colors <- c(
  "IPF.L"    = "#E63946",
  "IPF.U"    = "#F4A261",
  "MS"       = "#2A9D8F",
  "RA"       = "#457B9D",
  "SjD"      = "#8338EC",
  "SS"       = "#FB5607",
  "SScILD.L" = "#FFBE0B",
  "SScILD.U" = "#3A86FF",
  "T1D"      = "#06D6A0",
  "LN"       = "#6D6875"
)

# exclude the same cell types filtered earlier, using the SAME consistent
# color palette used throughout, before any plotting happens
excluded_celltypes <- c(
  "Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
  "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
  "Plasma Cells"
)

celltype_colors <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

sce@metadata$tensor_res$senders <- sce@metadata$tensor_res$senders %>%
  filter(!celltype %in% excluded_celltypes)
sce@metadata$tensor_res$receivers <- sce@metadata$tensor_res$receivers %>%
  filter(!celltype %in% excluded_celltypes)

cat("Cell types kept in senders:\n")
print(sort(unique(sce@metadata$tensor_res$senders$celltype)))
cat("Cell types kept in receivers:\n")
print(sort(unique(sce@metadata$tensor_res$receivers$celltype)))

missing_colors <- setdiff(
  unique(c(
    sce@metadata$tensor_res$senders$celltype,
    sce@metadata$tensor_res$receivers$celltype
  )),
  names(celltype_colors)
)
if (length(missing_colors) > 0) {
  cat("WARNING - these cell types have no assigned color:\n")
  print(missing_colors)
}

# overview plot: contexts, interactions, senders and receivers per factor
factors <- paste0("Factor.", 1:12)

plots <- lapply(factors, function(f) {
  p_ctx <- sce@metadata$tensor_res$contexts %>%
    ggplot(aes(x = context, y = .data[[f]], fill = context)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = condition_colors) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") +
    labs(x = "", y = f, title = if (f == "Factor.1") "Contexts" else "")

  p_int <- sce@metadata$tensor_res$interactions %>%
    ggplot(aes(x = lr, y = .data[[f]])) +
    geom_bar(stat = "identity", fill = "grey40") +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") +
    labs(x = "", y = "", title = if (f == "Factor.1") "Interactions" else "")

  p_snd <- sce@metadata$tensor_res$senders %>%
    ggplot(aes(x = celltype, y = .data[[f]], fill = celltype)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = celltype_colors) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") +
    labs(x = "", y = "", title = if (f == "Factor.1") "Senders" else "")

  p_rcv <- sce@metadata$tensor_res$receivers %>%
    ggplot(aes(x = celltype, y = .data[[f]], fill = celltype)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = celltype_colors) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") +
    labs(x = "", y = "", title = if (f == "Factor.1") "Receivers" else "")

  p_ctx | p_int | p_snd | p_rcv
})

leg_combined <- cowplot::plot_grid(
  cowplot::get_legend(
    ggplot(sce@metadata$tensor_res$contexts, aes(x = context, y = Factor.1, fill = context)) +
      geom_bar(stat = "identity") + scale_fill_manual(values = condition_colors) + labs(fill = "Condition")
  ),
  cowplot::get_legend(
    ggplot(sce@metadata$tensor_res$senders, aes(x = celltype, y = Factor.1, fill = celltype)) +
      geom_bar(stat = "identity") + scale_fill_manual(values = celltype_colors) + labs(fill = "Cell type")
  ),
  ncol = 2
)

final_plot <- wrap_plots(plots, ncol = 1) | leg_combined
ggsave("CCC/Fusion/output/tensor_overview_12factors.pdf", final_plot, width = 22, height = 40)

# downstream visualization: per-condition loading across factors
sample_loadings <- sce@metadata$tensor_res$contexts %>%
  pivot_longer(-c("Condition", "context"), names_to = "factor", values_to = "loadings") %>%
  mutate(factor = factor(factor))

ggbarplot(sample_loadings, x = condition_col, y = "loadings", fill = condition_col) +
  scale_fill_manual(values = condition_colors) +
  facet_wrap(~factor, ncol = 4) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), legend.position = "right") +
  labs(fill = "Condition")

ggsave("CCC/Fusion/output/tensor_downstream_visualization.pdf", width = 20, height = 15)
```

```{r}
# shared and unique programmes across conditions
test_plot <- sce@metadata$tensor_res$contexts %>%
  column_to_rownames("context") %>%
  select(-Condition) %>%
  as.matrix()

pdf("CCC/Fusion/output/tensor_shared_and_unique_programs.pdf", width = 10, height = 8)
print(
  pheatmap::pheatmap(
    test_plot,
    color = colorRampPalette(c("white", "#E63946"))(100),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    main = "Shared and unique programmes per condition"
  )
)
dev.off()
```

## 9. Shared and unique interactions between factors

```{r}
# test a few loading thresholds before choosing one
thresholds <- c(0.046, 0.059, 0.07) # median, 3rd quartile, stringent custom

threshold_results <- lapply(thresholds, function(threshold) {
  lr_assign <- sce@metadata$tensor_res$interactions %>%
    pivot_longer(-lr, names_to = "factor", values_to = "loading") %>%
    filter(loading > threshold) %>%
    group_by(lr) %>%
    summarise(n_factors = n_distinct(factor), .groups = "drop")

  data.frame(
    threshold = threshold,
    unique = sum(lr_assign$n_factors == 1),
    shared = sum(lr_assign$n_factors > 1),
    total = nrow(lr_assign)
  )
})

threshold_summary <- do.call(rbind, threshold_results)
print(threshold_summary)

# apply the chosen threshold
threshold <- 0.059

lr_factor_assignment <- sce@metadata$tensor_res$interactions %>%
  pivot_longer(-lr, names_to = "factor", values_to = "loading") %>%
  filter(loading > threshold) %>%
  group_by(lr) %>%
  summarise(
    n_factors = n_distinct(factor),
    factors = paste(unique(factor), collapse = ", "),
    .groups = "drop"
  )

unique_int <- lr_factor_assignment %>% filter(n_factors == 1)
shared_int <- lr_factor_assignment %>% filter(n_factors > 1)

cat("Threshold:", threshold, "\n")
cat("Unique interactions:", nrow(unique_int), "\n")
cat("Shared interactions:", nrow(shared_int), "\n")

write.csv(threshold_summary, "CCC/Fusion/output/threshold_test.csv", row.names = FALSE)
write.csv(unique_int, "CCC/Fusion/output/unique_interactions_per_factor.csv", row.names = FALSE)
write.csv(shared_int, "CCC/Fusion/output/shared_interactions_between_factors.csv", row.names = FALSE)

shared_int <- read.csv("CCC/Fusion/output/shared_interactions_between_factors.csv")
```

```{r}
# plot the top 40 shared ligand-receptor (LR) pairs across factors
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

factor_levels <- paste0("Factor.", 1:12)

top40_shared <- sce@metadata$tensor_res$interactions %>%
  pivot_longer(-lr, names_to = "factor", values_to = "int_loading") %>%
  filter(lr %in% shared_int$lr) %>%
  group_by(lr) %>%
  summarise(max_loading = max(int_loading), .groups = "drop") %>%
  slice_max(max_loading, n = 40) %>%
  pull(lr)

shared_expanded <- shared_int %>%
  tidyr::separate_rows(factors, sep = ", ")
shared_expanded <- data.frame(
  lr = shared_expanded$lr,
  factor = trimws(shared_expanded$factors)
)
lr_factor <- shared_expanded %>%
  filter(lr %in% top40_shared)

# barplot of condition contribution per factor
ctx_long <- sce@metadata$tensor_res$contexts %>%
  pivot_longer(-c(context, Condition), names_to = "factor", values_to = "loading") %>%
  mutate(factor = factor(factor, levels = factor_levels))

p_bar <- ggplot(ctx_long, aes(x = factor, y = loading, fill = context)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = condition_colors) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "right"
  ) +
  labs(x = "Factor", y = "Context loading", fill = "Condition")

# top sender / receiver cell type per factor
top_sender <- sce@metadata$tensor_res$senders %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "sender_loading") %>%
  group_by(factor) %>%
  slice_max(sender_loading, n = 1) %>%
  ungroup()

top_receiver <- sce@metadata$tensor_res$receivers %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "receiver_loading") %>%
  group_by(factor) %>%
  slice_max(receiver_loading, n = 1) %>%
  ungroup()

# one shared size scale, same name in both plots, so it merges into a
# single "Loading" legend instead of two separate ones
shared_size_scale <- scale_size(name = "Loading", range = c(1, 6))

# senders dotplot - the only plot showing the Cell type and Loading legends
sender_dot <- lr_factor %>%
  left_join(top_sender, by = "factor") %>%
  mutate(factor = factor(factor, levels = factor_levels))

p_snd <- ggplot(sender_dot, aes(
  x = factor, y = lr,
  size = sender_loading,
  color = celltype
)) +
  geom_point() +
  scale_color_manual(values = celltype_colors, limits = names(celltype_colors), drop = FALSE) +
  shared_size_scale +
  guides(
    size = guide_legend(order = 1),
    color = guide_legend(order = 2)
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10, face = "bold", color = "black")
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    color = "Cell type",
    title = "Senders"
  )

# receivers dotplot - color/size legends suppressed (already shown by Senders)
receiver_dot <- lr_factor %>%
  left_join(top_receiver, by = "factor") %>%
  mutate(factor = factor(factor, levels = factor_levels))

p_rcv <- ggplot(receiver_dot, aes(
  x = factor, y = lr,
  size = receiver_loading,
  color = celltype
)) +
  geom_point() +
  scale_color_manual(
    values = celltype_colors, limits = names(celltype_colors),
    drop = FALSE, guide = "none"
  ) +
  shared_size_scale +
  guides(size = "none") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10, face = "bold", color = "black")
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    color = "Cell type",
    title = "Receivers"
  )

# combine plots - guides="collect" merges/deduplicates legends across all
# three subplots into one consistent legend column, on the right
final_plot <- (p_bar / (p_snd | p_rcv)) +
  plot_layout(heights = c(1, 6), guides = "collect") &
  theme(legend.position = "right")

dir.create("CCC/Fusion/output", recursive = TRUE, showWarnings = FALSE)

ggsave(
  "CCC/Fusion/output/detailed_dotplot_factor_shared_interactions.pdf",
  final_plot,
  width = 24, height = 20
)
```

```{r}
# same dotplot style, but for interactions unique to a single factor
factor_levels <- paste0("Factor.", 1:12)

unique_int <- read.csv("CCC/Fusion/output/unique_interactions_per_factor.csv")

top_sender <- sce@metadata$tensor_res$senders %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "sender_loading") %>%
  group_by(factor) %>%
  slice_max(sender_loading, n = 1) %>%
  ungroup()

top_receiver <- sce@metadata$tensor_res$receivers %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "receiver_loading") %>%
  group_by(factor) %>%
  slice_max(receiver_loading, n = 1) %>%
  ungroup()

unique_long <- sce@metadata$tensor_res$interactions %>%
  pivot_longer(-lr, names_to = "factor", values_to = "int_loading") %>%
  inner_join(unique_int %>% select(lr, factors), by = "lr") %>%
  filter(factor == factors) %>%
  mutate(factor = factor(factor, levels = factor_levels))

sender_dot <- unique_long %>%
  left_join(top_sender, by = "factor")

p_snd <- ggplot(sender_dot, aes(
  x = factor, y = lr,
  size = sender_loading,
  color = celltype
)) +
  geom_point() +
  scale_color_manual(values = celltype_colors) +
  scale_size(range = c(1, 6)) +
  guides(
    size = guide_legend(order = 1),
    color = guide_legend(order = 2)
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "bold"),
    legend.position = "right"
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    size = "Sender loading", color = "Sender cell type",
    title = "Senders"
  )

receiver_dot <- unique_long %>%
  left_join(top_receiver, by = "factor")

p_rcv <- ggplot(receiver_dot, aes(
  x = factor, y = lr,
  size = receiver_loading,
  color = celltype
)) +
  geom_point() +
  scale_color_manual(values = celltype_colors) +
  scale_size(range = c(1, 6)) +
  guides(
    size = guide_legend(order = 1),
    color = guide_legend(order = 2)
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "bold"),
    legend.position = "right"
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    size = "Receiver loading", color = "Receiver cell type",
    title = "Receivers"
  )

ctx_long <- sce@metadata$tensor_res$contexts %>%
  pivot_longer(-c(context, Condition), names_to = "factor", values_to = "loading") %>%
  mutate(factor = factor(factor, levels = factor_levels))

p_bar <- ggplot(ctx_long, aes(x = factor, y = loading, fill = context)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = condition_colors) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  labs(x = "Factor", y = "Context loading", fill = "Condition")

final_plot <- p_bar / (p_snd | p_rcv) + plot_layout(heights = c(1, 6))

ggsave(
  "CCC/Fusion/output/unique_interactions_per_factor.pdf",
  final_plot,
  width = 24, height = 20
)
```

## 10. Top 5% interactions per factor

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

# top 5% of interactions, computed SEPARATELY for each factor - a fixed
# percentage per factor, not a global percentage across all LR pairs
top_pct <- 0.05
factors_all <- paste0("Factor.", 1:12)

# dominant sender/receiver cell type per factor
top_sender_per_factor <- sce@metadata$tensor_res$senders %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "loading") %>%
  group_by(factor) %>%
  slice_max(order_by = loading, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(factor, top_sender = celltype)

top_receiver_per_factor <- sce@metadata$tensor_res$receivers %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "loading") %>%
  group_by(factor) %>%
  slice_max(order_by = loading, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(factor, top_receiver = celltype)

# top 5% interactions per factor's own loading distribution
interactions_long <- sce@metadata$tensor_res$interactions %>%
  pivot_longer(-lr, names_to = "factor", values_to = "loading")

top_interactions <- interactions_long %>%
  group_by(factor) %>%
  filter(loading >= quantile(loading, 1 - top_pct, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(top_sender_per_factor, by = "factor") %>%
  left_join(top_receiver_per_factor, by = "factor") %>%
  mutate(factor = factor(factor, levels = factors_all)) %>%
  arrange(factor, desc(loading))

cat("Total dots in the plot (top", top_pct * 100, "% per factor):", nrow(top_interactions), "\n")

# LR pairs shared across more than one factor's top 5%
lr_factor_counts <- top_interactions %>%
  group_by(lr) %>%
  summarise(n_factors = n_distinct(factor), .groups = "drop")

cat("LR pairs unique to one factor:", sum(lr_factor_counts$n_factors == 1), "\n")
cat("LR pairs shared across >=2 factors:", sum(lr_factor_counts$n_factors > 1), "\n")

dir.create("CCC/Fusion/output", recursive = TRUE, showWarnings = FALSE)
write.csv(top_interactions, "CCC/Fusion/output/top5pct_lr_per_factor.csv", row.names = FALSE)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

condition_colors <- c(
  "IPF.L"    = "#E63946",
  "IPF.U"    = "#F4A261",
  "MS"       = "#2A9D8F",
  "RA"       = "#457B9D",
  "SjD"      = "#8338EC",
  "SS"       = "#FB5607",
  "SScILD.L" = "#FFBE0B",
  "SScILD.U" = "#3A86FF",
  "T1D"      = "#06D6A0",
  "LN"       = "#6D6875"
)

# disease contribution barplot on top
context_long <- sce@metadata$tensor_res$contexts %>%
  pivot_longer(-c(context, Condition), names_to = "factor", values_to = "loading") %>%
  mutate(factor = factor(factor, levels = factors_all)) %>%
  group_by(factor) %>%
  mutate(proportion = loading / sum(loading)) %>%
  ungroup()

p_bar <- ggplot(context_long, aes(x = factor, y = proportion, fill = Condition)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = condition_colors) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "right"
  ) +
  labs(y = "Disease\ncontribution", fill = "Condition")

# senders dotplot
p_snd <- ggplot(top_interactions, aes(
  x = factor, y = lr,
  size = loading, color = top_sender
)) +
  geom_point() +
  scale_color_manual(values = celltype_cols) +
  scale_size(range = c(1, 6)) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.position = "right"
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    size = "Interaction\nscore", color = "Sender cell type",
    title = "Senders"
  )

# receivers dotplot
p_rcv <- ggplot(top_interactions, aes(
  x = factor, y = lr,
  size = loading, color = top_receiver
)) +
  geom_point() +
  scale_color_manual(values = celltype_cols) +
  scale_size(range = c(1, 6)) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.position = "right"
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    size = "Interaction\nscore", color = "Receiver cell type",
    title = "Receivers"
  )

# one barplot spanning the full width, Senders and Receivers below it
final_plot <- p_bar / (p_snd | p_rcv) +
  plot_layout(heights = c(1, 5))

n_lr <- length(unique(top_interactions$lr))

ggsave(
  "CCC/Fusion/output/Top_5per_lr_per_factor.pdf",
  final_plot,
  width = 20, height = max(10, n_lr * 0.28), limitsize = FALSE
)
ggsave(
  "CCC/Fusion/output/Top_5per_lr_per_factor.png",
  final_plot,
  width = 20, height = max(10, n_lr * 0.28), dpi = 300, limitsize = FALSE
)

cat("Saved: CCC/Fusion/output/Top_5per_lr_per_factor.pdf and .png\n")
```

```{r}
# Factors 6 and 7 share several top LR pairs - check whether they are also
# driven by similar diseases and cell types.
# Result: different diseases dominate each (SS/IPF.U for Factor 6,
# IPF.U/MS for Factor 7), but they share the same top sender (Mono/Mac,
# Mural Cells) and receiver (Proliferating Cells, B Cells).
sce@metadata$tensor_res$contexts %>%
  select(Condition, Factor.6, Factor.7)

top_sender_per_factor %>% filter(factor %in% c("Factor.6", "Factor.7"))
top_receiver_per_factor %>% filter(factor %in% c("Factor.6", "Factor.7"))
```

## 11. Top 5% interactions, highlighting the most-similar factor pair

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(ggtext)

top_pct <- 0.05
factors_all <- paste0("Factor.", 1:12)

top_sender_per_factor <- sce@metadata$tensor_res$senders %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "loading") %>%
  group_by(factor) %>%
  slice_max(order_by = loading, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(factor, top_sender = celltype)

top_receiver_per_factor <- sce@metadata$tensor_res$receivers %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "loading") %>%
  group_by(factor) %>%
  slice_max(order_by = loading, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(factor, top_receiver = celltype)

# top 5% interactions per factor - an LR pair present in more than one
# factor's top 5% is genuinely shared (independently ranked strong in each)
interactions_long <- sce@metadata$tensor_res$interactions %>%
  pivot_longer(-lr, names_to = "factor", values_to = "loading")

top_interactions <- interactions_long %>%
  group_by(factor) %>%
  filter(loading >= quantile(loading, 1 - top_pct, na.rm = TRUE)) %>%
  ungroup() %>%
  left_join(top_sender_per_factor, by = "factor") %>%
  left_join(top_receiver_per_factor, by = "factor") %>%
  mutate(factor = factor(factor, levels = factors_all)) %>%
  arrange(factor, desc(loading))

cat("Total dots in the plot (top", top_pct * 100, "% per factor):", nrow(top_interactions), "\n")

lr_factor_counts <- top_interactions %>%
  group_by(lr) %>%
  summarise(n_factors = n_distinct(factor), .groups = "drop")

cat("LR pairs unique to one factor:", sum(lr_factor_counts$n_factors == 1), "\n")
cat("LR pairs shared across >=2 factors:", sum(lr_factor_counts$n_factors > 1), "\n")

# find the pair of factors sharing the most LR pairs, to highlight together
lr_sets <- split(top_interactions$lr, top_interactions$factor)
lr_sets <- lr_sets[sapply(lr_sets, length) > 0] # drop empty factors, if any

factor_pairs <- combn(names(lr_sets), 2, simplify = FALSE)

pairwise_shared <- sapply(factor_pairs, function(pair) {
  length(intersect(lr_sets[[pair[1]]], lr_sets[[pair[2]]]))
})

best_pair_idx <- which.max(pairwise_shared)
highlight_factors <- factor_pairs[[best_pair_idx]]
n_shared_best <- pairwise_shared[best_pair_idx]

cat(
  "\nFactor pair sharing the MOST LR pairs:",
  paste(highlight_factors, collapse = " & "),
  "-", n_shared_best, "shared LR pairs\n"
)

dir.create("CCC/Fusion/output", recursive = TRUE, showWarnings = FALSE)
write.csv(top_interactions, "CCC/Fusion/output/top5pct_lr_per_factor.csv", row.names = FALSE)
cat("Saved: CCC/Fusion/output/top5pct_lr_per_factor.csv\n")

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "darksalmon"
)

condition_colors <- c(
  "IPF.L"    = "#E63946",
  "IPF.U"    = "#F4A261",
  "MS"       = "#2A9D8F",
  "RA"       = "#457B9D",
  "SjD"      = "#8338EC",
  "SS"       = "#FB5607",
  "SScILD.L" = "#FFBE0B",
  "SScILD.U" = "#3A86FF",
  "T1D"      = "#06D6A0",
  "LN"       = "#6D6875"
)

highlight_positions <- match(highlight_factors, factors_all)
highlight_color <- "#B2182B"

highlight_rects <- lapply(highlight_positions, function(pos) {
  annotate("rect",
    xmin = pos - 0.5, xmax = pos + 0.5,
    ymin = -Inf, ymax = Inf,
    fill = highlight_color, alpha = 0.15
  )
})

# colored/bold labels only for the two highlighted factors, plain for the rest
axis_label_colors <- ifelse(
  factors_all %in% highlight_factors,
  paste0("<b style='color:", highlight_color, "'>", factors_all, "</b>"),
  factors_all
)
names(axis_label_colors) <- factors_all

context_long <- sce@metadata$tensor_res$contexts %>%
  pivot_longer(-c(context, Condition), names_to = "factor", values_to = "loading") %>%
  mutate(factor = factor(factor, levels = factors_all)) %>%
  group_by(factor) %>%
  mutate(proportion = loading / sum(loading)) %>%
  ungroup()

p_bar <- ggplot(context_long, aes(x = factor, y = proportion, fill = Condition)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = condition_colors) +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "right"
  ) +
  labs(y = "Disease\ncontribution", fill = "Condition")

p_snd <- ggplot(top_interactions, aes(
  x = factor, y = lr,
  size = loading, color = top_sender
)) +
  highlight_rects +
  geom_point() +
  scale_color_manual(values = celltype_cols) +
  scale_size(range = c(1, 6)) +
  scale_x_discrete(labels = axis_label_colors) +
  theme_bw() +
  theme(
    axis.text.x = ggtext::element_markdown(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.position = "right"
  ) +
  labs(
    x = "Factor", y = "Ligand-Receptor pair",
    size = "Interaction\nscore", color = "Sender cell type",
    title = "Senders"
  )

p_rcv <- ggplot(top_interactions, aes(
  x = factor, y = lr,
  size = loading, color = top_receiver
)) +
  highlight_rects +
  geom_point() +
  scale_color_manual(values = celltype_cols) +
  scale_size(range = c(1, 6)) +
  scale_x_discrete(labels = axis_label_colors) +
  theme_bw() +
  theme(
    axis.text.x = ggtext::element_markdown(angle = 90, hjust = 1),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "right"
  ) +
  labs(
    x = "Factor", y = NULL,
    size = "Interaction\nscore", color = "Receiver cell type",
    title = "Receivers"
  )

final_plot <- p_bar / (p_snd | p_rcv) +
  plot_layout(heights = c(1, 5))

n_lr <- length(unique(top_interactions$lr))

ggsave(
  "CCC/Fusion/output/Top_5per_factor_highlight.pdf",
  final_plot,
  width = 20, height = max(10, n_lr * 0.28), limitsize = FALSE
)
```

## 12. Disease similarity by factor loading

```{r}
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(circlize)
library(grid)

# disease x factor loading matrix from the contexts table
factor_levels <- paste0("Factor.", 1:12)

ctx_wide <- sce@metadata$tensor_res$contexts %>%
  select(Condition, all_of(factor_levels)) %>%
  distinct(Condition, .keep_all = TRUE)

disease_factor_mat <- as.matrix(ctx_wide[, factor_levels])
rownames(disease_factor_mat) <- ctx_wide$Condition

cat("Disease x Factor matrix dimensions:", dim(disease_factor_mat), "\n")
print(round(disease_factor_mat, 3))

# dominant sender / receiver cell type per factor
top_sender_per_factor <- sce@metadata$tensor_res$senders %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "loading") %>%
  group_by(factor) %>%
  slice_max(order_by = loading, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(factor, top_sender = celltype)

top_receiver_per_factor <- sce@metadata$tensor_res$receivers %>%
  pivot_longer(-celltype, names_to = "factor", values_to = "loading") %>%
  group_by(factor) %>%
  slice_max(order_by = loading, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(factor, top_receiver = celltype)

factor_annotation <- data.frame(factor = factor_levels) %>%
  left_join(top_sender_per_factor, by = "factor") %>%
  left_join(top_receiver_per_factor, by = "factor")

rownames(factor_annotation) <- factor_annotation$factor
factor_annotation <- factor_annotation[colnames(disease_factor_mat), ]

cat("Top sender/receiver per factor:\n")
print(factor_annotation)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

# column annotation: dominant sender / receiver cell type per factor
col_ha <- HeatmapAnnotation(
  `Top sender`   = factor_annotation$top_sender,
  `Top receiver` = factor_annotation$top_receiver,
  col = list(
    `Top sender`   = celltype_cols,
    `Top receiver` = celltype_cols
  ),
  show_legend = FALSE,
  annotation_name_gp = gpar(fontsize = 9, fontface = "bold"),
  simple_anno_size = unit(4, "mm")
)

present_celltypes <- intersect(
  names(celltype_cols),
  unique(c(factor_annotation$top_sender, factor_annotation$top_receiver))
)

lgd_celltype <- Legend(
  labels = present_celltypes,
  legend_gp = gpar(fill = celltype_cols[present_celltypes]),
  title = "Top senders\nand receivers",
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

# heatmap: diseases (rows) clustered by factor loading profile, factors
# (columns) clustered by which diseases activate them similarly
col_fun <- colorRamp2(
  c(min(disease_factor_mat), median(disease_factor_mat), max(disease_factor_mat)),
  c("white", "orange", "firebrick")
)

ht <- Heatmap(
  disease_factor_mat,
  name = "Loading",
  col = col_fun,
  top_annotation = col_ha,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_names_gp = gpar(fontsize = 11, fontface = "bold"),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 45,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(sprintf("%.2f", disease_factor_mat[i, j]),
      x, y,
      gp = gpar(fontsize = 7)
    )
  },
  border = TRUE,
  heatmap_legend_param = list(title = "Factor\nloading")
)

dir.create("output/Fusion", recursive = TRUE, showWarnings = FALSE)

pdf("output/Fusion/disease_factor_similarity.pdf", width = 12, height = 8)
draw(ht,
  annotation_legend_list = list(lgd_celltype),
  heatmap_legend_side = "right", annotation_legend_side = "right"
)
dev.off()
```

## 13. Save the processed object

```{r}
saveRDS(sce, file = "data/process/SeuratObjects/Fusion/sce_processed.rds")
sce <- readRDS("data/process/SeuratObjects/Fusion/sce_processed.rds")
```
