---
title: "R Notebook"
output: html_notebook
---
---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
--------------------Fusion of all autoimmune diseases------------------------------------------------ ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------



```{r setup, include=FALSE}

##################PACKAGES
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")


```


##1load all the data
```{r}
#--------process_data

MS = readRDS("data/process/SeuratObjects/MS/MS_process.rds")  
RA = readRDS("data/process/SeuratObjects/RA/RA_process.rds")        
SS = readRDS("data/process/SeuratObjects/SS/SS_process.rds")
IPF.U = readRDS("data/process/SeuratObjects/IPF/Upper/IPF.U_process.rds")
IPF.L = readRDS("data/process/SeuratObjects/IPF/Lower/IPF.L_Final.rds")
T1D = readRDS("data/process/SeuratObjects/T1D/T1D_process.rds")
SScILD.U = readRDS("data/process/SeuratObjects/SScILD/Upper/SS.U_process.rds")
SScILD.L = readRDS("data/process/SeuratObjects/SScILD/Low/SS.L_process.rds")
SjD  = readRDS("data/process/SeuratObjects/SjD/SjD.rds")
#DefaultAssay(SjD) = "RNA"
#SjD = DietSeurat(SjD, assays = "RNA") #delete the SCT assay




```



##2Merge all 
```{r}
#--------fusion

#create list
objs <- list(
  MS = MS,
  RA = RA,
  SS = SS,
  IPF.U = IPF.U,
  IPF.L = IPF.L,
  T1D = T1D,
  SS.U = SS.U,
  SS.L = SS.L,
  SjD = SjD
)
#
sample_ids = c("MS","RA","SS","IPF.U","IPF.L",
  "T1D","SS.U","SS.L","SjD")
#fusion
Fu = merge(
  x = objs[[1]],
  y = objs[-1],           
  add.cell.ids = sample_ids,  
  assay = "RNA",         
  merge.data = FALSE      
)
Fu$orig.ident <- recode( Fu$orig.ident, "1U" = "IPF.U", "2U" = "IPF.U", "3U" = "IPF.U" )
saveRDS(Fu, file = "data/process/SeuratObjects/Fusion/F_raw.rds")
Fu=readRDS("data/process/SeuratObjects/Fusion/F_raw.rds")
#save memory
 #rm(MS,RA,SS,IPF.U,IPF.L,
  #T1D,SS.U,SS.L,cd45neg,cd45pos,objs,sample_ids)
 #gc(full = TRUE)
 


```






##3.Statistics
```{r}
#-------
#Normalize,fvf,pca,scale
Fu=NormalizeData(Fu)
Fu=FindVariableFeatures(Fu)
Fu[["percent.mt"]] <- PercentageFeatureSet(Fu, pattern = "^MT-")
sum(is.na(Fu$percent.mt))  # should now print 0
Fu=ScaleData(Fu, vars.to.regress = c("nCount_RNA", "percent.mt"),features = VariableFeatures(object = Fu))
Fu=RunPCA(Fu,features = VariableFeatures(object = Fu))
ElbowPlot(Fu, ndims = 50, reduction = "pca")
Fu=RunUMAP(Fu,dims = 1:30)


```


##4 integration
```{r}
#------integration
##integration
library(harmony)
Fu=RunHarmony(Fu,group.by.vars = "orig.ident",orig.reduction = "pca",dims.use = 1:20  )
Fu[["RNA"]]=JoinLayers(Fu[["RNA"]])
Fu = RunUMAP(Fu, reduction = "harmony", dims= 1:20, reduction.name = "umap.harmony")

#plot

p = DimPlot(Fu, reduction = "umap.harmony", group.by = "orig.ident", label = FALSE)
p = p + labs(x = "UMAP.h", y = "UMAP.h",title = NULL)
ggsave(filename = "output/Fusion/UMAP.pdf", width=10, height = 8, plot = p)


```



```{r}

#---plot---with_the_DISEASE_type---------

# Convert to factor without a name
Fu$orig.ident <- factor(Fu$orig.ident)
names(attributes(Fu$orig.ident)$levels) <- NULL   # removes the legend title

# Set Idents
Idents(Fu) <- "orig.ident"

# Define colors
disease_colors <- c(
  "MS"        = "darkolivegreen3",
  "RA"        = "red3",
  "SS"        = "orangered",
  "IPF.U"     = "cyan4",
  "IPF.L"     = "cyan",
  "T1D"      = "darkred",
  "SScILD.U" = "gold",
  "SScILD.L" = "magenta",
  "SjD"      = "darkblue"
)

# UMAP plot
P <- DimPlot(
  Fu,
  reduction = "umap.harmony",
  group.by = "orig.ident",
  label = FALSE,
  cols = disease_colors,
  raster = TRUE
) +
  theme(
    aspect.ratio = 1,
    legend.text = element_text(size = 12),
    legend.position = "right"
  ) +
  labs(title = NULL) # Removes the default title

P = P + labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/Fusion/disease.pdf", plot = P, height = 15, width = 10)




#---plot---with_the_DISEASE_type---------

# Convert to factor without a name
Fu$orig.ident <- factor(Fu$orig.ident)
names(attributes(Fu$orig.ident)$levels) <- NULL   # removes the legend title

# Set Idents
Idents(Fu) <- "orig.ident"

# Define colors
disease_colors <- c(
  "MS"        = "darkolivegreen3",
  "RA"        = "red3",
  "SS"        = "orangered",
  "IPF.U"     = "cyan",
  "IPF.L"     = "cyan",
  "T1D"      = "darkred",
  "SScILD.U" = "gold",
  "SScILD.L" = "gold",
  "SjD"      = "darkblue"
)

# UMAP plot
P <- DimPlot(
  Fu,
  reduction = "umap.harmony",
  group.by = "orig.ident",
  label = FALSE,
  cols = disease_colors,
  raster = TRUE
) +
  theme(
    aspect.ratio = 1,
    legend.text = element_text(size = 12),
    legend.position = "right"
  ) +
  labs(title = NULL) # Removes the default title

P = P + labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/Fusion/disease2.pdf", plot = P, height = 8, width = 12)



```


###5.plot
```{r}
library(Seurat)
library(ggplot2)

#-----extract all the celltype

# List all groups in your dataset
orig_groups <- unique(Fu@meta.data$orig.ident)

# extract all celltype
celltypes_by_group <- lapply(orig_groups, function(group) {
  unique(Fu@meta.data$Major_Celltype[Fu@meta.data$orig.ident == group])
})

# name the list by disease
names(celltypes_by_group) <- orig_groups

# print
for (group in names(celltypes_by_group)) {
  cat("====", group, "====\n")
  print(celltypes_by_group[[group]])
  cat("\n")
}


#0. exclude "Immune Cells" - this is an annotation error, not a real
# cell type category, so it's dropped before plotting

excluded_celltypes <- c("Immune Cells")

Fu_plot <- subset(Fu, subset = Major_Celltype %in% setdiff(
  unique(Fu@meta.data$Major_Celltype), excluded_celltypes
))

cat("Cell types kept for the plot:\n")
print(sort(unique(Fu_plot@meta.data$Major_Celltype)))


#1. color palette (full list you provided, one color per cell type)

Idents(Fu_plot) <- "Major_Celltype"

celltype_cols <- c(
  #---- MS ----
  "Oligodendrocytes"           = "tomato4",
  "OPC"                        = "firebrick3",
  "Astrocytes"                 = "salmon3",
  "Microglia"                  = "sienna",
  "Neurons"                    = "mistyrose3",
  "Endothelial Cells"          = "olivedrab3",
  "CD8"                        = "tomato",
  "Fibroblasts"                = "maroon2",
  "Mural Cells"                = "paleturquoise3",
  "CD4"                        = "purple",
  #---- RA ----
  "Macrophages"                = "sienna2",
  "Mast Cells"                 = "palevioletred2",
  "Venous Endothelial"         = "chocolate1",
  "Arterial Endothelial"       = "cadetblue",
  "B Cells"                    = "yellow3",
  #---- SS ----
  "Secretory Cells"            = "orangered",
  "Smooth Muscle Cells"        = "slateblue2",
  "Suprabasal Keratinocyte"    = "deeppink",
  "Schwann Cells"              = "plum3",
  "Basal Keratinocyte"         = "olivedrab",
  "Melanocytes"                = "orangered4",
  "Dendritic Cells"            = "olivedrab2",
  "Lymphatic Endothelial"      = "brown3",
  #---- IPF.U ----
  "Goblet Cells"               = "cornflowerblue",
  "NK Cells"                   = "brown2",
  "Peribronchial Endothelial"  = "darkgoldenrod2",
  "Proliferating Cells"        = "coral1",
  "Monocytes"                  = "darksalmon",
  "Ciliated Cells"             = "saddlebrown",
  "Alveolar Cells"             = "aquamarine3",
  #---- IPF.L ----
  "Basal Cells"                = "indianred3",
  #---- T1D ----
  "Ductal Cells"               = "magenta",
  "Acinar Cells"               = "mediumorchid3",
  "Alpha Cells"                = "mediumorchid1",
  "Delta Cells"                = "mediumpurple2",
  #---- SScILD.U ----
  "Capillary Endothelial"      = "red3",
  "Myeloid Cells"              = "tomato1",
  "Plasmacytoid Dendritic"     = "red",
  #---- SjD ----
  "Mono/Mac"                   = "peru",
  "gd/NK"                      = "steelblue2",
  "Plasma Cells"               = "lightpink2"
)

# safety check: every kept cell type must have an assigned color
missing_colors <- setdiff(unique(Fu_plot@meta.data$Major_Celltype), names(celltype_cols))
if (length(missing_colors) > 0) {
  cat("WARNING - these cell types have no assigned color:\n")
  print(missing_colors)
}


#2. plain UMAP (no labels yet)

p <- DimPlot(
  Fu_plot,
  reduction = "umap.harmony",
  group.by = "Major_Celltype",
  label = FALSE,
  cols = celltype_cols,
  raster = TRUE
) +
  NoLegend() +
  theme(aspect.ratio = 1) +
  xlab("UMAP.h") +
  ylab("UMAP.h")


#3. add cluster labels

p_labeled <- LabelClusters(
  plot = p,
  id = "Major_Celltype",
  repel = TRUE,
  box = TRUE,
  max.overlaps = 1000,
  color = "black",
  fontface = "bold",
  size = 4,
  force = 6,
  force_pull = 0.5,
  box.padding = 0.8,
  point.padding = 0.4,
  min.segment.length = 0
) +
  labs(title = NULL)


#4. save

pdf("output/Fusion/Major_Celltype_all.pdf",
    width = 12, height = 10)
print(p_labeled)
dev.off()

cat("Saved: output/Fusion/Major_Celltype_all.pdf\n")


```




##.shared celltype

```{r}
library(dplyr)
library(tidyr)
library(pheatmap)
library(grid)

#0. exclude the same cell types used throughout the project, starting
# directly from Fu's metadata
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
                         "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
                         "Plasma Cells")
meta_filtered <- Fu@meta.data %>%
  filter(!General %in% excluded_celltypes)
cat("Cell types remaining after exclusion:\n")
print(sort(unique(meta_filtered$General)))

#1. count how many diseases per cell type (using the FILTERED metadata)
tab <- table(meta_filtered$General, meta_filtered$orig.ident)
n_diseases <- rowSums(tab > 0)
celltype_disease_table <- data.frame(
  General = rownames(tab),
  n_diseases = n_diseases,
  diseases = apply(tab > 0, 1, function(x) {
    paste(colnames(tab)[x], collapse = ", ")
  }),
  row.names = NULL
)

#2. extract only shared cell types (present in >=2 diseases)
shared_celltypes <- celltype_disease_table$General[
  celltype_disease_table$n_diseases >= 2
]
shared_celltype_table <- celltype_disease_table[
  celltype_disease_table$n_diseases >= 2,
]
write.csv(
  shared_celltype_table,
  file = "output/Fusion/shared_celltype.csv",
  row.names = FALSE
)
cat("Shared cell types (>=2 diseases):\n")
print(shared_celltypes)

#3. count cells per shared cell type per disease
count_tbl_shared <- meta_filtered %>%
  filter(General %in% shared_celltypes) %>%
  dplyr::count(General, orig.ident) %>%
  filter(n >= 2)

#4. convert to matrix
count_mat_shared <- count_tbl_shared %>%
  pivot_wider(
    names_from = orig.ident,
    values_from = n,
    values_fill = 0
  )
count_mat_shared <- as.data.frame(count_mat_shared)
rownames(count_mat_shared) <- count_mat_shared$General
count_mat_shared$General <- NULL
count_mat_shared <- as.matrix(count_mat_shared)

#5. convert counts to PERCENTAGE of cells per disease
# (each column = one disease; total is the FULL filtered cell count for
# that disease, across all kept cell types - not just the shared ones -
# so percentages reflect true composition, not just the shared subset)
total_per_disease <- meta_filtered %>%
  dplyr::count(orig.ident) %>%
  tibble::deframe()
# align to the same disease columns as count_mat_shared, in the same order
total_per_disease <- total_per_disease[colnames(count_mat_shared)]
percent_mat_shared <- sweep(count_mat_shared, 2, total_per_disease, FUN = "/") * 100
percent_mat_shared <- round(percent_mat_shared, 1)

#6. plot - color and displayed numbers both use PERCENTAGE now
# (only change from before: a "% of cells" title is added above the
# legend color bar. pheatmap has no built-in argument for that, so the
# heatmap is built once on a disposable scratch device to get its
# gtable, then drawn exactly once on the real output device with the
# title text added on top.)
out_width  <- 10
out_height <- 12

scratch_file <- tempfile(fileext = ".pdf")
pdf(scratch_file, width = out_width, height = out_height)
p <- pheatmap::pheatmap(
  percent_mat_shared,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  display_numbers = percent_mat_shared,
  number_color = "white",
  fontsize_row = 12,
  fontsize_col = 12,
  fontsize_number = 12,
  color = colorRampPalette(c("grey", "red", "red3"))(100),
  border_color = NA,
  silent = TRUE
)
dev.off()
unlink(scratch_file)

gt <- p$gtable

pdf("output/Fusion/HEATMAP_shared_celltype.pdf", width = out_width, height = out_height)
grid.newpage()   # start the page BEFORE any convertWidth/Height call below,
                  # otherwise that call silently creates its own blank page

w <- convertWidth(gt$widths, "npc", valueOnly = TRUE)
h <- convertHeight(gt$heights, "npc", valueOnly = TRUE)
leg_row <- gt$layout[gt$layout$name == "legend", ]
leg_x1 <- sum(w[seq_len(leg_row$r)])
leg_y1 <- 1 - sum(h[seq_len(leg_row$t - 1)])

grid.draw(gt)
grid.text(
  "% of cells",
  x = unit(leg_x1, "npc"),
  y = unit(leg_y1 + 0.015, "npc"),
  just = c("right", "bottom"),
  gp = gpar(fontsize = 10, fontface = "bold")
)
dev.off()


```









```{r}

library(harmony)
library(Seurat)
library(ggplot2)

#----Plot-----with_the_celltype--------

#1. extract unique cell types from General, and remove the same
# excluded cell types used throughout the rest of the project

all_general_celltypes <- sort(unique(Fu$General))
print(all_general_celltypes)

excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
                         "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
                         "Plasma Cells")

keep_celltypes <- setdiff(all_general_celltypes, excluded_celltypes)
cat("Cell types kept for the plot:\n")
print(keep_celltypes)
cat("Cell types excluded:\n")
print(intersect(all_general_celltypes, excluded_celltypes))

#2. subset the Seurat object to only the cells we're keeping
# (DimPlot/LabelClusters need the object itself filtered, not just the color list)

Fu_plot <- subset(Fu, subset = General %in% keep_celltypes)


#3. fixed color per cell type (same palette used across the project)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

# safety check: every kept cell type must have an assigned color
missing_colors <- setdiff(keep_celltypes, names(celltype_cols))
if (length(missing_colors) > 0) {
  cat("WARNING - these cell types have no assigned color yet:\n")
  print(missing_colors)
}


#4. Plain UMAP plot (no labels yet)

p <- DimPlot(
  Fu_plot,
  reduction = "umap.harmony",
  group.by = "General",
  label = FALSE,
  cols = celltype_cols,
  raster = TRUE
) +
  NoLegend() +
  theme(aspect.ratio = 1) +
  xlab("UMAP.h") +
  ylab("UMAP.h")

#5. add cluster labels (fixed typo: id = "General", not "Genarl")
# increased size/force/padding so labels in dense regions (e.g. Mono/Mac)
# are pushed further away and stay readable

p_labeled <- LabelClusters(
  plot = p,
  id = "General",
  repel = TRUE,
  box = TRUE,
  max.overlaps = 1000,
  color = "white",
  fontface = "bold",
  size = 8,
  force = 6,
  force_pull = 0.5,
  box.padding = 0.8,
  point.padding = 0.4,
  min.segment.length = 0
) +
  labs(title = NULL)

#6. save

pdf("output/Fusion/General_cellType.pdf",
    width = 12, height = 8)
print(p_labeled)
dev.off()

cat("Saved: output/Fusion/General_cellType.pdf\n")


```



##7Correlation

```{r}
library(Seurat)
library(pheatmap)

#------

#1. filter to only the cell types we're keeping (same exclusion list
# used throughout the project)

excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells", "Melanocytes",
                         "Keratinocytes", "Endocrine Cells", "Smooth Muscle Cells",
                         "Plasma Cells")

all_general_celltypes <- sort(unique(Fu$General))
keep_celltypes <- setdiff(all_general_celltypes, excluded_celltypes)

cat("Cell types kept for the correlation:\n")
print(keep_celltypes)

Fu_corr <- subset(Fu, subset = General %in% keep_celltypes)


#2. set identity to orig.ident (sample/dataset), on the FILTERED object

Idents(Fu_corr) <- "orig.ident"


#3. average expression per identity, computed only on kept cells

av.exp <- AverageExpression(
  Fu_corr,
  assays = "RNA",
  slot = "data"
)$RNA
# rows = genes, columns = identities (orig.ident)


#4. correlation matrix

correlation_matrix <- cor(
  as.matrix(av.exp),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrix[is.na(correlation_matrix)] <- 0


#5. heatmap
# NOTE: pheatmap() saves and closes the PNG device internally when given
# a filename - no separate dev.off() call is needed (or wanted) after it

mcol <- colorRampPalette(c("white", "darkblue", "red"))(100)

pheatmap(
  correlation_matrix,
  clustering_method = "ward.D2",
  color = mcol,
  border_color = NA,
  fontsize_row = 10,
  fontsize_col = 10,
  display_numbers = FALSE,
  filename = "output/Fusion/correlation.pdf",
  width = 12,
  height = 10
)

```



##9.shared pathways between diseases
```{r}

##1.load pathways
MS=read.csv("CCC/MS/output/PathwaysGO.csv")
RA=read.csv("CCC/RA/output/PathwaysGO.csv")
SS=read.csv("CCC/SS/output/PathwaysGO.csv")
T1D=read.csv("CCC/T1D/output/PathwaysGO.csv")
SjD=read.csv("CCC/SjD/output/PathwaysGO.csv")
IPF.U=read.csv("CCC/IPF/Upper/output/PathwaysGO.csv")
IPF.L=read.csv("CCC/IPF/Low/output/PathwaysGO.csv")
SScILD.U=read.csv("CCC/SScILD/Upper/output/PathwaysGO.csv")
SScILD.L=read.csv("CCC/SScILD/Low/output/PathwaysGO.csv")




```


```{r}

##2. Extract significant pathways per disease
MS.sig       <- MS[MS$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
RA.sig       <- RA[RA$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
SS.sig       <- SS[SS$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
T1D.sig      <- T1D[T1D$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
SjD.sig      <- SjD[SjD$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
IPF.U.sig    <- IPF.U[IPF.U$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
IPF.L.sig    <- IPF.L[IPF.L$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
SScILD.U.sig <- SScILD.U[SScILD.U$significant == TRUE, c("term_id", "term_name", "source", "p_value")]
SScILD.L.sig <- SScILD.L[SScILD.L$significant == TRUE, c("term_id", "term_name", "source", "p_value")]

##3. Get all unique term_id + term_name + source combinations
all_terms <- unique(rbind(
  MS.sig[,       c("term_id", "term_name", "source")],
  RA.sig[,       c("term_id", "term_name", "source")],
  SS.sig[,       c("term_id", "term_name", "source")],
  T1D.sig[,      c("term_id", "term_name", "source")],
  SjD.sig[,      c("term_id", "term_name", "source")],
  IPF.U.sig[,    c("term_id", "term_name", "source")],
  IPF.L.sig[,    c("term_id", "term_name", "source")],
  SScILD.U.sig[, c("term_id", "term_name", "source")],
  SScILD.L.sig[, c("term_id", "term_name", "source")]
))

##4. Function to get p_value for a term in a disease (NA if not present)
get_pval <- function(term_ids, disease_df) {
  pval <- disease_df$p_value[match(term_ids, disease_df$term_id)]
  return(pval)
}

##5. Build shared table with presence/absence + best_pval only
shared <- data.frame(
  term_id   = all_terms$term_id,
  term_name = all_terms$term_name,
  source    = all_terms$source,
  # presence/absence
  MS        = all_terms$term_id %in% MS.sig$term_id,
  RA        = all_terms$term_id %in% RA.sig$term_id,
  SS        = all_terms$term_id %in% SS.sig$term_id,
  T1D       = all_terms$term_id %in% T1D.sig$term_id,
  SjD       = all_terms$term_id %in% SjD.sig$term_id,
  IPF.U     = all_terms$term_id %in% IPF.U.sig$term_id,
  IPF.L     = all_terms$term_id %in% IPF.L.sig$term_id,
  SScILD.U  = all_terms$term_id %in% SScILD.U.sig$term_id,
  SScILD.L  = all_terms$term_id %in% SScILD.L.sig$term_id
)

##6. Count how many diseases share each pathway
shared$n_diseases <- rowSums(shared[, 4:12])

##7. Keep only pathways present in at least 2 diseases
shared <- shared[shared$n_diseases >= 2, ]

##8. Take the smallest p_value across all diseases directly
shared$best_pval <- apply(
  cbind(
    get_pval(shared$term_id, MS.sig),
    get_pval(shared$term_id, RA.sig),
    get_pval(shared$term_id, SS.sig),
    get_pval(shared$term_id, T1D.sig),
    get_pval(shared$term_id, SjD.sig),
    get_pval(shared$term_id, IPF.U.sig),
    get_pval(shared$term_id, IPF.L.sig),
    get_pval(shared$term_id, SScILD.U.sig),
    get_pval(shared$term_id, SScILD.L.sig)
  ),
  1, function(x) min(x, na.rm = TRUE))

##9. Sort by most shared then most significant
shared <- shared[order(-shared$n_diseases, shared$best_pval), ]

##10. Save
write.csv(shared, "CCC/Fusion/output/shared_pathways.csv", row.names = FALSE)

##11. Quick summary
cat("Total shared pathways:", nrow(shared), "\n")
cat("\nBreakdown by n_diseases:\n")
print(table(shared$n_diseases))
cat("\nBreakdown by source:\n")
print(table(shared$source))


```


```{r}

library(ComplexHeatmap)
library(circlize)
library(grid)
library(dplyr)

mat_full <- as.matrix(shared[, 4:12] + 0)
rownames(mat_full) <- shared$term_name

top_n <- 25
n_diseases_per_pathway <- rowSums(mat_full)
top_pathways <- names(sort(n_diseases_per_pathway, decreasing = TRUE))[1:top_n]

mat <- mat_full[top_pathways, , drop = FALSE]

mat_labels <- matrix(ifelse(mat == 1, "Present", "Absent"),
                      nrow = nrow(mat), dimnames = dimnames(mat))
col_map <- c("Absent" = "white", "Present" = "darkblue")

source_cols <- c("GO:BP" = "#1B9E77", "REAC" = "#D95F02")
row_ha <- rowAnnotation(
  Source = shared$source[match(top_pathways, shared$term_name)],
  col = list(Source = source_cols),
  show_annotation_name = FALSE
)

n_rows <- nrow(mat)
n_cols <- ncol(mat)

ht <- Heatmap(
  mat_labels,
  name = "Presence",
  col = col_map,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  row_dend_side = "left",
  row_dend_width = unit(30, "mm"),
  width  = unit(n_cols * 15, "mm"),   # fix column width too, so it doesn't stretch to fill a big page
  height = unit(n_rows * 10, "mm"),
  row_names_side = "right",
  row_names_gp = gpar(fontsize = 14),
  column_names_gp = gpar(fontsize = 12),
  column_names_rot = 45,
  row_names_max_width = max_text_width(rownames(mat), gp = gpar(fontsize = 14)) + unit(4, "mm"),
  left_annotation = row_ha,
  heatmap_legend_param = list(title = "Presence"),
  border = TRUE
)

# tighten the page to match the fixed component sizes instead of guessing big
pdf("CCC/Fusion/output/heatmap_shared_pathways.pdf",
    width  = 5 + n_cols * 0.6,   # dend + annotation + body + row names + legend, roughly
    height = 3 + n_rows * 0.42)  # column labels + body, roughly

draw(ht, padding = unit(c(2, 2, 2, 10), "mm"))  # bottom, left, top, right
dev.off()

```










```{r}

#.12 Presence/absence heatmap (58 pathways)
library(pheatmap)

mat <- as.matrix(shared[, 4:12] + 0)
rownames(mat) <- shared$term_name

pheatmap(mat,
         color          = c("white", "darkblue"),
         cluster_rows   = TRUE,
         cluster_cols   = FALSE,
         fontsize_row   = 8,
         fontsize_col   = 10,
         legend_breaks  = c(0, 1),
         legend_labels  = c("Absent", "Present"),
         filename       = "CCC/Fusion/output/heatmap_shared_pathways.pdf",
         width          = 14,
         height         = 14)

##13. UpSet plot
library(UpSetR)

upset_list <- list(
  MS       = MS.sig$term_id,
  RA       = RA.sig$term_id,
  SS       = SS.sig$term_id,
  T1D      = T1D.sig$term_id,
  SjD      = SjD.sig$term_id,
  IPF.U    = IPF.U.sig$term_id,
  IPF.L    = IPF.L.sig$term_id,
  SScILD.U = SScILD.U.sig$term_id,
  SScILD.L = SScILD.L.sig$term_id
)

pdf("CCC/Fusion/output/upset_shared_pathways.pdf", 
    width  = 16, 
    height = 12,
    onefile = FALSE)

suppressWarnings(
  upset(fromList(upset_list),
        nsets           = 9,
        order.by        = "freq",
        decreasing      = TRUE,
        mb.ratio        = c(0.6, 0.4),
        text.scale      = c(1.5, 1.5, 1.2, 1.2, 1.5, 1.2),
        point.size      = 4,
        line.size       = 1,
        mainbar.y.label = "Number of Shared Pathways",
        sets.x.label    = "Pathways per Disease",
        main.bar.color  = "#2166AC",
        sets.bar.color  = "#D6604D")
)

dev.off()

```




##10 shared interactioins between diseases


```{r}
library(dplyr)
#1. load
MS       = read.csv("CCC/MS/output/liana_filt.csv")
RA       = read.csv("CCC/RA/output/liana_filt.csv")
SS       = read.csv("CCC/SS/output/liana_filt.csv")
T1D      = read.csv("CCC/T1D/output/liana_filt.csv")
SjD      = read.csv("CCC/SjD/output/liana_filt.csv")
SScILD.U = read.csv("CCC/SScILD/Upper/output/liana_filt.csv")
SScILD.L = read.csv("CCC/SScILD/Low/output/liana_filt.csv")
IPF.U    = read.csv("CCC/IPF/Upper/output/liana_filt.csv")
IPF.L    = read.csv("CCC/IPF/Low/output/liana_filt.csv")
diseases_list <- list(
  MS = MS, RA = RA, SS = SS, T1D = T1D, SjD = SjD,
  SScILD.U = SScILD.U, SScILD.L = SScILD.L,
  IPF.U = IPF.U, IPF.L = IPF.L
)

# sanity check: confirm every disease's file actually has the columns
# step 3 depends on, BEFORE we try to build "interaction" from them
for (d in names(diseases_list)) {
  cols <- names(diseases_list[[d]])
  missing <- setdiff(c("ligand.complex", "receptor.complex", "magnitude_rank", "source", "target"), cols)
  if (length(missing) > 0) {
    cat("[WARNING]", d, "is missing columns:", paste(missing, collapse = ", "), "\n")
    cat("  actual columns in", d, ":", paste(cols, collapse = ", "), "\n")
  }
}

#2. rename cell types in source/target
celltype_rename_map <- c(
  "Astrocytes" = "Glial Cells",
  "Oligodendrocytes" = "Glial Cells",
  "OPC" = "Glial Cells",
  "Schwann Cells" = "Glial Cells",
  "Neurons" = "Neurons",
  "Endothelial Cells" = "Endothelial Cells",
  "Arterial Endothelial" = "Endothelial Cells",
  "Venous Endothelial" = "Endothelial Cells",
  "Lymphatic Endothelial" = "Endothelial Cells",
  "Capillary Endothelial" = "Endothelial Cells",
  "Peribronchial Endothelial" = "Endothelial Cells",
  "Macrophages" = "Mono/Mac",
  "Monocytes" = "Mono/Mac",
  "Myeloid Cells" = "Mono/Mac",
  "Mono/Mac" = "Mono/Mac",
  "Dendritic Cells" = "Mono/Mac",
  "Plasmacytoid Dendritic" = "Mono/Mac",
  "Microglia" = "Mono/Mac",
  "Alveolar Cells" = "Epithelial Cells",
  "Ciliated Cells" = "Epithelial Cells",
  "Goblet Cells" = "Epithelial Cells",
  "Basal Cells" = "Epithelial Cells",
  "Secretory Cells" = "Epithelial Cells",
  "Acinar Cells" = "Epithelial Cells",
  "Ductal Cells" = "Epithelial Cells",
  "Basal Keratinocyte" = "Keratinocytes",
  "Suprabasal Keratinocyte" = "Keratinocytes",
  "Alpha Cells" = "Endocrine Cells",
  "Delta Cells" = "Endocrine Cells",
  "CD4" = "CD4",
  "CD8" = "CD8",
  "B Cells" = "B Cells",
  "Fibroblasts" = "Fibroblasts",
  "Mast Cells" = "Mast Cells",
  "Melanocytes" = "Melanocytes",
  "Mural Cells" = "Mural Cells",
  "Smooth Muscle Cells" = "Smooth Muscle Cells",
  "Immune Cells" = "Immune Cells",
  "NK Cells" = "NK Cells",
  "Plasma Cells" = "Plasma Cells",
  "Proliferating Cells" = "Proliferating Cells"
)
# rename for SjD
celltype_rename_map <- c(celltype_rename_map, c(
  "Neuronal cells" = "Neurons",
  "Lymphatic Endothelial cells" = "Endothelial Cells",
  "CCL14 Venous Endothelial" = "Endothelial Cells",
  "Endothelial Progenitors" = "Endothelial Cells",
  "cDC1" = "Mono/Mac",
  "cDC2" = "Mono/Mac",
  "inflammatory cDC2" = "Mono/Mac",
  "Mo-DC" = "Mono/Mac",
  "M1 Macrophages" = "Mono/Mac",
  "M2 Macrophages" = "Mono/Mac",
  "Basal Epithelial cells" = "Epithelial Cells",
  "Epithelial Progenitor cells" = "Epithelial Cells",
  "Intercalated Ductal Epithelial cells" = "Epithelial Cells",
  "Ionocytes" = "Epithelial Cells",
  "Mucous Acinar Epithelial cells" = "Epithelial Cells",
  "Myoepithelial cells" = "Epithelial Cells",
  "Seromucous Acinar Epithelial cells" = "Epithelial Cells",
  "Serous Acinar Epithelial cells" = "Epithelial Cells",
  "Striated Ductal Epithelial cells" = "Epithelial Cells",
  "Activated Treg cells" = "CD4",
  "Resting Treg cells" = "CD4",
  "Cytotoxic Th cells" = "CD4",
  "DLGAP1+ Th cells" = "CD4",
  "LTB+ Th17 cells" = "CD4",
  "Memory Th cells" = "CD4",
  "Migrating Th17 cells" = "CD4",
  "Naive Tfh cells" = "CD4",
  "Naive Th cells" = "CD4",
  "Secretory Tph cells" = "CD4",
  "Tph17" = "CD4",
  "CD62L+ TPEx" = "CD8",
  "CD8+ Tcm" = "CD8",
  "CD8+ Tem" = "CD8",
  "CD8+ Trm" = "CD8",
  "Exhausted CD8+ T cells" = "CD8",
  "Exhausted IEG CD8+ T cells" = "CD8",
  "Infiltrating CD8+ T cells" = "CD8",
  "Inflammatory CD8+ T cells" = "CD8",
  "Naive CD8+ T cells" = "CD8",
  "Recruited Activated CD8+ T cells" = "CD8",
  "Recruiting Effector CD8+ Trm" = "CD8",
  "GC B cells" = "B Cells",
  "Naive B cells" = "B Cells",
  "Immunofibroblasts" = "Fibroblasts",
  "SPARC+ COL3A1+ Fibroblasts" = "Fibroblasts",
  "Capillary pericytes" = "Mural Cells",
  "Immunopericytes" = "Mural Cells",
  "VSMC" = "Smooth Muscle Cells",
  "gd-T cells" = "Immune Cells",
  "ILCs" = "Immune Cells",
  "CD16+ NK cells" = "NK Cells",
  "CD27+ NK cells" = "NK Cells",
  "CD56+ NK cells" = "NK Cells",
  "IgA Plasma cells" = "Plasma Cells",
  "IgG1 Plasma cells" = "Plasma Cells",
  "IgG3/4 Plasma cells" = "Plasma Cells",
  "IgM Plasma cells" = "Plasma Cells",
  "TNF+ Plasma cells" = "Plasma Cells",
  "Proliferating cells" = "Proliferating Cells"
))
rename_map_df <- data.frame(
  CellType     = names(celltype_rename_map),
  CellType_new = unname(celltype_rename_map),
  stringsAsFactors = FALSE
)
rename_vec <- function(x, map_df) {
  x <- gsub("(_[0-9]+)+$", "", x, perl = TRUE)
  x <- trimws(gsub("[[:space:]\u00A0]+", " ", x, perl = TRUE))
  idx <- match(x, map_df$CellType)
  ifelse(is.na(idx), x, map_df$CellType_new[idx])
}
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells","Melanocytes","Keratinocytes","Endocrine Cells","Smooth Muscle Cells","Plasma Cells")
diseases_list <- lapply(diseases_list, function(df) {
  df$source <- rename_vec(df$source, rename_map_df)
  df$target <- rename_vec(df$target, rename_map_df)
  df <- df[!(df$source %in% excluded_celltypes | df$target %in% excluded_celltypes), ]
  df
})

#3. build interaction column and combine all diseases
all_data <- dplyr::bind_rows(lapply(names(diseases_list), function(d) {
  df <- diseases_list[[d]]
  df$disease <- d
  df$interaction <- paste(df$ligand.complex, df$receptor.complex, sep = "^")
  df
}))
stopifnot("`interaction` did not survive bind_rows - check the per-disease columns above" = "interaction" %in% names(all_data))
cat("[checkpoint] all_data columns:", paste(names(all_data), collapse = ", "), "\n")
cat("[checkpoint] all_data rows:", nrow(all_data), "\n")

#4. find interactions present in at least 2 diseases
shared_int <- all_data %>%
  dplyr::group_by(interaction, ligand.complex, receptor.complex) %>%
  dplyr::summarise(
    n_diseases     = dplyr::n_distinct(disease),
    best_magnitude = min(magnitude_rank, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::filter(n_diseases >= 2) %>%
  dplyr::arrange(dplyr::desc(n_diseases), best_magnitude)
stopifnot("`interaction` missing from shared_int" = "interaction" %in% names(shared_int))
cat("[checkpoint] shared_int columns:", paste(names(shared_int), collapse = ", "), "\n")
cat("Total unique shared interactions:", nrow(shared_int), "\n")

#5. build long format table
long_int <- all_data %>%
  dplyr::filter(interaction %in% shared_int$interaction) %>%
  dplyr::group_by(interaction, ligand.complex, receptor.complex, disease) %>%
  dplyr::slice_min(order_by = magnitude_rank, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::rename(best_magnitude = magnitude_rank) %>%
  dplyr::left_join(
    shared_int %>% dplyr::select(interaction, n_diseases, group_best = best_magnitude),
    by = "interaction"
  )
stopifnot("`interaction` missing after left_join" = "interaction" %in% names(long_int))
cat("[checkpoint] post-join columns:", paste(names(long_int), collapse = ", "), "\n")

long_int <- long_int %>%
  # sort by the INTERACTION's overall best score first, so every row belonging
  # to the same interaction stays grouped together, and within each group,
  # order diseases by their own strength (best first)
  dplyr::arrange(dplyr::desc(n_diseases), group_best, interaction, best_magnitude) %>%
  dplyr::select(interaction, ligand.complex, receptor.complex, n_diseases, best_magnitude,
                disease, source, target)

#6. blank out repeated ligand/receptor/n_diseases for readability
interaction_id <- paste(long_int$ligand.complex, long_int$receptor.complex, sep = "_")
is_duplicate <- duplicated(interaction_id)
long_int$ligand.complex   <- as.character(long_int$ligand.complex)
long_int$receptor.complex <- as.character(long_int$receptor.complex)
long_int$n_diseases       <- as.character(long_int$n_diseases)
long_int$ligand.complex[is_duplicate]   <- ""
long_int$receptor.complex[is_duplicate] <- ""
long_int$n_diseases[is_duplicate]       <- ""

#7. save
dir.create("CCC/Fusion/output", recursive = TRUE, showWarnings = FALSE)
write.csv(long_int, "CCC/Fusion/output/shared_interactions_between_disease.csv", row.names = FALSE)
long_int <- read.csv("CCC/Fusion/output/shared_interactions_between_disease.csv")
  
   
  


```





```{r}
#plot
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)


disease_levels <- c("MS","RA","SS","T1D","SjD","IPF.U","IPF.L","SScILD.U","SScILD.L")

#1. fixed color palette per cell type (same categories used across the project)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)


#2. keep only interactions present in 8 or 9 of the 9 diseases,
# capped to the top N by strength (magnitude_rank)

top_n <- 40
n_per_group <- ceiling(top_n / 2)

shared_all_8_9 <- shared_int %>%
  filter(n_diseases %in% c(8, 9)) %>%
  mutate(lr = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  distinct(lr, .keep_all = TRUE) %>%
  arrange(best_magnitude)

cat("Total DISTINCT interactions present in 8-9 diseases:", nrow(shared_all_8_9), "\n")
print(table(shared_all_8_9$n_diseases))

top_9 <- shared_all_8_9 %>% filter(n_diseases == 9) %>% slice_head(n = n_per_group)
top_8 <- shared_all_8_9 %>% filter(n_diseases == 8) %>% slice_head(n = n_per_group)

shared_8_9 <- bind_rows(top_9, top_8) %>% arrange(desc(n_diseases), best_magnitude)

cat("Keeping", nrow(top_9), "pairs present in all 9 diseases,",
    nrow(top_8), "pairs present in 8 diseases (", nrow(shared_8_9), "total)\n")

top_lr <- shared_8_9$lr
stopifnot(!any(duplicated(top_lr)))


#3. lr x disease table

lr_disease <- all_data %>%
  mutate(lr = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  filter(lr %in% top_lr) %>%
  group_by(lr, ligand.complex, receptor.complex, disease) %>%
  slice_min(order_by = magnitude_rank, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    lr      = factor(lr, levels = rev(top_lr)),
    disease = factor(disease, levels = disease_levels),
    score   = -log10(magnitude_rank)
  )

cat("Rows in lr_disease (should be up to", length(top_lr), "LR pairs x 8-9 diseases each):",
    nrow(lr_disease), "\n")

stopifnot(!any(duplicated(paste(lr_disease$lr, lr_disease$disease))))

# separate dataframes with a common column name for the color mapping
lr_disease_snd <- lr_disease %>% mutate(celltype = source)
lr_disease_rcv <- lr_disease %>% mutate(celltype = target)


#4. shared size scale for BOTH plots

score_range <- range(lr_disease$score, na.rm = TRUE)
score_breaks <- seq(score_range[1], score_range[2], length.out = 5)
magnitude_labels <- signif(10^(-score_breaks), 2)

shared_size_scale <- scale_size(
  name   = "magnitude_rank\n(smaller = stronger)",
  range  = c(1, 6),
  limits = score_range,
  breaks = score_breaks,
  labels = as.character(magnitude_labels)
)


#5. Senders dotplot - this is the ONLY plot that shows the Cell type
# color legend. LR pair labels (y-axis) explicitly set to black.

p_snd <- ggplot(lr_disease_snd, aes(x = disease, y = lr,
                                     size = score, color = celltype)) +
  geom_point() +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols), drop = FALSE) +
  shared_size_scale +
  guides(size  = guide_legend(order = 1),
         color = guide_legend(order = 2)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 11, face = "bold"),
        axis.text.y = element_text(size = 11, face = "bold", color = "black"),
        axis.title = element_text(size = 12, face = "bold")) +
  labs(x = "Disease", y = "Ligand-Receptor pair",
       color = "Cell type",
       title = "Senders")


#6. Receivers dotplot - color legend explicitly suppressed (guide="none"),
# since Senders already shows the identical Cell type legend; y-axis
# (LR pair) labels moved to the RIGHT side, SAME explicit black color
# as Senders, so both panels' LR pair text matches exactly

p_rcv <- ggplot(lr_disease_rcv, aes(x = disease, y = lr,
                                     size = score, color = celltype)) +
  geom_point() +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols),
                      drop = FALSE, guide = "none") +
  scale_y_discrete(position = "right") +
  shared_size_scale +
  guides(size = guide_legend(order = 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 11, face = "bold"),
        axis.text.y = element_text(size = 11, face = "bold", color = "black"),
        axis.title = element_text(size = 12, face = "bold")) +
  labs(x = "Disease", y = "Ligand-Receptor pair",
       color = "Cell type",
       title = "Receivers")


#7. combine side by side, ONE merged legend

final_plot <- (p_snd | p_rcv) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

dir.create("CCC/Fusion/output", recursive = TRUE, showWarnings = FALSE)

plot_height <- max(10, length(top_lr) * 0.22)
cat("Number of LR pairs:", length(top_lr), "| plot height:", plot_height, "inches\n")

ggsave("CCC/Fusion/output/shared_interactions_8_9_diseases.pdf",
       final_plot, width = 18, height = plot_height, limitsize = FALSE)



cat("Saved: CCC/Fusion/output/shared_interactions_8_9_diseases.pdf and .png\n")




```




```{r}

#plot
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
disease_levels <- c("MS","RA","SS","T1D","SjD","IPF.U","IPF.L","SScILD.U","SScILD.L")
#1. fixed color palette per cell type (same categories used across the project)
celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)
#2. keep only interactions present in 8 or 9 of the 9 diseases,
# capped to the top N by strength (magnitude_rank)
top_n <- 15
n_per_group <- ceiling(top_n / 2)
shared_all_8_9 <- shared_int %>%
  filter(n_diseases %in% c(8, 9)) %>%
  mutate(lr = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  distinct(lr, .keep_all = TRUE) %>%
  arrange(best_magnitude)
cat("Total DISTINCT interactions present in 8-9 diseases:", nrow(shared_all_8_9), "\n")
print(table(shared_all_8_9$n_diseases))
top_9 <- shared_all_8_9 %>% filter(n_diseases == 9) %>% slice_head(n = n_per_group)
top_8 <- shared_all_8_9 %>% filter(n_diseases == 8) %>% slice_head(n = n_per_group)
shared_8_9 <- bind_rows(top_9, top_8) %>% arrange(desc(n_diseases), best_magnitude)
cat("Keeping", nrow(top_9), "pairs present in all 9 diseases,",
    nrow(top_8), "pairs present in 8 diseases (", nrow(shared_8_9), "total)\n")
top_lr <- shared_8_9$lr
stopifnot(!any(duplicated(top_lr)))
#3. lr x disease table
lr_disease <- all_data %>%
  mutate(lr = paste(ligand.complex, receptor.complex, sep = "^")) %>%
  filter(lr %in% top_lr) %>%
  group_by(lr, ligand.complex, receptor.complex, disease) %>%
  slice_min(order_by = magnitude_rank, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    lr      = factor(lr, levels = rev(top_lr)),
    disease = factor(disease, levels = disease_levels),
    score   = -log10(pmax(magnitude_rank, 1e-300))
  )
cat("Rows in lr_disease (should be up to", length(top_lr), "LR pairs x 8-9 diseases each):",
    nrow(lr_disease), "\n")
stopifnot(!any(duplicated(paste(lr_disease$lr, lr_disease$disease))))
# separate dataframes with a common column name for the color mapping
lr_disease_snd <- lr_disease %>% mutate(celltype = source)
lr_disease_rcv <- lr_disease %>% mutate(celltype = target)
#4. shared size scale for BOTH plots
score_range <- range(lr_disease$score, na.rm = TRUE)
score_breaks <- seq(score_range[1], score_range[2], length.out = 5)
magnitude_labels <- signif(10^(-score_breaks), 2)
shared_size_scale <- scale_size(
  name   = "magnitude_rank\n(smaller = stronger)",
  range  = c(1, 6),
  limits = score_range,
  breaks = score_breaks,
  labels = as.character(magnitude_labels)
)
#5. Senders dotplot - the Cell type legend is generated by a SEPARATE
# invisible layer, decoupled from the "score" size aesthetic entirely
# (see note below), so it can't be broken by size/color interactions in
# the legend-key rendering. One dummy row per category, placed at an
# existing x/y so no axis expansion happens, drawn with alpha = 0 so it
# never appears in the panel - it exists purely to drive the legend.
dummy_celltype <- data.frame(
  celltype = factor(names(celltype_cols), levels = names(celltype_cols)),
  disease  = lr_disease_snd$disease[1],
  lr       = lr_disease_snd$lr[1]
)
p_snd <- ggplot(lr_disease_snd, aes(x = disease, y = lr)) +
  geom_point(aes(size = score, color = celltype), show.legend = c(size = TRUE, colour = FALSE)) +
  geom_point(data = dummy_celltype, aes(x = disease, y = lr, color = celltype),
             size = 5, alpha = 0, inherit.aes = FALSE) +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols), drop = FALSE) +
  shared_size_scale +
  guides(size  = guide_legend(order = 1),
         color = guide_legend(order = 2, override.aes = list(alpha = 1, size = 5))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 11, face = "bold"),
        axis.text.y = element_text(size = 11, face = "bold", color = "black"),
        axis.title = element_text(size = 12, face = "bold")) +
  labs(x = "Disease", y = "Ligand-Receptor pair",
       color = "Cell type",
       title = "Senders")
#6. Receivers dotplot - color legend explicitly suppressed (guide="none"),
# since Senders already shows the identical Cell type legend; y-axis
# (LR pair) labels moved to the RIGHT side, SAME explicit black color
# as Senders, so both panels' LR pair text matches exactly
p_rcv <- ggplot(lr_disease_rcv, aes(x = disease, y = lr,
                                     size = score, color = celltype)) +
  geom_point() +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols),
                      drop = FALSE, guide = "none") +
  scale_y_discrete(position = "right") +
  shared_size_scale +
  guides(size = guide_legend(order = 1)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 11, face = "bold"),
        axis.text.y = element_text(size = 11, face = "bold", color = "black"),
        axis.title = element_text(size = 12, face = "bold")) +
  labs(x = "Disease", y = "Ligand-Receptor pair",
       color = "Cell type",
       title = "Receivers")
#7. combine side by side, ONE merged legend
final_plot <- (p_snd | p_rcv) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")
dir.create("CCC/Fusion/output", recursive = TRUE, showWarnings = FALSE)
plot_height <- max(10, length(top_lr) * 0.22)
cat("Number of LR pairs:", length(top_lr), "| plot height:", plot_height, "inches\n")
ggsave("CCC/Fusion/output/shared_interactions_8_9_diseases.pdf",
       final_plot, width = 18, height = plot_height, limitsize = FALSE)


```








##10.1 upset plot of shared interactions
```{r}


library(UpSetR)

#1. build interaction_list: one set of unique interactions per disease

interaction_list <- lapply(split(all_data$interaction, all_data$disease), unique)

cat("Interactions per disease:\n")
print(sapply(interaction_list, length))


#2. upset plot with all intersect combinations
pdf("CCC/Fusion/output/upset_plot_shared_interactions.pdf", 
    width = 20, height = 13,
    onefile = FALSE)

upset(
  fromList(interaction_list),
  nsets             = length(interaction_list),
  nintersects       = 30,
  order.by          = "freq",
  mb.ratio          = c(0.5, 0.5),
  text.scale        = c(1.8, 1.4, 1.4, 1.2, 1.5, 1.3),
  point.size        = 3.5,
  line.size         = 1.2,
  main.bar.color    = "#993404",
  sets.bar.color    = "#2c7fb8",
  matrix.color      = "black",
  shade.color       = "grey85",
  set_size.show     = TRUE,
  set_size.scale_max = max(sapply(interaction_list, length)) * 1.15
)

dev.off()




```






##11 GRN between diseases-:

##11.1compare regulons, shared and unique




```{r}

library(dplyr)
library(tidyr)
#1. load RSS files per disease
rss_MS       <- read.csv("GRN_SCENIC/MS/output/rss_cellType.csv")
rss_RA       <- read.csv("GRN_SCENIC/RA/output/rss_cellType.csv")
rss_SS       <- read.csv("GRN_SCENIC/SS/output/rss_cellType.csv")
rss_T1D      <- read.csv("GRN_SCENIC/T1D/output/rss_cellType.csv")
rss_SjD      <- read.csv("GRN_SCENIC/SjD/output/rss_cellType.csv")
rss_IPF.U    <- read.csv("GRN_SCENIC/IPF/Upper/output/rss_cellType.csv")
rss_IPF.L    <- read.csv("GRN_SCENIC/IPF/Low/output/rss_cellType.csv")
rss_SScILD.U <- read.csv("GRN_SCENIC/SScILD/Upper/output/rss_cellType.csv")
rss_SScILD.L <- read.csv("GRN_SCENIC/SScILD/Low/output/rss_cellType.csv")
#2. clean TF column names (pySCENIC "(+)" becomes "..." when read into R)
clean_tf_names <- function(df) {
  names(df) <- gsub("\\.+$", "", names(df))
  names(df)[1] <- "CellType"
  df
}
rss_list <- list(
  MS           = clean_tf_names(rss_MS),
  RA           = clean_tf_names(rss_RA),
  SS           = clean_tf_names(rss_SS),
  T1D          = clean_tf_names(rss_T1D),
  SjD          = clean_tf_names(rss_SjD),
  IPF.Upper    = clean_tf_names(rss_IPF.U),
  IPF.Low      = clean_tf_names(rss_IPF.L),
  SScILD.Upper = clean_tf_names(rss_SScILD.U),
  SScILD.Low   = clean_tf_names(rss_SScILD.L)
)
#3. rename cell types to standardized categories
celltype_rename_map <- c(
  # --- Neural / glial (MS) ---
  "Astrocytes"                = "Glial Cells",
  "Oligodendrocytes"          = "Glial Cells",
  "OPC"                       = "Glial Cells",
  "Schwann Cells"             = "Glial Cells",
  "Neurons"                   = "Neurons",
  # --- Endothelial subtypes -> Endothelial Cells ---
  "Endothelial Cells"         = "Endothelial Cells",
  "Arterial Endothelial"      = "Endothelial Cells",
  "Venous Endothelial"        = "Endothelial Cells",
  "Lymphatic Endothelial"     = "Endothelial Cells",
  "Capillary Endothelial"     = "Endothelial Cells",
  "Peribronchial Endothelial" = "Endothelial Cells",
  # --- Myeloid subtypes -> Mono/Mac ---
  "Macrophages"                = "Mono/Mac",
  "Monocytes"                  = "Mono/Mac",
  "Myeloid Cells"              = "Mono/Mac",
  "Mono/Mac"                   = "Mono/Mac",
  "Dendritic Cells"            = "Mono/Mac",
  "Plasmacytoid Dendritic"     = "Mono/Mac",
  "Microglia"                  = "Mono/Mac",
  # --- Epithelial subtypes -> Epithelial Cells ---
  "Alveolar Cells"             = "Epithelial Cells",
  "Ciliated Cells"             = "Epithelial Cells",
  "Goblet Cells"                = "Epithelial Cells",
  "Basal Cells"                 = "Epithelial Cells",
  "Secretory Cells"             = "Epithelial Cells",
  "Acinar Cells"                = "Epithelial Cells",
  "Ductal Cells"                = "Epithelial Cells",
  # --- Keratinocyte subtypes ---
  "Basal Keratinocyte"          = "Keratinocytes",
  "Suprabasal Keratinocyte"     = "Keratinocytes",
  # --- Endocrine (pancreas) ---
  "Alpha Cells"                  = "Endocrine Cells",
  "Delta Cells"                  = "Endocrine Cells",
  # match
  "CD4"                  = "CD4",
  "CD8"                  = "CD8",
  "B Cells"              = "B Cells",
  "Fibroblasts"          = "Fibroblasts",
  "Mast Cells"           = "Mast Cells",
  "Melanocytes"          = "Melanocytes",
  "Mural Cells"          = "Mural Cells",
  "Smooth Muscle Cells"  = "Smooth Muscle Cells",
  "Immune Cells"         = "Immune Cells",
  "NK Cells"             = "NK Cells",
  "Plasma Cells"         = "Plasma Cells",
  "Proliferating Cells"  = "Proliferating Cells"
)
rename_map_df <- data.frame(
  CellType     = names(celltype_rename_map),
  CellType_new = unname(celltype_rename_map),
  stringsAsFactors = FALSE
)
rename_celltypes <- function(df, map_df) {
  df$CellType <- trimws(gsub("[[:space:]\u00A0]+", " ", df$CellType, perl = TRUE))
  idx <- match(df$CellType, map_df$CellType)
  df$CellType <- ifelse(is.na(idx), df$CellType, map_df$CellType_new[idx])
  df
}
rss_list <- lapply(rss_list, rename_celltypes, map_df = rename_map_df)
#4. combine all diseases into one long table: CellType, TF, RSS, disease
rss_combined <- dplyr::bind_rows(lapply(names(rss_list), function(d) {
  df <- rss_list[[d]]
  df_long <- df %>% tidyr::pivot_longer(-CellType, names_to = "TF", values_to = "RSS")
  df_long$disease <- d
  df_long
}))
stopifnot("`TF` missing from rss_combined right after pivot_longer" = "TF" %in% names(rss_combined))
cat("[checkpoint] rss_combined columns:", paste(names(rss_combined), collapse = ", "), "\n")
cat("rss_combined built with", nrow(rss_combined), "rows\n")
#5. for each TF in each disease, keep only the top cell type (highest RSS score)
# RSS = Regulon Specificity Score: tells us how exclusively a TF's activity
# is concentrated in one cell type vs spread across all cell types
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells","Melanocytes","Keratinocytes","Endocrine Cells","Smooth Muscle Cells","Plasma Cells")
dominant_celltype_per_TF_disease <- rss_combined %>%
  dplyr::filter(!CellType %in% excluded_celltypes) %>%
  dplyr::group_by(disease, TF) %>%
  dplyr::slice_max(order_by = RSS, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::select(TF, disease, CellType, RSS)
stopifnot("`TF` missing from dominant_celltype_per_TF_disease" = "TF" %in% names(dominant_celltype_per_TF_disease))
cat("[checkpoint] dominant_celltype_per_TF_disease columns:", paste(names(dominant_celltype_per_TF_disease), collapse = ", "), "\n")
#6. figure out which diseases each TF is present in
# (a TF is "present" in a disease if it appears in that disease's RSS file,
#  regardless of which cell type ended up as its dominant one)
diseases_per_TF <- dominant_celltype_per_TF_disease %>%
  dplyr::group_by(TF) %>%
  dplyr::summarise(
    n_diseases = dplyr::n_distinct(disease),
    diseases   = paste(sort(unique(disease)), collapse = ", "),
    .groups = "drop"
  )
stopifnot("`TF` missing from diseases_per_TF" = "TF" %in% names(diseases_per_TF))
cat("[checkpoint] diseases_per_TF columns:", paste(names(diseases_per_TF), collapse = ", "), "\n")
#7. for each TF, get its best (highest) RSS score across all diseases it appears in
best_rss_per_TF <- dominant_celltype_per_TF_disease %>%
  dplyr::group_by(TF) %>%
  dplyr::slice_max(order_by = RSS, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::select(TF, best_disease = disease, Best_RSS = RSS)
stopifnot("`TF` missing from best_rss_per_TF" = "TF" %in% names(best_rss_per_TF))
cat("[checkpoint] best_rss_per_TF columns:", paste(names(best_rss_per_TF), collapse = ", "), "\n")
tf_summary <- diseases_per_TF %>%
  dplyr::left_join(best_rss_per_TF, by = "TF")
cat("[checkpoint] tf_summary columns:", paste(names(tf_summary), collapse = ", "), "\n")
#8. split into shared (>=2 diseases) vs unique (1 disease only)
shared_final <- tf_summary %>%
  dplyr::filter(n_diseases >= 2) %>%
  dplyr::transmute(TF, Diseases = diseases, Best_RSS = round(Best_RSS, 2)) %>%
  dplyr::arrange(TF)
unique_final <- tf_summary %>%
  dplyr::filter(n_diseases == 1) %>%
  dplyr::transmute(TF, Disease = diseases, Best_RSS = round(Best_RSS, 2)) %>%
  dplyr::arrange(TF)
cat("shared_final built with", nrow(shared_final), "rows\n")
cat("unique_final built with", nrow(unique_final), "rows\n")
#9. combine cell type and score into one text value per disease, e.g. "CD4: 0.90"
celltype_score_combined <- dominant_celltype_per_TF_disease %>%
  dplyr::mutate(CellType_Score = paste0(CellType, ": ", round(RSS, 2))) %>%
  dplyr::select(TF, disease, CellType_Score) %>%
  tidyr::pivot_wider(names_from = disease, values_from = CellType_Score)
#10. merge with the shared and unique TF tables
final_table_shared <- shared_final %>%
  dplyr::left_join(celltype_score_combined, by = "TF")
final_table_unique <- unique_final %>%
  dplyr::left_join(celltype_score_combined, by = "TF")
cat("Shared regulons table rows:", nrow(final_table_shared), "\n")
cat("Unique regulons table rows:", nrow(final_table_unique), "\n")
#11. save both tables
dir.create("output/Fusion", recursive = TRUE, showWarnings = FALSE)
write.csv(final_table_shared, "output/Fusion/shared_regulons_dominant_celltype.csv",
          row.names = FALSE)
write.csv(final_table_unique, "output/Fusion/unique_regulons_dominant_celltype.csv",
          row.names = FALSE)
final_table_unique <- read.csv("output/Fusion/unique_regulons_dominant_celltype.csv")
final_table_shared <- read.csv("output/Fusion/shared_regulons_dominant_celltype.csv")


```


```{r}
##plot
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(circlize)
library(grid)

# keep only the top 40 TFs by Best_RSS, then sort so regulons present in
# ALL diseases come first, Best_RSS as tiebreaker
top_n <- 40
final_table_top <- final_table_shared %>%
  arrange(desc(Best_RSS)) %>%
  slice_head(n = top_n) %>%
  mutate(n_diseases = lengths(strsplit(Diseases, ",\\s*"))) %>%
  arrange(desc(n_diseases), desc(Best_RSS))

# rename columns
names(final_table_top)[names(final_table_top) == "IPF.Upper"]    <- "IPF.U"
names(final_table_top)[names(final_table_top) == "IPF.Low"]      <- "IPF.L"
names(final_table_top)[names(final_table_top) == "SScILD.Upper"] <- "SScILD.U"
names(final_table_top)[names(final_table_top) == "SScILD.Low"]   <- "SScILD.L"

disease_cols <- c("MS","RA","SS","T1D","SjD","IPF.U","IPF.L",
                   "SScILD.U","SScILD.L")

# parse "CellType: Score" text into separate cell type and score matrices
parse_celltype <- function(x) {
  sapply(strsplit(x, ": "), function(v) if (length(v) == 2) v[1] else NA)
}
parse_score <- function(x) {
  sapply(strsplit(x, ": "), function(v) if (length(v) == 2) as.numeric(v[2]) else NA)
}

markers_present <- final_table_top$TF
dom_ct_mat <- matrix(NA, nrow = length(markers_present), ncol = length(disease_cols),
                      dimnames = list(markers_present, disease_cols))
score_mat  <- matrix(NA, nrow = length(markers_present), ncol = length(disease_cols),
                      dimnames = list(markers_present, disease_cols))

for (dcol in disease_cols) {
  dom_ct_mat[, dcol] <- parse_celltype(final_table_top[[dcol]])
  score_mat[, dcol]  <- parse_score(final_table_top[[dcol]])
}

# fixed color per cell type (Glial Cells, Neurons, Immune Cells excluded already)
celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

# defensive: strip backticks, newlines, and stray whitespace in case any
# survived a copy/paste - this is what caused "invalid color name '\n#E31A1C'"
celltype_cols <- trimws(gsub("[`\r\n]", "", celltype_cols))
names(celltype_cols) <- c("B Cells","CD4","CD8","NK Cells","Mono/Mac",
                           "Mast Cells","Proliferating Cells","Endothelial Cells",
                           "Fibroblasts","Mural Cells","Epithelial Cells")

stopifnot(all(grepl("^(#[0-9A-Fa-f]{6}|[a-zA-Z]+[0-9]*)$", celltype_cols)))

# matrix used only for clustering (missing values treated as 0)
cluster_mat <- score_mat
cluster_mat[is.na(cluster_mat)] <- 0

# only keep cell types actually present, for a clean legend
present_celltypes <- intersect(names(celltype_cols), unique(na.omit(as.vector(dom_ct_mat))))

# draw bubbles - solid color, no border, size scaled to RSS score
# where the TF is ABSENT for a disease (NA), draw a small grey "x" instead
# of leaving the cell completely blank, so absence is visually explicit
cell_fun <- function(j, i, x, y, w, h, fill) {
  ct <- dom_ct_mat[i, j]
  if (is.na(ct)) {
    grid::grid.text(
      "x", x = x, y = y,
      gp = grid::gpar(col = "grey60", fontsize = 9, fontface = "bold")
    )
    return(invisible(NULL))
  }
  sc <- score_mat[i, j]
  grid::grid.circle(
    x = x, y = y,
    r = sc * 0.42 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = celltype_cols[ct], col = celltype_cols[ct])
  )
}

# heatmap sized to fit exactly the number of rows/columns, avoiding empty space
ht <- Heatmap(
  cluster_mat,
  name = "blank",
  col = colorRamp2(c(0, 1), c("white", "white")),
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 9, fontface = "bold"),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "ward.D2",
  clustering_method_columns = "ward.D2",
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  row_dend_width = unit(10, "mm"),
  column_dend_height = unit(8, "mm"),
  row_gap = unit(0.5, "mm"),
  width = unit(ncol(cluster_mat) * 9, "mm"),
  height = unit(nrow(cluster_mat) * 5, "mm")
)

# cell type legend (only cell types actually shown)
lgd_celltype <- Legend(
  labels = present_celltypes,
  legend_gp = gpar(fill = celltype_cols[present_celltypes]),
  title = "Dominant cell type",
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

# size legend for RSS score
size_breaks <- c(0.2, 0.4, 0.6, 0.8, 1.0)
lgd_size <- Legend(
  labels = as.character(size_breaks),
  title = "RSS score",
  type = "points",
  pch = 16,
  size = unit(size_breaks * 0.42 * 4, "mm"),
  legend_gp = gpar(col = "grey40", fill = "grey70"),
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

# legend entry explaining the "x" symbol
lgd_absent <- Legend(
  labels = "Not present",
  type = "points",
  pch = 4,
  legend_gp = gpar(col = "grey60", fontsize = 9),
  title = "",
  labels_gp = gpar(fontsize = 9)
)

draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size, lgd_absent),
     heatmap_legend_side = "right", annotation_legend_side = "right")

# save, sized to match the actual data (top 40 rows, 9 disease columns)
plot_w <- 9
plot_h <- max(8, top_n * 0.22)

pdf("output/Fusion/shared_regulons_top40.pdf", width = plot_w, height = plot_h)
draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size, lgd_absent),
     heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size, lgd_absent),
     heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()


```




## GRN and TLS genes:##LOOK IF THE SHARED TF have TARGET that belong to the TLS genes list

```{r}
#1.⁠ ⁠count for each TF the number of targets it has from the TLS list

#look for TF and their target in each disease
library(dplyr)
library(purrr)
library(tidyr)

parse_target_genes <- function(x) {
  if (is.na(x) || x == "") return(character(0))
  m <- gregexpr("(?<=\\(')[^']+(?=')", x, perl = TRUE)
  regmatches(x, m)[[1]]
}

load_disease_regulons <- function(folder_parts, label) {
  path <- file.path("GRN_SCENIC", do.call(file.path, as.list(folder_parts)),
                     "output", "regulons.csv")
  df <- read.csv(path, skip = 1, header = TRUE)
  names(df)[1:2] <- c("TF", "MotifID")
  df %>%
    dplyr::filter(TargetGenes != "") %>%
    dplyr::transmute(TF, target = purrr::map(TargetGenes, parse_target_genes)) %>%
    tidyr::unnest(target) %>%
    dplyr::distinct(TF, target) %>%
    dplyr::mutate(disease = label)
}

disease_paths <- list(
  MS           = "MS",
  RA           = "RA",
  SS           = "SS",
  T1D          = "T1D",
  SjD          = "SjD",
  IPF.U        = c("IPF", "Upper"),
  IPF.L        = c("IPF", "Low"),
  SScILD.U     = c("SScILD", "Upper"),
  SScILD.L     = c("SScILD", "Low")
)

regulon_list <- list()
for (label in names(disease_paths)) {
  cat("loading:", label, "... ")
  result <- tryCatch(load_disease_regulons(disease_paths[[label]], label),
                      error = function(e) e)
  if (inherits(result, "error")) {
    cat("FAILED -", conditionMessage(result), "\n")
  } else {
    cat("ok,", nrow(result), "rows\n")
    regulon_list[[label]] <- result
  }
}
regulon_long <- dplyr::bind_rows(regulon_list)
cat("\ntotal rows:", nrow(regulon_long), "\n")
cat("diseases loaded:", paste(unique(regulon_long$disease), collapse = ", "), "\n")

## LOOK IF THE SHARED TF have TARGET that belong to the TLS genes list
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(stringr)

final_table_shared <- read.csv("output/Fusion/shared_regulons_dominant_celltype.csv")

core_threshold <- 6
tf_disease_counts <- regulon_long %>%
  dplyr::distinct(TF, disease) %>%
  dplyr::count(TF, name = "n_diseases")
core_tfs <- tf_disease_counts %>%
  dplyr::filter(n_diseases >= core_threshold) %>%
  dplyr::pull(TF)
cat("core_tfs (>=", core_threshold, "diseases, from regulon_long):", length(core_tfs), "TFs\n")

disease_cols <- c("IPF.Low", "IPF.Upper", "MS", "RA", "SS",
                   "SScILD.Low", "SScILD.Upper", "SjD", "T1D")
celltype_long <- final_table_shared %>%
  tidyr::pivot_longer(cols = dplyr::all_of(disease_cols),
                       names_to = "disease", values_to = "cell_rss") %>%
  dplyr::filter(!is.na(cell_rss), cell_rss != "") %>%
  dplyr::mutate(
    CellType = stringr::str_extract(cell_rss, "^[^:]+"),
    RSS      = as.numeric(stringr::str_extract(cell_rss, "[0-9.]+$"))
  ) %>%
  dplyr::select(TF, disease, CellType, RSS)

dominant_celltype_per_TF <- celltype_long %>%
  dplyr::group_by(TF) %>%
  dplyr::slice_max(order_by = RSS, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::select(TF, dominant_celltype = CellType, dominant_RSS = RSS)
cat("dominant cell type recovered for", nrow(dominant_celltype_per_TF), "TFs\n")

stopifnot("regulon_long not found - run the loader first" = exists("regulon_long"))

#1. TLS marker groups
germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
                      "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")
inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
                   "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
                   "ICAM1", "VCAM1", "SELE")
lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12",
                          "CXCR5", "CCR7", "CXCR4")
hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")
fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
         "CXCL13", "CD54", "ICAM1", "FCER2")

marker_groups <- list(
  germinal_center      = germinal_center,
  inflammation         = inflammation,
  lymphoid_chemokines  = lymphoid_chemokines,
  hev                  = hev,
  fdc                  = fdc
)

tls_genes_long <- purrr::imap_dfr(
  marker_groups,
  ~ data.frame(category = .y, gene = .x)
)

#2. restrict the regulon data to the shared/core TFs only
regulon_core <- regulon_long %>%
  dplyr::filter(TF %in% core_tfs)

#3. overlap: which TLS genes are targets of a shared TF, and in which disease(s)
tf_gene_overlap <- regulon_core %>%
  dplyr::inner_join(tls_genes_long, by = c("target" = "gene"),
                     relationship = "many-to-many") %>%
  dplyr::distinct(TF, category, target, disease) %>%
  dplyr::arrange(category, TF, target) %>%
  dplyr::left_join(dominant_celltype_per_TF, by = "TF")

cat("Matches (shared TF -> TLS gene, per disease, with dominant cell type):\n")
print(tf_gene_overlap)
cat("\nRows per disease (sanity check - should be non-zero for all 9):\n")
print(tf_gene_overlap %>% dplyr::count(disease))

#4. count, per TF, how many distinct TLS genes it regulates
tf_tls_target_counts <- tf_gene_overlap %>%
  dplyr::distinct(TF, target) %>%
  dplyr::count(TF, name = "n_tls_targets")

#5. final single output: one row per (TF, category, target), with the
# TF's dominant cell type/RSS and its total TLS-target count attached
tf_tls_full_table <- tf_gene_overlap %>%
  dplyr::distinct(TF, category, target, dominant_celltype, dominant_RSS) %>%
  dplyr::left_join(tf_tls_target_counts, by = "TF") %>%
  dplyr::arrange(dplyr::desc(n_tls_targets), TF, category, target)

cat("\nPreview of final table:\n")
print(head(tf_tls_full_table, 20))

write.csv(tf_tls_full_table, "GRN_SCENIC/TF_TLS_targets_by_category_celltype.csv", row.names = FALSE)



```




```{r}
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(circlize)
library(grid)

tf_disease_tls_counts <- tf_gene_overlap %>%
  dplyr::distinct(TF, target, disease) %>%
  dplyr::count(TF, disease, name = "n_tls_targets_in_disease")

top_n_tf <- 30
top_tfs <- tf_tls_target_counts %>%
  dplyr::arrange(dplyr::desc(n_tls_targets)) %>%
  dplyr::slice_head(n = top_n_tf) %>%
  dplyr::pull(TF)

disease_levels <- c("MS", "RA", "SS", "T1D", "SjD",
                     "IPF.U", "IPF.L", "SScILD.U", "SScILD.L")

heatmap_data <- tf_disease_tls_counts %>%
  dplyr::filter(TF %in% top_tfs) %>%
  tidyr::complete(TF = top_tfs, disease = disease_levels,
                   fill = list(n_tls_targets_in_disease = 0))

mat_df <- heatmap_data %>%
  tidyr::pivot_wider(names_from = disease, values_from = n_tls_targets_in_disease)

mat <- as.data.frame(mat_df)
rownames(mat) <- mat$TF
mat$TF <- NULL
mat <- as.matrix(mat)
mat <- mat[top_tfs, disease_levels]

col_fun <- circlize::colorRamp2(c(0, max(mat)), c("white", "#08519C"))

# cell_fun draws the actual count number in every cell, on top of the
# color fill - guaranteed to show regardless of pheatmap-translation quirks
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.text(mat[i, j], x, y, gp = grid::gpar(fontsize = 9,
                   col = ifelse(mat[i, j] > max(mat) * 0.5, "white", "black")))
}

ht <- Heatmap(
  mat,
  name = "Count",
  col = col_fun,
  cell_fun = cell_fun,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_dend = TRUE,
  row_names_side = "right",
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 11),
  column_names_rot = 45,
  border = TRUE
)

pdf("GRN_SCENIC/TF_TLS_count_heatmap.pdf", width = 10, height = 12)
draw(ht)
dev.off()


```



```{r}

#look for TF and their target in each disease
library(dplyr)
library(purrr)
library(tidyr)

parse_target_genes <- function(x) {
  if (is.na(x) || x == "") return(character(0))
  m <- gregexpr("(?<=\\(')[^']+(?=')", x, perl = TRUE)
  regmatches(x, m)[[1]]
}

load_disease_regulons <- function(folder_parts, label) {
  path <- file.path("GRN_SCENIC", do.call(file.path, as.list(folder_parts)),
                     "output", "regulons.csv")
  df <- read.csv(path, skip = 1, header = TRUE)
  names(df)[1:2] <- c("TF", "MotifID")
  df %>%
    dplyr::filter(TargetGenes != "") %>%
    dplyr::transmute(TF, target = purrr::map(TargetGenes, parse_target_genes)) %>%
    tidyr::unnest(target) %>%
    dplyr::distinct(TF, target) %>%
    dplyr::mutate(disease = label)
}

disease_paths <- list(
  MS           = "MS",
  RA           = "RA",
  SS           = "SS",
  T1D          = "T1D",
  SjD          = "SjD",
  IPF.U        = c("IPF", "Upper"),
  IPF.L        = c("IPF", "Low"),
  SScILD.U     = c("SScILD", "Upper"),
  SScILD.L     = c("SScILD", "Low")
)

regulon_list <- list()
for (label in names(disease_paths)) {
  cat("loading:", label, "... ")
  result <- tryCatch(load_disease_regulons(disease_paths[[label]], label),
                      error = function(e) e)
  if (inherits(result, "error")) {
    cat("FAILED -", conditionMessage(result), "\n")
  } else {
    cat("ok,", nrow(result), "rows\n")
    regulon_list[[label]] <- result
  }
}
regulon_long <- dplyr::bind_rows(regulon_list)
cat("\ntotal rows:", nrow(regulon_long), "\n")
cat("diseases loaded:", paste(unique(regulon_long$disease), collapse = ", "), "\n")

## LOOK IF THE SHARED TF have TARGET that belong to the TLS genes list
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(stringr)

final_table_shared <- read.csv("output/Fusion/shared_regulons_dominant_celltype.csv")

core_threshold <- 6
tf_disease_counts <- regulon_long %>%
  dplyr::distinct(TF, disease) %>%
  dplyr::count(TF, name = "n_diseases")
core_tfs <- tf_disease_counts %>%
  dplyr::filter(n_diseases >= core_threshold) %>%
  dplyr::pull(TF)
cat("core_tfs (>=", core_threshold, "diseases, from regulon_long):", length(core_tfs), "TFs\n")

disease_cols <- c("IPF.Low", "IPF.Upper", "MS", "RA", "SS",
                   "SScILD.Low", "SScILD.Upper", "SjD", "T1D")
celltype_long <- final_table_shared %>%
  tidyr::pivot_longer(cols = dplyr::all_of(disease_cols),
                       names_to = "disease", values_to = "cell_rss") %>%
  dplyr::filter(!is.na(cell_rss), cell_rss != "") %>%
  dplyr::mutate(
    CellType = stringr::str_extract(cell_rss, "^[^:]+"),
    RSS      = as.numeric(stringr::str_extract(cell_rss, "[0-9.]+$"))
  ) %>%
  dplyr::select(TF, disease, CellType, RSS) %>%
  # FIX (must stay - this is the pulmonary-diseases-missing bug):
  # relabel to match the short disease codes used everywhere else
  # (regulon_long, tf_gene_overlap). Without this, the plot's final
  # left_join silently fails for IPF.U/IPF.L/SScILD.U/SScILD.L,
  # producing NA RSS values that ggplot then drops as invisible points
  dplyr::mutate(disease = dplyr::case_when(
    disease == "IPF.Upper"    ~ "IPF.U",
    disease == "IPF.Low"      ~ "IPF.L",
    disease == "SScILD.Upper" ~ "SScILD.U",
    disease == "SScILD.Low"   ~ "SScILD.L",
    TRUE ~ disease
  ))

dominant_celltype_per_TF <- celltype_long %>%
  dplyr::group_by(TF) %>%
  dplyr::slice_max(order_by = RSS, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::select(TF, dominant_celltype = CellType, dominant_RSS = RSS)
cat("dominant cell type recovered for", nrow(dominant_celltype_per_TF), "TFs\n")

stopifnot("regulon_long not found - run the loader first" = exists("regulon_long"))

#1. TLS marker groups
germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
                      "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")
inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
                   "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
                   "ICAM1", "VCAM1", "SELE")
lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12",
                          "CXCR5", "CCR7", "CXCR4")
hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")
fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
         "CXCL13", "CD54", "ICAM1", "FCER2")

marker_groups <- list(
  germinal_center      = germinal_center,
  inflammation         = inflammation,
  lymphoid_chemokines  = lymphoid_chemokines,
  hev                  = hev,
  fdc                  = fdc
)

tls_genes_long <- purrr::imap_dfr(
  marker_groups,
  ~ data.frame(category = .y, gene = .x)
)

#2. restrict the regulon data to the shared/core TFs only
regulon_core <- regulon_long %>%
  dplyr::filter(TF %in% core_tfs)

#3. overlap: which TLS genes are targets of a shared TF, and in which disease(s)
tf_gene_overlap <- regulon_core %>%
  dplyr::inner_join(tls_genes_long, by = c("target" = "gene"),
                     relationship = "many-to-many") %>%
  dplyr::distinct(TF, category, target, disease) %>%
  dplyr::arrange(category, TF, target) %>%
  dplyr::left_join(dominant_celltype_per_TF, by = "TF")

cat("Rows per disease (sanity check - should be non-zero for all 9):\n")
print(tf_gene_overlap %>% dplyr::count(disease))

#4. count, per TF, how many distinct TLS genes it regulates
tf_tls_target_counts <- tf_gene_overlap %>%
  dplyr::distinct(TF, target) %>%
  dplyr::count(TF, name = "n_tls_targets")

#5. top 10 TFs by number of distinct TLS targets
top_n_tf <- 10
top10_tfs <- tf_tls_target_counts %>%
  dplyr::arrange(dplyr::desc(n_tls_targets)) %>%
  dplyr::slice_head(n = top_n_tf) %>%
  dplyr::pull(TF)
cat("\nTop", top_n_tf, "TFs by number of TLS targets:\n")
print(top10_tfs)

#6. for these 10 TFs, which TLS genes do they regulate
top10_tls_overlap <- tf_gene_overlap %>%
  dplyr::filter(TF %in% top10_tfs) %>%
  dplyr::arrange(category, TF, target)

#7. top 5 genes per category, by how many of these 10 TFs regulate each gene
top_n_gene <- 5
gene_tf_counts <- top10_tls_overlap %>%
  dplyr::distinct(TF, category, target) %>%
  dplyr::count(category, target, name = "n_tfs_targeting")

top5_genes_per_category <- gene_tf_counts %>%
  dplyr::group_by(category) %>%
  dplyr::slice_max(order_by = n_tfs_targeting, n = top_n_gene, with_ties = FALSE) %>%
  dplyr::ungroup()
cat("\nTop", top_n_gene, "genes per category:\n")
print(top5_genes_per_category)

#8. final plotting table
plot_data <- top10_tls_overlap %>%
  dplyr::inner_join(top5_genes_per_category %>% dplyr::select(category, target),
                     by = c("category", "target")) %>%
  dplyr::distinct(TF, category, target, disease) %>%
  dplyr::left_join(celltype_long %>% dplyr::select(TF, disease, CellType, RSS),
                    by = c("TF", "disease")) %>%
  dplyr::rename(RSS_in_disease = RSS, CellType_in_disease = CellType)

cat("\nRows per disease in plot_data (should be non-zero for all 9, including pulmonary ones):\n")
print(plot_data %>% dplyr::count(disease))

plot_data$TF <- factor(plot_data$TF, levels = top10_tfs)

disease_levels <- c("MS", "RA", "SS", "T1D", "SjD",
                     "IPF.U", "IPF.L", "SScILD.U", "SScILD.L")
plot_data$disease <- factor(plot_data$disease, levels = disease_levels)
plot_data$RSS_in_disease <- pmin(pmax(plot_data$RSS_in_disease, 0), 1)

#9. plot
celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

dummy_celltype <- data.frame(
  TF                  = plot_data$TF[1],
  target              = plot_data$target[1],
  category            = plot_data$category[1],
  disease             = plot_data$disease[1],
  CellType_in_disease = factor(names(celltype_cols), levels = names(celltype_cols))
)

p_final <- ggplot(plot_data, aes(x = TF, y = target)) +
  geom_point(aes(size = RSS_in_disease, color = CellType_in_disease),
             show.legend = c(size = TRUE, colour = FALSE)) +
  geom_point(data = dummy_celltype,
             aes(x = TF, y = target, color = CellType_in_disease),
             size = 5, alpha = 0, inherit.aes = FALSE) +
  facet_grid(category ~ disease, scales = "free_y", space = "free_y", switch = "y") +
  scale_color_manual(values = celltype_cols, limits = names(celltype_cols), drop = FALSE) +
  scale_size_continuous(
    name   = "RSS",
    range  = c(1.5, 6),
    breaks = c(0.2, 0.4, 0.6, 0.8, 1.0),
    limits = c(0, 1)
  ) +
  guides(
    size  = guide_legend(order = 1),
    color = guide_legend(order = 2, override.aes = list(alpha = 1, size = 5))
  ) +
  labs(x = "Transcription Factor", y = "Target Gene", color = "Cell type") +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
    panel.grid.minor = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(size = 14, face = "bold"),
    strip.text.y.left = element_text(angle = 0, size = 14, face = "bold")
  )

ggsave("GRN_SCENIC/TF_TLS_top10_by_targetcount_plot.pdf", p_final, width = 25, height = 12)
cat("\nplot saved to GRN_SCENIC/TF_TLS_top10_by_targetcount_plot.pdf\n")

```





##11.2 heatmap of shared regulons pairwise


```{r}

library(pheatmap)
library(dplyr)



tf_list <- lapply(rss_list, function(df) setdiff(names(df), "CellType"))

# rename to short .U / .L naming
names(tf_list)[names(tf_list) == "IPF.Upper"]    <- "IPF.U"
names(tf_list)[names(tf_list) == "IPF.Low"]      <- "IPF.L"
names(tf_list)[names(tf_list) == "SScILD.Upper"] <- "SScILD.U"
names(tf_list)[names(tf_list) == "SScILD.Low"]   <- "SScILD.L"

cat("TFs per disease:\n")
print(sapply(tf_list, length))


#11.2 heatmap of shared regulons pairwise

diseases <- names(tf_list)
n <- length(diseases)

pairwise_mat <- matrix(0, nrow = n, ncol = n, dimnames = list(diseases, diseases))

for (i in 1:n) {
  for (j in 1:n) {
    pairwise_mat[i, j] <- length(intersect(tf_list[[diseases[i]]], tf_list[[diseases[j]]]))
  }
}

write.csv(pairwise_mat, "output/Fusion/pairwise_shared_regulons_matrix.csv")

pdf("output/Fusion/pairwise_shared_regulons.pdf", width = 13, height = 12)
pheatmap(
  pairwise_mat,
  display_numbers = TRUE,
  number_format    = "%d",
  fontsize_number  = 16,
  fontsize         = 13,
  fontsize_row     = 13,
  fontsize_col     = 13,
  color            = colorRampPalette(c("white", "red3", "red"))(100),
  border_color     = "grey60",
  cellwidth        = 45,
  cellheight       = 45,
  angle_col        = "90"
)
dev.off()

cat("Done!\n")




```






##11.3upset plot with all intersect combinations
```{r}

library(UpSetR)


#1.3 upset plot with all intersect combinations

pdf("output/Fusion/upset_regulons_combinations.pdf", 
    width = 14, height = 12,
    onefile = FALSE)

upset(
  fromList(tf_list),
  nsets             = length(tf_list),
  nintersects       = 30,
  order.by          = "freq",
  mb.ratio          = c(0.5, 0.5),     # more room for set size bars on the left
  text.scale        = c(1.8, 1.4, 1.4, 1.2, 1.5, 1.3),
  point.size        = 3.5,
  line.size         = 1.2,
  main.bar.color    = "#993404",
  sets.bar.color    = "#2c7fb8",
  matrix.color      = "black",
  shade.color       = "grey85",
  set_size.show     = TRUE,
  set_size.scale_max = max(sapply(tf_list, length)) * 1.15   # add headroom so numbers aren't cut off
)

dev.off()





```



##12. TLS GLObal plots


```{r}
#without filtering.

library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(grid)
library(dplyr)

#1.load all seurat obj and put in a list
seurat_list <- list(
  MS = MS, RA = RA, SS = SS, T1D = T1D,
  IPF.U = IPF.U, IPF.L = IPF.L, SjD = SjD,
  SScILD.L = SScILD.L, SScILD.U = SScILD.U
)


#2.define marker groups

germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
                      "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")
inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
                   "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
                   "ICAM1", "VCAM1", "SELE")
lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12",
                          "CXCR5", "CCR7", "CXCR4")
hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")
fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
         "CXCL13", "CD54", "ICAM1", "FCER2")

markers_all <- unique(c(germinal_center, inflammation,
                         lymphoid_chemokines, hev, fdc))

marker_groups_all <- case_when(
  markers_all %in% germinal_center     ~ "Germinal Center",
  markers_all %in% inflammation        ~ "Inflammation & Cytokines",
  markers_all %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers_all %in% hev                 ~ "HEV",
  markers_all %in% fdc                 ~ "FDC"
)
names(marker_groups_all) <- markers_all

group_cols <- c(
  "Germinal Center"          = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines"      = "#35978F",
  "HEV"                      = "#8073AC",
  "FDC"                      = "#E08214"
)


#3.color for celltype

all_celltypes <- sort(unique(unlist(lapply(seurat_list, function(o) as.character(unique(o$General))))))
print(all_celltypes)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)

# cell types to exclude from being picked as the dominant cell type
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells","Melanocytes","Keratinocytes","Endocrine Cells","Smooth Muscle Cells","Plasma Cells")


#4.dominant celltypes (excluding the unwanted cell types BEFORE picking the max)

get_disease_dominance <- function(obj, disease_name, markers, exclude = excluded_celltypes) {
  Idents(obj) <- "General"
  DefaultAssay(obj) <- "RNA"
  present <- intersect(markers, rownames(obj))
  if (length(present) == 0) return(NULL)

  expr_mat <- GetAssayData(obj, layer = "data")

  percent_mat <- sapply(levels(Idents(obj)), function(ct) {
    cells <- WhichCells(obj, idents = ct)
    if (length(cells) == 0) return(rep(0, length(present)))
    Matrix::rowMeans(expr_mat[present, cells, drop = FALSE] > 0) * 100
  })
  percent_mat <- as.matrix(percent_mat)
  rownames(percent_mat) <- present

  avg_exp <- AverageExpression(obj, features = present, assays = "RNA",
                                layer = "data", return.seurat = FALSE)$RNA
  exp_mat <- as.matrix(avg_exp[present, colnames(percent_mat), drop = FALSE])
  exp_mat[is.infinite(exp_mat) | is.na(exp_mat)] <- 0
  mode(exp_mat) <- "numeric"

  # NOTE: no z-score scaling here - using raw log1p-transformed average
  # expression directly to determine the dominant cell type
  exp_scaled <- log1p(exp_mat)
  exp_scaled[is.na(exp_scaled)] <- 0

  # remove excluded cell types from consideration BEFORE picking the dominant one
  keep_cols <- setdiff(colnames(exp_scaled), exclude)
  exp_scaled  <- exp_scaled[, keep_cols, drop = FALSE]
  percent_mat <- percent_mat[, keep_cols, drop = FALSE]

  dom_idx <- apply(exp_scaled, 1, which.max)
  dom_ct  <- colnames(exp_scaled)[dom_idx]
  dom_pct <- percent_mat[cbind(seq_len(nrow(percent_mat)), dom_idx)]

  data.frame(
    marker = present,
    disease = disease_name,
    dominant_celltype = dom_ct,
    pct_exp = dom_pct,
    stringsAsFactors = FALSE
  )
}


#5.loop over disease

results <- bind_rows(lapply(names(seurat_list), function(d) {
  get_disease_dominance(seurat_list[[d]], d, markers_all)
}))


#6.marker and disease matrix

diseases <- names(seurat_list)
markers_present <- markers_all[markers_all %in% unique(results$marker)]

dom_ct_mat <- matrix(NA, nrow = length(markers_present), ncol = length(diseases),
                      dimnames = list(markers_present, diseases))
pct_mat <- matrix(0, nrow = length(markers_present), ncol = length(diseases),
                   dimnames = list(markers_present, diseases))

for (i in seq_len(nrow(results))) {
  r <- results[i, ]
  dom_ct_mat[r$marker, r$disease] <- r$dominant_celltype
  pct_mat[r$marker, r$disease]    <- r$pct_exp
}

marker_groups_final <- marker_groups_all[markers_present]

cat("Total markers plotted (no filtering applied):", length(markers_present), "\n")


#7.bubble drawing function - solid color dots, no black border
# NOTE: no pct_exp filtering step here - every marker in markers_all is
# kept and plotted, regardless of how low its expression is

cell_fun <- function(j, i, x, y, w, h, fill) {
  ct <- dom_ct_mat[i, j]
  if (is.na(ct)) return(invisible(NULL))
  grid::grid.circle(
    x = x, y = y,
    r = (pct_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = celltype_cols[ct], col = celltype_cols[ct])
  )
}


#8.marker annotation

row_ha <- rowAnnotation(
  Group = marker_groups_final,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(title = "TLS genes",
                 title_gp = gpar(fontsize = 10, fontface = "bold"),
                 labels_gp = gpar(fontsize = 9))
  )
)


#9.heatmap - clustering enabled (rows cluster WITHIN each TLS group, columns cluster too)

ht <- Heatmap(
  pct_mat,
  name = "blank",
  col = colorRamp2(c(0, 100), c("white", "white")),
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = gpar(fontsize = 15, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  left_annotation = row_ha,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",
  show_row_dend = FALSE,
  show_column_dend = TRUE,
  row_split = marker_groups_final,
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)


#10.legend

lgd_celltype <- Legend(
  labels = names(celltype_cols)[!(names(celltype_cols) %in% excluded_celltypes)],
  legend_gp = gpar(fill = celltype_cols[!(names(celltype_cols) %in% excluded_celltypes)]),
  title = "Celltypes",
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

# size legend - full 0-100% range since no floor/threshold is applied
pct_breaks <- c(0, 25, 50, 75, 100)
lgd_size <- Legend(
  labels = paste0(pct_breaks, "%"),
  title = "% of cells expressing gene",
  type = "points",
  pch = 16,
  size = unit((pct_breaks / 100) * 0.45 * 4, "mm"),
  legend_gp = gpar(col = "grey40", fill = "grey70"),
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size))

pdf("output/Fusion/TLS_plot_nofilter.pdf", width = 10, height = 12)
draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size))
dev.off()




```





```{r}
##with filter


library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(grid)
library(dplyr)

#1.load all seurat obj and put in a list
seurat_list <- list(
  MS = MS, RA = RA, SS = SS, T1D = T1D,
  IPF.U = IPF.U, IPF.L = IPF.L, SjD = SjD,
  SScILD.L = SScILD.L, SScILD.U = SScILD.U
)


#2.define marker groups

germinal_center <- c("AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
                      "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5")
inflammation <- c("TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
                   "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
                   "ICAM1", "VCAM1", "SELE")
lymphoid_chemokines <- c("CXCL13", "CCL21", "CCL19", "CXCL12",
                          "CXCR5", "CCR7", "CXCR4")
hev <- c("NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
         "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C")
fdc <- c("CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
         "CXCL13", "CD54", "ICAM1", "FCER2")

markers_all <- unique(c(germinal_center, inflammation,
                         lymphoid_chemokines, hev, fdc))

marker_groups_all <- case_when(
  markers_all %in% germinal_center     ~ "Germinal Center",
  markers_all %in% inflammation        ~ "Inflammation & Cytokines",
  markers_all %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers_all %in% hev                 ~ "HEV",
  markers_all %in% fdc                 ~ "FDC"
)
names(marker_groups_all) <- markers_all

group_cols <- c(
  "Germinal Center"          = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines"      = "#35978F",
  "HEV"                      = "#8073AC",
  "FDC"                      = "#E08214"
)


#3.color for celltype

all_celltypes <- sort(unique(unlist(lapply(seurat_list, function(o) as.character(unique(o$General))))))
print(all_celltypes)

celltype_cols <- c(
  "B Cells"             = "gold",
  "CD4"                 = "olivedrab2",
  "CD8"                 = "tomato",
  "NK Cells"            = "magenta",
  "Mono/Mac"            = "darkblue",
  "Mast Cells"          = "#FDBF6F",
  "Proliferating Cells" = "#B15928",
  "Endothelial Cells"   = "#E31A1C",
  "Fibroblasts"         = "cyan",
  "Mural Cells"         = "#1F78B4",
  "Epithelial Cells"    = "mediumpurple2"
)


# cell types to exclude from being picked as the dominant cell type
excluded_celltypes <- c("Glial Cells", "Neurons", "Immune Cells","Melanocytes","Keratinocytes","Endocrine Cells","Smooth Muscle Cells","Plasma Cells")


#4.dominant celltypes 


get_disease_dominance <- function(obj, disease_name, markers, exclude = excluded_celltypes) {
  Idents(obj) <- "General"
  DefaultAssay(obj) <- "RNA"
  present <- intersect(markers, rownames(obj))
  if (length(present) == 0) return(NULL)

  expr_mat <- GetAssayData(obj, layer = "data")

  percent_mat <- sapply(levels(Idents(obj)), function(ct) {
    cells <- WhichCells(obj, idents = ct)
    if (length(cells) == 0) return(rep(0, length(present)))
    Matrix::rowMeans(expr_mat[present, cells, drop = FALSE] > 0) * 100
  })
  percent_mat <- as.matrix(percent_mat)
  rownames(percent_mat) <- present

  avg_exp <- AverageExpression(obj, features = present, assays = "RNA",
                                layer = "data", return.seurat = FALSE)$RNA
  exp_mat <- as.matrix(avg_exp[present, colnames(percent_mat), drop = FALSE])
  exp_mat[is.infinite(exp_mat) | is.na(exp_mat)] <- 0
  mode(exp_mat) <- "numeric"

  # z-score scaling per gene, across cell types (fair comparison between genes)
  exp_scaled <- log1p(exp_mat)
  exp_scaled <- t(scale(t(exp_scaled)))
  exp_scaled[is.na(exp_scaled)] <- 0

  # remove excluded cell types from consideration BEFORE picking the dominant one
  keep_cols <- setdiff(colnames(exp_scaled), exclude)
  exp_scaled  <- exp_scaled[, keep_cols, drop = FALSE]
  percent_mat <- percent_mat[, keep_cols, drop = FALSE]

  dom_idx <- apply(exp_scaled, 1, which.max)
  dom_ct  <- colnames(exp_scaled)[dom_idx]
  dom_pct <- percent_mat[cbind(seq_len(nrow(percent_mat)), dom_idx)]
  dom_exp <- exp_scaled[cbind(seq_len(nrow(exp_scaled)), dom_idx)]  # scaled expression at the winning cell type

  data.frame(
    marker = present,
    disease = disease_name,
    dominant_celltype = dom_ct,
    pct_exp = dom_pct,
    dom_exp = dom_exp,
    stringsAsFactors = FALSE
  )
}


#5.loop over disease

results <- bind_rows(lapply(names(seurat_list), function(d) {
  get_disease_dominance(seurat_list[[d]], d, markers_all)
}))


#6.marker and disease matrix

diseases <- names(seurat_list)
markers_present <- markers_all[markers_all %in% unique(results$marker)]

dom_ct_mat <- matrix(NA, nrow = length(markers_present), ncol = length(diseases),
                      dimnames = list(markers_present, diseases))
pct_mat <- matrix(0, nrow = length(markers_present), ncol = length(diseases),
                   dimnames = list(markers_present, diseases))
dom_exp_mat <- matrix(NA, nrow = length(markers_present), ncol = length(diseases),
                       dimnames = list(markers_present, diseases))

for (i in seq_len(nrow(results))) {
  r <- results[i, ]
  dom_ct_mat[r$marker, r$disease]  <- r$dominant_celltype
  pct_mat[r$marker, r$disease]     <- r$pct_exp
  dom_exp_mat[r$marker, r$disease] <- r$dom_exp
}

marker_groups_final <- marker_groups_all[markers_present]


#7.filter out markers with negligible expression everywhere

pct_threshold <- 10   # adjust if needed

keep_markers <- rownames(pct_mat)[apply(pct_mat, 1, function(x) any(x >= pct_threshold, na.rm = TRUE))]

dom_ct_mat   <- dom_ct_mat[keep_markers, , drop = FALSE]
pct_mat      <- pct_mat[keep_markers, , drop = FALSE]
dom_exp_mat  <- dom_exp_mat[keep_markers, , drop = FALSE]
marker_groups_final <- marker_groups_final[keep_markers]
markers_present <- keep_markers

cat("Markers kept after filtering (pct_exp >=", pct_threshold,
    "% in at least one disease):", length(keep_markers), "\n")


#8.row-wise (per-marker) z-score ACROSS DISEASES


row_scaled_mat <- t(scale(t(dom_exp_mat)))
row_scaled_mat[is.na(row_scaled_mat)] <- 0

# normalize to 0-1 for dot size (min-max per the whole matrix, with a
# floor so even below-average diseases still show a small visible dot)
min_val <- min(row_scaled_mat, na.rm = TRUE)
max_val <- max(row_scaled_mat, na.rm = TRUE)
size_mat <- 0.15 + 0.85 * (row_scaled_mat - min_val) / (max_val - min_val)


#9.bubble drawing function

cell_fun <- function(j, i, x, y, w, h, fill) {
  ct <- dom_ct_mat[i, j]
  if (is.na(ct)) return(invisible(NULL))
  grid::grid.circle(
    x = x, y = y,
    r = size_mat[i, j] * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = celltype_cols[ct], col = celltype_cols[ct])
  )
}


#10.marker annotation

row_ha <- rowAnnotation(
  Group = marker_groups_final,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(title = "TLS genes",
                 title_gp = gpar(fontsize = 10, fontface = "bold"),
                 labels_gp = gpar(fontsize = 9))
  )
)


#11.heatmap 

ht <- Heatmap(
  pct_mat,
  name = "blank",
  col = colorRamp2(c(0, 100), c("white", "white")),
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  show_heatmap_legend = FALSE,
  row_names_gp = gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = gpar(fontsize = 15, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  left_annotation = row_ha,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",
  show_row_dend = FALSE,
  show_column_dend = TRUE,
  row_split = marker_groups_final,
  row_title_gp = gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)


#12.legend

lgd_celltype <- Legend(
  labels = names(celltype_cols)[!(names(celltype_cols) %in% excluded_celltypes)],
  legend_gp = gpar(fill = celltype_cols[!(names(celltype_cols) %in% excluded_celltypes)]),
  title = "Celltypes",
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

# size legend - now represents relative (row-scaled) expression across
# diseases, not raw pct_exp - so breaks are shown in SD units
size_breaks_sd <- c(-1.5, -0.5, 0.5, 1.5, 2.5)
size_breaks_norm <- 0.15 + 0.85 * (size_breaks_sd - min_val) / (max_val - min_val)
size_breaks_norm <- pmax(0.15, pmin(1, size_breaks_norm))

lgd_size <- Legend(
  labels = paste0(size_breaks_sd, " SD"),
  title = "Relative expression\n(z-score across diseases)",
  type = "points",
  pch = 16,
  size = unit(size_breaks_norm * 0.45 * 4, "mm"),
  legend_gp = gpar(col = "grey40", fill = "grey70"),
  title_gp = gpar(fontsize = 10, fontface = "bold"),
  labels_gp = gpar(fontsize = 9)
)

draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size))

pdf("output/Fusion/TLS_plot_filtered.pdf", width = 10, height = 12)
draw(ht, annotation_legend_list = list(lgd_celltype, lgd_size))
dev.off()



```




```{r}

#--------save
saveRDS(Fu, file = "data/process/SeuratObjects/Fusion/F_process.rds")
#load-----later
Fu=readRDS("data/process/SeuratObjects/Fusion/F_process.rds")

```
















