---
title: "R Notebook"
output: html_notebook

---
##############################################################################################################################
###ANALYSIS OF SYSTEMATIC SCLEROSIS DATA######################################################################################
###############################################################################################################################

###LOAD PACKAGES#####
```{r setup, include=FALSE}
##################PACKAGES
library(pacman)
  p_load(Seurat, SeuratData, Matrix, dplyr, DropletUtils, presto,gprofiler2, ggplot2, patchwork, gplots, cowplot, openxlsx,plyr, scDblFinder, BiocParallel,devtools,openxlsx,reticulate,hdf5r,DropletUtils)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")

```



#####STEP_1
###LOAD the data
```{r}
##load data using a function for all 12obj, create seurat_obj,merge to one and save it without any process
#folder
data.dir="../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Systemic Sclerosis/GSE138669_RAW/"
# List all .h5 files
h5.files=list.files( path = data.dir,pattern = "_feature_bc_matrix\\.h5$",full.names = TRUE)

# Assign sample names SS1–SS12
sample.names=paste0("SS", seq_along(h5.files))

 # Loop: read, filter, create Seurat object, save
for (i in seq_along(h5.files)) {
  message("Processing ", sample.names[i])
 # Read raw counts
  SS.x=Read10X_h5(h5.files[i], use.names = TRUE, unique.features = TRUE) 
# defaultDrops
  iscell_dd <- defaultDrops(SS.x, expected = 5000)
 # emptyDrops
  eout=emptyDrops(SS.x, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed=eout$FDR <= 0.01
  # metadata
  meta=data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    row.names = colnames(SS.x)
  )
  # keep cells
  keep=iscell_dd | iscell_ed
  # Create Seurat object
  SS=CreateSeuratObject(
    counts = SS.x[, keep],
    meta.data = meta[keep, ]
  )
  # Add sample name inside Seurat object
  SS$sample=sample.names[i]   
  
# Save Seurat object
  saveRDS(
    SS,
    file = paste0("data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/", sample.names[i], ".rds")
  )
}

##merge in one seurat obj
# Folder with Seurat objects
seurat.dir="data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/"
rds.files=list.files(seurat.dir,pattern = "^SS[0-9]+\\.rds$",full.names = TRUE)
# Read them into a list
SS_list=lapply(rds.files, readRDS)
# Name the list (SS1–SS12)
names(SS_list)=sub("\\.rds$", "", basename(rds.files))

# Merge
SS_obj=merge(
  x = SS_list[[1]],
  y = SS_list[-1],
  add.cell.ids = names(SS_list),
  project = "Systemic_Sclerosis"
)
saveRDS(SS_obj,file = "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_raw.rds")
rm(eout,meta,SS,SS_list,SS.x,h5.files,i,iscell_ed,iscell_dd,keep,rds.files,sample.names)
gc(full = TRUE)
```

###STEP_2:process and analyse

###1_the qc
```{r, fig.width=10,fig.height=8}
##qc
SS_obj[["percent.mt"]]=PercentageFeatureSet(SS_obj,pattern = "^MT-")
VlnPlot(SS_obj,features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3,log = TRUE)
SS_obj=subset(SS_obj,subset=nFeature_RNA>200 & nFeature_RNA<6000 & percent.mt<15)

```


###2-statistics
```{r}
##norm, fvf,scale,runpca
SS_obj=NormalizeData(SS_obj)
SS_obj=FindVariableFeatures(SS_obj)
SS_obj=ScaleData(SS_obj,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = SS_obj))
SS_obj=RunPCA(SS_obj, features = VariableFeatures(object = SS_obj))
ElbowPlot(SS_obj, ndims = 50, reduction = "pca")
SS_obj=RunUMAP(SS_obj, dims = 1:30)
DimPlot(SS_obj,reduction = "umap", label = TRUE)

```


###3-integration
```{r}
##integration
library(harmony)
SS_obj =RunHarmony(SS_obj,group.by.vars = "sample",orig.reduction = "pca",dims.use = 1:30  )
SS_obj[["RNA"]]=JoinLayers(SS_obj[["RNA"]])
SS_obj = RunUMAP(SS_obj, reduction = "harmony", dims= 1:30, reduction.name = "umap.harmony")
p=DimPlot(SS_obj,reduction = "umap.harmony",group.by = c("sample"),label.size = 2, shuffle = T)
ggsave(filename = "output/Systemic_Sclerosis_in_Skin/SS.integration.png", plot=p, width = 12, height = 8)


#for the resolution
SS_obj=FindNeighbors(SS_obj,reduction= "harmony",dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  SS_obj=FindClusters(SS_obj, resolution = res)
  colname=paste0("res_", res)
  SS_obj[[colname]]=SS_obj$seurat_clusters
}
res_cols=grep("^res_", colnames(SS_obj@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(SS_obj, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}
#save 
p=DimPlot(SS_obj,group.by = "res_0.3",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.3")
ggsave(filename = "output/Systemic_Sclerosis_in_Skin/SSresolution.png",plot = p,width = 12,height = 8,dpi = 300)
```


###4-clusters and markers
```{r}
###clusters

SS_obj=FindNeighbors(SS_obj,reduction= "harmony",dims = 1:30)
SS_obj=FindClusters(SS_obj, resolution = 0.3,cluster.name = "harmony_clusters")
#markers
SS_markers=FindAllMarkers(SS_obj,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
SS_markers=SS_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
write.xlsx(SS_markers,file = "output/Systemic_Sclerosis_in_Skin/SS_markers.xlsx")
View(SS_markers)


```







###5-Annotation

  | Cell type                              | Marker genes                        |
| -------------------------------------- | ----------------------------------- |
| **Basal keratinocytes**                | KRT1, KRT14, KRT5, KRT17            |
| **Keratinocytes (cornified envelope)** | LOR, SPRR2G                         |
| **Fibroblasts**                        | SFRP2, COMP, APOE                   |
| **Leukocytes (T / NK)**                | CD3D, CD4, CD8, GNLY                |
| **Macrophages**                        | CCL3, CCL19, APOC1                  |
| **Mast cells**                         | TPSAB1, FTH1                        |
| **Plasma / B cells**                   | IGHG3, MS4A1                        |
| **Pericytes**                          | RGS5, STEAP4, RERGL, CXCR4          |
| **Smooth muscle cells**                | DES                                 |
| **Neural cells**                       | PLP1, PMP22                         |
| **Schwann cells**                      | MPZ, MBP                            |
| **Proliferating cells**                | MKI67                               |
| **Melanocytes**                        | TYRP1, PMEL                         |
| **Preadipocytes**                      | APOC1, APMAP                        |
| **Endothelial cells**                  | VWF, RAMP2 *(AQP1, IFI27 optional)* |
| **Secretory cells – CA6⁺**             | KRT19, SNORC                        |
| **Secretory cells – SCGB2A2⁺**         | SCGB2A2, MUCL1                      |
| **Secretory cells – SCGB1D2⁺**         | CALML5                              |
                       

```{r, fig.width=10,fig.height=8}
##based on markers
genes = c("KRT1","KRT14","KRT5","KRT17","RGS5","KRT6A","KRT6B","KRT2","SFRP2","COMP","APOD","HLA-DRA","APOE","POSTN","ANGPTL7","COCH","ASPN","CD3D","GNLY","CD4","CD1C","CCL3","CCL19","TPSAB1","FTH1","IGHG3","MS4A1","STEAP4","GEM","TAGLN","RERGL","CXCR4","DES","PLP1","PMP22","MKI67","TYRP1","PMEL","APOC1","APMAP","LOR","SPRR2G","MPZ","MBP","VWF","RAMP2","CCL21","CLDN5","KRT19","SNORC","CALML5")
  
##1celltype
Idents(SS_obj)="harmony_clusters"
cluster_name=c(
  "0"  = "Terminal Keratinocytes" ,
  "1"  = "Fibroblasts"    ,
  "2"  = "Basal Keratinocytes"            ,
  "3"  = "Endothelial cells"      ,
  "4"  = "Basal Keratinocytes"    ,
  "5"  = "Pericytes"            ,
  "6"  = "Macrophages/DC"          ,
  "7"  = "NK/T cells"               ,
  "8"  = "Pericytes"                ,
  "9"  = "Fibroblasts"              ,
  "10" = "Basal Keratinocytes"        ,
  "11" = "Secretory cells"           ,
  "12" = "Smooth Muscle cells"    ,
  "13" = "Neural cells"           ,
  "14" = "Mast cells"             ,
  "15" = "Keratinocytes"            ,
  "16" = "Melanocytes"          ,         
  "17" ="Endothelial cells"    
)

SS_obj=RenameIdents(SS_obj, cluster_name)
SS_obj$celltype=Idents(SS_obj)
##2color
celltype_cols=c(
  "Terminal Keratinocytes"  = "#D73027",  
  "Fibroblasts"             = "#4575B4", 
  "Keratinocytes"           = "#FDAE61",  
  "Endothelial cells"       = "#1B9E77", 
  "Basal Keratinocytes"     = "#66A61E", 
  "NK/T cells"              = "#000000",  
  "Pericytes"               = "#762A83", 
  "Smooth Muscle cells"     = "#74ADD1",  
  "Melanocytes"             = "#A6761D",  
  "Secretory cells"         = "#E78AC3",  
  "Neural cells"            = "#E69F00",
  "Mast cells"              = "#E7298A"
  )

##3 center
umap_df=as.data.frame(Embeddings(SS_obj, "umap"))
colnames(umap_df) <- c("umap_1", "umap_2")
umap_df$celltype <- SS_obj$celltype
centers <- aggregate(
  cbind(umap_1, umap_2) ~ celltype,
  data = umap_df,
  FUN = median
)

##supp duplicate and plot
library(ggrepel)
suppressWarnings({
  fig1a=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = umap_1, y = umap_2, label = celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 3,
      box.padding = 0.4,
      label.size = 0.25,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "umap 1", y = "umap 2")
})
fig1a
ggsave(filename = "output/Systemic_Sclerosis_in_Skin/SScellT.png", plot = fig1a, width = 12, height = 8)

```



```{r,fig.width=13,fig.height=8}
##with markers 
SS_obj$celltype = Idents(SS_obj)
p=DotPlot(
  SS_obj,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = T, 
  group.by = "celltype") + RotatedAxis() 
p
ggsave(filename = "output/Systemic_Sclerosis_in_Skin/SSmarkers.png", plot = p, width = 14, height = 8)
```




###6-The proportion of cells
```{r}
library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(SS_obj@meta.data) %>%
  dplyr::count(`sample`, celltype) %>%
  dplyr::group_by(`sample`) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
SS_obj_meta <- dplyr::as_tibble(SS_obj@meta.data) %>%
  dplyr::rename(sample_id = sample)   # rename to avoid conflicts
prop_df=SS_obj_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/Systemic_Sclerosis_in_Skin/SSproportion.png", plot = fig1c, width = 12, height = 10)
```




###7-proportion of Donors
```{r}
library(dplyr)
library(ggplot2)
library(scales)

#change colname 
SS_obj$Donors = SS_obj$sample

#create the meta data
meta = SS_obj@meta.data[, c("harmony_clusters", "Donors")]
table(meta$Donors)

# Make sure dplyr functions are used explicitly
meta = SS_obj@meta.data

#the  percentage 
donors_counts = meta %>%
  as_tibble() %>%
  dplyr::group_by(harmony_clusters,Donors) %>% #count cells each donor use in each cluster
  dplyr::summarise( n= dplyr::n(),.groups = "drop") %>% #numb of cells
  dplyr::group_by(harmony_clusters) %>%
  dplyr::mutate(proportion = n /sum(n))  #fraction of cells per donor
```



```{r}

###plot
p = ggplot(donors_counts, aes(
  x = harmony_clusters,
  y = proportion,
  group = Donors,
  color = Donors
)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_viridis_d(option = "turbo", name = "Donor") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Donors Proportion",
    x = "Cluster",
    y = "Proportion (%)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
p

ggsave(filename = "output/Systemic_Sclerosis_in_Skin/SSDonor.png", plot = p, width = 12, height = 8)

```



```{r}
##correlation with the cells
# 1️⃣ Set identities to cell types
Idents(SS_obj) <- "celltype"

# sanity check
levels(Idents(SS_obj))
# should show cell type names, NOT numbers

# 2️⃣ Average expression per cell type
av.expSS <- AverageExpression(
  SS_obj,
  assays = "RNA",
  slot = "data"
)$RNA
# rows = genes, columns = cell types

# 3️⃣ Cell-type correlation matrix
correlation_matrixSS <- cor(
  as.matrix(av.expSS),
  use = "pairwise.complete.obs"
)

# safety cleanup
correlation_matrixSS[is.na(correlation_matrixSS)]=0

# 4️⃣ Save heatmap
png(
  filename = "output/Systemic_Sclerosis_in_Skin/SScellcor.png",
  width = 3000,
  height = 2400,
  res = 300
)

pheatmap(
  correlation_matrixSS,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)

dev.off()
```









###save the final meta.data
```{r}
##save
# Keep your existing metadata names
SS_obj$orig.ident="Systemic Sclerosis in Skin"

# Remove unwanted columns 
cols_to_remove=c(
  grep("res_|RNA_snn_res", colnames(SS_obj@meta.data), value = TRUE),
  "cell_type",
  "iscell_ed",
  "iscell_dd",
  "Donors"
)
SS_obj@meta.data=SS_obj@meta.data[, !(colnames(SS_obj@meta.data) %in% cols_to_remove)]

# Ensure active identity reflects your celltype
Idents(SS_obj)=SS_obj$celltype

# Save 
saveRDS(SS_obj, file = "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_Final.rds")

# Save only the final metadata 
saveRDS(SS_obj@meta.data, file = "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_obj.metadata.rds")

# Optional CSV 
write.csv(SS_obj@meta.data, 
          file = "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_obj.metadata.csv",
          row.names = TRUE)
View(SS_obj@meta.data)

```





##save
```{r}
##save
#saveRDS(SS_obj, file = "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_process.rds")

## run it later to load all 
SS_obj = readRDS("data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_process.rds")
```
























