---
title: "Systemic Sclerosis (Skin) - Processing and Annotation"
output: html_notebook
---



## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

## 2. Load raw data and create per-sample Seurat objects

```{r}
# DropletUtils provides defaultDrops() and emptyDrops() used below
library(DropletUtils)

# raw 10x .h5 files, one per sample
data.dir <- "../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Systemic Sclerosis/GSE138669_RAW/"
h5.files <- list.files(path = data.dir, pattern = "_feature_bc_matrix\\.h5$", full.names = TRUE)
sample.names <- paste0("SS", seq_along(h5.files))

# for each sample: read counts, call cells, create and save a Seurat object
for (i in seq_along(h5.files)) {
  message("Processing ", sample.names[i])
  SS.x <- Read10X_h5(h5.files[i], use.names = TRUE, unique.features = TRUE)

  # combine two cell-calling methods to decide which barcodes are real cells
  iscell_dd <- defaultDrops(SS.x, expected = 5000)
  eout <- emptyDrops(SS.x, lower = 200)
  eout$FDR[is.na(eout$FDR)] <- 1
  iscell_ed <- eout$FDR <= 0.01

  meta <- data.frame(
    iscell_dd = iscell_dd,
    iscell_ed = iscell_ed,
    row.names = colnames(SS.x)
  )
  keep <- iscell_dd | iscell_ed

  SS <- CreateSeuratObject(
    counts = SS.x[, keep],
    meta.data = meta[keep, ]
  )
  SS$sample <- sample.names[i]

  saveRDS(SS, file = paste0("data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/", sample.names[i], ".rds"))
}

# reload all per-sample objects and merge into one
seurat.dir <- "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/"
rds.files <- list.files(seurat.dir, pattern = "^SS[0-9]+\\.rds$", full.names = TRUE)
SS_list <- lapply(rds.files, readRDS)
names(SS_list) <- sub("\\.rds$", "", basename(rds.files))

SS <- merge(
  x = SS_list[[1]],
  y = SS_list[-1],
  add.cell.ids = names(SS_list),
  project = "Systemic_Sclerosis"
)
saveRDS(SS, file = "data/process/SeuratObjects/Systemic_Sclerosis_in_Skin/SS_raw.rds")

DefaultAssay(SS)

# free memory - keep SS, it is used in the following chunks
rm(eout, meta, SS_list, SS.x, h5.files, i, iscell_ed, iscell_dd, keep, rds.files, sample.names)
gc(full = TRUE)
```

## 3. Quality control

```{r, fig.width=10, fig.height=8}
SS[["percent.mt"]] <- PercentageFeatureSet(SS, pattern = "^MT-")
VlnPlot(SS, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE)

# QC thresholds set from this dataset's own distributions
SS <- subset(SS, subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 15)
```

## 4. Normalization, scaling, and PCA

```{r}
SS <- NormalizeData(SS)
SS <- FindVariableFeatures(SS)
SS <- ScaleData(SS, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = SS))
SS <- RunPCA(SS, features = VariableFeatures(object = SS))
ElbowPlot(SS, ndims = 50, reduction = "pca")
SS <- RunUMAP(SS, dims = 1:30)
DimPlot(SS, reduction = "umap", label = TRUE)
```

## 5. Harmony integration and clustering resolutions

```{r}
library(harmony)
SS <- RunHarmony(SS, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:30)
SS[["RNA"]] <- JoinLayers(SS[["RNA"]])
SS <- RunUMAP(SS, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")
SS <- FindNeighbors(SS, reduction = "harmony", dims = 1:30)

# scan a range of resolutions before picking one
for (res in seq(0.2, 1, by = 0.1)) {
  SS <- FindClusters(SS, resolution = res)
  colname <- paste0("res_", res)
  SS[[colname]] <- SS$seurat_clusters
}
res_cols <- grep("^res_", colnames(SS@meta.data), value = TRUE)

# save all resolutions to a single PDF for visual comparison
pdf("output/SS/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(SS, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}
dev.off()

# resolution 0.8 was chosen for downstream clustering
p <- DimPlot(SS, group.by = "res_0.8", label = TRUE, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.8") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/SS/SS_resolution.pdf", plot = p, width = 13, height = 8, dpi = 300)
```

## 6. Clustering and marker genes

```{r}
SS <- FindNeighbors(SS, reduction = "harmony", dims = 1:30)
SS <- FindClusters(SS, resolution = 0.8, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
SS_markers <- FindAllMarkers(SS, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
SS_markers <- SS_markers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 7. Cell type annotation

```{r, fig.width=10, fig.height=8}
genes <- c(
  # Differentiating keratinocytes (suprabasal)
  "KRT1", "KRT2",

  # Basal / activated keratinocytes
  "KRT14", "KRT5", "KRT17", "KRT6A", "KRT6B",

  # Terminal keratinocytes
  "LOR", "FLG", "IVL",

  # Cornified envelope
  "SPRR2G",

  # Fibroblasts
  "SFRP2", "COMP", "APOD", "APOE", "POSTN", "ANGPTL7", "COCH", "ASPN",

  # CD4 T cells
  "CD4", "TCF7", "SELL", "CD3D", "CD3E", "TRAC",

  # CD8 T cells
  "CD8A", "CD8B", "GZMB",

  # B cells
  "MS4A1", "CD79A", "CD79B",

  # plasma cells
  "MZB1", "JCHAIN",

  # mon/mac
  "CD14", "CD36", "FCGR1A", "LYZ", "CD68", "CLEC9A", "CSF1R",

  # NK cells
  "GNLY", "NKG7",

  # Dendritic
  "CD1C", "HLA-DRA",

  # Macrophages
  "CCL3", "APOC1",

  # Mast cells
  "TPSAB1", "FTH1",

  # Pericytes / Mural cells
  "RGS5", "STEAP4", "RERGL", "CXCR4",

  # Smooth muscle cells
  "DES", "TAGLN",

  # Neural / Schwann-related
  "PLP1", "PMP22", "MPZ", "MBP",

  # Proliferating cells
  "MKI67",

  # Melanocytes
  "TYRP1", "PMEL",

  # Preadipocytes
  "APMAP",

  # Endothelial cells
  "PECAM1", "VWF", "RAMP2", "CLDN5",

  # Lymphatic endothelial cells
  "CCL21", "PROX1", "LYVE1",

  # Secretory cells
  "KRT19", "SNORC", "CALML5"
)

p <- DotPlot(
  SS,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  group.by = "harmony_clusters"
) + RotatedAxis()

p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SS/SS_cluster.pdf", plot = p, width = 13, height = 8)
```

```{r}
l <- FeaturePlot(SS, features = unique(genes), combine = FALSE, order = TRUE, reduction = "umap.harmony")
names(l) <- unique(genes)
```


```{r}
#cell type labels
Idents(SS) <- "harmony_clusters"
cluster_name_x <- c(
  "0"  = "Suprabasal Keratinocyte_0",
  "1"  = "Basal Keratinocyte_1",
  "2"  = "Basal Keratinocyte_2",
  "3"  = "Mural Cells_3",
  "4"  = "Fibroblasts_4",
  "5"  = "Endothelial Cells_5",
  "6"  = "Fibroblasts_6",
  "7"  = "Smooth Muscle Cells_7",
  "8"  = "Endothelial Cells_8",
  "9"  = "Fibroblasts_9",
  "10" = "CD4_10",
  "11" = "Basal Keratinocyte_11",
  "12" = "Suprabasal Keratinocyte_12",
  "13" = "Dendritic Cells_13",
  "14" = "Suprabasal Keratinocyte_14",
  "15" = "Fibroblasts_15",
  "16" = "Macrophages_16",
  "17" = "Basal Keratinocyte_17",
  "18" = "Secretory Cells_18",
  "19" = "Smooth Muscle Cells_19",
  "20" = "CD8_20",
  "21" = "Schwann Cells_21",
  "22" = "Suprabasal Keratinocyte_22",
  "23" = "Mast Cells_23",
  "24" = "Melanocytes_24",
  "25" = "Lymphatic Endothelial_25"
)

SS <- RenameIdents(SS, cluster_name_x)
SS$CellType <- Idents(SS)

p <- DotPlot(
  SS,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  group.by = "CellType"
) + RotatedAxis()

p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SS/SS_cluster_name.pdf", plot = p, width = 14, height = 10)

```

```{r}
# major cell types
Idents(SS) <- "harmony_clusters"
cluster_name <- c(
  "0"  = "Suprabasal Keratinocyte",
  "1"  = "Basal Keratinocyte",
  "2"  = "Basal Keratinocyte",
  "3"  = "Mural Cells",
  "4"  = "Fibroblasts",
  "5"  = "Endothelial Cells",
  "6"  = "Fibroblasts",
  "7"  = "Smooth Muscle Cells",
  "8"  = "Endothelial Cells",
  "9"  = "Fibroblasts",
  "10" = "CD4",
  "11" = "Basal Keratinocyte",
  "12" = "Suprabasal Keratinocyte",
  "13" = "Dendritic Cells",
  "14" = "Suprabasal Keratinocyte",
  "15" = "Fibroblasts",
  "16" = "Macrophages",
  "17" = "Basal Keratinocyte",
  "18" = "Secretory Cells",
  "19" = "Smooth Muscle Cells",
  "20" = "CD8",
  "21" = "Schwann Cells",
  "22" = "Suprabasal Keratinocyte",
  "23" = "Mast Cells",
  "24" = "Melanocytes",
  "25" = "Lymphatic Endothelial"
)

SS <- RenameIdents(SS, cluster_name)
SS$Major_Celltype <- Idents(SS)

celltype_cols <- c(
  "Basal Keratinocyte"      = "#9467bd",
  "Suprabasal Keratinocyte" = "brown2",
  "Fibroblasts"             = "cyan",
  "Lymphatic Endothelial"   = "mediumorchid1",
  "Mural Cells"             = "#1F78B4",
  "Smooth Muscle Cells"     = "lightpink2",
  "CD4"                     = "olivedrab2",
  "CD8"                     = "tomato",
  "Dendritic Cells"         = "yellow3",
  "Macrophages"             = "orangered",
  "Mast Cells"              = "#FDBF6F",
  "Schwann Cells"           = "plum3",
  "Melanocytes"             = "orangered4",
  "Secretory Cells"         = "salmon3",
  "Endothelial Cells"       = "#E31A1C"
)

# UMAP with a label placed at each cell type's centroid
library(ggrepel)
colnames(Embeddings(SS, "umap.harmony")) # check column names (dot in reduction name is stripped)
umap_df <- as.data.frame(Embeddings(SS, "umap.harmony"))
umap_df$Major_Celltype <- SS$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 5,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")
fig1a

ggsave(filename = "output/SS/SS_CellType.pdf", plot = fig1a, width = 12, height = 10)
```

```{r}
p <- DotPlot(
  SS,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  group.by = "Major_Celltype"
) + RotatedAxis()

p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 9, face = "bold"),
  axis.text.y = element_text(size = 9, face = "bold")
)
ggsave(filename = "output/SS/SS_markers.pdf", plot = p, width = 14, height = 8)
```

## 8. Correlation between cell types

```{r}
Idents(SS) <- "Major_Celltype"
levels(Idents(SS)) # sanity check - should show cell type names, not numbers

av.expSS <- AverageExpression(
  SS,
  assays = "RNA",
  slot = "data"
)$RNA

correlation_matrixSS <- cor(
  as.matrix(av.expSS),
  use = "pairwise.complete.obs"
)
correlation_matrixSS[is.na(correlation_matrixSS)] <- 0

pdf(
  file = "output/SS/SS_correlation.pdf",
  width = 13,
  height = 10
)
library(pheatmap)
pheatmap(
  correlation_matrixSS,
  clustering_method = "ward.D2",
  fontsize_row = 9,
  fontsize_col = 9
)
dev.off()
```

## 9. cell-type grouping (General)

```{r}
unique(SS$Major_Celltype)

SS$General <- recode(SS$Major_Celltype,
  "Fibroblasts"             = "Fibroblasts",
  "Mast Cells"              = "Mast Cells",
  "Schwann Cells"           = "Glial Cells",
  "Dendritic Cells"         = "Mono/Mac",
  "Secretory Cells"         = "Epithelial Cells",
  "Mural Cells"             = "Mural Cells",
  "CD8"                     = "CD8",
  "CD4"                     = "CD4",
  "Smooth Muscle Cells"     = "Smooth Muscle Cells",
  "Endothelial Cells"       = "Endothelial Cells",
  "Lymphatic Endothelial"   = "Endothelial Cells",
  "Melanocytes"             = "Melanocytes",
  "Suprabasal Keratinocyte" = "Keratinocytes",
  "Basal Keratinocyte"      = "Keratinocytes",
  "Macrophages"             = "Mono/Mac"
)
```

## 10. TLS (tertiary lymphoid structure) marker heatmap

```{r}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(
  germinal_center, inflammation,
  lymphoid_chemokines, hev, fdc
))
markers <- markers[markers %in% rownames(SS)]

marker_groups <- case_when(
  markers %in% germinal_center     ~ "Germinal Center",
  markers %in% inflammation        ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev                 ~ "HEV",
  markers %in% fdc                 ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center"          = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines"      = "#35978F",
  "HEV"                      = "#8073AC",
  "FDC"                      = "#E08214"
)

# percent of cells expressing each marker, per cell type
Idents(SS) <- "General"
DefaultAssay(SS) <- "RNA"
expr_mat <- GetAssayData(SS, layer = "data")

percent_mat <- sapply(levels(Idents(SS)), function(ct) {
  cells <- WhichCells(SS, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  SS,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/SS/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 11. Export for pySCENIC

```{r}
expr <- GetAssayData(SS, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SS/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SS/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/SS/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

## 12. Save final metadata and processed object

```{r}
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General"
)

SS@meta.data <- SS@meta.data[, keep_cols]

write.csv(SS@meta.data,
  file = "data/process/SeuratObjects/SS/SS_obj.metadata.csv",
  row.names = TRUE
)

saveRDS(SS, file = "data/process/SeuratObjects/SS/SS_process.rds")
SS <- readRDS("data/process/SeuratObjects/SS/SS_process.rds")
```
