---
title: "Type 1 Diabetes - Processing and Annotation"
output: html_notebook
---



## 1. Packages and setup

```{r setup, include=FALSE}
library(Seurat)
library(Matrix)
library(dplyr)
library(gprofiler2)
library(ggplot2)
library(patchwork)
library(gplots)
library(cowplot)
library(BiocManager)
library(devtools)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```

## 2. Load raw data and subset to Type 1 Diabetes samples

```{r}
# the pancreas dataset contains several disease groups - keep only T1D
T1D <- readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Pancreas Type 1 Diabetes/Pancreas Type 1 Diabetes.rds")

# check default assay, set to RNA if not already
DefaultAssay(T1D)
DefaultAssay(T1D) <- "RNA"

setT1D <- c("T1D_ 1", "T1D_ 2", "T1D_ 3", "T1D_ 4", "T1D_ 5", "T1D_ 6", "T1D_ 7", "T1D_ 8", "T1D_ 9")
T1D <- subset(T1D, subset = disease_id %in% setT1D)
saveRDS(T1D, file = "data/process/SeuratObjects/T1D/T1D_raw.rds")
```

## 3. Quality control

```{r}
# QC thresholds set based on this dataset's own distribution
T1D[["percent.mt"]] <- PercentageFeatureSet(T1D, pattern = "^MT-")
Idents(T1D) <- "x"
T1D$x <- "All"
VlnPlot(T1D, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol = 3, log = TRUE, group.by = "x")
T1D <- subset(T1D, subset = nFeature_RNA > 300 & nFeature_RNA < 7000 & percent.mt < 10)
```

## 4. Normalization, scaling, and PCA

```{r}
T1D <- NormalizeData(T1D)
T1D <- FindVariableFeatures(T1D)
T1D <- ScaleData(T1D, vars.to.regress = c("nCount_RNA", "percent.mt"), features = VariableFeatures(object = T1D))
T1D <- RunPCA(T1D, features = VariableFeatures(object = T1D))
ElbowPlot(T1D, ndims = 50, reduction = "pca")
T1D <- RunUMAP(T1D, dims = 1:30)
DimPlot(T1D, reduction = "umap", label = TRUE)
```

## 5. Harmony integration and clustering resolutions

```{r}
library(harmony)
T1D <- RunHarmony(T1D, group.by.vars = "sample", orig.reduction = "pca", dims.use = 1:30)
T1D <- RunUMAP(T1D, reduction = "harmony", dims = 1:30, reduction.name = "umap.harmony")

# scan a range of resolutions before picking one
T1D <- FindNeighbors(T1D, reduction = "harmony", dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  T1D <- FindClusters(T1D, resolution = res)
  colname <- paste0("res_", res)
  T1D[[colname]] <- T1D$seurat_clusters
}
res_cols <- grep("^res_", colnames(T1D@meta.data), value = TRUE)

for (res in res_cols) {
  p <- DimPlot(T1D, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste(res))
  print(p)
}

# save all resolutions to a single PDF for visual comparison
pdf("output/T1D/All_resolutions_UMAP.pdf", width = 6, height = 5)
for (res in res_cols) {
  p <- DimPlot(T1D, group.by = res, label = TRUE, reduction = "umap.harmony") +
    ggtitle(paste("Resolution:", res))
  print(p)
}
dev.off()

# resolution 0.6 was chosen for downstream clustering
p <- DimPlot(T1D, group.by = "res_0.6", label = TRUE, reduction = "umap.harmony") +
  ggtitle("Resolution: 0.6") +
  NoLegend() +
  labs(x = "UMAP.h", y = "UMAP.h")
ggsave(filename = "output/T1D/T1D_resolution.pdf", plot = p, width = 10, height = 8, dpi = 300)
```

## 6. Clustering and marker genes

```{r}
T1D <- FindNeighbors(T1D, reduction = "harmony", dims = 1:30)
T1D <- FindClusters(T1D, resolution = 0.6, cluster.name = "harmony_clusters")

# markers per cluster, used below to name the cell types
T1Dmarkers <- FindAllMarkers(T1D, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
T1Dmarkers <- T1Dmarkers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
```

## 7. Cell type annotation

```{r}
genes <- c(
  # Acinar cells
  "PRSS1", "CPA1", "PTF1A",

  # Alpha cells
  "GCG", "IRX2", "ARX",

  # Beta cells
  "INS", "MAFA", "NKX6-1",

  # Delta cells
  "SST", "LEPR",

  # Ductal cells
  "KRT19", "SOX9", "CFTR",

  # Endothelial cells
  "VWF",

  # Epsilon cells
  "GHRL",

  # Immune cells
  "NCF2", "PTPRC", "CD14", "FCGR3A", "CD68", "LYZ", "CSF1R",

  # CD4 T cells
  "CD4", "TCF7", "SELL", "CD3D", "CD3E", "TRAC",

  # CD8 T cells
  "CD8A", "CD8B", "GZMB",

  # B cells
  "MS4A1", "CD79A", "CD79B",

  # Plasma cells
  "MZB1", "JCHAIN",

  # PP (gamma) cells
  "PPY",

  # Pericytes
  "RGS5",

  # Stellate cells
  "COL1A1",

  # Fibroblasts
  "DCN"
)

p <- DotPlot(
  T1D,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "harmony_clusters"
) + RotatedAxis()
p
ggsave(filename = "output/T1D/T1D_cluster.pdf", plot = p, width = 12, height = 8)
```

```{r}
# per-gene expression maps, for interactive inspection of individual markers
l <- FeaturePlot(T1D, features = unique(genes), combine = FALSE, order = TRUE, reduction = "umap.harmony")
names(l) <- unique(genes)
```

```{r}
# cell type labels
Idents(T1D) <- "harmony_clusters"
cluster_name_x <- c(
  "0" = "Ductal Cells_0",
  "1" = "Acinar Cells_1",
  "2" = "Alpha Cells_2",
  "3" = "Acinar Cells_3",
  "4" = "Alpha Cells_4",
  "5" = "Delta Cells_5",
  "6" = "Acinar Cells_6",
  "7" = "Endothelial Cells_7",
  "8" = "Fibroblasts_8",
  "9" = "Acinar Cells_9",
  "10" = "Alpha Cells_10",
  "11" = "Ductal Cells_11",
  "12" = "Mural Cells_12",
  "13" = "Macrophages_13",
  "14" = "Acinar Cells_14",
  "15" = "Ductal Cells_15",
  "16" = "Acinar Cells_16",
  "17" = "Acinar Cells_17",
  "18" = "Fibroblasts_18"
)
T1D <- RenameIdents(T1D, cluster_name_x)
T1D$CellType <- Idents(T1D)

p <- DotPlot(
  T1D,
  features = genes,
  cols = c("lightgrey", "blue"),
  col.min = -2.5,
  col.max = 2.5,
  dot.min = 0,
  dot.scale = 6,
  split.by = NULL,
  scale = TRUE,
  scale.by = "radius",
  scale.min = NA,
  scale.max = NA,
  cluster.idents = TRUE,
  group.by = "CellType"
) + RotatedAxis()
p
ggsave(filename = "output/T1D/T1D_cluster_name.pdf", plot = p, width = 13, height = 10)
```

```{r}
# major cell types (clusters of the same type merged, no cluster number)
Idents(T1D) <- "harmony_clusters"
cluster_name <- c(
  "0" = "Ductal Cells",
  "1" = "Acinar Cells",
  "2" = "Alpha Cells",
  "3" = "Acinar Cells",
  "4" = "Alpha Cells",
  "5" = "Delta Cells",
  "6" = "Acinar Cells",
  "7" = "Endothelial Cells",
  "8" = "Fibroblasts",
  "9" = "Acinar Cells",
  "10" = "Alpha Cells",
  "11" = "Ductal Cells",
  "12" = "Mural Cells",
  "13" = "Macrophages",
  "14" = "Acinar Cells",
  "15" = "Ductal Cells",
  "16" = "Acinar Cells",
  "17" = "Acinar Cells",
  "18" = "Fibroblasts"
)
T1D <- RenameIdents(T1D, cluster_name)
T1D$Major_Celltype <- Idents(T1D)

celltype_cols <- c(
  "Acinar Cells" = "coral1",
  "Alpha Cells" = "mediumorchid1",
  "Ductal Cells" = "mistyrose3",
  "Delta Cells" = "#762A83",
  "Endothelial Cells" = "#E31A1C",
  "Stellate Cells" = "yellow",
  "Macrophages" = "orangered",
  "Fibroblasts" = "cyan",
  "Mural Cells" = "#1F78B4"
)

# UMAP with a label placed at each cell type's centroid
colnames(Embeddings(T1D, "umap.harmony"))
umap_df <- as.data.frame(Embeddings(T1D, "umap.harmony"))
umap_df$Major_Celltype <- T1D$Major_Celltype

centers <- aggregate(
  cbind(umapharmony_1, umapharmony_2) ~ Major_Celltype,
  data = umap_df,
  FUN = median
)

library(ggrepel)
fig1a <- ggplot(umap_df, aes(umapharmony_1, umapharmony_2, color = Major_Celltype)) +
  geom_point(size = 0.3) +
  scale_color_manual(values = celltype_cols) +
  geom_label_repel(
    data = centers,
    aes(x = umapharmony_1, y = umapharmony_2, label = Major_Celltype),
    inherit.aes = FALSE,
    fill = "white",
    color = "black",
    size = 4.5,
    box.padding = 0.4,
    label.size = 0.25,
    segment.color = NA
  ) +
  theme_classic() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none"
  ) +
  labs(x = "UMAP.h", y = "UMAP.h")
fig1a
ggsave(filename = "output/T1D/T1D_cellType.pdf", plot = fig1a, width = 12, height = 10)
```

## 8. Correlation and marker expression heatmaps

```{r}
library(circlize)
library(ComplexHeatmap)
library(viridis)
library(grid)
library(plyr)
library(pheatmap)

# average expression per cluster, then correlation between clusters
av.expSS <- AverageExpression(
  T1D,
  assays = "RNA",
  slot = "data"
)$RNA

correlation_matrixSS <- cor(
  as.matrix(av.expSS),
  use = "pairwise.complete.obs"
)
correlation_matrixSS[is.na(correlation_matrixSS)] <- 0
correlation_matrixSS[is.nan(correlation_matrixSS)] <- 0

# small jitter so identical values don't break pheatmap's clustering
set.seed(1)
epsilon <- 1e-10
correlation_matrixSS_jitter <- correlation_matrixSS +
  matrix(runif(length(correlation_matrixSS), 0, epsilon),
    nrow = nrow(correlation_matrixSS)
  )

png(
  filename = "output/T1D/T1D_correlation.png",
  width = 3100,
  height = 2200,
  res = 300
)
pheatmap(
  correlation_matrixSS_jitter,
  clustering_method = "ward.D2",
  fontsize_row = 8,
  fontsize_col = 8
)
dev.off()

# marker dot-heatmap: dot size = percent of cells expressing, color = scaled expression
T1D$celltype <- mapvalues(as.character(Idents(T1D)),
  from = names(cluster_name),
  to = cluster_name
)
Idents(T1D) <- "celltype"

DefaultAssay(T1D) <- "RNA"
expr_mat <- GetAssayData(T1D, layer = "data")

percent_mat <- sapply(levels(Idents(T1D)), function(ct) {
  cells <- WhichCells(T1D, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(T1D,
  features = genes, assays = "RNA",
  layer = "data", return.seurat = FALSE
)$RNA
exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

q <- quantile(exp_mat, c(0.05, 0.5, 0.95))
col_fun <- colorRamp2(q, c("navy", "white", "firebrick"))

cell_fun <- function(j, i, x, y, w, h, fill) {
  grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(unit.c(w, h)),
    gp = gpar(fill = col_fun(exp_mat[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat)
cols <- viridis(length(celltypes))
names(cols) <- celltypes
column_ha <- HeatmapAnnotation(celltype = celltypes, col = list(celltype = cols))

# log1p + per-gene z-score, to bring expression to a comparable range
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

ht <- Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, rot = 45),
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

pdf(
  file = "output/T1D/T1D_markers.pdf",
  width = 12, height = 10
)
draw(ht)
dev.off()
```

## 9.cell-type grouping (General)

```{r}
unique(T1D$Major_Celltype)

# broader lineage grouping, used as the identity level for downstream CCC analysis
T1D$General <- recode(T1D$Major_Celltype,
  "Ductal Cells" = "Epithelial Cells",
  "Acinar Cells" = "Epithelial Cells",
  "Alpha Cells" = "Endocrine Cells",
  "Fibroblasts" = "Fibroblasts",
  "Endothelial Cells" = "Endothelial Cells",
  "Delta Cells" = "Endocrine Cells",
  "Mural Cells" = "Mural Cells",
  "Macrophages" = "Mono/Mac"
)
```

## 10. TLS (tertiary lymphoid structure) marker heatmap

```{r, fig.width=12, fig.height=8}
# marker sets per functional group, from genes_TLS.csv
germinal_center <- c(
  "AID", "BCL6", "PAX5", "CD40LG", "AICDA", "MYC",
  "IRF8", "S1PR2", "ICOS", "CXCR4", "CXCR5"
)

inflammation <- c(
  "TNF", "IL1B", "IL6", "IL7", "IL8", "IL12", "IL17",
  "IFNG", "TGFB", "CCL2", "CCL3", "CCL4", "CCL5",
  "ICAM1", "VCAM1", "SELE"
)

lymphoid_chemokines <- c(
  "CXCL13", "CCL21", "CCL19", "CXCL12",
  "CXCR5", "CCR7", "CXCR4"
)

hev <- c(
  "NTAN1", "MADCAM1", "CHST4", "GLYCAM1", "FUT7", "CHST2",
  "MEOX2", "TSPAN7", "ANKRD43", "ZNF280C"
)

fdc <- c(
  "CD21", "CD23", "CR1", "CR2", "CD35", "FDCSP",
  "CXCL13", "CD54", "ICAM1", "FCER2"
)

markers <- unique(c(
  germinal_center, inflammation,
  lymphoid_chemokines, hev, fdc
))
markers <- markers[markers %in% rownames(T1D)]

marker_groups <- case_when(
  markers %in% germinal_center ~ "Germinal Center",
  markers %in% inflammation ~ "Inflammation & Cytokines",
  markers %in% lymphoid_chemokines ~ "Lymphoid Chemokines",
  markers %in% hev ~ "HEV",
  markers %in% fdc ~ "FDC"
)
names(marker_groups) <- markers

group_cols <- c(
  "Germinal Center" = "#2166AC",
  "Inflammation & Cytokines" = "#D6604D",
  "Lymphoid Chemokines" = "#35978F",
  "HEV" = "#8073AC",
  "FDC" = "#E08214"
)

# percent of cells expressing each marker, per cell type
Idents(T1D) <- "General"
DefaultAssay(T1D) <- "RNA"
expr_mat <- GetAssayData(T1D, layer = "data")

percent_mat <- sapply(levels(Idents(T1D)), function(ct) {
  cells <- WhichCells(T1D, idents = ct)
  Matrix::rowMeans(expr_mat[markers, cells, drop = FALSE] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

# average expression per marker, per cell type
avg_exp <- AverageExpression(
  T1D,
  features = markers,
  assays = "RNA",
  layer = "data",
  return.seurat = FALSE
)$RNA

exp_mat <- as.matrix(avg_exp[markers, colnames(percent_mat)])
exp_mat[is.infinite(exp_mat)] <- NA

# drop markers with no expression in any cell type
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
marker_groups <- marker_groups[rownames(exp_mat)]

exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"
exp_mat_scaled <- log1p(exp_mat)
exp_mat_scaled <- t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)] <- 0

q <- quantile(exp_mat_scaled, c(0.05, 0.5, 0.95))
col_fun <- circlize::colorRamp2(q, c("navy", "white", "firebrick"))

# bubble size = percent expressing, color = scaled expression
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid::grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(grid::unit.c(w, h)),
    gp = grid::gpar(fill = col_fun(exp_mat_scaled[i, j]), col = "black")
  )
}

celltypes <- colnames(exp_mat_scaled)
cols <- grDevices::colorRampPalette(
  RColorBrewer::brewer.pal(8, "Set2")
)(length(celltypes))
names(cols) <- celltypes

column_ha <- ComplexHeatmap::HeatmapAnnotation(
  celltype = celltypes,
  col = list(celltype = cols),
  show_annotation_name = FALSE
)

row_ha <- ComplexHeatmap::rowAnnotation(
  Group = marker_groups,
  col = list(Group = group_cols),
  annotation_width = unit(3, "mm"),
  show_annotation_name = FALSE,
  annotation_legend_param = list(
    Group = list(
      title = "TLS genes",
      title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
      labels_gp = grid::gpar(fontsize = 9)
    )
  )
)

ht <- ComplexHeatmap::Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,
  rect_gp = grid::gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = grid::gpar(fontsize = 12, fontface = "bold"),
  column_names_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 90,
  border = TRUE,
  top_annotation = column_ha,
  left_annotation = row_ha,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = marker_groups,
  row_title_gp = grid::gpar(fontsize = 10, fontface = "bold"),
  row_gap = unit(2, "mm")
)

pdf("output/T1D/TLS_markers.pdf", width = 8, height = 10)
ComplexHeatmap::draw(ht, merge_legend = TRUE)
dev.off()
```

## 11. Export for pySCENIC

```{r}
# raw counts as a sparse matrix + gene/barcode lists, for pySCENIC in the terminal
expr <- GetAssayData(T1D, assay = "RNA", layer = "counts")

writeMM(expr, "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/T1D/input/matrix.mtx")

write.table(rownames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/T1D/input/genes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)

write.table(colnames(expr),
  file = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS/GRN_SCENIC/T1D/input/barcodes.tsv",
  quote = FALSE,
  row.names = FALSE,
  col.names = FALSE
)
```

## 12. Save final metadata and processed object

```{r}
keep_cols <- c(
  "orig.ident",
  "nCount_RNA",
  "nFeature_RNA",
  "sample",
  "harmony_clusters",
  "CellType",
  "Major_Celltype",
  "General"
)
T1D@meta.data <- T1D@meta.data[, keep_cols]

write.csv(T1D@meta.data, file = "data/process/SeuratObjects/T1D/T1D.metadata.csv", row.names = TRUE)

saveRDS(T1D, file = "data/process/SeuratObjects/T1D/T1D_process.rds")

```
