---
title: "R Notebook"
output: html_notebook
---
################################################################################################################################
#####ANALYSE OF PANCREAS TYPE 1 DIABETES#######################################################################################
################################################################################################################################

```{r,setup, include=FALSE}
##################PACKAGES
library(pacman)
  p_load(Seurat, SeuratData, Matrix, dplyr, DropletUtils, presto,gprofiler2, ggplot2, patchwork, gplots, cowplot, openxlsx,plyr, scDblFinder, BiocParallel,devtools,openxlsx,reticulate,hdf5r,DropletUtils)
 #### the wd
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/mnt/nfs/WORKSP/aminata.sow/LAB_DOCS")
```


#####STEP_1
###LOAD the data
```{r}
##load the data and set only the T1D
T1D=readRDS("../../tomas.gomes/processed.data/autoimm_scrna_TomasGuerreiro/Datasets/Pancreas Type 1 Diabetes/Pancreas Type 1 Diabetes.rds")
setT1D = c("T1D_ 1","T1D_ 2","T1D_ 3","T1D_ 4","T1D_ 5","T1D_ 6","T1D_ 7","T1D_ 8","T1D_ 9")
T1D = subset (T1D, subset = disease_id %in% setT1D)
saveRDS (T1D,file = "data/process/SeuratObjects/T1Diabetes/T1D_raw.rds")

```

####STEP_2:PROCESS AND ANALYSE THE RAW_FILE

###1-FOR THE QC
```{r}
##load and process
T1D[["percent.mt"]]=PercentageFeatureSet(T1D, pattern = "^MT-")
Idents(T1D)="x"
T1D$x="All"
VlnPlot(T1D, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3, log = TRUE, group.by = "x")
T1D=subset(T1D, subset=nFeature_RNA >300 & nFeature_RNA <7000 & percent.mt<10)


```


###2 statistics
```{r}
##norm, fvf,scale,runpca

T1D=NormalizeData(T1D)
T1D=FindVariableFeatures(T1D)
T1D=ScaleData(T1D,vars.to.regress = c("nCount_RNA","percent.mt"),features = VariableFeatures(object = T1D))
T1D=RunPCA(T1D, features = VariableFeatures(object = T1D))
ElbowPlot(T1D, ndims = 50, reduction = "pca")
T1D=RunUMAP(T1D, dims = 1:30)
DimPlot(T1D,reduction = "umap", label = TRUE)

```


###3.integration
```{r}
##integration
library(harmony)
T1D=RunHarmony(T1D,group.by.vars = "sample_id",orig.reduction = "pca",dims.use = 1:30  )
DefaultAssay(T1D)="SCT"

T1D = RunUMAP(T1D, reduction = "harmony", dims= 1:30, reduction.name = "umap.harmony")
p=DimPlot(T1D,reduction = "umap.harmony",group.by = c("sample_id"),label.size = 2, shuffle = T)
ggsave(filename = "output/T1Diabetes/T1D.integration.png", plot=p, width = 12, height = 8)
p

#for the resolution
T1D=FindNeighbors(T1D,reduction= "harmony",dims = 1:30)
for (res in seq(0.2, 1, by = 0.1)) {
  T1D=FindClusters(T1D, resolution = res)
  colname=paste0("res_", res)
  T1D[[colname]]=T1D$seurat_clusters
}
res_cols=grep("^res_", colnames(T1D@meta.data), value = TRUE)
for (res in res_cols) {
  p=DimPlot(T1D, group.by = res, label = TRUE, reduction = "umap.harmony") +
       ggtitle(paste("Resolution:", res))
  print(p)
  readline(prompt="Press [Enter] to see the next resolution...")
}
#save 
p=DimPlot(T1D,group.by = "res_0.6",label = TRUE,reduction = "umap.harmony" )+ ggtitle("Resolution: 0.6")
ggsave(filename = "output/T1Diabetes/T1Dresolution.png",plot = p,width = 12,height = 8,dpi = 300)
```


##4-cluster and markers
```{r}

###clusters
T1D=FindNeighbors(T1D,reduction= "harmony",dims = 1:30)
T1D=FindClusters(T1D, resolution = 0.6,cluster.name = "harmony_clusters")
#markers
T1Dmarkers=FindAllMarkers(T1D,only.pos = TRUE,min.pct = 0.25, logfc.threshold = 0.25)
T1Dmarkers=T1Dmarkers %>% filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.25)
write.xlsx(T1Dmarkers,file = "output/T1Diabetes/T1Dmarkers.xlsx")
View(T1Dmarkers)



```


Acinar cells (PRSS1 high), Alpha cells (GCG high), Beta cells (INS high), Delta cells (SST high), Ductal
cells (KRT19 high), Endothelial cells (VWF high), Epsilon cells (GHRL high), Immune (NCF2 high), PP
cells (PPY high), Stellates cells (COL1A1 high)


###5 annotation

```{r}

#--------markers------
genes=c("PRSS1","GCG","INS","SST","KRT19","VWF","GHRL","NCF2")
##1celltype
Idents(T1D)="harmony_clusters"
cluster_name=c(
  "0"  = "Acinar cells" ,
  "1"  = "Alpha cells"    ,
  "2"  = "Ductal cells"            ,
  "3"  = "Acinar cells"      ,
  "4"  = "Alpha cells"    ,
  "5"  = "Acinar cells"            ,
  "6"  = "Delta cells"            ,
  "7"  = "Endothelial cells"            ,
  "8"  = "Beta cells"             ,
  "9"  = "Stellates cells"               ,
  "10" = "Immune cells"        ,
  "11" = "Acinar cells"    ,
  "12" = "Endothelial cells"    ,
  "13" = "Acinar cells"           ,
  "14" = "Ductal cells"             ,
  "15" = "Endothelial cells"            ,
  "16" = "Ductal cells"          ,
  "17" = "Acinar cells"      
  
)

T1D=RenameIdents(T1D, cluster_name)
T1D$celltype=Idents(T1D)
##2color
celltype_cols=c(
  "Acinar cells"            = "#D73027",  
  "Alpha cells"             = "#4575B4", 
  "Ductal cells"            = "#FDAE61",  
  "Delta cells"             = "#1B9E77", 
  "Endothelial cells"       = "#66A61E", 
  "Beta cells"              = "#000000",  
  "Stellates cells"         = "#762A83", 
  "Immune cells"            = "#74ADD1"  

  )

##3 center
##3 center
umap_df=as.data.frame(Embeddings(T1D, "umap.harmony"))
colnames(umap_df)=c("umap_1", "umap_2")
umap_df$celltype=T1D$celltype
centers=aggregate(
  cbind(umap_1, umap_2) ~ celltype,
  data = umap_df,
  FUN = median
)

##supp duplicate and plot
library(ggrepel)
suppressWarnings({
  fig1a=ggplot(umap_df, aes(umap_1, umap_2, color = celltype)) +
    geom_point(size = 0.3) +
    scale_color_manual(values = celltype_cols) +
    geom_label_repel(
      data = centers,
      aes(x = umap_1, y = umap_2, label = celltype),
      inherit.aes = FALSE,
      fill = "white",
      color = "black",
      size = 4,
      box.padding = 0.4,
      label.size = 0.25,
      segment.color = NA,
      max.overlaps = Inf
    ) +
    theme_classic() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    ) +
    labs(x = "umap 1", y = "umap 2")
})
fig1a
ggsave(ggsave(filename = "output/T1Diabetes/T1DcellT.png", plot = fig1a, width = 12, height = 8))

```








##6Plot with markers and Correlation
```{r}

## ===============================
## ComplexHeatmap Dot Heatmap
## ===============================

## ===============================
## ComplexHeatmap Dot Heatmap
## ===============================

library(Seurat)
library(dplyr)
library(Matrix)
library(circlize)
library(ComplexHeatmap)
library(viridis)
library(grid)
library(plyr)
library(pheatmap)

# folder
dir.create("output/T1Diabetes", recursive = TRUE, showWarnings = FALSE)

# -------------------------------
# 1️⃣ Correlation heatmap (optional)
# -------------------------------

# Average expression per cluster (SCT)
av.expSS <- AverageExpression(
  T1D,
  assays = "SCT",
  slot = "data"
)$SCT

# Correlation matrix
correlation_matrixSS <- cor(
  as.matrix(av.expSS),
  use = "pairwise.complete.obs"
)

# Replace NA / NaN
correlation_matrixSS[is.na(correlation_matrixSS)] <- 0
correlation_matrixSS[is.nan(correlation_matrixSS)] <- 0

# Add tiny jitter to avoid identical breaks
set.seed(1)
epsilon <- 1e-10
correlation_matrixSS_jitter <- correlation_matrixSS +
  matrix(runif(length(correlation_matrixSS), 0, epsilon),
         nrow = nrow(correlation_matrixSS))

# Save correlation heatmap as PNG

png(
  filename = "output/T1Diabetes/T1Dcorr_matrix.png",
  width = 3100,
  height = 2200,
  res = 300
)
#---
pheatmap(
  correlation_matrixSS_jitter,
  clustering_method = "ward.D2",
  main = "Correlation between Cell Types",
  fontsize_row = 8,
  fontsize_col = 8,
  filename = "output/T1Diabetes/T1Dcorr_matrix.png"
)
dev.off()
# -------------------------------
# 2️⃣ Dot heatmap (ComplexHeatmap)
# -------------------------------

# Make sure you have defined these:
# genes <- c(...)  # your genes of interest
# cluster_name <- c(...)  # mapping of cluster IDs to cell types

T1D$celltype <- mapvalues(as.character(Idents(T1D)), 
                          from = names(cluster_name), 
                          to = cluster_name)
Idents(T1D) <- "celltype"

# Expression and percent matrices
DefaultAssay(T1D) <- "RNA"
expr_mat <- GetAssayData(T1D, layer = "data")

percent_mat <- sapply(levels(Idents(T1D)), function(ct) {
  cells <- WhichCells(T1D, idents = ct)
  Matrix::rowMeans(expr_mat[genes, cells] > 0) * 100
})
percent_mat <- as.matrix(percent_mat)

avg_exp <- AverageExpression(T1D, features = genes, assays = "RNA",
                             layer = "data", return.seurat = FALSE)$RNA
exp_mat <- as.matrix(avg_exp[genes, colnames(percent_mat)])

# Clean Inf / NA
exp_mat[is.infinite(exp_mat)] <- NA
keep <- rowSums(is.na(exp_mat)) < ncol(exp_mat)
exp_mat <- exp_mat[keep, ]
percent_mat <- percent_mat[keep, ]
exp_mat[is.na(exp_mat)] <- 0
mode(exp_mat) <- "numeric"

# Color function
q <- quantile(exp_mat, c(0.05, 0.5, 0.95))
col_fun <- colorRamp2(q, c("navy", "white", "firebrick"))

# Dot function
cell_fun <- function(j, i, x, y, w, h, fill) {
  grid.circle(
    x = x, y = y,
    r = (percent_mat[i, j] / 100) * 0.45 * min(unit.c(w, h)),
    gp = gpar(fill = col_fun(exp_mat[i, j]), col = "black")
  )
}

# Column annotation
celltypes <- colnames(exp_mat)
cols <- viridis(length(celltypes))
names(cols) <- celltypes
column_ha <- HeatmapAnnotation(celltype = celltypes, col = list(celltype = cols))

# Create heatmap object

# log1p transform (log(1+x)) to bring values to reasonable range
exp_mat_scaled=log1p(exp_mat)

# optional: z-score per gene (row)
exp_mat_scaled=t(scale(t(exp_mat_scaled)))
exp_mat_scaled[is.na(exp_mat_scaled)]= 0

ht=Heatmap(
  exp_mat_scaled,
  name = "Expression",
  col = col_fun,   # you can regenerate col_fun for scaled values
  rect_gp = gpar(type = "none"),
  cell_fun = cell_fun,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, rot = 45),
  border = TRUE,
  top_annotation = column_ha,
  row_km = 4
)

# -------------------------------
# 3️⃣ Save ComplexHeatmap as PNG
# -------------------------------

png(filename = "output/T1Diabetes/T1Dcorr_dotheatmap.png",
    width = 3100, height = 2200, res = 300)
draw(ht)
dev.off()


```



###7 proportion of Donors
```{r}
library(dplyr)
library(ggplot2)
library(scales)

#change colname 
T1D$Donors = T1D$sample_id

#create the meta data
meta = T1D@meta.data[, c("harmony_clusters", "Donors")]
table(meta$Donors)

# Make sure dplyr functions are used explicitly
meta = T1D@meta.data

#the  percentage 
donors_counts = meta %>%
  as_tibble() %>%
  dplyr::group_by(harmony_clusters,Donors) %>% #count cells each donor use in each cluster
  dplyr::summarise( n= dplyr::n(),.groups = "drop") %>% #numb of cells
  dplyr::group_by(harmony_clusters) %>%
  dplyr::mutate(proportion = n /sum(n))  #fraction of cells per donor
```



```{r}

###plot
p = ggplot(donors_counts, aes(
  x = harmony_clusters,
  y = proportion,
  group = Donors,
  color = Donors
)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(size = 2.8, alpha = 0.9) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_viridis_d(option = "turbo", name = "Donor") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Donors Proportion",
    x = "Cluster",
    y = "Proportion (%)"
  ) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
p

ggsave(filename = "output/T1Diabetes/T1DDonor.png", plot = p, width = 12, height = 8)
```




```{r}
##proportion of cells

library(dplyr)
library(ggplot2)
# Use explicit dplyr:: functions to avoid masking issues
prop_df=dplyr::as_tibble(T1D@meta.data) %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()

##rename col
T1D_meta <- dplyr::as_tibble(T1D@meta.data) %>%
  dplyr::rename(sample_id = sample_id)   # rename to avoid conflicts
prop_df=T1D_meta %>%
  dplyr::count(sample_id, celltype) %>%
  dplyr::group_by(sample_id) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup()
##plot
fig1c=ggplot(
  prop_df,
  aes(x = celltype, y = proportion, color = celltype)
) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 2) +
  scale_color_manual(values = celltype_cols) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  ylab("Proportion") +
  xlab("")
fig1c
ggsave(filename = "output/T1Diabetes/T1Dproportion.png", plot = fig1c, width = 12, height = 10)

```


```{r}
#genes
###with the markers another plot type


# Create DotPlot but prevent automatic flipping
p1=DotPlot(T1D, features = genes, assay = "RNA") + 
  labs(x = "Genes", y = "Cell types") +
  scale_color_gradientn(colours = c('#330066','#336699','#66CC66','#FFCC33')) +
  guides(size = guide_legend("Percent Expression")) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )
p1=p1 + aes(x = features.plot, y = id)  # 'id' = cell type, 'features.plot' = gene
p1

```




###save the final meta.data
```{r}

##save
# Keep your existing metadata names
T1D$orig.ident="Type 1 Diabete"

# Remove unwanted columns 
cols_to_remove=c(
  grep("res_|RNA_snn_res", colnames(T1D@meta.data), value = TRUE),
  "cell_type",
  "iscell_ed",
  "iscell_dd",
  "Donors"
)
T1D@meta.data=T1D@meta.data[, !(colnames(T1D@meta.data) %in% cols_to_remove)]

# Ensure active identity reflects your celltype
Idents(T1D)=T1D$celltype

# Save 
saveRDS(T1D, file = "data/process/SeuratObjects//T1D_Final.rds")

# Save only the final metadata 
saveRDS(T1D@meta.data, file = "data/process/SeuratObjects/T1Diabetes/T1D_obj.metadata.rds")

# Optional CSV 
write.csv(T1D@meta.data, 
          file = "data/process/SeuratObjects/T1Diabetes/T1D.metadata.csv",
          row.names = TRUE)
View(T1D@meta.data)
```





###save
```{r}
##save
#saveRDS(T1D, file="data/process/SeuratObjects/T1Diabetes/T1D_process.rds")

#load it
T1D = readRDS("data/process/SeuratObjects/T1Diabetes/T1D_process.rds")
```





















